### 2024-03-27 01-Manuscript/Manuscript.qmd
Adding MS file

---
title: "Computational methods in landscape ecology"
bibliography: CLER.bib
---

Landscape ecology is fun [@Turner1989].

### 2024-03-28 01-Manuscript/Manuscript.qmd
Update .gitignore, Project and MS

@@
 title: "Computational methods in landscape ecology"
+author:
+  - name: Maximilian H.K. Hesselbarth
+    orcid: 0000-0003-1125-9918
+    email: mhk.hesselbarth@gmail.com
+    affiliations:
+      - name: International Institute for Applied Systems Analysis (IIASA) 
+        city: Laxenburg
 bibliography: CLER.bib
+format:
+  pdf
 ---

### 2024-03-29 04-Various/Outline.qmd
Add outline draft

---
format: html
---

**Journal**: Current Landscape Ecology Reports

**Section**: Methodological Developments in Landscape Ecology

**Article Topic**: Computational methods in landscape ecology

**Editors**: Dr. Dimitri Justeau-Allaire (Section Editor); Dr. Sara Gagné (Editor-in-Chief)

**Due Date**: August 16, 2024

Most recent update: `r format(Sys.time(), "%B %d, %Y // %H:%M:%S")`

# General

1. Please feel free to add topics to the outline. This is just based on some brief brain storming and more a conversation starter
2. Please feel free to add your name to a section as lead. Of course multiple people can work on one section
3. Please feel free to add general comments, remarks, etc.

# Outline

- Introduction to landscape ecology (*Hesselbarth*)
  - Definition of landscape ecology
  - Brief history of landscape ecology
  - Computational aspects of landscape ecology
- Open source vs. proprietary software
- GUI vs. script based software
- Programming language
  - R (*Hesselbarth*, *Nowosad*)
  - Python (*Bosch*)
  - Julia (*Schouten*)
- Classic/traditional tools
- Upcoming/modern tools
  - AI/ML tools
  - Virtual reality
- Spatial planning perspective (*Jung*)
- Spatial patterns (*Nowosad*, *Hesselbarth?*)
- Landscape genetics (*Breitbart*)

# Open Questions

-	Remote sensing?
-	General spatial data developments?

# Literature to look at
[10.1007/s10980-021-01366-9](www.doi.org/10.1007/s10980-021-01366-9)

### 2024-03-29 04-Various/Authors.qmd
Update authors

---
title: Authors
format: 
  html:
    page-layout: full
---

Table with authors information (in random order)

| Name | Affiliation | E-Mail | ORCID |
|:------|-------------|--------|-------:|
| Jakub Nowosad | Adam Mickiewicz Uni | nowosad.jakub@gmail.com | 0000-0002-1057-3721 |
| Martin Jung | IIASA | jung@iiasa.ac.at | 0000-0002-7569-1390 |
| Marti Bosch | EPFL | marti.bosch@epfl.ch | 0000-0001-8735-9144 |
| Rafael Schouten | University of Copenhagen | rafael.schouten@sund.ku.dk | 0000-0002-8380-0884 |
| Sophie Breitbart | University of Toronto | sophie.breitbart@mail.utoronto.ca | 0000-0001-9641-9786 |
| Maximilian Hesselbarth | IIASA | mhk.hesselbarth@gmail.com | 0000-0003-1125-9918 |

### 2024-03-29 01-Manuscript/Manuscript.qmd
Adding MS

--- 2024-03-27 01-Manuscript/Manuscript.qmd
@@
 title: "Computational methods in landscape ecology"
+author:
+  - name: Maximilian H.K. Hesselbarth
+    orcid: 0000-0003-1125-9918
+    email: mhk.hesselbarth@gmail.com
+    affiliations:
+      - name: International Institute for Applied Systems Analysis (IIASA) 
+        city: Laxenburg
 bibliography: CLER.bib
+format:
+  pdf:
+    documentclass: scrartcl
+    number-sections: true
+    include-in-header: 
+      text: '\usepackage{lineno}\linenumbers'
 ---
 
+# Introduction
+
 Landscape ecology is fun [@Turner1989].
+There are many things to explore [@With2019].
+
+\printbibliography

### 2024-04-05 04-Various/Literature.qmd
Adding literature doc

---
title: Interesting literature
format: html
---

Most recent update: `r format(Sys.time(), "%B %d, %Y // %H:%M:%S")`

# Literature to look at
[10.1007/s10980-021-01366-9](www.doi.org/10.1007/s10980-021-01366-9)

### 2024-04-10 04-Various/Outline.qmd
Update some docs
Update outline

@@
 ---
+title: Outline
 format: html
@@
 2. Please feel free to add your name to a section as lead. Of course multiple people can work on one section
-3. Please feel free to add general comments, remarks, etc.
+3. Please feel free to add general comments, remarks, etc. (using your initals)
+4. Knit and push changes
 
@@
 - Introduction to landscape ecology (*Hesselbarth*)
-  - Definition of landscape ecology
-  - Brief history of landscape ecology
+  - Brief definition of landscape ecology
   - Computational aspects of landscape ecology
+- Software and programming language for landscape ecology
   - Open source vs. proprietary software
@@
     - Julia (*Schouten*)
-- Classic/traditional tools
-- Upcoming/modern tools
-  - AI/ML tools
-  - Virtual reality
+- Computational methods
+  - Visualization (i.e., maps)
+  - Spatial patterns (*Nowosad*, *Hesselbarth?*)
+  - Connectivity
+  - Simulations (e.g., neutral landscapes) (*Hesselbarth*)
+  - Optimization
   - Spatial planning perspective (*Jung*)
-- Spatial patterns (*Nowosad*, *Hesselbarth?*)
-- Landscape genetics (*Breitbart*)
+  - Landscape genetics (*Breitbart?*)
+- Discussion
+  - General spatial data developments
+  - AI impacts on landscape ecology
 
@@
 
--	Remote sensing?
--	General spatial data developments?
-
-# Literature to look at
-[10.1007/s10980-021-01366-9](www.doi.org/10.1007/s10980-021-01366-9)
+-	Remote sensing developments?

### 2024-04-10 03-RScripts/dummy.R
Adding dummy files

@@ -1,0 +1,4 @@
+# This is just a dummy script
+
+x <- 5
+x * 2

### 2024-05-15 04-Various/Literature.qmd
Update literature
Update engines various

@@
 format: html
+engine: knitr
 ---
 
-Most recent update: `r format(Sys.time(), "%B %d, %Y // %H:%M:%S")`
+Most recent update: `r format(Sys.time(), "%B %d, %Y // %H:%M")`
 
@@
 [10.1007/s10980-021-01366-9](www.doi.org/10.1007/s10980-021-01366-9)
+[https://doi.org/10.1111/2041-210X.14061](https://doi.org/10.1111/2041-210X.14061)
+[https://sustainability.fb.com/blog/2024/04/22/using-artificial-intelligence-to-map-the-earths-forests/](https://sustainability.fb.com/blog/2024/04/22/using-artificial-intelligence-to-map-the-earths-forests/)

### 2024-05-15 01-Manuscript/Manuscript.qmd
Update some docs
Update MS

--- 2024-03-27 01-Manuscript/Manuscript.qmd
@@
 title: "Computational methods in landscape ecology"
+author:
+  - name: Maximilian H.K. Hesselbarth
+    orcid: 0000-0003-1125-9918
+    email: mhk.hesselbarth@gmail.com
+    affiliations:
+      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg (Austria)
 bibliography: CLER.bib
+format:
+  pdf:
+    keep-tex: true
+    number-sections: true
+    template-partials: 
+      - title.tex
+    include-in-header: 
+      text: | 
+        \usepackage{lineno}\linenumbers
+        \usepackage[noblocks]{authblk}
+        \renewcommand*{\Authsep}{, }
+        \renewcommand*{\Authand}{, }
+        \renewcommand*{\Authands}{, }
+        \renewcommand\Affilfont{\small}
 ---
 
+# Introduction
+
 Landscape ecology is fun [@Turner1989].
+There are many things to explore [@With2019].
+
+{{< pagebreak >}}

### 2024-05-15 Manuscript.qmd
Starting to move things around

---
title: "Computational methods in landscape ecology"
author:
  - name: Maximilian H.K. Hesselbarth
    orcid: 0000-0003-1125-9918
    email: mhk.hesselbarth@gmail.com
    affiliations:
      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg (Austria)
bibliography: CLER.bib
format:
  pdf:
    keep-tex: true
    template-partials: 
      - title.tex
    include-in-header: 
      text: | 
        \usepackage{lineno}\linenumbers
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
---

<!-- FIXME: Add Authors -->
<!-- FIXME: Add sections -->

{{< include Sections/introduction.qmd >}}
{{< include Sections/spatial-planning.qmd >}}

{{< pagebreak >}}

### References

::: {#refs}
:::

### 2024-05-15 Manuscript.tex
Starting to move things around

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{lineno}\linenumbers
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Computational methods in landscape ecology},
  pdfauthor={Maximilian H.K. Hesselbarth},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Computational methods in landscape ecology}

  \author{Maximilian H.K. Hesselbarth}
            \affil{%
                  International Institute for Applied Systems Analysis,
                  Biodiversity, Ecology, and Conservation Group,
                  Laxenburg (Austria)
              }

\date{}
\begin{document}
\maketitle

\section{Introduction}\label{introduction}

Landscape ecology is fun (Turner 1989). There are many things to explore
(With 2019).

\section{Spatial planning}\label{spatial-planning}

Landscape ecology is fun (Turner 1989). There are many things to explore
(With 2019).

\newpage{}

\subsubsection{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-Turner1989}
Turner, M. G. 1989. {``Landscape Ecology: The Effect of Pattern on
Process.''} \emph{Annual Review of Ecology and Systematics} 20 (1):
171--97. \url{https://doi.org/10.1146/annurev.es.20.110189.001131}.

\bibitem[\citeproctext]{ref-With2019}
With, K. A. 2019. \emph{Essentials of Landscape Ecology}. 1st ed.
Oxford, {UK}: Oxford University Press.

\end{CSLReferences}

\end{document}

### 2024-05-15 Sections/introduction.qmd
Starting to move things around

--- 2024-03-28 01-Manuscript/Manuscript.qmd
@@
 ---
-title: "Computational methods in landscape ecology"
-author:
-  - name: Maximilian H.K. Hesselbarth
-    orcid: 0000-0003-1125-9918
-    email: mhk.hesselbarth@gmail.com
-    affiliations:
-      - name: International Institute for Applied Systems Analysis (IIASA) 
-        city: Laxenburg
 bibliography: CLER.bib
-format:
-  pdf
 ---
 
+<!-- Person responsible: Hesselbarth -->
+
+# Introduction
+
 Landscape ecology is fun [@Turner1989].
+There are many things to explore [@With2019].

### 2024-05-17 Manuscript.qmd
Add new MS and Sections structure

@@
 
-<!-- FIXME: Add Authors -->
+<!-- FIXME: Update Authors -->
 <!-- FIXME: Add sections -->
@@
 {{< include Sections/introduction.qmd >}}
+{{< include Sections/spatial-pattern.qmd >}}
 {{< include Sections/spatial-planning.qmd >}}

### 2024-05-17 Sections/simulations.qmd
Add new MS and Sections structure

---
bibliography: CLER.bib
---

<!-- Person responsible: Hesselbarth, Schouten -->

# Simulation models

## Neutral landscapes
<!-- Hesselbarth -->

## Process-based simulation
<!-- Schouten, Hesselbarth -->

### 2024-05-17 Sections/software.qmd
Add new MS and Sections structure

---
bibliography: CLER.bib
---

<!-- Person responsible: Hesselbarth, Nowosad, Bosch, Jung, Schouten  -->

# Software and programming languages

## Open source vs. proprietary software

## GUI vs. script-based software

## Programing languages

### R
<!-- Hesselbarth -->

### Python
<!-- Bosch -->

### Julia
<!-- Schouten -->

### 2024-05-17 Sections/spatial-patterns.qmd
Add new MS and Sections structure

---
bibliography: CLER.bib
---

<!-- Person responsible: Nowosad, Bosch, Hesselbarth -->

# Spatial patterns

Landscape ecology is fun [@Turner1989].
There are many things to explore [@With2019].

## Landscape metrics
<!-- Hesselbarth -->

## Vector metrics
<!-- Nowosad, Hesselbarth -->

## Surface metrics
<!-- Bosch, Hesselbarth -->

## Entropy
<!-- Nowosad, Bosch -->

## Landscape Mosaic
<!-- Bosch -->

### 2024-05-17 Sections/introduction.qmd
Start to work on intro

@@
 
-Landscape ecology is fun [@Turner1989].
-There are many things to explore [@With2019].
+Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986,@Forman1995] - with empathize on existing heterogeneity in at least one factor of interest [@Turner2015].
+The link of spatial heterogeneity and abiotic as well as biotic processes is the fundamental concept of landscape ecology.
+This includes potential feedbacks between processes and heterogeneity, spatial and temporal flows across heterogenous landscapes, or the management of heterogeneity in landscapes [@Risser1984,@Turner1989].
+However, due to the spatial context of heterogeneity, the pattern-process link is scale dependent [@Levin1992,@Wiens1989,@Wu2013].
+<!-- MH: Add remote sensing data paper -->
+Because landscape ecology is a cross-disciplinary field including, e.g., social and political sciences, geography, or ecology and evolution [@Wiens1997,With2019] and the general increase of available data [@Jaric2020,@Nathan2022,@Wuest2020], there is the need for sophisticated computational methods.
+
+<!-- Computational methods refer to techniques and procedures used to solve problems and perform calculations using computers -->
+<!-- Computational methods in landscape ecology refer to the use of computer-based algorithms and models to analyze and understand the patterns and processes of landscapes. -->

### 2024-05-20 Sections/spatial-patterns.qmd
adds various comments

@@
 
-# Spatial patterns
-
 Landscape ecology is fun [@Turner1989].
@@
 
+# Spatial patterns
+
+spatial patterns definitions in the literature (general)
+?existance of spatial pattterns in many contexts, from geography, through biology and urban planning to medical studies
+?relation of spatial patterns to spatial processes
+
+spatial patterns in data (computational approach); its limitations and considerations
+raster data
+vector data: point, line, polygon
+point clouds 
+time series data
+
+one variable vs. multivariable spatial patterns
+categorical vs. continuous spatial patterns
+
+resolution, scale, and extent in spatial patterns
+
+short intro to the subsequent sections
+
+!to conclusions that relation of spatial patterns to spatial processes is a golden grail of landscape ecology
+!what with fractals?
+!ai and deep learning still seems to be underused in landscape ecology
+
 ## Landscape metrics
 <!-- Hesselbarth -->
-
-## Vector metrics
-<!-- Nowosad, Hesselbarth -->
 
@@
 
-## Landscape Mosaic
+intro to entropy
+various concepts: thermodynamic (Boltzmann), information (Shannon), Kolmogorov, etc.
+how they are related to each other
+how they are related to spatial patterns and landscape ecology
+
+entropy measures are still related to the underlining data representation
+
+claims of relations between entropy and thermodynamics
+
+## Landscape mosaic
 <!-- Bosch -->
+
+## Vector metrics
+<!-- Nowosad, Hesselbarth -->
+
+an intro to the section
+what is similar to landscape metrics
+
+add an info about shape metrics

### 2024-05-21 Sections/spatial-patterns.qmd
adds a bunch of ieas

@@
 # Spatial patterns
+
+@daleSpatialPatternAnalysis2000
+@wiegandUsingPatternOriented2003
 
@@
 intro to entropy
-various concepts: thermodynamic (Boltzmann), information (Shannon), Kolmogorov, etc.
+various concepts:
+
+- information (Shannon)/Shannon diversity
+@Shannon1948
+@nowosadInformationTheoryConsistent2019a
+
+- thermodynamic (Boltzmann)
+
+@Cushman2016
+@cushmanGeneralizingBoltzmannConfigurational2021
+@gaoComputationBoltzmannEntropy2019
+@zhangCalculatingWassersteinMetricBased2020a
+@stepinskiCurbYourEnthusiasm2022
+
+- Kolmogorov, etc.; Juhász-Nagy’s information theory-based functions, Simpson index, Rao index, Pielou, Renyu, ...
+@konradAdaptingMultiscaleApproach2023; 
+@simpsonMeasurementDiversity1949
+
 how they are related to each other
@@
 claims of relations between entropy and thermodynamics
+
+<!-- https://r-spatialecology.github.io/landscapemetrics/ -->
+<!-- https://jakubnowosad.com/bespatial/ -->
+<!-- https://cran.r-project.org/web/packages/LandComp/index.html -->
 
@@
 add an info about shape metrics
+@basaranerPerformanceShapeIndices2017
+
+https://github.com/r-spatialecology/vectormetrics
+
+<!--misc-->
+<!-- the need for open-source software in landscape ecology -->
+<!-- every proposed method should be open-source and available to the community -->
+
+## Operations on spatial patterns
+
+@nowosadMotifOpensourceTool2021
+
+<!-- FOSS4G Europe pap? -->

### 2024-05-21 Sections/introduction.qmd
Work on intro

--- 2024-05-15 Sections/introduction.qmd
@@
 
-Landscape ecology is fun [@Turner1989].
-There are many things to explore [@With2019].
+<!--MH: Flesh-out all paragraphs a little bit adding some details and examples-->
+
+Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986;@Forman1995] - with empathize on existing heterogeneity in at least one factor of interest [@Turner2015].
+Linking the spatial heterogeneity and abiotic as well as biotic processes is the fundamental concept of landscape ecology.
+This includes potential feedbacks between heterogeneity and processes, spatial and temporal flows across heterogenous landscapes, or the management of heterogeneity in landscapes [@Risser1984;@Turner1989].
+However, due to the spatial context of heterogeneity, the pattern-process link is scale dependent [@Levin1992;@Wiens1989,@Wu2013].
+
+Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms. 
+Thus, computational ecology can be defined if computational science is used to address ecological research questions with focus on complex adaptive systems and data-driven approaches [@Poisot2019].
+Computational ecology in general and simulations specifically are important for ecology because ecological data is often context- and scale-dependent and infeasible to study in reproducible, replicable and controlled experimental settings [@Petrovskii2012].
+
+Because landscape ecology is a cross-disciplinary field including [e.g., social sciences, geography, or ecology and evolution @Wiens1997;With2019], the general increase of available data [@Chi2016;@Jaric2020;@Nathan2022;@Wuest2020], or the complex systems characteristics of landscapes and processes [@Newman2019], there is the need for sophisticated computational methods in order to link patterns and processes.
+
+This paper aims to introduce the latest developments of computation methods in landscape ecology.
+However, it is not a general introduction into (computational) landscape ecology. 
+For a more comprehensive introductions to landscape ecology in general, please see e.g. Turner and Gardner [-@Turner2015] or Gergel and Turner [-@Gergel2017].

### 2024-05-22 Manuscript.qmd
Update some meta stuff

@@
 title: "Computational methods in landscape ecology"
+keywords: keyword1, keyword2
 author:
@@
       - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg (Austria)
-bibliography: CLER.bib
 format:
@@
         \renewcommand\Affilfont{\small}
+papersize: a4paper
+fontsize: 12pt
+geometry: margin=25mm
+indent: 2m
+linestretch: 1.75
+bibliography: CLER.bib
+csl: springer-basic-author-date.csl
 ---
@@
 {{< include Sections/introduction.qmd >}}
-{{< include Sections/spatial-pattern.qmd >}}
+{{< include Sections/spatial-patterns.qmd >}}
 {{< include Sections/spatial-planning.qmd >}}

### 2024-05-22 Manuscript.tex
Update some meta stuff

@@
 \documentclass[
+  12pt,
+  a4paperpaper,
 ]{article}
@@
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
 \usepackage{xcolor}
+\usepackage[margin=25mm]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   pdfauthor={Maximilian H.K. Hesselbarth},
+  pdfkeywords={keyword1, keyword2},
   colorlinks=true,
@@
 
+\setstretch{1.75}
 \section{Introduction}\label{introduction}
+
+Landscapes are typically defined as mosaics of different land covers,
+habitats, ecosystems, or land use systems (Forman and Godron 1986;
+Forman 1995) - with empathize on existing heterogeneity in at least one
+factor of interest (Turner and Gardner 2015). Linking the spatial
+heterogeneity and abiotic as well as biotic processes is the fundamental
+concept of landscape ecology. This includes potential feedbacks between
+heterogeneity and processes, spatial and temporal flows across
+heterogenous landscapes, or the management of heterogeneity in
+landscapes (Risser et al. 1984; Turner 1989). However, due to the
+spatial context of heterogeneity, the pattern-process link is scale
+dependent (Wiens 1989; Levin 1992; Wu 2013).
+
+Computational science involves analyzing abstracted core mechanisms of
+research questions using data and algorithms. Thus, computational
+ecology can be defined if computational science is used to address
+ecological research questions with focus on complex adaptive systems and
+data-driven approaches (Poisot et al. 2019). Computational ecology in
+general and simulations specifically are important for ecology because
+ecological data is often context- and scale-dependent and infeasible to
+study in reproducible, replicable and controlled experimental settings
+(Petrovskii and Petrovskaya 2012).
+
+Because landscape ecology is a cross-disciplinary field including (e.g.,
+social sciences, geography, or ecology and evolution Wiens 1997; With
+2019), the general increase of available data (Chi et al. 2016; Jarić et
+al. 2020; Wüest et al. 2020; Nathan et al. 2022), or the complex systems
+characteristics of landscapes and processes (Newman et al. 2019), there
+is the need for sophisticated computational methods in order to link
+patterns and processes.
+
+This paper aims to introduce the latest developments of computation
+methods in landscape ecology. However, it is not a general introduction
+into (computational) landscape ecology. For a more comprehensive
+introductions to landscape ecology in general, please see e.g.~Turner
+and Gardner (2015) or Gergel and Turner (2017).
+
+\section{Spatial patterns}\label{spatial-patterns}
 
@@
 (With 2019).
+
+\subsection{Landscape metrics}\label{landscape-metrics}
+
+\subsection{Vector metrics}\label{vector-metrics}
+
+\subsection{Surface metrics}\label{surface-metrics}
+
+\subsection{Entropy}\label{entropy}
+
+\subsection{Landscape Mosaic}\label{landscape-mosaic}
 
@@
 \phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{1}{1}
+\bibitem[\citeproctext]{ref-Chi2016}
+Chi M, Plaza A, Benediktsson JA, et al (2016) Big data for remote
+sensing: Challenges and opportunities. Proceedings of the IEEE
+104:2207--2219. \url{https://doi.org/10.1109/JPROC.2016.2598228}
+
+\bibitem[\citeproctext]{ref-Forman1995}
+Forman RTT (1995) Land mosaics: The ecology of landscapes and regions.
+Cambridge University Press, Cambridge, UK
+
+\bibitem[\citeproctext]{ref-Forman1986}
+Forman RTT, Godron M (1986) Landscape ecology. Wiley \& Sons,
+Chichester, UK
+
+\bibitem[\citeproctext]{ref-Gergel2017}
+Gergel SE, Turner MG (eds) (2017)
+\href{https://doi.org/10.1007/978-1-4939-6374-4}{Learning landscape
+ecology}. Springer New York, New York, NY
+
+\bibitem[\citeproctext]{ref-Jaric2020}
+Jarić I, Correia RA, Brook BW, et al (2020) iEcology: Harnessing large
+online resources to generate ecological insights. Trends in Ecology \&
+Evolution 35:630--639. \url{https://doi.org/10.1016/j.tree.2020.03.003}
+
+\bibitem[\citeproctext]{ref-Levin1992}
+Levin SA (1992) The problem of pattern and scale in ecology. Ecology
+73:1943--1967. \url{https://doi.org/10.2307/1941447}
+
+\bibitem[\citeproctext]{ref-Nathan2022}
+Nathan R, Monk CT, Arlinghaus R, et al (2022) Big-data approaches lead
+to an increased understanding of the ecology of animal movement. Science
+375:eabg1780. \url{https://doi.org/10.1126/science.abg1780}
+
+\bibitem[\citeproctext]{ref-Newman2019}
+Newman EA, Kennedy MC, Falk DA, McKenzie D (2019) Scaling and complexity
+in landscape ecology. Frontiers in Ecology and Evolution 7:293.
+\url{https://doi.org/10.3389/fevo.2019.00293}
+
+\bibitem[\citeproctext]{ref-Petrovskii2012}
+Petrovskii S, Petrovskaya N (2012) Computational ecology as an emerging
+science. Interface Focus 2:241--254.
+\url{https://doi.org/10.1098/rsfs.2011.0083}
+
+\bibitem[\citeproctext]{ref-Poisot2019}
+Poisot T, LaBrie R, Larson E, et al (2019) Data-based, synthesis-driven:
+Setting the agenda for computational ecology. Ideas in Ecology and
+Evolution 12: \url{https://doi.org/10.24908/iee.2019.12.2.e}
+
+\bibitem[\citeproctext]{ref-Risser1984}
+Risser PG, Karr JR, Forman RTT (1984) Landscape ecology: Directions and
+approaches. Illinois Natural History Survery Special Publication 2:7--14
+
 \bibitem[\citeproctext]{ref-Turner1989}
-Turner, M. G. 1989. {``Landscape Ecology: The Effect of Pattern on
-Process.''} \emph{Annual Review of Ecology and Systematics} 20 (1):
-171--97. \url{https://doi.org/10.1146/annurev.es.20.110189.001131}.
+Turner MG (1989) Landscape ecology: The effect of pattern on process.
+Annual Review of Ecology and Systematics 20:171--197.
+\url{https://doi.org/10.1146/annurev.es.20.110189.001131}
+
+\bibitem[\citeproctext]{ref-Turner2015}
+Turner MG, Gardner RH (2015) Landscape ecology in theory and practice:
+Pattern and process, 2nd edition. Springer, New York
+
+\bibitem[\citeproctext]{ref-Wiens1989}
+Wiens JA (1989) Spatial scaling in ecology. Functional Ecology
+3:385--397. \url{https://doi.org/10.2307/2389612}
+
+\bibitem[\citeproctext]{ref-Wiens1997}
+Wiens JA (1997)
+\href{https://doi.org/10.1016/B978-012323445-2/50005-5}{Metapopulation
+dynamics and landscape ecology}. In: Metapopulation biology. Elsevier,
+pp 43--62
 
 \bibitem[\citeproctext]{ref-With2019}
-With, K. A. 2019. \emph{Essentials of Landscape Ecology}. 1st ed.
-Oxford, {UK}: Oxford University Press.
+With KA (2019) Essentials of landscape ecology, 1st edn. Oxford
+University Press, Oxford, UK
+
+\bibitem[\citeproctext]{ref-Wu2013}
+Wu J (2013) Key concepts and research topics in landscape ecology
+revisited: 30 years after the allerton park workshop. Landscape Ecology
+28:1--11. \url{https://doi.org/10.1007/s10980-012-9836-y}
+
+\bibitem[\citeproctext]{ref-Wuest2020}
+Wüest RO, Zimmermann NE, Zurell D, et al (2020) Macroecology in the age
+of big data -- where to go from here? Journal of Biogeography 47:1--12.
+\url{https://doi.org/10.1111/jbi.13633}
 

### 2024-05-22 Manuscript.qmd
Minor changes intro

@@
 indent: 2m
-linestretch: 1.75
+linestretch: 1.5
 bibliography: CLER.bib

### 2024-05-22 Manuscript.tex
Minor changes intro

@@
 
-\setstretch{1.75}
+\setstretch{1.5}
 \section{Introduction}\label{introduction}
@@
 habitats, ecosystems, or land use systems (Forman and Godron 1986;
-Forman 1995) - with empathize on existing heterogeneity in at least one
-factor of interest (Turner and Gardner 2015). Linking the spatial
-heterogeneity and abiotic as well as biotic processes is the fundamental
-concept of landscape ecology. This includes potential feedbacks between
-heterogeneity and processes, spatial and temporal flows across
-heterogenous landscapes, or the management of heterogeneity in
-landscapes (Risser et al. 1984; Turner 1989). However, due to the
-spatial context of heterogeneity, the pattern-process link is scale
-dependent (Wiens 1989; Levin 1992; Wu 2013).
+Forman 1995) - with empathize on heterogeneity of at least one factor of
+interest (Turner and Gardner 2015). Linking the spatial heterogeneity
+and ecological processes, including potential feedbacks, is the
+fundamental concept of landscape ecology (Risser et al. 1984; Turner
+1989, 2005). Typical research topics include ecological flows in
+landscape mosaics, drivers and consequences of land use and land cover
+change, nonlinear dynamics and landscape complexity, scale and scaling,
+spatial analysis and landscape modeling, linking landscape metrics to
+ecological processes, integrating humans and their activities into
+landscape ecology, landscape conservation and sustainability, or
+accuracy assessment and uncertainty analysis (Wu and Hobbs 2002; Wu
+2013).
 
@@
 research questions using data and algorithms. Thus, computational
-ecology can be defined if computational science is used to address
-ecological research questions with focus on complex adaptive systems and
-data-driven approaches (Poisot et al. 2019). Computational ecology in
-general and simulations specifically are important for ecology because
-ecological data is often context- and scale-dependent and infeasible to
-study in reproducible, replicable and controlled experimental settings
-(Petrovskii and Petrovskaya 2012).
-
-Because landscape ecology is a cross-disciplinary field including (e.g.,
-social sciences, geography, or ecology and evolution Wiens 1997; With
-2019), the general increase of available data (Chi et al. 2016; Jarić et
-al. 2020; Wüest et al. 2020; Nathan et al. 2022), or the complex systems
-characteristics of landscapes and processes (Newman et al. 2019), there
-is the need for sophisticated computational methods in order to link
-patterns and processes.
+ecology can be defined as computational science that is used to address
+ecological research questions with focus on complex, adaptive systems
+and data-driven approaches (Poisot et al. 2019). Computational ecology
+in general and simulations specifically are of importance for ecology
+because ecological data is often context- and scale-dependent and
+unfeasible to study in reproducible, replicable, and controllable
+experimental settings (Petrovskii and Petrovskaya 2012; but see Wiersma
+2022). Because landscape ecology is a cross-disciplinary field
+including, e.g., social sciences, geography, or ecology and evolution
+(Wiens 1997; With 2019), the general increase of available data (Chi et
+al. 2016; Jarić et al. 2020; Wüest et al. 2020; Nathan et al. 2022), or
+the complex landscape systems (Newman et al. 2019), there is the need
+for sophisticated computational methods in order to link patterns and
+processes. Thus, computational methods are one of the most important
+tools of modern scientific research (Prlić and Procter 2012; Wilson et
+al. 2014).
 
 This paper aims to introduce the latest developments of computation
-methods in landscape ecology. However, it is not a general introduction
-into (computational) landscape ecology. For a more comprehensive
-introductions to landscape ecology in general, please see e.g.~Turner
-and Gardner (2015) or Gergel and Turner (2017).
+methods in landscape ecology. However, it is neither a systematic review
+of the literature nor general introduction to (computational) landscape
+ecology. For a more comprehensive introductions to landscape ecology in
+general, please see Turner and Gardner (2015) or Gergel and Turner
+(2017).
 
@@
 
-\bibitem[\citeproctext]{ref-Levin1992}
-Levin SA (1992) The problem of pattern and scale in ecology. Ecology
-73:1943--1967. \url{https://doi.org/10.2307/1941447}
-
 \bibitem[\citeproctext]{ref-Nathan2022}
@@
 
+\bibitem[\citeproctext]{ref-Prlic2012}
+Prlić A, Procter JB (2012) Ten simple rules for the open development of
+scientific software. PLoS Computational Biology 8:e1002802.
+\url{https://doi.org/10.1371/journal.pcbi.1002802}
+
 \bibitem[\citeproctext]{ref-Risser1984}
@@
 
+\bibitem[\citeproctext]{ref-Turner2005}
+Turner MG (2005) Landscape ecology: What is the state of the science?
+Annual Review of Ecology, Evolution, and Systematics 36:319--344.
+\url{https://doi.org/10.1146/annurev.ecolsys.36.102003.152614}
+
 \bibitem[\citeproctext]{ref-Turner2015}
@@
 Pattern and process, 2nd edition. Springer, New York
-
-\bibitem[\citeproctext]{ref-Wiens1989}
-Wiens JA (1989) Spatial scaling in ecology. Functional Ecology
-3:385--397. \url{https://doi.org/10.2307/2389612}
 
@@
 pp 43--62
+
+\bibitem[\citeproctext]{ref-Wiersma2022}
+Wiersma YF (2022) A review of landscape ecology experiments to
+understand ecological processes. Ecological Processes 11:57.
+\url{https://doi.org/10.1186/s13717-022-00401-0}
+
+\bibitem[\citeproctext]{ref-Wilson2014}
+Wilson G, Aruliah DA, Brown CT, et al (2014) Best practices for
+scientific computing. PLoS Biology 12:e1001745.
+\url{https://doi.org/10.1371/journal.pbio.1001745}
 
@@
 
+\bibitem[\citeproctext]{ref-Wu2002}
+Wu J, Hobbs R (2002) Key issues and research priorities in landscape
+ecology: An idiosyncratic synthesis. Landscape Ecology 17:355--365.
+\url{https://doi.org/10.1023/A:1020561630963}
+
 \bibitem[\citeproctext]{ref-Wuest2020}

### 2024-05-22 Sections/introduction.qmd
Minor changes intro

--- 2024-05-15 Sections/introduction.qmd
@@
 
-Landscape ecology is fun [@Turner1989].
-There are many things to explore [@With2019].
+Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
+Linking the spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
+Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].
+
+<!-- MH: Currently using "computational science", "computational ecology", and "computational methods". Streamline terms?--> 
+Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
+Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
+Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in reproducible, replicable, and controllable experimental settings [@Petrovskii2012; but see @Wiersma2022].
+Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997; @With2019], the general increase of available data [@Chi2016; @Jaric2020; @Nathan2022; @Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link patterns and processes.
+Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].
+
+This paper aims to introduce the latest developments of computation methods in landscape ecology.
+However, it is neither a systematic review of the literature nor general introduction to (computational) landscape ecology.
+For a more comprehensive introductions to landscape ecology in general, please see Turner and Gardner [-@Turner2015] or Gergel and Turner [-@Gergel2017].

### 2024-05-28 Sections/spatial-patterns.qmd
Adding some notes to spatial-patterns

@@
 
-Landscape ecology is fun [@Turner1989].
-There are many things to explore [@With2019].
+# Spatial patterns
 
-# Spatial patterns
+[Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022)
+[Where are we now, and where are we going?](https://doi.org/10.1007/s10980-019-00889-6)
 
@@
 
-spatial patterns definitions in the literature (general)
-?existance of spatial pattterns in many contexts, from geography, through biology and urban planning to medical studies
-?relation of spatial patterns to spatial processes
+- Spatial patterns definitions in the literature (general; Dale 2000)
+- Spatial patterns in many contexts (e.g. geography, biology, urban planning, medical studies) <!-- MH: I would at least cut the medical bit -->
+- Spatial patterns in landscape ecology
+  - Resolution, scale, and extent in spatial patterns
+- Relation of spatial patterns and spatial (ecological?) processes
+- Relation of spatial patterns to spatial processes is a golden grail of landscape ecology
 
-spatial patterns in data (computational approach); its limitations and considerations
-raster data
-vector data: point, line, polygon
-point clouds 
-time series data
+- Spatial patterns in data (computational approach)
+  - Limitations and considerations
+- Data types <!-- MH: Could be moved to the individual sub-sections? -->
+  - Raster data
+  - Vector data: point, line, polygon
+  - Point clouds 
+  - Time series data
+  - One variable vs. multivariable spatial patterns
+  - Categorical vs. continuous spatial patterns
 
-one variable vs. multivariable spatial patterns
-categorical vs. continuous spatial patterns
-
-resolution, scale, and extent in spatial patterns
-
-short intro to the subsequent sections
-
-!to conclusions that relation of spatial patterns to spatial processes is a golden grail of landscape ecology
 !what with fractals?
@@
 
+- Traditional approach to describe patterns
+- Quantify categorical raster data (patch mosaic model)
+- Quantify composition and configuration
+- Widely applied since FRAGSTATS release
+- Several alternative tools available and still widely used
+
 ## Surface metrics
 <!-- Bosch, Hesselbarth -->
+[Surface metrics])(https://doi.org/10.1007/s10980-018-0685-1)
+
+- More recent development
+- Quantify continuous raster data (gradient surface models)
+- More realistic representation of landscapes
+- Based on molecular physics/medicine
+- Not widely adopted yet (due to missing tools?)
 
@@
 <!-- Nowosad, Bosch -->
-
 intro to entropy

### 2024-05-28 Sections/introduction.qmd
Several minor cosmetic changes and some progress to simulation section

@@
 Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
-Linking the spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
+Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
 Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].
 
-<!-- MH: Currently using "computational science", "computational ecology", and "computational methods". Streamline terms?--> 
 Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
 Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
-Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in reproducible, replicable, and controllable experimental settings [@Petrovskii2012; but see @Wiersma2022].
-Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997; @With2019], the general increase of available data [@Chi2016; @Jaric2020; @Nathan2022; @Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link patterns and processes.
+Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012; but see @Wiersma2022].
+Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997; @With2019], the general increase of available data [@Chi2016; @Jaric2020; @Nathan2022; @Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
 Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].
 
-This paper aims to introduce the latest developments of computation methods in landscape ecology.
-However, it is neither a systematic review of the literature nor general introduction to (computational) landscape ecology.
-For a more comprehensive introductions to landscape ecology in general, please see Turner and Gardner [-@Turner2015] or Gergel and Turner [-@Gergel2017].
+Here, we aim to introduce the latest developments of computational methods in landscape ecology.
+However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
+For a more comprehensive, general introductions to landscape ecology, please see Turner and Gardner [-@Turner2015], or Gergel and Turner [-@Gergel2017].
+First, we will introduce recent computational developments of the most important topics of landscape ecology.
+This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations.
+Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
+However, due to its many advantages, we will focus on open-source software [@vonKrogh2006], and specifically scripting languages, i.e., R, Python, and Julia.
+Last, we will discuss potential future needs developments of computational methods in landscape ecology..

### 2024-05-28 Sections/simulations.qmd
Several minor cosmetic changes and some progress to simulation section

@@
 # Simulation models
-
-## Neutral landscapes
 <!-- Hesselbarth -->
 
-## Process-based simulation
+Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005; @Petrovskii2012].
+Thus, they can be seen as experimental systems that allow, unlike to nature, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004].
+Due to the general temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006; @Synes2016].
+In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004;@Synes2016].
+Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.
+
+## Pattern based
+<!-- Hesselbarth -->
+
+### Neutral landscapes models
+Neutral landscapes models simulate landscape patterns without assuming any underlying abiotic or biotic processes.
+Thus, they are often used as null hypothesis or baselines for comparisons with real landscape patterns, while controlling certain aspects of the landscape [@Li2004;@Wang2008].
+The first neutral models simply assigned land cover types randomly to cells in the landscape and were based on percolation theory.
+Further developments included hierarchical models that considered different spatial scales while assigning land cover types.
+While these first two classes can only simulate landscape with discrete land cover classes, fractal neutral models can also simulate continuous land cover maps by utilizing e.g., Brownian motion.
+Last, more complex neutral models were developed that are based on contagion, autocorrelation, or spatial aggregation concepts to simulate more realistic landscapes [@Wang2008].
+In order to make multiple neutral landscape model algorithms available several software libraries are available in prominent programming languages including Python [NLMpy; @Etherington2015] or R [NLMR; @Sciaini2018].
+
+### Cellular Automata
+
+## Process based
 <!-- Schouten, Hesselbarth -->
+
+[RangeShifter](https://doi.org/10.1111/ecog.05687)
+
+### Landscape Generator
+<!-- Hesselbarth -->
+
+[G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
+[LG](https://doi.org/10.1002/ece3.2145)
+
+### Coupled lattice
+
+### IBMs/ABMs
+
+## Misc
+[Marine SDMS](https://doi.org/10.1016/j.ecolmodel.2019.108837)
+[SDM past present future](https://doi.org/10.1111/jbi.14617)
+
+[Simulations in landscape genetics](https://doi.org/10.1111/j.1365-294X.2010.04678.x)
+[HexSim modeling environment](https://doi.org/10.1007/s10980-017-0605-9)
+[Digital twins?](https://doi.org/10.1016/j.tree.2023.04.010)

### 2024-05-28 Manuscript.qmd
Adding unsorted author list to Manuscript

@@
     affiliations:
-      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg (Austria)
+      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
+    attributes:
+      - corresponding: boolean
+  - name: Jakub Nowosad
+    orcid: 0000-0002-1057-3721
+    email: nowosad.jakub@gmail.com
+    affiliations:
+      - name: Adam Mickiewicz University, Poznan, Poland
+  - name: Marti Bosch
+    orcid: 0000-0001-8735-9144
+    email: marti.bosch@epfl.ch
+    affiliations:
+      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
+  - name: Martin Jung
+    orcid: 0000-0002-7569-1390
+    email: jung@iiasa.ac.at
+    affiliations:
+      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
+  - name: Rafael Schouten
+    orcid: 0000-0002-8380-0884
+    email: rafael.schouten@sund.ku.dk
+    affiliations:
+      - name: Globe Institute, University of Copenhagen, Copenhagen, Denmark
 format:
@@
 
-<!-- FIXME: Update Authors -->
-<!-- FIXME: Add sections -->
+<!-- MH: Author order is not fixed yet. I just started to list everyone for now. -->
 
 {{< include Sections/introduction.qmd >}}
-{{< include Sections/spatial-patterns.qmd >}}
-{{< include Sections/spatial-planning.qmd >}}
+{{< include Sections/simulations.qmd >}}
 

### 2024-05-28 Manuscript.tex
Adding unsorted author list to Manuscript

@@
   pdftitle={Computational methods in landscape ecology},
-  pdfauthor={Maximilian H.K. Hesselbarth},
+  pdfauthor={Maximilian H.K. Hesselbarth; Jakub Nowosad; Marti Bosch; Martin Jung; Rafael Schouten},
   pdfkeywords={keyword1, keyword2},
@@
                   Biodiversity, Ecology, and Conservation Group,
-                  Laxenburg (Austria)
+                  Laxenburg, Austria
+              }
+        \author{Jakub Nowosad}
+            \affil{%
+                  Adam Mickiewicz University, Poznan, Poland
+              }
+        \author{Marti Bosch}
+            \affil{%
+                  École polytechnique fédérale de Lausanne, Lausanne,
+                  Switzerland
+              }
+        \author{Martin Jung}
+            \affil{%
+                  International Institute for Applied Systems Analysis,
+                  Biodiversity, Ecology, and Conservation Group,
+                  Laxenburg, Austria
+              }
+        \author{Rafael Schouten}
+            \affil{%
+                  Globe Institute, University of Copenhagen, Copenhagen,
+                  Denmark
               }
@@
 Forman 1995) - with empathize on heterogeneity of at least one factor of
-interest (Turner and Gardner 2015). Linking the spatial heterogeneity
-and ecological processes, including potential feedbacks, is the
-fundamental concept of landscape ecology (Risser et al. 1984; Turner
-1989, 2005). Typical research topics include ecological flows in
-landscape mosaics, drivers and consequences of land use and land cover
-change, nonlinear dynamics and landscape complexity, scale and scaling,
-spatial analysis and landscape modeling, linking landscape metrics to
-ecological processes, integrating humans and their activities into
-landscape ecology, landscape conservation and sustainability, or
-accuracy assessment and uncertainty analysis (Wu and Hobbs 2002; Wu
-2013).
+interest (Turner and Gardner 2015). Linking spatial heterogeneity and
+ecological processes, including potential feedbacks, is the fundamental
+concept of landscape ecology (Risser et al. 1984; Turner 1989, 2005).
+Typical research topics include ecological flows in landscape mosaics,
+drivers and consequences of land use and land cover change, nonlinear
+dynamics and landscape complexity, scale and scaling, spatial analysis
+and landscape modeling, linking landscape metrics to ecological
+processes, integrating humans and their activities into landscape
+ecology, landscape conservation and sustainability, or accuracy
+assessment and uncertainty analysis (Wu and Hobbs 2002; Wu 2013).
 
@@
 because ecological data is often context- and scale-dependent and
-unfeasible to study in reproducible, replicable, and controllable
+unfeasible to study in controllable, reproducible, and replicable
 experimental settings (Petrovskii and Petrovskaya 2012; but see Wiersma
@@
 the complex landscape systems (Newman et al. 2019), there is the need
-for sophisticated computational methods in order to link patterns and
-processes. Thus, computational methods are one of the most important
-tools of modern scientific research (Prlić and Procter 2012; Wilson et
-al. 2014).
+for sophisticated computational methods in order to link heterogeneity
+patterns and ecological processes. Thus, computational methods are one
+of the most important tools of modern scientific research (Prlić and
+Procter 2012; Wilson et al. 2014).
 
-This paper aims to introduce the latest developments of computation
-methods in landscape ecology. However, it is neither a systematic review
-of the literature nor general introduction to (computational) landscape
-ecology. For a more comprehensive introductions to landscape ecology in
-general, please see Turner and Gardner (2015) or Gergel and Turner
-(2017).
+Here, we aim to introduce the latest developments of computational
+methods in landscape ecology. However, we do not provide a systematic
+literature review, or a general introduction to (computational)
+landscape ecology. For a more comprehensive, general introductions to
+landscape ecology, please see Turner and Gardner (2015), or Gergel and
+Turner (2017). First, we will introduce recent computational
+developments of the most important topics of landscape ecology. This
+includes, for example, spatial patterns quantification, connectivity and
+movement, or model simulations. Second, we will introduce common
+software tools that implement or are potentially capable of implementing
+these developments. However, due to its many advantages, we will focus
+on open-source software (von Krogh and von Hippel 2006), and
+specifically scripting languages, i.e., R, Python, and Julia. Last, we
+will discuss potential future needs developments of computational
+methods in landscape ecology..
 
-\section{Spatial patterns}\label{spatial-patterns}
+\section{Simulation models}\label{simulation-models}
 
-Landscape ecology is fun (Turner 1989). There are many things to explore
-(With 2019).
+Simulation models are a powerful tool to study complex adaptive
+ecological systems in controllable, reproducible, and replicable
+settings (Pascual 2005; Petrovskii and Petrovskaya 2012). Thus, they can
+be seen as experimental systems that allow, unlike to nature, all
+imaginable manipulations in order to advance theoretical developments or
+test hypotheses (Peck 2004). Due to the general temporal and spatial
+scales, complex interactions and feedbacks, or scale mismatches between
+patterns and processes, simulation models are one of the major
+approaches in landscape ecology (Schröder and Seppelt 2006; Synes et al.
+2016). In general, simulations models can be classified using two major
+divisions, namely as \emph{i)} predictive or exploratory models, and
+\emph{ii)} pattern- or process-based models (Peck 2004; Synes et al.
+2016). Here, we are going to focus mainly on exploratory models, but
+include both pattern- and process-based models.
 
-\subsection{Landscape metrics}\label{landscape-metrics}
+\subsection{Pattern based}\label{pattern-based}
 
-\subsection{Vector metrics}\label{vector-metrics}
+\subsubsection{Neutral landscapes
+models}\label{neutral-landscapes-models}
 
-\subsection{Surface metrics}\label{surface-metrics}
+Neutral landscapes models simulate landscape patterns without assuming
+any underlying abiotic or biotic processes. Thus, they are often used as
+null hypothesis or baselines for comparisons with real landscape
+patterns, while controlling certain aspects of the landscape (Li et al.
+2004; Wang and Malanson 2008). The first neutral models simply assigned
+land cover types randomly to cells in the landscape and were based on
+percolation theory. Further developments included hierarchical models
+that considered different spatial scales while assigning land cover
+types. While these first two classes can only simulate landscape with
+discrete land cover classes, fractal neutral models can also simulate
+continuous land cover maps by utilizing e.g., Brownian motion. Last,
+more complex neutral models were developed that are based on contagion,
+autocorrelation, or spatial aggregation concepts to simulate more
+realistic landscapes (Wang and Malanson 2008). In order to make multiple
+neutral landscape model algorithms available several software libraries
+are available in prominent programming languages including Python
+{[}NLMpy; Etherington et al. (2015){]} or R {[}NLMR; Sciaini et al.
+(2018){]}.
 
-\subsection{Entropy}\label{entropy}
+\subsubsection{Cellular Automata}\label{cellular-automata}
 
-\subsection{Landscape Mosaic}\label{landscape-mosaic}
+\subsection{Process based}\label{process-based}
 
-\section{Spatial planning}\label{spatial-planning}
+\href{https://doi.org/10.1111/ecog.05687}{RangeShifter}
 
-Landscape ecology is fun (Turner 1989). There are many things to explore
-(With 2019).
+\subsubsection{Landscape Generator}\label{landscape-generator}
+
+\href{https://doi.org/10.1371/journal.pone.0064968}{G-RaFFE}
+\href{https://doi.org/10.1002/ece3.2145}{LG}
+
+\subsubsection{Coupled lattice}\label{coupled-lattice}
+
+\subsubsection{IBMs/ABMs}\label{ibmsabms}
+
+\subsection{Misc}\label{misc}
+
+\href{https://doi.org/10.1016/j.ecolmodel.2019.108837}{Marine SDMS}
+\href{https://doi.org/10.1111/jbi.14617}{SDM past present future}
+
+\href{https://doi.org/10.1111/j.1365-294X.2010.04678.x}{Simulations in
+landscape genetics}
+\href{https://doi.org/10.1007/s10980-017-0605-9}{HexSim modeling
+environment} \href{https://doi.org/10.1016/j.tree.2023.04.010}{Digital
+twins?}
 
@@
 104:2207--2219. \url{https://doi.org/10.1109/JPROC.2016.2598228}
+
+\bibitem[\citeproctext]{ref-Etherington2015}
+Etherington TR, Holland EP, O'Sullivan D (2015) NLMpy: A python software
+package for the creation of neutral landscape models within a general
+numerical framework. Methods in Ecology and Evolution 6:164--168.
+\url{https://doi.org/10.1111/2041-210X.12308}
 
@@
 
+\bibitem[\citeproctext]{ref-Li2004}
+Li X, He HS, Wang X, et al (2004) Evaluating the effectiveness of
+neutral landscape models to represent a real landscape. Landscape and
+Urban Planning 69:137--148.
+\url{https://doi.org/10.1016/j.landurbplan.2003.10.037}
+
 \bibitem[\citeproctext]{ref-Nathan2022}
@@
 \url{https://doi.org/10.3389/fevo.2019.00293}
+
+\bibitem[\citeproctext]{ref-Pascual2005}
+Pascual M (2005) Computational ecology: From the complex to the simple
+and back. PLoS Computional Biology 1:e18.
+\url{https://doi.org/10.1371/journal.pcbi.0010018}
+
+\bibitem[\citeproctext]{ref-Peck2004}
+Peck SL (2004) Simulation as experiment: A philosophical reassessment
+for biological modeling. Trends in Ecology \& Evolution 19:530--534.
+\url{https://doi.org/10.1016/j.tree.2004.07.019}
 
@@
 
+\bibitem[\citeproctext]{ref-Schroder2006}
+Schröder B, Seppelt R (2006) Analysis of pattern--process interactions
+based on landscape models - overview, general concepts, and
+methodological issues. Ecological Modelling 199:505--516.
+\url{https://doi.org/10.1016/j.ecolmodel.2006.05.036}
+
+\bibitem[\citeproctext]{ref-Sciaini2018}
+Sciaini M, Fritsch M, Scherer C, Simpkins CE (2018) NLMR and
+landscapetools: An integrated environment for simulating and modifying
+neutral landscape models in r. Methods in Ecology and Evolution
+9:2240--2248. \url{https://doi.org/10.1111/2041-210X.13076}
+
+\bibitem[\citeproctext]{ref-Synes2016}
+Synes NW, Brown C, Watts K, et al (2016) Emerging opportunities for
+landscape ecological modelling. Current Landscape Ecology Reports
+1:146--167
+
 \bibitem[\citeproctext]{ref-Turner1989}
@@
 Pattern and process, 2nd edition. Springer, New York
+
+\bibitem[\citeproctext]{ref-vonKrogh2006}
+von Krogh G, von Hippel E (2006) The promise of research on open source
+software. Management Science 52:975--983.
+\url{https://doi.org/10.1287/mnsc.1060.0560}
+
+\bibitem[\citeproctext]{ref-Wang2008}
+Wang Q, Malanson GP (2008) Neutral landscapes: Bases for exploration in
+landscape ecology. Geography Compass 2:319--339.
+\url{https://doi.org/10.1111/j.1749-8198.2008.00090.x}
 

### 2024-05-29 Sections/spatial-patterns.qmd
Update spatial-patterns.qmd
starts working on the spatial patterns section

@@
 ---
-bibliography: CLER.bib
+bibliography: ../CLER.bib
 ---
@@
 
-[Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022)
+Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. 
+@daleSpatialPatternAnalysis2000 defined them as the scale-dependent predictability of the physical arrangement of observations, while a more general definition by @grimmPatternOrientedModellingPopulation1996 describes them as a characteristic, clearly identifiable structure in nature itself or in the data extracted from nature, indicating an underlying process that generates this pattern.
+Both definitions highlight several important aspects of spatial patterns: they are related to processes, depending on the scale of observation, and can be found in data.
+
+The relation of spatial patterns to spatial processes is a golden grail of landscape ecology.
+Patterns observed in nature contain information about the memory of the system, including its demographic processes, dispersal characteristics, and constraining factors[@wiegandUsingPatternOriented2003].
+Moreover, spatial patterns are not only a result of processes but could also be a driver of them.
+Thus, untangling the complex history of a landscape from its spatial patterns is a hard challenge.
+
+The scale dependency of spatial patterns may relate to either the temporal or spatial scale.
+Different scales of the processes could lead to different patterns.
+<!-- ... -->
+
+Here, our focus is on the computational approaches to spatial patterns.
+Thus, we will discuss the spatial patterns that could be found in data, the metrics that could be used to quantify them, and the operations that could be performed on them.
+
+Computational approaches to spatial patterns open many new possibilities (as compared to traditional methods). 
+They allow not only to quantify the patterns but also to model them, simulate them, and predict them.
+Moreover, they allow us to analyze large datasets and apply complex algorithms.
+At the same time, the computational approaches have considerations and limitations that should be taken into account.
+
+Every data type is a simplified model of the real world, and the spatial patterns found in the data are a simplified representation of the real spatial patterns.
+Thus, the choice of the data type, the scale of the data, and the scale of the analysis are crucial for the results of the analysis.
+
+Traditionally and still the most prominent, computational approaches to spatial patterns in landscape ecology revolve around "patch mosaic" model, where the landscape is represented as a raster data with different categories of land cover [@costanzaDescribingAnalyzingLandscape2019].
+Another approach is the "surface model," where the landscape is represented as continuous raster data with different values of some variable, such as elevation or temperature <!--@costanzaDescribingAnalyzingLandscape2019 for more refs-->.
+Importantly, both these models are based on raster data.
+
+However, the raster data is not the only data type that could be used to represent spatial patterns.
+Another important data type is vector data, which is used to represent point, line, and polygon features.
+Here, the spatial patterns could be related to densities of points, ..., or shapes of polygons.
+Moreover, the vector data could be used to represent the point clouds, which are the data type that is becoming more and more popular with the development of remote sensing technologies (i.e., LiDAR).
+Spatial patterns in point clouds could be related to the shapes of the point clouds or the distances between points.<!--ref-->
+
+<!-- - Resolution, scale, and extent in spatial patterns -->
+
+<!-- The scale dependency of spatial patterns may relate to the scale of the processes, the scale of the data, or the scale of the analysis. -->
+<!-- All of the above aspects could also relate to either the temporal or spatial scale. -->
+
+<!-- [Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022)
 [Where are we now, and where are we going?](https://doi.org/10.1007/s10980-019-00889-6)
+[Pattern metrics](https://doi.org/10.1007/s10980-018-0755-4) -->
 
-@daleSpatialPatternAnalysis2000
-@wiegandUsingPatternOriented2003
+  <!-- - Time series data
+  - One variable vs. multivariable spatial patterns
+  - Categorical vs. continuous spatial patterns -->
 
-- Spatial patterns definitions in the literature (general; Dale 2000)
-- Spatial patterns in many contexts (e.g. geography, biology, urban planning, medical studies) <!-- MH: I would at least cut the medical bit -->
-- Spatial patterns in landscape ecology
-  - Resolution, scale, and extent in spatial patterns
-- Relation of spatial patterns and spatial (ecological?) processes
-- Relation of spatial patterns to spatial processes is a golden grail of landscape ecology
-
-- Spatial patterns in data (computational approach)
-  - Limitations and considerations
-- Data types <!-- MH: Could be moved to the individual sub-sections? -->
-  - Raster data
-  - Vector data: point, line, polygon
-  - Point clouds 
-  - Time series data
-  - One variable vs. multivariable spatial patterns
-  - Categorical vs. continuous spatial patterns
-
-!what with fractals?
-!ai and deep learning still seems to be underused in landscape ecology
+<!-- !what with fractals?
+!ai and deep learning still seems to be underused in landscape ecology -->
 

### 2024-05-31 Sections/spatial-patterns.qmd
adds more data and scale discussion

@@
 bibliography: ../CLER.bib
+format: pdf
 ---
@@
 # Spatial patterns
+
+<!-- [Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022)
+[Where are we now, and where are we going?](https://doi.org/10.1007/s10980-019-00889-6)
+[Pattern metrics](https://doi.org/10.1007/s10980-018-0755-4) -->
 
@@
 
-The scale dependency of spatial patterns may relate to either the temporal or spatial scale.
-Different scales of the processes could lead to different patterns.
-<!-- ... -->
-
 Here, our focus is on the computational approaches to spatial patterns.
@@
 
-Computational approaches to spatial patterns open many new possibilities (as compared to traditional methods). 
+Computational approaches to spatial patterns open many new possibilities as compared to traditional methods. 
 They allow not only to quantify the patterns but also to model them, simulate them, and predict them.
@@
 
+<!-- datatype -->
 Traditionally and still the most prominent, computational approaches to spatial patterns in landscape ecology revolve around "patch mosaic" model, where the landscape is represented as a raster data with different categories of land cover [@costanzaDescribingAnalyzingLandscape2019].
@@
 
-<!-- - Resolution, scale, and extent in spatial patterns -->
+<!-- data sources??+SAR -->
+Currently, the most common data sources in landscape ecology are remote sensing data, either in a direct form (satellite images) or in an indirect form (spectral indices, land cover maps).
+The number of available spatial data products is still increasing, with new sensors, new satellites, and new data processing techniques.
+This trend is seen, for example, with the frequent release of new global land cover maps, such as the ESA WorldCover, Dynamic World, or the Esri LULC.
+<!-- refs -->
+Other examples are the increasing availability of high-resolution data and the growing popularity of LiDAR and SAR data.
 
-<!-- The scale dependency of spatial patterns may relate to the scale of the processes, the scale of the data, or the scale of the analysis. -->
-<!-- All of the above aspects could also relate to either the temporal or spatial scale. -->
+<!-- scale of data -->
+In computational approaches, the issue of scale is closely related to the used data. 
+Here, an important distinction should be made between the scale of the data and the data resolution.
+The data scale refers to the degree of generalization of the data, while the data resolution refers to the smallest unit of the data.
+Thus, it is possible to have data with a high resolution but a small scale, or data with a low resolution but a large scale.
 
-<!-- [Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022)
-[Where are we now, and where are we going?](https://doi.org/10.1007/s10980-019-00889-6)
-[Pattern metrics](https://doi.org/10.1007/s10980-018-0755-4) -->
+<!-- data cats -->
+Similary, it is crucial to understand the thematic resolution of the data, which in landscape ecology often refers to the number and definition of the categories in the data.
+<!-- ... -->
 
-  <!-- - Time series data
-  - One variable vs. multivariable spatial patterns
-  - Categorical vs. continuous spatial patterns -->
+<!-- scale of analysis -->
+Related to the scale of the data is the scale of the analysis.
+It specifies the extent of the study area and the way the data is analyzed.
+Given that the categorical rasters consist of patches, the analysis could be performed at the patch, class, or landscape level. 
+However, it is possible to analyze these levels for a fixed area, using moving windows or buffers around the sampling points.
+
+<!-- time -->
+Early approaches to spatial patterns in landscape ecology were mostly static, focusing on the spatial patterns at a single point in time.
+However, the development of remote sensing technologies and the availability of time series data have allowed for the analysis of spatial patterns over time.
+This opportunity further highlights the importance of understanding of the used data: is it a snapshot of a single moment, a composite of multiple moments, or a temporal generalization?
+The properties of the used data should be in line with the research question and the scale of the analysis.
+
+<!-- data quality -->
+It should be noted that the quality of the data is not constant and the data products are not perfect: they could contain errors and biases.
+For example, the global land cover products usually have an overall accuracy of 70-80%, which means that 20-30% of the pixels are misclassified.
+Moreover, the misclassification is not random, but it is often related to the land cover type, the region, and the time of the year.
+<!-- refs -->
+However, most landscape ecology studies treat the data as a given, without considering its accuracy or uncertainty.
+
+<!-- data projection -->
+Less discussed but often crucial is the spatial reference system of the data.
+Most methods in landscape ecology assume that the data is in a Cartesian coordinate reference system, where the distance between two points is the Euclidean distance.
+This assumption has an impact on measures related to distances and areas.
+However, the data could be in a geographic coordinate reference system, where the distance between two points is the great-circle distance, which would render the results of the analysis incorrect.
+On the other hand, many projected coordinate reference systems exist, each with its own properties -- some distort areas, some distances, some angles, and some distort all of them.
 
@@
 <!-- Bosch, Hesselbarth -->
-[Surface metrics])(https://doi.org/10.1007/s10980-018-0685-1)
+[Surface metrics](https://doi.org/10.1007/s10980-018-0685-1)
 
@@
 - Quantify continuous raster data (gradient surface models)
-- More realistic representation of landscapes
+- More realistic representation of landscapes (?)
 - Based on molecular physics/medicine

### 2024-06-01 Sections/software.qmd
adds software ideas

@@
 <!-- Person responsible: Hesselbarth, Nowosad, Bosch, Jung, Schouten  -->
+
+<!-- "The future applications of landscape pattern metrics will almost certainly depend on the software and analysis tools available" -- @costanzaDescribingAnalyzingLandscape2019 -->
 
@@
 <!-- Hesselbarth -->
+landscapemetrics
+geodiv
+vectormetrics
+bespatial
+motif
+?GLCMTextures
+?rasterdiv
+?terra
 
@@
 <!-- Schouten -->
+
+## Other open source software
+
+GuidosToolbox
+?GRASS GIS
+?QGIS: LecoS plugin

### 2024-06-01 Sections/spatial-patterns.qmd
starts entropy part

@@
 ---
-bibliography: ../CLER.bib
+bibliography: 
+    - ../CLER.bib
+    - ../2024-cler.bib
 format: pdf
@@
 <!-- Person responsible: Nowosad, Bosch, Hesselbarth -->
+<!-- jn: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
 
@@
 <!-- Nowosad, Bosch -->
-intro to entropy
-various concepts:
+<!-- jn: should we use equations here? -->
 
-- information (Shannon)/Shannon diversity
-@Shannon1948
-@nowosadInformationTheoryConsistent2019a
+<!-- information (Shannon)/Shannon diversity -->
+One of the most commonly used metrics in landscape ecology is the Shannon index of diversity (SHDI, @Shannon1948).
+It quantifies the richness and evenness of the categories in the landscape, with higher values indicating higher diversity (more different classes and more even distribution of them).
+SHDI comes from the information theory.
+It is an adaptation of the Shannon entropy, which measures of the uncertainty in a random variable.
 
-- thermodynamic (Boltzmann)
+The SHDI metric is based on the composition of the landscape only, omitting the spatial configuration of the categories.
+<!--composition is the most important -- Ritters-->
+Thus, other measures from the information theory were adapted to landscape ecology to quantify the spatial configuration and the total complexity of the landscape.
+@nowosadInformationTheoryConsistent2019a proposed to treat a categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
+This approach allows to compress the information about the composition and the configuration of the landscape into a co-occurrence matrix, which could be used to calculate various entropy measures.
+It includes the conditional entropy (representing configurational complexity), the joint entropy (representing overall spatio-thematic complexity), the mutual information , and the relative mutual information (representing the degree of spatial autocorrelation).
+<!-- https://r-spatialecology.github.io/landscapemetrics/ -->
 
-@Cushman2016
-@cushmanGeneralizingBoltzmannConfigurational2021
-@gaoComputationBoltzmannEntropy2019
-@zhangCalculatingWassersteinMetricBased2020a
+<!-- thermodynamic (Boltzmann) -->
+The last few years have seen a surge in the development of entropy-based metrics for landscape ecology based on the notions from thermodynamics initatied by the work of @vrankenReviewUseEntropy2015a and @cushmanThermodynamicsLandscapeEcology2015a.
+The overreaching goal of such studies is to link landscape ecology concepts to the thermodynamic principles, such as the second law of thermodynamics.
+They often use the Boltzmann entropy, which is a measure of the disorder in a system, to quantify the complexity of the landscape.
+In late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics, using the concepts of macrostate (general state of a system) and microstate (configuration of the system elements).
+
+@Cushman2016 proposed to relate the edge lenght to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic in raster data.
+Simoultaneously, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface in raster data by transforming the input grid into a series of landscape sufraces with different levels of detail and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
+@zhaoCalculatingSpatialConfigurational2019 extended the original definition of Boltzmann entropy to incorporate the information about the adjancency of the same categories in the landscape mosaic.
+It is also worth noting that later, Cushman refined and generalized this approach for calculations based on raster's surface model (based on mean slope) and point patterns (based on the average distance between points) [@cushmanGeneralizingBoltzmannConfigurational2021].
+<!-- https://jakubnowosad.com/bespatial/ -->
+
+claims of relations between entropy and thermodynamics
 @stepinskiCurbYourEnthusiasm2022
+entropy measures are still related to the underlining data representation (!)
+@zhaoCalculatingSpatialConfigurational2019
+
+<!-- "nstead of Shannon entropy, what should be used is argued to be Boltzmann (1872) entropy, and calls have recently been made to revisit it" @gaoComputationBoltzmannEntropy2019 -->
+
+how they are related to each other
+how they are related to spatial patterns and landscape ecology
 
@@
 
-how they are related to each other
-how they are related to spatial patterns and landscape ecology
+<!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->
 
-entropy measures are still related to the underlining data representation
+<!-- https://cran.r-project.org/web/packages/LandComp/index.html -->
 
-claims of relations between entropy and thermodynamics
-
-<!-- https://r-spatialecology.github.io/landscapemetrics/ -->
-<!-- https://jakubnowosad.com/bespatial/ -->
-<!-- https://cran.r-project.org/web/packages/LandComp/index.html -->
+<!-- unification for various data types, e.g., Cushman, vectormetrics -->
 

### 2024-06-02 Sections/software.qmd
adds new ideas
adds new comments

@@
 <!-- "The future applications of landscape pattern metrics will almost certainly depend on the software and analysis tools available" -- @costanzaDescribingAnalyzingLandscape2019 -->
+
+<!--jn: misc-->
+<!-- the need for open-source software in landscape ecology -->
+<!-- every proposed method should be open-source and available to the community -->
 
@@
 geodiv
-vectormetrics
+vectormetrics (also: VecLI -- they mention open source in their paper a few times, but the software seems to be closed source -- https://urbancomp.net/archives/vecliv3beta)
 bespatial
+LandComp
 motif

### 2024-06-02 Sections/spatial-patterns.qmd
starts vectormetrics section

@@
 <!-- Nowosad, Hesselbarth -->
+<!-- jn: ask T. Matuszek for review of this section later -->
 
-an intro to the section
-what is similar to landscape metrics
+As already mentioned, the initial computaional approaches in landscape ecology were based on raster data.
+However, the decision on the used data model was not theorethical, but rather practical as at the time the most comonnly used software was raster-based, such as GRASS GIS.
+In fact, early studies were based on polygonal vector data that was later converted to raster data for the analysis [@costanzaDescribingAnalyzingLandscape2019]<!--jn: we could also cite actual examples from that paper, but the list of the references would be long-->
+<!-- jn: maybe we could also mention that it is still the case -->
 
-add an info about shape metrics
-@basaranerPerformanceShapeIndices2017
+The spatial vector data model is based on the representation of the Earth's surface as a set of points, lines, and polygons, where each feature is defined by its geometry and attributes.
+Here, we will focus on the polygonal vector data, which is the most common in landscape ecology.
+The polygons are discrete areas, with well-defined boundaries.
+Thus, converting the polygons to raster data (*rasterization*) is a lossy process, where the information about the intricates of the polygon shapes is lost.
+Moreover, the rasterization process largely depends on the decided resolution of the raster outcome.
 
-https://github.com/r-spatialecology/vectormetrics
+Most landscape ecology studies are still based on raster data, due to the familarity with the techiniques, the availability of the data, and the availability of the software.
+For example, most land cover maps (as LULC products are often related to undelining remote sensing products), and all on the global scale, are provided as rasters. 
+At the same time, nowadays, vector data of land cover on regional or local scales is also available and most geospatial software support also vector data.
 
-<!--misc-->
-<!-- the need for open-source software in landscape ecology -->
-<!-- every proposed method should be open-source and available to the community -->
+As hundreds of metrics have been already developed for raster data, the most straightforward approach is to reimplement the same metrics for vector data.
+@yaoVecLIFrameworkCalculating2022 showed that this is possible for most of the metrics, with only several exceptions of metrics (of the shape and aggregation group), which parameters are related to the raster-format structure.
+Their results highlighted that about 80% of the raster and vector-based metrics are highly correlated (for their study area), and the higher the raster resolution, the stronger the correlation is.
+
+<!-- metrics from other fields -->
+<!-- add an info about shape metrics -->
+<!-- @basaranerPerformanceShapeIndices2017 -->
+
+<!-- https://github.com/r-spatialecology/vectormetrics -->
+
+<!-- vectormetrics limitations?? 
+surely, large data performance is one of them
+(e.g., could we do landscapemetrics vs vectormetrics benchmarking?) -->
+
+<!-- polygon vs multipolygon technical distinction -->
 

### 2024-06-03 Sections/spatial-patterns.qmd
adds more vector-metrics text

@@
 
-As already mentioned, the initial computaional approaches in landscape ecology were based on raster data.
+As already mentioned, the initial computational approaches in landscape ecology were based on raster data.
 However, the decision on the used data model was not theorethical, but rather practical as at the time the most comonnly used software was raster-based, such as GRASS GIS.
@@
 
-<!-- metrics from other fields -->
-<!-- add an info about shape metrics -->
-<!-- @basaranerPerformanceShapeIndices2017 -->
+While landscape ecology studies have focussed on using raster data, the vector data model has been widely used in other fields.
+@maceachrenCompactnessGeographicShape1985 presented four categories of compactness measures for vector data: perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes,
+@angelTenCompactnessProperties2010 introduced a unified theoretical foundation for measuring shape compactness.
+They observed that the circle is the most compact of shapes, proposed a set of ten distinct properties of the circle, and then defined metrics associated with each of these properties.
+Vector-based metrics are now popular in urban planning, as they allow to quantify the shapes of urban areas, e.g., @basaranerPerformanceShapeIndices2017a used 18 existing metrics and two new ones to characterize the complexity of building footprints.
 

### 2024-06-03 Sections/spatial-patterns.qmd
adds first draft of the  Operations on spatial patterns section

@@
 
-@nowosadMotifOpensourceTool2021
+Various metrics are often used to describe spatial patterns in landscape ecology.
+<!-- limits, correlations?? -->
+At the same time, the spatial patterns could be analyzed using various operations, such as comparing, searching, or grouping [@turnerSpatialTemporalAnalysis1990;@wickhamMappingAnalyzingLandscape1994; @jasiewiczGeoPATToolboxPatternbased2015; @remmelDistributionsHyperLocalConfiguration2020b;@nowosadMotifOpensourceTool2021].
+The above operations are based on two main ideas: the spatial signatures and the dissimilarity measures.
+A spatial signature is a multi-numerical representation of a landscape pattern for a given area.
+It may be created based on a set of (preferably uncorrelated) landscape metrics, or be derived from the data itself, e.g., by using the co-occurrence matrix [@nowosadPatternbasedIdentificationMapping2021].
+The dissimilarity measure is a function that quantifies the difference between two spatial signatures, with low values indicating similar patterns.
+
+The most basic operation, the comparison of spatial patterns, could be performed for just a pair of areas or the same area in two different moments.
+It can be also extended in a straightforward way by dividing the landscape into multiple smaller areas and comparing them for two moments in time.
+The comparison of spatial patterns is often used to analyze the landscape dynamics, e.g., to detect the changes in the landscape structure over time [@duncanMonitoringSpatialPatterns2023].
+The pattern-based search is a one to many operation, in which spatial signature of a query area is compared to the spatial signatures of all other areas.
+This allows us to find areas with similar patterns to the query area, which could be used to identify areas with similar environmental conditions [@jasiewiczLandscapeSimilarityRetrieval2014].
+Finally, the grouping of spatial patterns is a many to many operation, in which spatial signatures are calculated for all areas and then these areas are clustered based on their similarity [@nowosadPatternbasedIdentificationMapping2021].
+
+<!-- motif -->
+<!-- spquery -->
 

### 2024-06-04 Sections/spatial-patterns.qmd
adds entropy content

@@
 Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. 
-@daleSpatialPatternAnalysis2000 defined them as the scale-dependent predictability of the physical arrangement of observations, while a more general definition by @grimmPatternOrientedModellingPopulation1996 describes them as a characteristic, clearly identifiable structure in nature itself or in the data extracted from nature, indicating an underlying process that generates this pattern.
+@daleSpatialPatternAnalysis2000 defined them as the scale-dependent predictability of the physical arrangement of observations, while a more general definition by @grimmPatternOrientedModellingPopulation1996 describes them as a characteristic, clearly identifiable structure in nature itself or the data extracted from nature, indicating an underlying process that generates this pattern.
 Both definitions highlight several important aspects of spatial patterns: they are related to processes, depending on the scale of observation, and can be found in data.
@@
 The relation of spatial patterns to spatial processes is a golden grail of landscape ecology.
-Patterns observed in nature contain information about the memory of the system, including its demographic processes, dispersal characteristics, and constraining factors[@wiegandUsingPatternOriented2003].
+Patterns observed in nature contain information about a system's memory, including its demographic processes, dispersal characteristics, and constraining factors[@wiegandUsingPatternOriented2003].
 Moreover, spatial patterns are not only a result of processes but could also be a driver of them.
-Thus, untangling the complex history of a landscape from its spatial patterns is a hard challenge.
+Thus, untangling the complex history of a landscape from its spatial patterns is a complex challenge.
 
 Here, our focus is on the computational approaches to spatial patterns.
-Thus, we will discuss the spatial patterns that could be found in data, the metrics that could be used to quantify them, and the operations that could be performed on them.
+Thus, we will discuss the spatial patterns found in data, the metrics that could be used to quantify them, and the operations that could be performed on them.
 
-Computational approaches to spatial patterns open many new possibilities as compared to traditional methods. 
-They allow not only to quantify the patterns but also to model them, simulate them, and predict them.
+Computational approaches to spatial patterns open many new possibilities compared to traditional methods. 
+They allow not only quantifying the patterns but also modeling, simulating, and predicting them.
 Moreover, they allow us to analyze large datasets and apply complex algorithms.
@@
 
-Every data type is a simplified model of the real world, and the spatial patterns found in the data are a simplified representation of the real spatial patterns.
-Thus, the choice of the data type, the scale of the data, and the scale of the analysis are crucial for the results of the analysis.
+Every data type is a simplified model of the natural world, and the spatial patterns found in the data are a simplified representation of the actual spatial patterns.
+Thus, the choice of the data type, the scale of the data, and the scale of the analysis are crucial for the analysis's results.
 
@@
 Another approach is the "surface model," where the landscape is represented as continuous raster data with different values of some variable, such as elevation or temperature <!--@costanzaDescribingAnalyzingLandscape2019 for more refs-->.
-Importantly, both these models are based on raster data.
+Notably, both these models are based on raster data.
 
 However, the raster data is not the only data type that could be used to represent spatial patterns.
-Another important data type is vector data, which is used to represent point, line, and polygon features.
+Another important data type is vector data, representing point, line, and polygon features.
 Here, the spatial patterns could be related to densities of points, ..., or shapes of polygons.
-Moreover, the vector data could be used to represent the point clouds, which are the data type that is becoming more and more popular with the development of remote sensing technologies (i.e., LiDAR).
+Moreover, vector data could be used to represent point clouds, which are data types that are becoming increasingly popular with the development of remote sensing technologies (i.e., LiDAR).
 Spatial patterns in point clouds could be related to the shapes of the point clouds or the distances between points.<!--ref-->
@@
 Currently, the most common data sources in landscape ecology are remote sensing data, either in a direct form (satellite images) or in an indirect form (spectral indices, land cover maps).
-The number of available spatial data products is still increasing, with new sensors, new satellites, and new data processing techniques.
+The number of available spatial data products is still increasing, with new sensors, satellites, and data processing techniques.
 This trend is seen, for example, with the frequent release of new global land cover maps, such as the ESA WorldCover, Dynamic World, or the Esri LULC.
@@
 Here, an important distinction should be made between the scale of the data and the data resolution.
-The data scale refers to the degree of generalization of the data, while the data resolution refers to the smallest unit of the data.
+The data scale refers to the degree of data generalization, while the data resolution refers to the smallest data unit.
 Thus, it is possible to have data with a high resolution but a small scale, or data with a low resolution but a large scale.
@@
 <!-- data cats -->
-Similary, it is crucial to understand the thematic resolution of the data, which in landscape ecology often refers to the number and definition of the categories in the data.
+Similarly, it is crucial to understand the data's thematic resolution, which in landscape ecology often refers to the number and definition of the categories in the data.
 <!-- ... -->
@@
 <!-- scale of analysis -->
-Related to the scale of the data is the scale of the analysis.
+The scale of the data is related to the scale of the analysis.
 It specifies the extent of the study area and the way the data is analyzed.
@@
 <!-- time -->
-Early approaches to spatial patterns in landscape ecology were mostly static, focusing on the spatial patterns at a single point in time.
+Early approaches to spatial patterns in landscape ecology were primarily static, focusing on the spatial patterns at a single point in time.
 However, the development of remote sensing technologies and the availability of time series data have allowed for the analysis of spatial patterns over time.
 This opportunity further highlights the importance of understanding of the used data: is it a snapshot of a single moment, a composite of multiple moments, or a temporal generalization?
-The properties of the used data should be in line with the research question and the scale of the analysis.
+The properties of the used data should align with the research question and the scale of the analysis.
 
 <!-- data quality -->
-It should be noted that the quality of the data is not constant and the data products are not perfect: they could contain errors and biases.
+It should be noted that the quality of the data is not constant, and the data products are not perfect: they could contain errors and biases.
 For example, the global land cover products usually have an overall accuracy of 70-80%, which means that 20-30% of the pixels are misclassified.
@@
 <!-- jn: should we use equations here? -->
+<!-- jn: ask T. Stepinski for review of this section later -->
+
+Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity) and less often unpredictability (temporal heterogeneity) and scale dependence (@vrankenReviewUseEntropy2015a).
+It is derived using various methods, mainly from the information theory and thermodynamics, including Shannon entropy (and its analogous forms, such as Simpson index<!-- Simpson index @simpsonMeasurementDiversity1949 -->
+), Boltzmann entropy, Renyi entropy, and more.
 
 <!-- information (Shannon)/Shannon diversity -->
-One of the most commonly used metrics in landscape ecology is the Shannon index of diversity (SHDI, @Shannon1948).
+One of the most commonly used metrics in landscape ecology is the Shannon diversity index (SHDI, @Shannon1948).
 It quantifies the richness and evenness of the categories in the landscape, with higher values indicating higher diversity (more different classes and more even distribution of them).
@@
 
-The SHDI metric is based on the composition of the landscape only, omitting the spatial configuration of the categories.
+The SHDI metric is based on the composition of the landscape only, omitting the categories' spatial configuration.
 <!--composition is the most important -- Ritters-->
@@
 @nowosadInformationTheoryConsistent2019a proposed to treat a categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
-This approach allows to compress the information about the composition and the configuration of the landscape into a co-occurrence matrix, which could be used to calculate various entropy measures.
-It includes the conditional entropy (representing configurational complexity), the joint entropy (representing overall spatio-thematic complexity), the mutual information , and the relative mutual information (representing the degree of spatial autocorrelation).
+This approach allows us to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
+It includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (representing the degree of spatial autocorrelation).
+Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances, as shown by @claramuntSpatiotemporalFormEntropy2012.
 <!-- https://r-spatialecology.github.io/landscapemetrics/ -->
@@
 <!-- thermodynamic (Boltzmann) -->
-The last few years have seen a surge in the development of entropy-based metrics for landscape ecology based on the notions from thermodynamics initatied by the work of @vrankenReviewUseEntropy2015a and @cushmanThermodynamicsLandscapeEcology2015a.
-The overreaching goal of such studies is to link landscape ecology concepts to the thermodynamic principles, such as the second law of thermodynamics.
-They often use the Boltzmann entropy, which is a measure of the disorder in a system, to quantify the complexity of the landscape.
-In late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics, using the concepts of macrostate (general state of a system) and microstate (configuration of the system elements).
+<!-- "instead of Shannon entropy, what should be used is argued to be Boltzmann (1872) entropy, and calls have recently been made to revisit it" @gaoComputationBoltzmannEntropy2019 -->
+The last few years have seen a surge in the development of entropy-based metrics for landscape ecology based on the notions from thermodynamics initiated by the work of @vrankenReviewUseEntropy2015a and @cushmanThermodynamicsLandscapeEcology2015a.
+The overreaching goal of such studies is to link landscape ecology concepts to thermodynamic principles, such as the second law of thermodynamics.
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics, using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements).
+These studies often use Boltzmann entropy, a measure of a system's disorder, to quantify the landscape's complexity.
 
-@Cushman2016 proposed to relate the edge lenght to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic in raster data.
+@Cushman2016 proposed to relate the edge length to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic in raster data.
+It is also worth noting that later, Cushman refined and generalized this approach for calculations based on raster's surface model (based on mean slope) and point patterns (based on the average distance between points) [@cushmanGeneralizingBoltzmannConfigurational2021].
 Simoultaneously, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface in raster data by transforming the input grid into a series of landscape sufraces with different levels of detail and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
-@zhaoCalculatingSpatialConfigurational2019 extended the original definition of Boltzmann entropy to incorporate the information about the adjancency of the same categories in the landscape mosaic.
-It is also worth noting that later, Cushman refined and generalized this approach for calculations based on raster's surface model (based on mean slope) and point patterns (based on the average distance between points) [@cushmanGeneralizingBoltzmannConfigurational2021].
+@zhaoCalculatingSpatialConfigurational2019 extended the original definition of Boltzmann entropy to incorporate the information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
 <!-- https://jakubnowosad.com/bespatial/ -->
 
-claims of relations between entropy and thermodynamics
-@stepinskiCurbYourEnthusiasm2022
-entropy measures are still related to the underlining data representation (!)
-@zhaoCalculatingSpatialConfigurational2019
+The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of an open discussion.
+@stepinskiCurbYourEnthusiasm2022 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
+He also argues that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
+Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.
 
-<!-- "nstead of Shannon entropy, what should be used is argued to be Boltzmann (1872) entropy, and calls have recently been made to revisit it" @gaoComputationBoltzmannEntropy2019 -->
+Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
+The Renyi [@renyiMeasuresEntropyInformation1961] and Gibbs entropies, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
+Rao quadratic entropy [@raoDiversityDissimilarityCoefficients1982a] has also been used more in recent years, as it measures not only the relative abundances of elements but also the pairwise dissimilarities or distances between them.<!--https://doi.org/10.1016/j.ecolind.2016.07.039-->
+Thus, it can be useful in cases, for example, when the dissimilarities between land cover types, and not only their categories, are relevant.
 
-how they are related to each other
-how they are related to spatial patterns and landscape ecology
+Importantly, the exisiting studies on entropy-based measures clearly show that the outcomes do not only relay on the various formulations of the entropy, but also on the underlining data representation [@zhaoCalculatingSpatialConfigurational2019].
+The same entropy equation could provide different insights when applied just to the composition of the categories, co-occurrence matrix, or on the landscape with multiple spatial scales.
 
-- Kolmogorov, etc.; Juhász-Nagy’s information theory-based functions, Simpson index, Rao index, Pielou, Renyu, ...
-@konradAdaptingMultiscaleApproach2023; 
-@simpsonMeasurementDiversity1949
-
+<!-- Kolmogorov, etc. -->
+<!--@konradAdaptingMultiscaleApproach2023; Juhász-Nagy’s information theory-based functions;  https://cran.r-project.org/web/packages/LandComp/index.html -->
 <!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->
-
-<!-- https://cran.r-project.org/web/packages/LandComp/index.html -->
-
-<!-- unification for various data types, e.g., Cushman, vectormetrics -->
 

### 2024-06-04 Sections/spatial-patterns.qmd
improves lang

@@
 
-As already mentioned, the initial computational approaches in landscape ecology were based on raster data.
-However, the decision on the used data model was not theorethical, but rather practical as at the time the most comonnly used software was raster-based, such as GRASS GIS.
+As mentioned, the initial computational approaches in landscape ecology were based on raster data.
+However, the decision on the data model used was not theoretical but rather practical, as at the time, the most commonly used software was raster-based, such as GRASS GIS.
 In fact, early studies were based on polygonal vector data that was later converted to raster data for the analysis [@costanzaDescribingAnalyzingLandscape2019]<!--jn: we could also cite actual examples from that paper, but the list of the references would be long-->
@@
 The polygons are discrete areas, with well-defined boundaries.
-Thus, converting the polygons to raster data (*rasterization*) is a lossy process, where the information about the intricates of the polygon shapes is lost.
+Thus, converting the polygons to raster data (*rasterization*) is a lossy process, in which information about the intricate details of the polygon shapes is lost.
 Moreover, the rasterization process largely depends on the decided resolution of the raster outcome.
@@
 Most landscape ecology studies are still based on raster data, due to the familarity with the techiniques, the availability of the data, and the availability of the software.
-For example, most land cover maps (as LULC products are often related to undelining remote sensing products), and all on the global scale, are provided as rasters. 
-At the same time, nowadays, vector data of land cover on regional or local scales is also available and most geospatial software support also vector data.
+For example, most land cover maps (as LULC products are often related to underlying remote sensing products), and all on the global scale, are provided as rasters. 
+At the same time, nowadays, vector data of land cover on regional or local scales is also available, and most geospatial software supports vector data.
 
-As hundreds of metrics have been already developed for raster data, the most straightforward approach is to reimplement the same metrics for vector data.
-@yaoVecLIFrameworkCalculating2022 showed that this is possible for most of the metrics, with only several exceptions of metrics (of the shape and aggregation group), which parameters are related to the raster-format structure.
+Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
+@yaoVecLIFrameworkCalculating2022 showed that this is possible for most of the metrics, with only several exceptions of metrics (of the shape and aggregation group), whose parameters are related to the raster-format structure.
 Their results highlighted that about 80% of the raster and vector-based metrics are highly correlated (for their study area), and the higher the raster resolution, the stronger the correlation is.
@@
 The most basic operation, the comparison of spatial patterns, could be performed for just a pair of areas or the same area in two different moments.
-It can be also extended in a straightforward way by dividing the landscape into multiple smaller areas and comparing them for two moments in time.
-The comparison of spatial patterns is often used to analyze the landscape dynamics, e.g., to detect the changes in the landscape structure over time [@duncanMonitoringSpatialPatterns2023].
-The pattern-based search is a one to many operation, in which spatial signature of a query area is compared to the spatial signatures of all other areas.
+It can also be extended straightforwardly by dividing the landscape into multiple smaller areas and comparing them at two moments in time.
+Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@duncanMonitoringSpatialPatterns2023].
+Pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the spatial signatures of all other areas.
 This allows us to find areas with similar patterns to the query area, which could be used to identify areas with similar environmental conditions [@jasiewiczLandscapeSimilarityRetrieval2014].
-Finally, the grouping of spatial patterns is a many to many operation, in which spatial signatures are calculated for all areas and then these areas are clustered based on their similarity [@nowosadPatternbasedIdentificationMapping2021].
+Finally, the grouping of spatial patterns is a many-to-many operation, in which spatial signatures are calculated for all areas, and then these areas are clustered based on their similarity [@nowosadPatternbasedIdentificationMapping2021].
 

### 2024-06-06 Sections/spatial-patterns.qmd
cleans and expands the content

@@
 Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. 
-@daleSpatialPatternAnalysis2000 defined them as the scale-dependent predictability of the physical arrangement of observations, while a more general definition by @grimmPatternOrientedModellingPopulation1996 describes them as a characteristic, clearly identifiable structure in nature itself or the data extracted from nature, indicating an underlying process that generates this pattern.
+@daleSpatialPatternAnalysis2000 defined them as the scale-dependent predictability of the physical arrangement of observations.
+A more general definition by @grimmPatternOrientedModellingPopulation1996 describes patterns as a characteristic, clearly identifiable structure in nature itself or the data extracted from nature, indicating an underlying process that generates this pattern.
 Both definitions highlight several important aspects of spatial patterns: they are related to processes, depending on the scale of observation, and can be found in data.
 
-The relation of spatial patterns to spatial processes is a golden grail of landscape ecology.
-Patterns observed in nature contain information about a system's memory, including its demographic processes, dispersal characteristics, and constraining factors[@wiegandUsingPatternOriented2003].
+Patterns observed in nature contain information about a system's memory, including its demographic processes, dispersal characteristics, and constraining factors [@wiegandUsingPatternOriented2003].
 Moreover, spatial patterns are not only a result of processes but could also be a driver of them.
-Thus, untangling the complex history of a landscape from its spatial patterns is a complex challenge.
+Thus, untangling the complex history of a landscape from its spatial patterns is a a golden grail of landscape ecology.
 
 Here, our focus is on the computational approaches to spatial patterns.
-Thus, we will discuss the spatial patterns found in data, the metrics that could be used to quantify them, and the operations that could be performed on them.
-
-Computational approaches to spatial patterns open many new possibilities compared to traditional methods. 
+Computational approaches to spatial patterns open many new possibilities compared to traditional methods based on visual interpretation.
 They allow not only quantifying the patterns but also modeling, simulating, and predicting them.
@@
 Traditionally and still the most prominent, computational approaches to spatial patterns in landscape ecology revolve around "patch mosaic" model, where the landscape is represented as a raster data with different categories of land cover [@costanzaDescribingAnalyzingLandscape2019].
-Another approach is the "surface model," where the landscape is represented as continuous raster data with different values of some variable, such as elevation or temperature <!--@costanzaDescribingAnalyzingLandscape2019 for more refs-->.
+Another approach is the "surface model" [@mcgarigalGradientConceptLandscape2005a], where the landscape is represented as continuous raster data with different values of some variable, such as elevation or temperature <!--@costanzaDescribingAnalyzingLandscape2019 for more refs-->.
 Notably, both these models are based on raster data.
@@
 However, the raster data is not the only data type that could be used to represent spatial patterns.
-Another important data type is vector data, representing point, line, and polygon features.
-Here, the spatial patterns could be related to densities of points, ..., or shapes of polygons.
+Another noteworthy data type is vector data, representing point, line, and polygon features.
+Here, the spatial patterns could be related to densities of points, orientation of lines, or shapes of polygons.
 Moreover, vector data could be used to represent point clouds, which are data types that are becoming increasingly popular with the development of remote sensing technologies (i.e., LiDAR).
@@
 <!-- data sources??+SAR -->
-Currently, the most common data sources in landscape ecology are remote sensing data, either in a direct form (satellite images) or in an indirect form (spectral indices, land cover maps).
+Currently, the most common data sources in landscape ecology are remote sensing-based data, either in a direct form (satellite images) or in an indirect form (spectral indices, land cover maps).
 The number of available spatial data products is still increasing, with new sensors, satellites, and data processing techniques.
 This trend is seen, for example, with the frequent release of new global land cover maps, such as the ESA WorldCover, Dynamic World, or the Esri LULC.
-<!-- refs -->
-Other examples are the increasing availability of high-resolution data and the growing popularity of LiDAR and SAR data.
+<!-- refs?-->
+Other examples are the increasing availability of high-resolution data (e.g., Sentinel-2) and the growing popularity of LiDAR and SAR data.
 
 <!-- scale of data -->
+<!-- grain -->
 In computational approaches, the issue of scale is closely related to the used data. 
@@
 Similarly, it is crucial to understand the data's thematic resolution, which in landscape ecology often refers to the number and definition of the categories in the data.
+Many classification systems exists that are relevant on a local, regional, or global scale, such as the NLCD, the CORINE, or the FAO systems [@neddSynthesisLandUse2021].
+Their definition of the categories could be different, and the same category could have different meanings in different systems.
 <!-- ... -->
@@
 <!-- scale of analysis -->
+<!-- extent -->
 The scale of the data is related to the scale of the analysis.
@@
 For example, the global land cover products usually have an overall accuracy of 70-80%, which means that 20-30% of the pixels are misclassified.
-Moreover, the misclassification is not random, but it is often related to the land cover type, the region, and the time of the year.
-<!-- refs -->
+Moreover, the misclassification is not random, but it is often related to the land cover type, the region, and the time of the year [@liuFinerResolutionMappingGlobal2021].
 However, most landscape ecology studies treat the data as a given, without considering its accuracy or uncertainty.
@@
 Less discussed but often crucial is the spatial reference system of the data.
-Most methods in landscape ecology assume that the data is in a Cartesian coordinate reference system, where the distance between two points is the Euclidean distance.
-This assumption has an impact on measures related to distances and areas.
-However, the data could be in a geographic coordinate reference system, where the distance between two points is the great-circle distance, which would render the results of the analysis incorrect.
-On the other hand, many projected coordinate reference systems exist, each with its own properties -- some distort areas, some distances, some angles, and some distort all of them.
+Most methods and software tools in landscape ecology assume that the input data is in a Cartesian coordinate reference system, where the distance between two points is the Euclidean distance (most often in meters).
+This assumption a significant impact, especially on measures related to distances and areas.
+However, when the provided data is in a geographic coordinate reference system (with units in degrees), where the distance between two points is the great-circle distance, the results of the analysis could be incorrect.
+At the same time, many projected coordinate reference systems (with units in, e.g., meters) exist, each with its own properties -- some distort areas, some distances, some angles, and some distort all of them.
+Thus, the data may require reprojection before other calculations and the decision on the spatial reference system should be made based on the research question.
 
@@
 
-Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity) and less often unpredictability (temporal heterogeneity) and scale dependence (@vrankenReviewUseEntropy2015a).
-It is derived using various methods, mainly from the information theory and thermodynamics, including Shannon entropy (and its analogous forms, such as Simpson index<!-- Simpson index @simpsonMeasurementDiversity1949 -->
-), Boltzmann entropy, Renyi entropy, and more.
+Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity) and less often unpredictability (temporal heterogeneity) and scale dependence (spatio-temporal heterogeneity) (@vrankenReviewUseEntropy2015a).
+It is derived using various methods, mainly stemmed from the information theory and thermodynamics, including Shannon entropy (and its analogous forms, such as Simpson index<!-- Simpson index @simpsonMeasurementDiversity1949 -->), Boltzmann entropy, Renyi entropy, and more.
 
 <!-- information (Shannon)/Shannon diversity -->
-One of the most commonly used metrics in landscape ecology is the Shannon diversity index (SHDI, @Shannon1948).
-It quantifies the richness and evenness of the categories in the landscape, with higher values indicating higher diversity (more different classes and more even distribution of them).
-SHDI comes from the information theory.
-It is an adaptation of the Shannon entropy, which measures of the uncertainty in a random variable.
-
+The Shannon diversity index (SHDI, @Shannon1948) it quantifies the richness and evenness of the categories in the landscape, with higher values indicating higher diversity (more different classes and more even distribution of them).
+<!-- SHDI comes from the information theory.
+It is an adaptation of the Shannon entropy, which measures of the uncertainty in a random variable. -->
 The SHDI metric is based on the composition of the landscape only, omitting the categories' spatial configuration.
 <!--composition is the most important -- Ritters-->
-Thus, other measures from the information theory were adapted to landscape ecology to quantify the spatial configuration and the total complexity of the landscape.
+However, other measures from the information theory can be adapted to landscape ecology to quantify the spatial configuration and the total complexity of the landscape.
 @nowosadInformationTheoryConsistent2019a proposed to treat a categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
@@
 It includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (representing the degree of spatial autocorrelation).
-Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances, as shown by @claramuntSpatiotemporalFormEntropy2012.
+Original Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances, as shown by @claramuntSpatiotemporalFormEntropy2012.
 <!-- https://r-spatialecology.github.io/landscapemetrics/ -->
@@
 The last few years have seen a surge in the development of entropy-based metrics for landscape ecology based on the notions from thermodynamics initiated by the work of @vrankenReviewUseEntropy2015a and @cushmanThermodynamicsLandscapeEcology2015a.
-The overreaching goal of such studies is to link landscape ecology concepts to thermodynamic principles, such as the second law of thermodynamics.
+The overreaching goal of such studies is to link landscape ecology concepts to thermodynamic principles, mainly the second law of thermodynamics.
 In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics, using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements).
-These studies often use Boltzmann entropy, a measure of a system's disorder, to quantify the landscape's complexity.
+<!-- These studies often use Boltzmann entropy, a measure of a system's disorder, to quantify the landscape's complexity. -->
 
 @Cushman2016 proposed to relate the edge length to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic in raster data.
-It is also worth noting that later, Cushman refined and generalized this approach for calculations based on raster's surface model (based on mean slope) and point patterns (based on the average distance between points) [@cushmanGeneralizingBoltzmannConfigurational2021].
-Simoultaneously, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface in raster data by transforming the input grid into a series of landscape sufraces with different levels of detail and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
-@zhaoCalculatingSpatialConfigurational2019 extended the original definition of Boltzmann entropy to incorporate the information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
+It is also worth noting that later, he refined and generalized this approach for calculations based on raster's surface model (based on mean slope) and point patterns (based on the average distance between points) [@cushmanGeneralizingBoltzmannConfigurational2021].
+Subsequently, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface in raster data by transforming the input grid into a series of landscape sufraces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
+In later attempt, @zhaoCalculatingSpatialConfigurational2019 extended the original definition of Boltzmann entropy to incorporate the information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
 <!-- https://jakubnowosad.com/bespatial/ -->
 
-The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of an open discussion.
+The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
 @stepinskiCurbYourEnthusiasm2022 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
@@
 Thus, it can be useful in cases, for example, when the dissimilarities between land cover types, and not only their categories, are relevant.
+<!-- more -->
 
-Importantly, the exisiting studies on entropy-based measures clearly show that the outcomes do not only relay on the various formulations of the entropy, but also on the underlining data representation [@zhaoCalculatingSpatialConfigurational2019].
-The same entropy equation could provide different insights when applied just to the composition of the categories, co-occurrence matrix, or on the landscape with multiple spatial scales.
+Importantly, the exisiting studies on entropy-based measures clearly show that the outcomes do not only depend on the various formulations of the entropy, but also on the underlining data representation [@zhaoCalculatingSpatialConfigurational2019].
+The same entropy equation could provide different insights when applied just to the composition of the categories, co-occurrence matrix, or on the landscape transformed into a series of data with different levels of detail.
+Thus, more work in this area is needed to understand the relations between the entropy measures and the landscape patterns they represent.
 
@@
 
-As mentioned, the initial computational approaches in landscape ecology were based on raster data.
-However, the decision on the data model used was not theoretical but rather practical, as at the time, the most commonly used software was raster-based, such as GRASS GIS.
-In fact, early studies were based on polygonal vector data that was later converted to raster data for the analysis [@costanzaDescribingAnalyzingLandscape2019]<!--jn: we could also cite actual examples from that paper, but the list of the references would be long-->
-<!-- jn: maybe we could also mention that it is still the case -->
-
 The spatial vector data model is based on the representation of the Earth's surface as a set of points, lines, and polygons, where each feature is defined by its geometry and attributes.
@@
 
-Most landscape ecology studies are still based on raster data, due to the familarity with the techiniques, the availability of the data, and the availability of the software.
-For example, most land cover maps (as LULC products are often related to underlying remote sensing products), and all on the global scale, are provided as rasters. 
+As mentioned, the initial computational approaches in landscape ecology were based on raster data.
+However, the decision on the data model used was not theoretical but rather practical, as at the time, the most commonly used related software was raster-based, such as GRASS GIS.
+In fact, early studies were based on polygonal vector data that was later converted to raster data for the analysis [@costanzaDescribingAnalyzingLandscape2019].<!--jn: we could also cite actual examples from that paper, but the list of the references would be long-->
+<!-- jn: maybe we could also mention that it is still the case -->
+
+Most landscape ecology studies are still based on raster data, probably due to the familarity with the techiniques, the availability of the data, and the availability of the software.
+For example, most land cover maps (as LULC products are often related to underlying remote sensing products), including all on the global scale, are provided as rasters. 
 At the same time, nowadays, vector data of land cover on regional or local scales is also available, and most geospatial software supports vector data.
@@
 Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
-@yaoVecLIFrameworkCalculating2022 showed that this is possible for most of the metrics, with only several exceptions of metrics (of the shape and aggregation group), whose parameters are related to the raster-format structure.
+@yaoVecLIFrameworkCalculating2022 showed that this is possible for most of the metrics, with only several exceptions (of the shape and aggregation group), whose parameters are related to the raster-format structure.
 Their results highlighted that about 80% of the raster and vector-based metrics are highly correlated (for their study area), and the higher the raster resolution, the stronger the correlation is.
 
-While landscape ecology studies have focussed on using raster data, the vector data model has been widely used in other fields.
-@maceachrenCompactnessGeographicShape1985 presented four categories of compactness measures for vector data: perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes,
+While landscape ecology studies have focussed on using raster data, the vector data model has been widely used in other fields, such as geography and urban studies.
+@maceachrenCompactnessGeographicShape1985 presented four categories of compactness measures for vector data: perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes.
 @angelTenCompactnessProperties2010 introduced a unified theoretical foundation for measuring shape compactness.
 They observed that the circle is the most compact of shapes, proposed a set of ten distinct properties of the circle, and then defined metrics associated with each of these properties.
-Vector-based metrics are now popular in urban planning, as they allow to quantify the shapes of urban areas, e.g., @basaranerPerformanceShapeIndices2017a used 18 existing metrics and two new ones to characterize the complexity of building footprints.
+Vector-based metrics are now popular, for example, in urban planning, as they allow to quantify the shapes of urban areas, e.g., @basaranerPerformanceShapeIndices2017a used 18 existing metrics and two new ones to characterize the complexity of building footprints.
+<!-- more -->
+
+The vector-based metrics also have their limitations.
+The most important is the computational complexity, which makes the calculation of vector metrics slower than rasters ones, especially for large datasets.
+Another technical issue is the need for the data to be topologically correct, which is not always the case, especially for data from different sources.
+Vector-based metrics are also much less understood than raster-based ones, and their relations to landscape patterns still need to be explored.
+<!-- (e.g., could we do landscapemetrics vs vectormetrics benchmarking?) -->
+<!-- polygon vs multipolygon technical distinction -->
 
 <!-- https://github.com/r-spatialecology/vectormetrics -->
-
-<!-- vectormetrics limitations?? 
-surely, large data performance is one of them
-(e.g., could we do landscapemetrics vs vectormetrics benchmarking?) -->
-
-<!-- polygon vs multipolygon technical distinction -->
 
@@
 <!-- limits, correlations?? -->
-At the same time, the spatial patterns could be analyzed using various operations, such as comparing, searching, or grouping [@turnerSpatialTemporalAnalysis1990;@wickhamMappingAnalyzingLandscape1994; @jasiewiczGeoPATToolboxPatternbased2015; @remmelDistributionsHyperLocalConfiguration2020b;@nowosadMotifOpensourceTool2021].
+These spatial patterns can then be further analyzed and explored through a range of computational operations, such as comparing, searching, or grouping [@turnerSpatialTemporalAnalysis1990;@wickhamMappingAnalyzingLandscape1994; @jasiewiczGeoPATToolboxPatternbased2015; @remmelDistributionsHyperLocalConfiguration2020b;@nowosadMotifOpensourceTool2021].
 The above operations are based on two main ideas: the spatial signatures and the dissimilarity measures.
@@
 
-The most basic operation, the comparison of spatial patterns, could be performed for just a pair of areas or the same area in two different moments.
+The most basic operation, the comparison of spatial patterns, can be performed for just a pair of areas or the same area in two different moments.
 It can also be extended straightforwardly by dividing the landscape into multiple smaller areas and comparing them at two moments in time.
@@
 
+<!-- one/two par about other approaches -->
+
 <!-- motif -->

### 2024-06-07 Sections/software.qmd
adds small improvements

--- 2024-06-01 Sections/software.qmd
@@
 <!-- Hesselbarth -->
-landscapemetrics
+<!-- landscapemetrics
 geodiv
-vectormetrics
+vectormetrics (also: VecLI -- they mention open source in their paper a few times, but the software seems to be closed source -- https://urbancomp.net/archives/vecliv3beta)
 bespatial
+LandComp
 motif
@@
 ?rasterdiv
-?terra
+?terra -->
 
@@
 
-GuidosToolbox
+<!-- GuidosToolbox
 ?GRASS GIS
-?QGIS: LecoS plugin
+?QGIS: LecoS plugin -->

### 2024-06-07 Sections/spatial-patterns.qmd
adds small improvements

@@
 Thus, it can be useful in cases, for example, when the dissimilarities between land cover types, and not only their categories, are relevant.
-<!-- more -->
+Another recent developement is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of the difference between two probability distributions, to describe patterns across scales [@huckebaMultiscaleSpatialEcology2023].
 
@@
 <!-- one/two par about other approaches -->
+<!-- e.g., modeling, simulation, prediction, etc. -->
 

### 2024-06-08 Sections/spatial-patterns.qmd
adds new ref

@@
     - ../2024-cler.bib
-format: pdf
+format: docx
 ---
@@
 The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
-@stepinskiCurbYourEnthusiasm2022 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
+@stepinskiCurbYourEnthusiasm2022 and @huckebaMultiscaleSpatialEcology2023 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
 He also argues that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.

### 2024-06-19 Manuscript.tex
Some updates on intro and spatial-patterns

@@
 will discuss potential future needs developments of computational
-methods in landscape ecology..
+methods in landscape ecology.
+
+\section{Spatial patterns}\label{spatial-patterns}
+
+Spatial patterns can be defined as the scale-dependent predictability of
+the physical arrangement of observations (Dale 1999), or as clearly
+identifiable structures in nature itself or data extracted from nature
+(Grimm et al. 1996). Importantly, patterns observed in nature contain
+information about the history of the system, such as demographic
+processes, dispersal characteristics, or climatic patterns (Wiegand et
+al. 2003). However, spatial patterns are not only a result of processes,
+but could also be a driver of them. Thus, untangling the history of a
+landscape from its spatial patterns is one of the core concepts of
+landscape ecology.
+
+In order to analyze spatial patterns they must be translated into
+simplified and quantifiable representations of the natural world. Thus,
+the choice of the data types, spatial and temporal scales, or the
+thematic resolution are crucial for the analysis. Potential data types
+include raster or vector data to represent landscape characteristics
+using grid cells or points, lines and polygons, respectively. For the
+spatial scale the most important features are the resolution and extent.
+The increasing availability of temporal time series data allows analysis
+of spatial patterns over time, introducing an additional temporal scale.
+For the thematic resolution, which often refers to the number or
+definitions of the categories (e.g, LULC classes), many classification
+systems exists, such as the NLCD, the CORINE, or the FAO systems (Nedd
+et al. 2021).
+
+\subsection{Landscape metrics}\label{landscape-metrics}
+
+Traditionally and still the most prominent, computational approaches to
+spatial patterns in landscape ecology revolve around the patch mosaic
+model, where the landscape is represented as raster data using
+categorical raster data (Costanza et al. 2019).
+
+\begin{itemize}
+\tightlist
+\item
+  Quantify composition and configuration
+\item
+  Widely applied since FRAGSTATS release
+\item
+  Several alternative tools available and still widely used
+\end{itemize}
+
+\subsection{Surface metrics}\label{surface-metrics}
+
+\href{https://doi.org/10.1007/s10980-018-0685-1}{Surface metrics}
+
+\begin{itemize}
+\tightlist
+\item
+  More recent development
+\item
+  Quantify continuous raster data (gradient surface models)
+\item
+  More realistic representation of landscapes (?)
+\item
+  Based on molecular physics/medicine
+\item
+  Not widely adopted yet (due to missing tools?)
+\end{itemize}
+
+\subsection{Entropy}\label{entropy}
+
+Entropy in landscape ecology is mainly used to quantify the complexity
+of the landscape (spatial heterogeneity), and less often
+unpredictability (temporal heterogeneity) and scale dependence
+(spatio-temporal heterogeneity; (Vranken et al. 2015)). It is mainly
+derived from information theory and thermodynamics, including besides
+others Shannon entropy (and its analogous forms), Boltzmann entropy, or
+Renyi entropy. However, existing studies show that insights do not only
+depend on the formulation of the used entropy, but also on the
+underlining data representation, for example the composition of
+categories or the co-occurrence matrix (Zhao and Zhang 2019).
+
+The Shannon diversity index (Shannon 1948) quantifies the richness and
+evenness of categories in the landscape. Shannon's entropy formula can
+also be modified to include the landscape's spatial configuration, e.g.,
+by weights calculated from intra- and inter-class distances (Claramunt
+2012). However, to quantify the spatial configuration and the total
+complexity of the landscape, other measures from information theory can
+be adapted. Nowosad and Stepinski (2019) proposed to treat categorical
+raster as a bivariate random variable \((x,y)\) where \(x\) is the
+category of the focal cell and \(y\) is the category of the neighboring
+cell. This approach allows us to compress information about the
+landscape's composition and configuration into a co-occurrence matrix,
+which could be used to calculate various entropy measures. This includes
+conditional entropy (representing configurational complexity), joint
+entropy (representing overall spatial-thematic complexity), mutual
+information, and relative mutual information (representing the degree of
+spatial autocorrelation).
+
+The last few years have seen a surge in the development of entropy-based
+metrics for landscape ecology based on notions from thermodynamics
+initiated by the work of Vranken et al. (2015) and Cushman (2015). In
+the late 1800s, Boltzmann formulated a probabilistic interpretation of
+the second law of thermodynamics, using the concepts of macrostate (the
+general state of a system) and microstate (the configuration of the
+system elements). Cushman (2016) proposed to relate the edge length to
+the microstate of the landscape, and use the proportion of microstates
+to compute the relative Boltzmann entropy of a landscape mosaic. This
+approach was generalized for calculations based on raster's surface
+model and point patterns (Cushman 2021). Subsequently, Gao et al. (2017)
+proposed to use the Boltzmann entropy to quantify the complexity of a
+landscape surface by transforming the input raster into a series of
+landscape surfaces with different levels of detail (microstate) and
+calculating the Boltzmann entropy based on the total number of
+microstates that are able to generate the observed macrostate. Recently,
+Zhang et al. (2020) extended the original definition of Boltzmann
+entropy to incorporate the information about the adjacency of the same
+categories in the landscape mosaic by using the number of contiguous
+patches of the same category.
+
+Moreover, many other entropy-based metrics have been proposed for use in
+landscape ecology. The Renyi (Rényi 1961) and Gibbs entropies, which are
+both generalizations of the Shannon entropy, are applied to quantify the
+complexity of the landscape. The Rao quadratic entropy (Rao 1982) has
+also been applied recently (Rocchini et al. 2017), as it measures not
+only the relative abundances of elements, but also the pairwise
+dissimilarities or distances between them. Thus, it can be useful in
+cases, for example, when the dissimilarities between land cover types,
+and not only their categories, are relevant. Another recent development
+is the use of Kullback-Leibler divergence (also known as relative
+entropy), which is a measure of differences between two probability
+distributions to describe patterns across scales (Huckeba et al. 2023).
+
+\subsection{Landscape mosaic}\label{landscape-mosaic}
+
+\subsection{Vector metrics}\label{vector-metrics}
+
+Many landscape ecology studies are based on raster data. However, the
+decision on the data model is often driven by the availability of the
+data and software, or the familiarity with the approach (Costanza et al.
+2019). For example, several early landscape ecology studies were based
+on polygonal vector data that was converted to raster data . Yet,
+nowadays, vector data is increasingly available, and most geospatial
+software supports it.
+
+Hundreds of metrics have already been developed for raster data, so the
+most straightforward approach is to reimplement them for vector data.
+This is possible for most of the metrics, with only several exceptions
+of metrics that are specifically related to raster data (Yao et al.
+2022). Yet, the vector data model has been widely used in other fields,
+such as geography or urban studies. For example, in urban planning,
+vector metrics are applied to quantify the shapes of urban areas and
+characterize the complexity of building footprints (Basaraner and
+Cetinkaya 2017). Earlier approaches included four categories of
+compactness measures for vector data, namely perimeter-area, single
+parameters of related circles, dispersion of elements of the area around
+a centroid, and direct comparison to standard shapes (Maceachren 1985).
+More recently, a unified theoretical foundation for measuring shape
+compactness was introduced using a set of ten distinct properties of a
+circle and metrics associated with each of these properties (Angel et
+al. 2010)
+
+Vector-based metrics also have their limitations. The most important is
+the computational complexity, which makes the calculation of vector
+metrics slower than their raster equivalents. Another technical issue is
+the need for the data to be topologically correct, which is not always
+the case, especially for data from different sources. Last, the
+pattern-process link is still less explored for vector-based metrics
+compared to raster-based metrics.
+
+\subsection{Operations on spatial
+patterns}\label{operations-on-spatial-patterns}
+
+Spatial patterns can be analyzed through a range of computational
+operations, such as comparing, searching, or grouping (Turner 1990;
+Wickham and Norton 1994; Jasiewicz et al. 2015; Remmel 2020; Nowosad
+2021). These operations are based on spatial signatures which are
+multi-numerical representations of landscape pattern, and dissimilarity
+measures which are functions that quantify differences between the
+signatures.
+
+Spatial patterns can be compared between different areas, or for the
+same area at two different moments in time. Comparing spatial patterns
+is often used to analyze landscape dynamics, e.g., to detect changes in
+landscape structure over time (Duncan and Boruff 2023). Furthermore,
+pattern-based search is a one-to-many operation in which the spatial
+signature of a query area is compared to the signatures of all other
+areas. This allows to find areas with similar patterns compared to the
+query area, e.g., areas with similar environmental conditions (Jasiewicz
+et al. 2014). Finally, the grouping of spatial patterns is a
+many-to-many operation in which spatial signatures are calculated for
+all areas and clustered based on their similarity (Nowosad and Stepinski
+2021).
 
@@
 imaginable manipulations in order to advance theoretical developments or
-test hypotheses (Peck 2004). Due to the general temporal and spatial
+test hypotheses (Peck 2004). Due to the usual temporal and spatial
 scales, complex interactions and feedbacks, or scale mismatches between
@@
 autocorrelation, or spatial aggregation concepts to simulate more
-realistic landscapes (Wang and Malanson 2008). In order to make multiple
-neutral landscape model algorithms available several software libraries
-are available in prominent programming languages including Python
-{[}NLMpy; Etherington et al. (2015){]} or R {[}NLMR; Sciaini et al.
-(2018){]}.
+realistic landscapes (Wang and Malanson 2008).
+
+doi:10.12942/lrlr-2014-2 Perlin noise (Etherington 2022) Binary space
+partitioning {[}Etherington2022a{]} Generating in comparison to real
+landscapes https://doi.org/10.1016/j.ecoinf.2017.08.003
+https://doi.org/10.1016/j.ecolmodel.2018.12.010
+
+In order to make multiple neutral landscape model algorithms accessible,
+several software libraries are available in prominent programming
+languages including Python {[}NLMpy; Etherington et al. (2015){]}, R
+{[}NLMR; Sciaini et al. (2018){]}, or Julia
+(https://github.com/EcoJulia/NeutralLandscapes.jl).
 
@@
 \href{https://doi.org/10.1002/ece3.2145}{LG}
+https://journals.plos.org/plosone/article/file?id=10.1371/journal.pone.0222949\&type=printable
 
@@
 \begin{CSLReferences}{1}{1}
+\bibitem[\citeproctext]{ref-Angel2010}
+Angel S, Parent J, Civco DL (2010) Ten compactness properties of
+circles: Measuring shape in geography. Canadian Geographies /
+G{é}ographies canadiennes 54:441--461.
+\url{https://doi.org/10.1111/j.1541-0064.2009.00304.x}
+
+\bibitem[\citeproctext]{ref-Basaraner2017}
+Basaraner M, Cetinkaya S (2017) Performance of shape indices and
+classification schemes for characterising perceptual shape complexity of
+building footprints in GIS. International Journal of Geographical
+Information Science 31:1952--1977.
+\url{https://doi.org/10.1080/13658816.2017.1346257}
+
 \bibitem[\citeproctext]{ref-Chi2016}
@@
 104:2207--2219. \url{https://doi.org/10.1109/JPROC.2016.2598228}
+
+\bibitem[\citeproctext]{ref-Claramunt2012}
+Claramunt C (2012)
+\href{https://doi.org/10.1007/978-3-642-33999-8_28}{Towards a
+spatio-temporal form of entropy}. In: Advances in conceptual modeling.
+Springer Berlin Heidelberg, Berlin, Heidelberg, pp 221--230
+
+\bibitem[\citeproctext]{ref-Costanza2019}
+Costanza JK, Riitters K, Vogt P, Wickham J (2019) Describing and
+analyzing landscape patterns: Where are we now, and where are we going?
+Landscape Ecology 34:2049--2055.
+\url{https://doi.org/10.1007/s10980-019-00889-6}
+
+\bibitem[\citeproctext]{ref-Cushman2015}
+Cushman SA (2015) Thermodynamics in landscape ecology: The importance of
+integrating measurement and modeling of landscape entropy. Landscape
+Ecology 30:7--10. \url{https://doi.org/10.1007/s10980-014-0108-x}
+
+\bibitem[\citeproctext]{ref-Cushman2016}
+Cushman SA (2016) Calculating the configurational entropy of a landscape
+mosaic. Landscape Ecology 31:481--489.
+\url{https://doi.org/10.1007/s10980-015-0305-2}
+
+\bibitem[\citeproctext]{ref-Cushman2021a}
+Cushman SA (2021) Generalizing boltzmann configurational entropy to
+surfaces, point patterns and landscape mosaics. Entropy 23:1616.
+\url{https://doi.org/10.3390/e23121616}
+
+\bibitem[\citeproctext]{ref-Dale1999}
+Dale MRT (1999) \href{https://doi.org/10.1017/CBO9780511612589}{Spatial
+pattern analysis in plant ecology}, 1st edn. Cambridge University Press
+
+\bibitem[\citeproctext]{ref-Duncan2023}
+Duncan JMA, Boruff B (2023) Monitoring spatial patterns of urban
+vegetation: A comparison of contemporary high-resolution datasets.
+Landscape and Urban Planning 233:104671.
+\url{https://doi.org/10.1016/j.landurbplan.2022.104671}
+
+\bibitem[\citeproctext]{ref-Etherington2022}
+Etherington TR (2022) Perlin noise as a hierarchical neutral landscape
+model. Web Ecology 22:1--6. \url{https://doi.org/10.5194/we-22-1-2022}
 
@@
 
+\bibitem[\citeproctext]{ref-Gao2017}
+Gao P, Zhang H, Li Z (2017) A hierarchy-based solution to calculate the
+configurational entropy of landscape gradients. Landscape Ecology.
+\url{https://doi.org/10.1007/s10980-017-0515-x}
+
 \bibitem[\citeproctext]{ref-Gergel2017}
@@
 
+\bibitem[\citeproctext]{ref-Grimm1996}
+Grimm V, Frank K, Jeltsch F, et al (1996) Pattern-oriented modelling in
+population ecology. The Science of the Total Environment 183:151--166.
+\url{https://doi.org/10.1016/0048-9697(95)04966-5}
+
+\bibitem[\citeproctext]{ref-Huckeba2023}
+Huckeba G, Andresen B, Roach TNF (2023) Multi-scale spatial ecology
+analyses: A kullback information approach. Landscape Ecology
+38:645--657. \url{https://doi.org/10.1007/s10980-022-01514-9}
+
 \bibitem[\citeproctext]{ref-Jaric2020}
@@
 Evolution 35:630--639. \url{https://doi.org/10.1016/j.tree.2020.03.003}
+
+\bibitem[\citeproctext]{ref-Jasiewicz2015}
+Jasiewicz J, Netzel P, Stepinski T (2015) GeoPAT: A toolbox for
+pattern-based information retrieval from large geospatial databases.
+Computers \& Geosciences 80:62--73.
+\url{https://doi.org/10.1016/j.cageo.2015.04.002}
+
+\bibitem[\citeproctext]{ref-Jasiewicz2014}
+Jasiewicz J, Netzel P, Stepinski TF (2014) Landscape similarity,
+retrieval, and machine mapping of physiographic units. Geomorphology
+221:104--112. \url{https://doi.org/10.1016/j.geomorph.2014.06.011}
 
@@
 
+\bibitem[\citeproctext]{ref-Maceachren1985}
+Maceachren AM (1985) Compactness of geographic shape: Comparison and
+evaluation of measures. Geografiska Annaler: Series B, Human Geography
+67:53--67. \url{https://doi.org/10.1080/04353684.1985.11879515}
+
 \bibitem[\citeproctext]{ref-Nathan2022}
@@
 
+\bibitem[\citeproctext]{ref-Nedd2021}
+Nedd R, Light K, Owens M, et al (2021) A synthesis of land use/land
+cover studies: Definitions, classification systems, meta-studies,
+challenges and knowledge gaps on a global landscape. Land 10:994.
+\url{https://doi.org/10.3390/land10090994}
+
 \bibitem[\citeproctext]{ref-Newman2019}
@@
 \url{https://doi.org/10.3389/fevo.2019.00293}
+
+\bibitem[\citeproctext]{ref-Nowosad2021a}
+Nowosad J (2021) Motif: An open-source r tool for pattern-based spatial
+analysis. Landscape Ecology 36:29--43.
+\url{https://doi.org/10.1007/s10980-020-01135-0}
+
+\bibitem[\citeproctext]{ref-Nowosad2019a}
+Nowosad J, Stepinski TF (2019) Information theory as a consistent
+framework for quantification and classification of landscape patterns.
+Landscape Ecology 34:2091--2101.
+\url{https://doi.org/10.1007/s10980-019-00830-x}
+
+\bibitem[\citeproctext]{ref-Nowosad2021b}
+Nowosad J, Stepinski TF (2021) Pattern-based identification and mapping
+of landscape types using multi-thematic data. International Journal of
+Geographical Information Science 35:1634--1649.
+\url{https://doi.org/10.1080/13658816.2021.1893324}
 
@@
 
+\bibitem[\citeproctext]{ref-Rao1982}
+Rao CR (1982) Diversity and dissimilarity coefficients: A unified
+approach. Theoretical Population Biology 21:24--43.
+\url{https://doi.org/10.1016/0040-5809(82)90004-1}
+
+\bibitem[\citeproctext]{ref-Remmel2020}
+Remmel TK (2020) Distributions of hyper-local configuration elements to
+characterize, compare, and assess landscape-level spatial patterns.
+Entropy 22:420. \url{https://doi.org/10.3390/e22040420}
+
+\bibitem[\citeproctext]{ref-Renyi1961}
+Rényi A (1961) On measures of entropy and information. In: Proceedings
+of the fourth berkeley symposium on mathematical statistics and
+probability, volume 1: Contributions to the theory of statistics.
+University of California Press, pp 547--562
+
 \bibitem[\citeproctext]{ref-Risser1984}
@@
 approaches. Illinois Natural History Survery Special Publication 2:7--14
+
+\bibitem[\citeproctext]{ref-Rocchini2017}
+Rocchini D, Marcantonio M, Ricotta C (2017) Measuring rao's q diversity
+index from remote sensing: An open source solution. Ecological
+Indicators 72:234--238.
+\url{https://doi.org/10.1016/j.ecolind.2016.07.039}
 
@@
 
+\bibitem[\citeproctext]{ref-Shannon1948}
+Shannon CE (1948) A mathematical theory of communication. Bell System
+Technical Journal 27:379--423.
+\url{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}
+
 \bibitem[\citeproctext]{ref-Synes2016}
@@
 \url{https://doi.org/10.1146/annurev.es.20.110189.001131}
+
+\bibitem[\citeproctext]{ref-Turner1990}
+Turner MG (1990) Spatial and temporal analysis of landscape patterns.
+Landscape Ecology 4:21--30. \url{https://doi.org/10.1007/BF02573948}
 
@@
 
+\bibitem[\citeproctext]{ref-Vranken2015}
+Vranken I, Baudry J, Aubinet M, et al (2015) A review on the use of
+entropy in landscape ecology: Heterogeneity, unpredictability, scale
+dependence and their links with thermodynamics. Landscape Ecology
+30:51--65. \url{https://doi.org/10.1007/s10980-014-0105-0}
+
 \bibitem[\citeproctext]{ref-Wang2008}
@@
 \url{https://doi.org/10.1111/j.1749-8198.2008.00090.x}
+
+\bibitem[\citeproctext]{ref-Wickham1994}
+Wickham JD, Norton DJ (1994) Mapping and analyzing landscape patterns.
+Landscape Ecology 9:7--23. \url{https://doi.org/10.1007/BF00135075}
+
+\bibitem[\citeproctext]{ref-Wiegand2003}
+Wiegand T, Jeltsch F, Hanski I, Grimm V (2003) Using pattern-oriented
+modeling for revealing hidden information: A key for reconciling
+ecological theory and application. Oikos 100:209--222.
+\url{https://doi.org/10.1034/j.1600-0706.2003.12027.x}
 
@@
 
+\bibitem[\citeproctext]{ref-Yao2022}
+Yao Y, Cheng T, Sun Z, et al (2022) VecLI: A framework for calculating
+vector landscape indices considering landscape fragmentation.
+Environmental Modelling and Software 149:105325.
+\url{https://doi.org/10.1016/j.envsoft.2022.105325}
+
+\bibitem[\citeproctext]{ref-Zhang2020}
+Zhang H, Wu Z, Lan T, et al (2020) Calculating the wasserstein
+metric-based boltzmann entropy of a landscape mosaic. Entropy 22:381.
+\url{https://doi.org/10.3390/e22040381}
+
+\bibitem[\citeproctext]{ref-Zhao2019}
+Zhao Y, Zhang X (2019) Calculating spatial configurational entropy of a
+landscape mosaic based on the wasserstein metric. Landscape Ecology
+34:1849--1858. \url{https://doi.org/10.1007/s10980-019-00876-x}
+
 \end{CSLReferences}

### 2024-06-19 Sections/introduction.qmd
Some updates on intro and spatial-patterns

@@
 
-Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
-Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
-Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].
+Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986;@Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
+Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984;@Turner1989;@Turner2005].
+Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002;@Wu2013].
 
@@
 Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
-Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012; but see @Wiersma2022].
-Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997; @With2019], the general increase of available data [@Chi2016; @Jaric2020; @Nathan2022; @Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
-Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].
+Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012;but see @Wiersma2022].
+Because landscape ecology is a cross-disciplinary field including, e.g., social sciences, geography, or ecology and evolution [@Wiens1997;@With2019], the general increase of available data [@Chi2016;@Jaric2020;@Nathan2022;@Wuest2020], or the complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
+Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012;@Wilson2014].
+
+For all computational methods in landscape ecology the natural world must translated into a data model.
+The spatial vector data model is based on the representation of the natural world as a set of points, lines, and polygons, where each feature is defined by its geometry and attributes.
+Contrastingly, the raster data model is based on grid cells representing features of the natural worlds.
+<!-- MH: Flesh out a bit also using PPM and gradient model -->
 
@@
 However, due to its many advantages, we will focus on open-source software [@vonKrogh2006], and specifically scripting languages, i.e., R, Python, and Julia.
-Last, we will discuss potential future needs developments of computational methods in landscape ecology..
+Last, we will discuss potential future needs developments of computational methods in landscape ecology.

### 2024-06-19 Sections/simulations.qmd
Some updates on intro and spatial-patterns

@@
 # Simulation models
-<!-- Hesselbarth -->
 
@@
 Thus, they can be seen as experimental systems that allow, unlike to nature, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004].
-Due to the general temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006; @Synes2016].
+Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006; @Synes2016].
 In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004;@Synes2016].
@@
 ### Neutral landscapes models
+
 Neutral landscapes models simulate landscape patterns without assuming any underlying abiotic or biotic processes.
@@
 Last, more complex neutral models were developed that are based on contagion, autocorrelation, or spatial aggregation concepts to simulate more realistic landscapes [@Wang2008].
-In order to make multiple neutral landscape model algorithms available several software libraries are available in prominent programming languages including Python [NLMpy; @Etherington2015] or R [NLMR; @Sciaini2018].
+
+doi:10.12942/lrlr-2014-2
+Perlin noise [@Etherington2022]
+Binary space partitioning [Etherington2022a]
+Generating in comparison to real landscapes https://doi.org/10.1016/j.ecoinf.2017.08.003
+https://doi.org/10.1016/j.ecolmodel.2018.12.010
+
+In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy; @Etherington2015], R [NLMR; @Sciaini2018], or Julia (https://github.com/EcoJulia/NeutralLandscapes.jl).
 
@@
 [LG](https://doi.org/10.1002/ece3.2145)
+https://journals.plos.org/plosone/article/file?id=10.1371/journal.pone.0222949&type=printable
 

### 2024-06-19 Sections/spatial-patterns.qmd
Some updates on intro and spatial-patterns

--- 2024-06-06 Sections/spatial-patterns.qmd
@@
 ---
-bibliography: 
-    - ../CLER.bib
-    - ../2024-cler.bib
-format: pdf
+bibliography: CLER.bib
 ---
@@
 <!-- Person responsible: Nowosad, Bosch, Hesselbarth -->
+
 <!-- jn: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
+<!-- MH: I think we decided to not reference software here but in a seperate section -->
 
@@
 
-<!-- [Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022)
-[Where are we now, and where are we going?](https://doi.org/10.1007/s10980-019-00889-6)
-[Pattern metrics](https://doi.org/10.1007/s10980-018-0755-4) -->
+<!-- [Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022) -->
+<!-- [Pattern metrics](https://doi.org/10.1007/s10980-018-0755-4) -->
+<!--[Emerging trajectories for spatial pattern](https://link.springer.com/article/10.1007/s10980-019-00880-1) -->
 
-Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. 
-@daleSpatialPatternAnalysis2000 defined them as the scale-dependent predictability of the physical arrangement of observations.
-A more general definition by @grimmPatternOrientedModellingPopulation1996 describes patterns as a characteristic, clearly identifiable structure in nature itself or the data extracted from nature, indicating an underlying process that generates this pattern.
-Both definitions highlight several important aspects of spatial patterns: they are related to processes, depending on the scale of observation, and can be found in data.
+<!-- Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. -->
+Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996].
+Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003].
+However, spatial patterns are not only a result of processes, but could also be a driver of them.
+Thus, untangling the history of a landscape from its spatial patterns is one of the core concepts of landscape ecology.
 
-Patterns observed in nature contain information about a system's memory, including its demographic processes, dispersal characteristics, and constraining factors [@wiegandUsingPatternOriented2003].
-Moreover, spatial patterns are not only a result of processes but could also be a driver of them.
-Thus, untangling the complex history of a landscape from its spatial patterns is a a golden grail of landscape ecology.
-
-Here, our focus is on the computational approaches to spatial patterns.
-Computational approaches to spatial patterns open many new possibilities compared to traditional methods based on visual interpretation.
-They allow not only quantifying the patterns but also modeling, simulating, and predicting them.
-Moreover, they allow us to analyze large datasets and apply complex algorithms.
-At the same time, the computational approaches have considerations and limitations that should be taken into account.
-
-Every data type is a simplified model of the natural world, and the spatial patterns found in the data are a simplified representation of the actual spatial patterns.
-Thus, the choice of the data type, the scale of the data, and the scale of the analysis are crucial for the analysis's results.
-
-<!-- datatype -->
-Traditionally and still the most prominent, computational approaches to spatial patterns in landscape ecology revolve around "patch mosaic" model, where the landscape is represented as a raster data with different categories of land cover [@costanzaDescribingAnalyzingLandscape2019].
-Another approach is the "surface model" [@mcgarigalGradientConceptLandscape2005a], where the landscape is represented as continuous raster data with different values of some variable, such as elevation or temperature <!--@costanzaDescribingAnalyzingLandscape2019 for more refs-->.
-Notably, both these models are based on raster data.
-
-However, the raster data is not the only data type that could be used to represent spatial patterns.
-Another noteworthy data type is vector data, representing point, line, and polygon features.
-Here, the spatial patterns could be related to densities of points, orientation of lines, or shapes of polygons.
-Moreover, vector data could be used to represent point clouds, which are data types that are becoming increasingly popular with the development of remote sensing technologies (i.e., LiDAR).
-Spatial patterns in point clouds could be related to the shapes of the point clouds or the distances between points.<!--ref-->
-
-<!-- data sources??+SAR -->
-Currently, the most common data sources in landscape ecology are remote sensing-based data, either in a direct form (satellite images) or in an indirect form (spectral indices, land cover maps).
-The number of available spatial data products is still increasing, with new sensors, satellites, and data processing techniques.
-This trend is seen, for example, with the frequent release of new global land cover maps, such as the ESA WorldCover, Dynamic World, or the Esri LULC.
-<!-- refs?-->
-Other examples are the increasing availability of high-resolution data (e.g., Sentinel-2) and the growing popularity of LiDAR and SAR data.
-
-<!-- scale of data -->
-<!-- grain -->
-In computational approaches, the issue of scale is closely related to the used data. 
-Here, an important distinction should be made between the scale of the data and the data resolution.
-The data scale refers to the degree of data generalization, while the data resolution refers to the smallest data unit.
-Thus, it is possible to have data with a high resolution but a small scale, or data with a low resolution but a large scale.
-
-<!-- data cats -->
-Similarly, it is crucial to understand the data's thematic resolution, which in landscape ecology often refers to the number and definition of the categories in the data.
-Many classification systems exists that are relevant on a local, regional, or global scale, such as the NLCD, the CORINE, or the FAO systems [@neddSynthesisLandUse2021].
-Their definition of the categories could be different, and the same category could have different meanings in different systems.
-<!-- ... -->
-
-<!-- scale of analysis -->
-<!-- extent -->
-The scale of the data is related to the scale of the analysis.
-It specifies the extent of the study area and the way the data is analyzed.
-Given that the categorical rasters consist of patches, the analysis could be performed at the patch, class, or landscape level. 
-However, it is possible to analyze these levels for a fixed area, using moving windows or buffers around the sampling points.
-
-<!-- time -->
-Early approaches to spatial patterns in landscape ecology were primarily static, focusing on the spatial patterns at a single point in time.
-However, the development of remote sensing technologies and the availability of time series data have allowed for the analysis of spatial patterns over time.
-This opportunity further highlights the importance of understanding of the used data: is it a snapshot of a single moment, a composite of multiple moments, or a temporal generalization?
-The properties of the used data should align with the research question and the scale of the analysis.
-
-<!-- data quality -->
-It should be noted that the quality of the data is not constant, and the data products are not perfect: they could contain errors and biases.
-For example, the global land cover products usually have an overall accuracy of 70-80%, which means that 20-30% of the pixels are misclassified.
-Moreover, the misclassification is not random, but it is often related to the land cover type, the region, and the time of the year [@liuFinerResolutionMappingGlobal2021].
-However, most landscape ecology studies treat the data as a given, without considering its accuracy or uncertainty.
-
-<!-- data projection -->
-Less discussed but often crucial is the spatial reference system of the data.
-Most methods and software tools in landscape ecology assume that the input data is in a Cartesian coordinate reference system, where the distance between two points is the Euclidean distance (most often in meters).
-This assumption a significant impact, especially on measures related to distances and areas.
-However, when the provided data is in a geographic coordinate reference system (with units in degrees), where the distance between two points is the great-circle distance, the results of the analysis could be incorrect.
-At the same time, many projected coordinate reference systems (with units in, e.g., meters) exist, each with its own properties -- some distort areas, some distances, some angles, and some distort all of them.
-Thus, the data may require reprojection before other calculations and the decision on the spatial reference system should be made based on the research question.
-
-<!-- !what with fractals?
-!ai and deep learning still seems to be underused in landscape ecology -->
+<!-- MH: This section could use 1,2 references -->
+In order to analyze spatial patterns they must be translated into simplified and quantifiable representations of the natural world.
+Thus, the choice of the data types, spatial and temporal scales, or the thematic resolution are crucial for the analysis.
+Potential data types include raster or vector data to represent landscape characteristics using grid cells or points, lines and polygons, respectively.
+For the spatial scale the most important features are the resolution and extent.
+The increasing availability of temporal time series data allows analysis of spatial patterns over time, introducing an additional temporal scale.
+For the thematic resolution, which often refers to the number or definitions of the categories (e.g, LULC classes), many classification systems exists, such as the NLCD, the CORINE, or the FAO systems [@Nedd2021].
 
@@
 
-- Traditional approach to describe patterns
-- Quantify categorical raster data (patch mosaic model)
+Traditionally and still the most prominent, computational approaches to spatial patterns in landscape ecology revolve around the patch mosaic model, where the landscape is represented as raster data using categorical raster data [@Costanza2019].
+
 - Quantify composition and configuration
@@
 - Several alternative tools available and still widely used
+
+<!-- https://github.com/r-spatialecology/landcapemetrics -->
+<!-- https://fragstats.org/ -->
 
@@
 <!-- Nowosad, Bosch -->
+
 <!-- jn: should we use equations here? -->
+<!-- MH: Personally, I think if possible not -->
+
 <!-- jn: ask T. Stepinski for review of this section later -->
 
-Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity) and less often unpredictability (temporal heterogeneity) and scale dependence (spatio-temporal heterogeneity) (@vrankenReviewUseEntropy2015a).
-It is derived using various methods, mainly stemmed from the information theory and thermodynamics, including Shannon entropy (and its analogous forms, such as Simpson index<!-- Simpson index @simpsonMeasurementDiversity1949 -->), Boltzmann entropy, Renyi entropy, and more.
+<!-- MH: Do we need some references here for the entropies? -->
+Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence (spatio-temporal heterogeneity; [@Vranken2015]).
+It is mainly derived from information theory and thermodynamics, including besides others Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
+However, existing studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data representation, for example the composition of categories or the co-occurrence matrix [@Zhao2019].
 
-<!-- information (Shannon)/Shannon diversity -->
-The Shannon diversity index (SHDI, @Shannon1948) it quantifies the richness and evenness of the categories in the landscape, with higher values indicating higher diversity (more different classes and more even distribution of them).
-<!-- SHDI comes from the information theory.
-It is an adaptation of the Shannon entropy, which measures of the uncertainty in a random variable. -->
-The SHDI metric is based on the composition of the landscape only, omitting the categories' spatial configuration.
-<!--composition is the most important -- Ritters-->
-However, other measures from the information theory can be adapted to landscape ecology to quantify the spatial configuration and the total complexity of the landscape.
-@nowosadInformationTheoryConsistent2019a proposed to treat a categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
+<!-- (Shannon)/Shannon diversity -->
+The Shannon diversity index [@Shannon1948] quantifies the richness and evenness of categories in the landscape.
+<!-- SHDI comes from the information theory. -->
+<!-- It is an adaptation of the Shannon entropy, which measures of the uncertainty in a random variable. -->
+<!-- The Shannon diversity is based on the composition of the landscape only, omitting the categories' spatial configuration. -->
+Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012].
+However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory can be adapted.
+@Nowosad2019a proposed to treat categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
 This approach allows us to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
-It includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (representing the degree of spatial autocorrelation).
-Original Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances, as shown by @claramuntSpatiotemporalFormEntropy2012.
-<!-- https://r-spatialecology.github.io/landscapemetrics/ -->
+This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (representing the degree of spatial autocorrelation).
 
-<!-- thermodynamic (Boltzmann) -->
-<!-- "instead of Shannon entropy, what should be used is argued to be Boltzmann (1872) entropy, and calls have recently been made to revisit it" @gaoComputationBoltzmannEntropy2019 -->
-The last few years have seen a surge in the development of entropy-based metrics for landscape ecology based on the notions from thermodynamics initiated by the work of @vrankenReviewUseEntropy2015a and @cushmanThermodynamicsLandscapeEcology2015a.
-The overreaching goal of such studies is to link landscape ecology concepts to thermodynamic principles, mainly the second law of thermodynamics.
+<!-- https://github.com/r-spatialecology/landcapemetrics -->
+
+<!-- Thermodynamic (Boltzmann) -->
+<!-- "instead of Shannon entropy, what should be used is argued to be Boltzmann (1872) entropy, and calls have recently been made to revisit it" @Gao2019 -->
+The last few years have seen a surge in the development of entropy-based metrics for landscape ecology based on notions from thermodynamics initiated by the work of @Vranken2015 and @Cushman2015.
+<!-- The overreaching goal is to link landscape ecology concepts to thermodynamic principles, mainly the second law of thermodynamics. -->
 In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics, using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements).
 <!-- These studies often use Boltzmann entropy, a measure of a system's disorder, to quantify the landscape's complexity. -->
+@Cushman2016 proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
+This approach was generalized for calculations based on raster's surface model and point patterns [@Cushman2021a].
+Subsequently, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
+Recently, @Zhang2020 extended the original definition of Boltzmann entropy to incorporate the information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
 
-@Cushman2016 proposed to relate the edge length to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic in raster data.
-It is also worth noting that later, he refined and generalized this approach for calculations based on raster's surface model (based on mean slope) and point patterns (based on the average distance between points) [@cushmanGeneralizingBoltzmannConfigurational2021].
-Subsequently, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface in raster data by transforming the input grid into a series of landscape sufraces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
-In later attempt, @zhaoCalculatingSpatialConfigurational2019 extended the original definition of Boltzmann entropy to incorporate the information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
 <!-- https://jakubnowosad.com/bespatial/ -->
 
+<!--
 The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
-@stepinskiCurbYourEnthusiasm2022 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
-He also argues that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
+@Stepinski2022 and @Huckeba2023 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
+They argue that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
 Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.
+-->
 
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@renyiMeasuresEntropyInformation1961] and Gibbs entropies, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
-Rao quadratic entropy [@raoDiversityDissimilarityCoefficients1982a] has also been used more in recent years, as it measures not only the relative abundances of elements but also the pairwise dissimilarities or distances between them.<!--https://doi.org/10.1016/j.ecolind.2016.07.039-->
+The Renyi [@Renyi1961] and Gibbs entropies, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
+The Rao quadratic entropy [@Rao1982] has also been applied recently [@Rocchini2017], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
 Thus, it can be useful in cases, for example, when the dissimilarities between land cover types, and not only their categories, are relevant.
-<!-- more -->
-
-Importantly, the exisiting studies on entropy-based measures clearly show that the outcomes do not only depend on the various formulations of the entropy, but also on the underlining data representation [@zhaoCalculatingSpatialConfigurational2019].
-The same entropy equation could provide different insights when applied just to the composition of the categories, co-occurrence matrix, or on the landscape transformed into a series of data with different levels of detail.
-Thus, more work in this area is needed to understand the relations between the entropy measures and the landscape patterns they represent.
+Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023].
 
 <!-- Kolmogorov, etc. -->
-<!--@konradAdaptingMultiscaleApproach2023; Juhász-Nagy’s information theory-based functions;  https://cran.r-project.org/web/packages/LandComp/index.html -->
+<!--@Konrad2023; Juhász-Nagy’s information theory-based functions;  https://cran.r-project.org/web/packages/LandComp/index.html -->
 <!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->
@@
 <!-- Nowosad, Hesselbarth -->
+
 <!-- jn: ask T. Matuszek for review of this section later -->
 
-The spatial vector data model is based on the representation of the Earth's surface as a set of points, lines, and polygons, where each feature is defined by its geometry and attributes.
-Here, we will focus on the polygonal vector data, which is the most common in landscape ecology.
-The polygons are discrete areas, with well-defined boundaries.
-Thus, converting the polygons to raster data (*rasterization*) is a lossy process, in which information about the intricate details of the polygon shapes is lost.
-Moreover, the rasterization process largely depends on the decided resolution of the raster outcome.
+Many landscape ecology studies are based on raster data.
+However, the decision on the data model is often driven by the availability of the data and software, or the familiarity with the approach [@Costanza2019].
+<!-- Most LULC maps are provided as raster as they often related to underlying remote sensing products. -->
+For example, several early landscape ecology studies were based on polygonal vector data that was converted to raster data <!-- MH: Add one eg ref -->.
+<!-- jn: maybe we could also mention that it is still the case -->
+Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.
 
-As mentioned, the initial computational approaches in landscape ecology were based on raster data.
-However, the decision on the data model used was not theoretical but rather practical, as at the time, the most commonly used related software was raster-based, such as GRASS GIS.
-In fact, early studies were based on polygonal vector data that was later converted to raster data for the analysis [@costanzaDescribingAnalyzingLandscape2019].<!--jn: we could also cite actual examples from that paper, but the list of the references would be long-->
-<!-- jn: maybe we could also mention that it is still the case -->
+Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement them for vector data.
+This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to raster data [@Yao2022].
+Yet, the vector data model has been widely used in other fields, such as geography or urban studies.
+For example, in urban planning, vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017].
+Earlier approaches included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985].
+More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010]
 
-Most landscape ecology studies are still based on raster data, probably due to the familarity with the techiniques, the availability of the data, and the availability of the software.
-For example, most land cover maps (as LULC products are often related to underlying remote sensing products), including all on the global scale, are provided as rasters. 
-At the same time, nowadays, vector data of land cover on regional or local scales is also available, and most geospatial software supports vector data.
-
-Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
-@yaoVecLIFrameworkCalculating2022 showed that this is possible for most of the metrics, with only several exceptions (of the shape and aggregation group), whose parameters are related to the raster-format structure.
-Their results highlighted that about 80% of the raster and vector-based metrics are highly correlated (for their study area), and the higher the raster resolution, the stronger the correlation is.
-
-While landscape ecology studies have focussed on using raster data, the vector data model has been widely used in other fields, such as geography and urban studies.
-@maceachrenCompactnessGeographicShape1985 presented four categories of compactness measures for vector data: perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes.
-@angelTenCompactnessProperties2010 introduced a unified theoretical foundation for measuring shape compactness.
-They observed that the circle is the most compact of shapes, proposed a set of ten distinct properties of the circle, and then defined metrics associated with each of these properties.
-Vector-based metrics are now popular, for example, in urban planning, as they allow to quantify the shapes of urban areas, e.g., @basaranerPerformanceShapeIndices2017a used 18 existing metrics and two new ones to characterize the complexity of building footprints.
-<!-- more -->
-
-The vector-based metrics also have their limitations.
-The most important is the computational complexity, which makes the calculation of vector metrics slower than rasters ones, especially for large datasets.
+Vector-based metrics also have their limitations.
+The most important is the computational complexity, which makes the calculation of vector metrics slower than their raster equivalents.
 Another technical issue is the need for the data to be topologically correct, which is not always the case, especially for data from different sources.
-Vector-based metrics are also much less understood than raster-based ones, and their relations to landscape patterns still need to be explored.
+<!-- MH: What does topologically correct mean?-->
+Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
 <!-- (e.g., could we do landscapemetrics vs vectormetrics benchmarking?) -->
@@
 
-Various metrics are often used to describe spatial patterns in landscape ecology.
-<!-- limits, correlations?? -->
-These spatial patterns can then be further analyzed and explored through a range of computational operations, such as comparing, searching, or grouping [@turnerSpatialTemporalAnalysis1990;@wickhamMappingAnalyzingLandscape1994; @jasiewiczGeoPATToolboxPatternbased2015; @remmelDistributionsHyperLocalConfiguration2020b;@nowosadMotifOpensourceTool2021].
-The above operations are based on two main ideas: the spatial signatures and the dissimilarity measures.
-A spatial signature is a multi-numerical representation of a landscape pattern for a given area.
-It may be created based on a set of (preferably uncorrelated) landscape metrics, or be derived from the data itself, e.g., by using the co-occurrence matrix [@nowosadPatternbasedIdentificationMapping2021].
-The dissimilarity measure is a function that quantifies the difference between two spatial signatures, with low values indicating similar patterns.
+Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990;@Wickham1994;@Jasiewicz2015;@Remmel2020;@Nowosad2021a].
+These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between the signatures.
 
-The most basic operation, the comparison of spatial patterns, can be performed for just a pair of areas or the same area in two different moments.
-It can also be extended straightforwardly by dividing the landscape into multiple smaller areas and comparing them at two moments in time.
-Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@duncanMonitoringSpatialPatterns2023].
-Pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the spatial signatures of all other areas.
-This allows us to find areas with similar patterns to the query area, which could be used to identify areas with similar environmental conditions [@jasiewiczLandscapeSimilarityRetrieval2014].
-Finally, the grouping of spatial patterns is a many-to-many operation, in which spatial signatures are calculated for all areas, and then these areas are clustered based on their similarity [@nowosadPatternbasedIdentificationMapping2021].
+Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
+Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023].
+Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
+This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014].
+Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021b].
 
 <!-- one/two par about other approaches -->
+<!-- e.g., modeling, simulation, prediction, etc. -->
 
-<!-- motif -->
-<!-- spquery -->
-
+<!-- https://github.com/Nowosad/motif/ -->
+<!-- https://jakubnowosad.com/spquery/ -->
 <!-- FOSS4G Europe pap? -->

### 2024-06-24 Sections/spatial-patterns.qmd
Update spatial pattern

@@
 
-<!-- [Methods for landscape characterization](https://doi.org/10.1016/j.landusepol.2018.04.022) -->
-<!-- [Pattern metrics](https://doi.org/10.1007/s10980-018-0755-4) -->
-<!--[Emerging trajectories for spatial pattern](https://link.springer.com/article/10.1007/s10980-019-00880-1) -->
-
 <!-- Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. -->
@@
 However, spatial patterns are not only a result of processes, but could also be a driver of them.
-Thus, untangling the history of a landscape from its spatial patterns is one of the core concepts of landscape ecology.
-
-<!-- MH: This section could use 1,2 references -->
-In order to analyze spatial patterns they must be translated into simplified and quantifiable representations of the natural world.
-Thus, the choice of the data types, spatial and temporal scales, or the thematic resolution are crucial for the analysis.
-Potential data types include raster or vector data to represent landscape characteristics using grid cells or points, lines and polygons, respectively.
-For the spatial scale the most important features are the resolution and extent.
-The increasing availability of temporal time series data allows analysis of spatial patterns over time, introducing an additional temporal scale.
-For the thematic resolution, which often refers to the number or definitions of the categories (e.g, LULC classes), many classification systems exists, such as the NLCD, the CORINE, or the FAO systems [@Nedd2021].
+Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989].
 
@@
 
-Traditionally and still the most prominent, computational approaches to spatial patterns in landscape ecology revolve around the patch mosaic model, where the landscape is represented as raster data using categorical raster data [@Costanza2019].
-
-- Quantify composition and configuration
-- Widely applied since FRAGSTATS release
-- Several alternative tools available and still widely used
+Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017; @Costanza2019].
+The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015], however, they are also critiqued for not being able to realistically represent continuous features of the natural word [@Frazier2017].
+Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019].
+Recently also several metrics describing landscape complexity were introduced, e.g., several entropy measures (see following sections).
 
@@
 <!-- Bosch, Hesselbarth -->
-[Surface metrics](https://doi.org/10.1007/s10980-018-0685-1)
 
-- More recent development
-- Quantify continuous raster data (gradient surface models)
-- More realistic representation of landscapes (?)
-- Based on molecular physics/medicine
-- Not widely adopted yet (due to missing tools?)
+Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009; @Cushman2010] and were adapted from microscopy and molecular physics [@McGarigal2009; @Kedron2018].
+The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010; @Kedron2018].
+Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010].
+Many surface metrics have analogous landscape metrics [@Cushman2010], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009; @Borthwick2020].
+Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018; @Frazier2019].
+<!-- MH: What about Spectral and Wavelet Analysis?
+
+<!-- https://www.digitalsurf.com/free-trial/ -->
+<!-- https://github.com/bioXgeo/geodiv -->
 
@@
 
-<!-- MH: Do we need some references here for the entropies? -->
-Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence (spatio-temporal heterogeneity; [@Vranken2015]).
-It is mainly derived from information theory and thermodynamics, including besides others Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
-However, existing studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data representation, for example the composition of categories or the co-occurrence matrix [@Zhao2019].
+It is mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
+Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015].
+However, studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix [@Zhao2019].
 
-<!-- (Shannon)/Shannon diversity -->
-The Shannon diversity index [@Shannon1948] quantifies the richness and evenness of categories in the landscape.
-<!-- SHDI comes from the information theory. -->
-<!-- It is an adaptation of the Shannon entropy, which measures of the uncertainty in a random variable. -->
-<!-- The Shannon diversity is based on the composition of the landscape only, omitting the categories' spatial configuration. -->
+The Shannon diversity index [@Shannon1948] quantifies the richness and evenness of categories in the landscape omitting spatial configuration.
+The index is based on information theory and is an adaptation of the Shannon entropy which measures uncertainty of a random variable.
 Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012].
@@
 
-<!-- Thermodynamic (Boltzmann) -->
-<!-- "instead of Shannon entropy, what should be used is argued to be Boltzmann (1872) entropy, and calls have recently been made to revisit it" @Gao2019 -->
-The last few years have seen a surge in the development of entropy-based metrics for landscape ecology based on notions from thermodynamics initiated by the work of @Vranken2015 and @Cushman2015.
-<!-- The overreaching goal is to link landscape ecology concepts to thermodynamic principles, mainly the second law of thermodynamics. -->
-In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics, using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements).
-<!-- These studies often use Boltzmann entropy, a measure of a system's disorder, to quantify the landscape's complexity. -->
+Based on notions from thermodynamics initiated by the work of @Cushman2015 or @Vranken2015, there was a surge in the development of entropy-based metrics for landscape ecology in the last few years.
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019].
+<!-- REF -->
 @Cushman2016 proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
@@
 Subsequently, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
-Recently, @Zhang2020 extended the original definition of Boltzmann entropy to incorporate the information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
+Recently, @Zhang2020 extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
 
@@
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961] and Gibbs entropies, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
+The Renyi [@Renyi1961] and Gibbs entropies <!-- REF -->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
 The Rao quadratic entropy [@Rao1982] has also been applied recently [@Rocchini2017], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
-Thus, it can be useful in cases, for example, when the dissimilarities between land cover types, and not only their categories, are relevant.
+Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
 Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023].
@@
 
-Many landscape ecology studies are based on raster data.
-However, the decision on the data model is often driven by the availability of the data and software, or the familiarity with the approach [@Costanza2019].
+Many landscape ecology studies are based on the raster data model.
+However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019].
 <!-- Most LULC maps are provided as raster as they often related to underlying remote sensing products. -->
-For example, several early landscape ecology studies were based on polygonal vector data that was converted to raster data <!-- MH: Add one eg ref -->.
-<!-- jn: maybe we could also mention that it is still the case -->
+For example, several early landscape ecology studies were based on polygonal vector data that was converted to raster data. <!-- REF -->
 Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.
 
-Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement them for vector data.
-This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to raster data [@Yao2022].
+Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
+This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022].
 Yet, the vector data model has been widely used in other fields, such as geography or urban studies.
-For example, in urban planning, vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017].
+For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017].
 Earlier approaches included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985].
@@
 Vector-based metrics also have their limitations.
-The most important is the computational complexity, which makes the calculation of vector metrics slower than their raster equivalents.
-Another technical issue is the need for the data to be topologically correct, which is not always the case, especially for data from different sources.
+The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
 <!-- MH: What does topologically correct mean?-->
+Another technical issue is the need for the data to be topologically correct, which is often problematic especially for data from different sources.
 Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
-<!-- (e.g., could we do landscapemetrics vs vectormetrics benchmarking?) -->
-<!-- polygon vs multipolygon technical distinction -->
 
@@
 
-Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990;@Wickham1994;@Jasiewicz2015;@Remmel2020;@Nowosad2021a].
+Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990; @Wickham1994; @Jasiewicz2015; @Remmel2020; @Nowosad2021a].
 These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between the signatures.
@@
 
-<!-- one/two par about other approaches -->
-<!-- e.g., modeling, simulation, prediction, etc. -->
-
 <!-- https://github.com/Nowosad/motif/ -->

### 2024-06-24 Sections/simulations.qmd
Update simulations

--- 2024-05-17 Sections/simulations.qmd
@@
 
-## Neutral landscapes
-<!-- Hesselbarth -->
+Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005; @Petrovskii2012].
+Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004].
+Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006; @Synes2016].
+In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004;@Synes2016].
+Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.
 
-## Process-based simulation
-<!-- Schouten, Hesselbarth -->
+## Neutral landscapes models
+Neutral landscapes models simulate landscape patterns without assuming any underlying abiotic or biotic processes.
+Thus, they are often used as null hypothesis, baselines, or scenarios for comparisons with real landscape patterns, while controlling certain aspects of the landscape [@Li2004; @Wang2008].
+The first neutral models simply assigned land cover types randomly to cells in the landscape and were based on percolation theory [@Gardner1987].
+Further developments included hierarchical models that considered different spatial scales while assigning land cover types [@ONeill1992].
+While these first two classes can only simulate landscape with discrete land cover classes, fractal neutral models can also simulate continuous land cover maps by utilizing e.g., Brownian motion [@Palmer1992].
+Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or Binary space partitioning [@Etherington2022a] hierarchical neutral landscapes.
+In order to simulate landscapes dominated by human activities approaches based on least-cost network movement exist [@Etherington2024].
+Neutral landscapes models can also be used to simulated landscapes similar to real landscapes [@Inkoom2017], or based on target values of landscape metrics [@VanStrien2016; @Justeau-Allaire2022].
+<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], or Julia (https://github.com/EcoJulia/NeutralLandscapes.jl).
+-->
+
+## Landscape Generator
+Processes-based landscape generators are related to neutral landscapes models because they can simulate virtual landscapes.
+However, in contrast to neutral landscape models, processes-based landscape generators are able to simulate structurally more realistic landscapes by including pattern creating processes [@Langhammer2019; @Salecker2019].
+
+<!--
+Cellular automaton
+[G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
+[GradientLand Software](https://doi.org/10.1016/j.ecoinf.2014.12.001)
+[PioLaG](https://doi.org/10.1007/s10980-020-01066-w)
+-->
+
+## IBMs /Coupled lattice
+
+## Digital twins
+
+<!-- 
+[RangeShifter](https://doi.org/10.1111/ecog.05687)
+[Marine SDMS](https://doi.org/10.1016/j.ecolmodel.2019.108837)
+[SDM past present future](https://doi.org/10.1111/jbi.14617)
+[Simulations in landscape genetics](https://doi.org/10.1111/j.1365-294X.2010.04678.x)
+[HexSim modeling environment](https://doi.org/10.1007/s10980-017-0605-9)
+[Digital twins?](https://doi.org/10.1016/j.tree.2023.04.010)
+-->

### 2024-06-24 Manuscript.qmd
Some updates on intro and spatial-patterns
Updat MS file

@@
 fontsize: 12pt
+mainfont: "Garamond"
 geometry: margin=25mm
@@
 {{< include Sections/introduction.qmd >}}
+{{< include Sections/spatial-patterns.qmd >}}
 {{< include Sections/simulations.qmd >}}

### 2024-06-24 Manuscript.tex
Updat MS file

@@
     % xetex/luatex font selection
+  \setmainfont[]{Garamond}
 \fi
@@
 Typical research topics include ecological flows in landscape mosaics,
-drivers and consequences of land use and land cover change, nonlinear
-dynamics and landscape complexity, scale and scaling, spatial analysis
-and landscape modeling, linking landscape metrics to ecological
+drivers and consequences of land use and land cover (LULC) change,
+nonlinear dynamics and landscape complexity, scale and scaling, spatial
+analysis and landscape modeling, linking landscape metrics to ecological
 processes, integrating humans and their activities into landscape
@@
 experimental settings (Petrovskii and Petrovskaya 2012; but see Wiersma
-2022). Because landscape ecology is a cross-disciplinary field
-including, e.g., social sciences, geography, or ecology and evolution
-(Wiens 1997; With 2019), the general increase of available data (Chi et
-al. 2016; Jarić et al. 2020; Wüest et al. 2020; Nathan et al. 2022), or
-the complex landscape systems (Newman et al. 2019), there is the need
-for sophisticated computational methods in order to link heterogeneity
-patterns and ecological processes. Thus, computational methods are one
-of the most important tools of modern scientific research (Prlić and
-Procter 2012; Wilson et al. 2014).
+2022). Because landscape ecology is a cross-disciplinary field including
+(e.g., social sciences, geography, or ecology and evolution; Wiens 1997;
+With 2019), the general increase of data availability (Chi et al. 2016;
+Jarić et al. 2020; Wüest et al. 2020; Nathan et al. 2022), or the
+studied complex landscape systems (Newman et al. 2019), there is the
+need for sophisticated computational methods in order to link
+heterogeneity patterns and ecological processes. Thus, computational
+methods are one of the most important tools of modern scientific
+research (Prlić and Procter 2012; Wilson et al. 2014).
+
+For all computational methods the natural world must be translated into
+representative data models. The spatial vector data model is based on
+points, lines, and polygons features that are defined by their geometry
+and additional attributes related to them. Contrastingly, the raster
+data model is based on evenly-spaced grid cells and discrete or
+continuous values related to each cell (Lovelace et al. 2019; With
+2019). The choice of the data models, spatial and temporal scales, or
+the thematic resolution are crucial for the analysis (Wade et al. 2003;
+Lechner and Rhodes 2016; Nedd et al. 2021).
 
@@
 landscape ecology. For a more comprehensive, general introductions to
-landscape ecology, please see Turner and Gardner (2015), or Gergel and
-Turner (2017). First, we will introduce recent computational
-developments of the most important topics of landscape ecology. This
-includes, for example, spatial patterns quantification, connectivity and
-movement, or model simulations. Second, we will introduce common
-software tools that implement or are potentially capable of implementing
-these developments. However, due to its many advantages, we will focus
-on open-source software (von Krogh and von Hippel 2006), and
-specifically scripting languages, i.e., R, Python, and Julia. Last, we
-will discuss potential future needs developments of computational
-methods in landscape ecology.
+landscape ecology, please see Turner and Gardner (2015), Gergel and
+Turner (2017), or With (2019). First, we will introduce recent
+computational developments of the most important topics of landscape
+ecology. This includes, for example, spatial patterns quantification,
+connectivity and movement, or model simulations. Second, we will
+introduce common software tools that implement or are potentially
+capable of implementing these developments. However, due to its many
+advantages (Powers and Hampton 2019), we will focus on open-source
+software , and specifically scripting languages, i.e., R, Python, and
+Julia. Last, we will discuss potential future needs developments of
+computational methods in landscape ecology.
 
@@
 but could also be a driver of them. Thus, untangling the history of a
-landscape from its spatial patterns is one of the core concepts of
-landscape ecology.
-
-In order to analyze spatial patterns they must be translated into
-simplified and quantifiable representations of the natural world. Thus,
-the choice of the data types, spatial and temporal scales, or the
-thematic resolution are crucial for the analysis. Potential data types
-include raster or vector data to represent landscape characteristics
-using grid cells or points, lines and polygons, respectively. For the
-spatial scale the most important features are the resolution and extent.
-The increasing availability of temporal time series data allows analysis
-of spatial patterns over time, introducing an additional temporal scale.
-For the thematic resolution, which often refers to the number or
-definitions of the categories (e.g, LULC classes), many classification
-systems exists, such as the NLCD, the CORINE, or the FAO systems (Nedd
-et al. 2021).
+landscape and linking its spatial patterns to ecological processes is
+one of the core concepts of landscape ecology (Turner 1989).
 
@@
 
-Traditionally and still the most prominent, computational approaches to
-spatial patterns in landscape ecology revolve around the patch mosaic
-model, where the landscape is represented as raster data using
-categorical raster data (Costanza et al. 2019).
-
-\begin{itemize}
-\tightlist
-\item
-  Quantify composition and configuration
-\item
-  Widely applied since FRAGSTATS release
-\item
-  Several alternative tools available and still widely used
-\end{itemize}
+Traditionally and still the most prominent approaches to quantify
+spatial patterns revolve around raster data using categorical categories
+based on the patch mosaic model (Frazier and Kedron 2017; Costanza et
+al. 2019). The strength of landscape metrics is that they are comparable
+easy to apply and communicate (Lausch et al. 2015), however, they are
+also critiqued for not being able to realistically represent continuous
+features of the natural word (Frazier and Kedron 2017). Metrics can be
+calculated on patch-level (describing each patch), class-level
+(describing all patches of the same class), or landscape level
+(describing the entire landscape) and are able to quantify
+characteristics related to e.g., patch areas and edge lengths, patch
+shapes, or class aggregation and diversity (Hesselbarth et al. 2019).
+Recently also several metrics describing landscape complexity were
+introduced, e.g., several entropy measures (see following sections).
 
@@
 
-\href{https://doi.org/10.1007/s10980-018-0685-1}{Surface metrics}
-
-\begin{itemize}
-\tightlist
-\item
-  More recent development
-\item
-  Quantify continuous raster data (gradient surface models)
-\item
-  More realistic representation of landscapes (?)
-\item
-  Based on molecular physics/medicine
-\item
-  Not widely adopted yet (due to missing tools?)
-\end{itemize}
+Surface metrics are based on the gradient surface model using raster
+data and continuous values (McGarigal et al. 2009; Cushman et al. 2010)
+and were adapted from microscopy and molecular physics (McGarigal et al.
+2009; Kedron et al. 2018). The gradient surface model can increase the
+resemblance of the data model and the natural world because it allows to
+include more heterogeneity within each grid cell (Cushman et al. 2010;
+Kedron et al. 2018). Surface metrics are able to quantify various
+characteristics, such as roughness, skewness and kurtosis, total and
+relative amplitudes, curvatures of local peaks, or surface bearing area
+ratios (Cushman et al. 2010). Many surface metrics have analogous
+landscape metrics (Cushman et al. 2010), however, using surface metrics
+allows to reveal different or additional patterns and potentially
+pattern-process links (McGarigal et al. 2009; Borthwick et al. 2020).
+Yet, software to calculate these metrics is still rare and further
+research is needed into the specific pattern-process links and
+ecological meaningful interpretations (Kedron et al. 2018; Frazier
+2019).
 
@@
 
-Entropy in landscape ecology is mainly used to quantify the complexity
-of the landscape (spatial heterogeneity), and less often
+It is mainly derived from information theory and thermodynamics, such as
+Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi
+entropy. Entropy in landscape ecology is mainly used to quantify the
+complexity of the landscape (spatial heterogeneity), and less often
 unpredictability (temporal heterogeneity) and scale dependence
-(spatio-temporal heterogeneity; (Vranken et al. 2015)). It is mainly
-derived from information theory and thermodynamics, including besides
-others Shannon entropy (and its analogous forms), Boltzmann entropy, or
-Renyi entropy. However, existing studies show that insights do not only
-depend on the formulation of the used entropy, but also on the
-underlining data representation, for example the composition of
-categories or the co-occurrence matrix (Zhao and Zhang 2019).
+(spatio-temporal heterogeneity; Vranken et al. 2015). However, studies
+show that insights do not only depend on the formulation of the used
+entropy, but also on the underlining data model, for example the
+composition of categories or the co-occurrence matrix (Zhao and Zhang
+2019).
 
 The Shannon diversity index (Shannon 1948) quantifies the richness and
-evenness of categories in the landscape. Shannon's entropy formula can
-also be modified to include the landscape's spatial configuration, e.g.,
-by weights calculated from intra- and inter-class distances (Claramunt
-2012). However, to quantify the spatial configuration and the total
-complexity of the landscape, other measures from information theory can
-be adapted. Nowosad and Stepinski (2019) proposed to treat categorical
-raster as a bivariate random variable \((x,y)\) where \(x\) is the
-category of the focal cell and \(y\) is the category of the neighboring
-cell. This approach allows us to compress information about the
-landscape's composition and configuration into a co-occurrence matrix,
-which could be used to calculate various entropy measures. This includes
-conditional entropy (representing configurational complexity), joint
-entropy (representing overall spatial-thematic complexity), mutual
-information, and relative mutual information (representing the degree of
-spatial autocorrelation).
+evenness of categories in the landscape omitting spatial configuration.
+The index is based on information theory and is an adaptation of the
+Shannon entropy which measures uncertainty of a random variable.
+Shannon's entropy formula can also be modified to include the
+landscape's spatial configuration, e.g., by weights calculated from
+intra- and inter-class distances (Claramunt 2012). However, to quantify
+the spatial configuration and the total complexity of the landscape,
+other measures from information theory can be adapted. Nowosad and
+Stepinski (2019) proposed to treat categorical raster as a bivariate
+random variable \((x,y)\) where \(x\) is the category of the focal cell
+and \(y\) is the category of the neighboring cell. This approach allows
+us to compress information about the landscape's composition and
+configuration into a co-occurrence matrix, which could be used to
+calculate various entropy measures. This includes conditional entropy
+(representing configurational complexity), joint entropy (representing
+overall spatial-thematic complexity), mutual information, and relative
+mutual information (representing the degree of spatial autocorrelation).
 
-The last few years have seen a surge in the development of entropy-based
-metrics for landscape ecology based on notions from thermodynamics
-initiated by the work of Vranken et al. (2015) and Cushman (2015). In
+Based on notions from thermodynamics initiated by the work of Cushman
+(2015) or Vranken et al. (2015), there was a surge in the development of
+entropy-based metrics for landscape ecology in the last few years. In
 the late 1800s, Boltzmann formulated a probabilistic interpretation of
-the second law of thermodynamics, using the concepts of macrostate (the
+the second law of thermodynamics using the concepts of macrostate (the
 general state of a system) and microstate (the configuration of the
-system elements). Cushman (2016) proposed to relate the edge length to
-the microstate of the landscape, and use the proportion of microstates
-to compute the relative Boltzmann entropy of a landscape mosaic. This
-approach was generalized for calculations based on raster's surface
-model and point patterns (Cushman 2021). Subsequently, Gao et al. (2017)
-proposed to use the Boltzmann entropy to quantify the complexity of a
-landscape surface by transforming the input raster into a series of
-landscape surfaces with different levels of detail (microstate) and
-calculating the Boltzmann entropy based on the total number of
-microstates that are able to generate the observed macrostate. Recently,
-Zhang et al. (2020) extended the original definition of Boltzmann
-entropy to incorporate the information about the adjacency of the same
-categories in the landscape mosaic by using the number of contiguous
-patches of the same category.
+system elements; Gao and Li 2019). Cushman (2016) proposed to relate the
+edge length to the microstate of the landscape, and use the proportion
+of microstates to compute the relative Boltzmann entropy of a landscape
+mosaic. This approach was generalized for calculations based on raster's
+surface model and point patterns (Cushman 2021). Subsequently, Gao et
+al. (2017) proposed to use the Boltzmann entropy to quantify the
+complexity of a landscape surface by transforming the input raster into
+a series of landscape surfaces with different levels of detail
+(microstate) and calculating the Boltzmann entropy based on the total
+number of microstates that are able to generate the observed macrostate.
+Recently, Zhang et al. (2020) extended the original definition of
+Boltzmann entropy to incorporate information about the adjacency of the
+same categories in the landscape mosaic by using the number of
+contiguous patches of the same category.
 
 Moreover, many other entropy-based metrics have been proposed for use in
-landscape ecology. The Renyi (Rényi 1961) and Gibbs entropies, which are
-both generalizations of the Shannon entropy, are applied to quantify the
-complexity of the landscape. The Rao quadratic entropy (Rao 1982) has
-also been applied recently (Rocchini et al. 2017), as it measures not
-only the relative abundances of elements, but also the pairwise
+landscape ecology. The Renyi (Rényi 1961) and Gibbs entropies , which
+are both generalizations of the Shannon entropy, are applied to quantify
+the complexity of the landscape. The Rao quadratic entropy (Rao 1982)
+has also been applied recently (Rocchini et al. 2017), as it measures
+not only the relative abundances of elements, but also the pairwise
 dissimilarities or distances between them. Thus, it can be useful in
-cases, for example, when the dissimilarities between land cover types,
-and not only their categories, are relevant. Another recent development
-is the use of Kullback-Leibler divergence (also known as relative
-entropy), which is a measure of differences between two probability
-distributions to describe patterns across scales (Huckeba et al. 2023).
+cases, when the dissimilarities between LULC types, and not only their
+categories, are relevant. Another recent development is the use of
+Kullback-Leibler divergence (also known as relative entropy), which is a
+measure of differences between two probability distributions to describe
+patterns across scales (Huckeba et al. 2023).
 
@@
 
-Many landscape ecology studies are based on raster data. However, the
-decision on the data model is often driven by the availability of the
-data and software, or the familiarity with the approach (Costanza et al.
-2019). For example, several early landscape ecology studies were based
-on polygonal vector data that was converted to raster data . Yet,
+Many landscape ecology studies are based on the raster data model.
+However, the decision on the data model is often driven by data and
+software availability, or the familiarity with the approach (Costanza et
+al. 2019). For example, several early landscape ecology studies were
+based on polygonal vector data that was converted to raster data. Yet,
 nowadays, vector data is increasingly available, and most geospatial
@@
 Hundreds of metrics have already been developed for raster data, so the
-most straightforward approach is to reimplement them for vector data.
-This is possible for most of the metrics, with only several exceptions
-of metrics that are specifically related to raster data (Yao et al.
-2022). Yet, the vector data model has been widely used in other fields,
-such as geography or urban studies. For example, in urban planning,
-vector metrics are applied to quantify the shapes of urban areas and
-characterize the complexity of building footprints (Basaraner and
-Cetinkaya 2017). Earlier approaches included four categories of
-compactness measures for vector data, namely perimeter-area, single
-parameters of related circles, dispersion of elements of the area around
-a centroid, and direct comparison to standard shapes (Maceachren 1985).
-More recently, a unified theoretical foundation for measuring shape
-compactness was introduced using a set of ten distinct properties of a
-circle and metrics associated with each of these properties (Angel et
-al. 2010)
+most straightforward approach is to reimplement the same metrics for
+vector data. This is possible for most of the metrics, with only several
+exceptions of metrics that are specifically related to the raster data
+model (Yao et al. 2022). Yet, the vector data model has been widely used
+in other fields, such as geography or urban studies. For example, in
+urban planning vector metrics are applied to quantify the shapes of
+urban areas and characterize the complexity of building footprints
+(Basaraner and Cetinkaya 2017). Earlier approaches included four
+categories of compactness measures for vector data, namely
+perimeter-area, single parameters of related circles, dispersion of
+elements of the area around a centroid, and direct comparison to
+standard shapes (Maceachren 1985). More recently, a unified theoretical
+foundation for measuring shape compactness was introduced using a set of
+ten distinct properties of a circle and metrics associated with each of
+these properties (Angel et al. 2010)
 
 Vector-based metrics also have their limitations. The most important is
-the computational complexity, which makes the calculation of vector
-metrics slower than their raster equivalents. Another technical issue is
-the need for the data to be topologically correct, which is not always
-the case, especially for data from different sources. Last, the
+computational complexity, which makes calculations of vector metrics
+slower than their raster equivalents. Another technical issue is the
+need for the data to be topologically correct, which is often
+problematic especially for data from different sources. Last, the
 pattern-process link is still less explored for vector-based metrics
@@
 settings (Pascual 2005; Petrovskii and Petrovskaya 2012). Thus, they can
-be seen as experimental systems that allow, unlike to nature, all
-imaginable manipulations in order to advance theoretical developments or
-test hypotheses (Peck 2004). Due to the usual temporal and spatial
-scales, complex interactions and feedbacks, or scale mismatches between
-patterns and processes, simulation models are one of the major
-approaches in landscape ecology (Schröder and Seppelt 2006; Synes et al.
-2016). In general, simulations models can be classified using two major
-divisions, namely as \emph{i)} predictive or exploratory models, and
-\emph{ii)} pattern- or process-based models (Peck 2004; Synes et al.
-2016). Here, we are going to focus mainly on exploratory models, but
-include both pattern- and process-based models.
+be seen as experimental systems that allow, unlike to the natural world,
+all imaginable manipulations in order to advance theoretical
+developments or test hypotheses (Peck 2004). Due to the usual temporal
+and spatial scales, complex interactions and feedbacks, or scale
+mismatches between patterns and processes, simulation models are one of
+the major approaches in landscape ecology (Schröder and Seppelt 2006;
+Synes et al. 2016). In general, simulations models can be classified
+using two major divisions, namely as \emph{i)} predictive or exploratory
+models, and \emph{ii)} pattern- or process-based models (Peck 2004;
+Synes et al. 2016). Here, we are going to focus mainly on exploratory
+models, but include both pattern- and process-based models.
 
-\subsection{Pattern based}\label{pattern-based}
-
-\subsubsection{Neutral landscapes
-models}\label{neutral-landscapes-models}
+\subsection{Neutral landscapes models}\label{neutral-landscapes-models}
 
@@
 any underlying abiotic or biotic processes. Thus, they are often used as
-null hypothesis or baselines for comparisons with real landscape
-patterns, while controlling certain aspects of the landscape (Li et al.
-2004; Wang and Malanson 2008). The first neutral models simply assigned
-land cover types randomly to cells in the landscape and were based on
-percolation theory. Further developments included hierarchical models
-that considered different spatial scales while assigning land cover
-types. While these first two classes can only simulate landscape with
-discrete land cover classes, fractal neutral models can also simulate
-continuous land cover maps by utilizing e.g., Brownian motion. Last,
-more complex neutral models were developed that are based on contagion,
-autocorrelation, or spatial aggregation concepts to simulate more
-realistic landscapes (Wang and Malanson 2008).
+null hypothesis, baselines, or scenarios for comparisons with real
+landscape patterns, while controlling certain aspects of the landscape
+(Li et al. 2004; Wang and Malanson 2008). The first neutral models
+simply assigned land cover types randomly to cells in the landscape and
+were based on percolation theory (Gardner et al. 1987). Further
+developments included hierarchical models that considered different
+spatial scales while assigning land cover types (O'Neill et al. 1992).
+While these first two classes can only simulate landscape with discrete
+land cover classes, fractal neutral models can also simulate continuous
+land cover maps by utilizing e.g., Brownian motion (Palmer 1992).
+Borrowing from computer graphics, more recent neutral landscape models
+make use of spectral synthesis (e.g., Perlin noise; Etherington 2022) or
+Binary space partitioning (Etherington et al. 2022) hierarchical neutral
+landscapes. In order to simulate landscapes dominated by human
+activities approaches based on least-cost network movement exist
+(Etherington et al. 2024). Neutral landscapes models can also be used to
+simulated landscapes similar to real landscapes (Inkoom et al. 2017), or
+based on target values of landscape metrics (Van Strien et al. 2016;
+Justeau-Allaire et al. 2022).
 
-doi:10.12942/lrlr-2014-2 Perlin noise (Etherington 2022) Binary space
-partitioning {[}Etherington2022a{]} Generating in comparison to real
-landscapes https://doi.org/10.1016/j.ecoinf.2017.08.003
-https://doi.org/10.1016/j.ecolmodel.2018.12.010
+\subsection{Landscape Generator}\label{landscape-generator}
 
-In order to make multiple neutral landscape model algorithms accessible,
-several software libraries are available in prominent programming
-languages including Python {[}NLMpy; Etherington et al. (2015){]}, R
-{[}NLMR; Sciaini et al. (2018){]}, or Julia
-(https://github.com/EcoJulia/NeutralLandscapes.jl).
+Processes-based landscape generators are related to neutral landscapes
+models because they can simulate virtual landscapes. However, in
+contrast to neutral landscape models, processes-based landscape
+generators are able to simulate structurally more realistic landscapes
+by including pattern creating processes (Langhammer et al. 2019;
+Salecker et al. 2019).
 
-\subsubsection{Cellular Automata}\label{cellular-automata}
+\subsection{IBMs /Coupled lattice}\label{ibms-coupled-lattice}
 
-\subsection{Process based}\label{process-based}
-
-\href{https://doi.org/10.1111/ecog.05687}{RangeShifter}
-
-\subsubsection{Landscape Generator}\label{landscape-generator}
-
-\href{https://doi.org/10.1371/journal.pone.0064968}{G-RaFFE}
-\href{https://doi.org/10.1002/ece3.2145}{LG}
-https://journals.plos.org/plosone/article/file?id=10.1371/journal.pone.0222949\&type=printable
-
-\subsubsection{Coupled lattice}\label{coupled-lattice}
-
-\subsubsection{IBMs/ABMs}\label{ibmsabms}
-
-\subsection{Misc}\label{misc}
-
-\href{https://doi.org/10.1016/j.ecolmodel.2019.108837}{Marine SDMS}
-\href{https://doi.org/10.1111/jbi.14617}{SDM past present future}
-
-\href{https://doi.org/10.1111/j.1365-294X.2010.04678.x}{Simulations in
-landscape genetics}
-\href{https://doi.org/10.1007/s10980-017-0605-9}{HexSim modeling
-environment} \href{https://doi.org/10.1016/j.tree.2023.04.010}{Digital
-twins?}
+\subsection{Digital twins}\label{digital-twins}
 
@@
 \url{https://doi.org/10.1080/13658816.2017.1346257}
+
+\bibitem[\citeproctext]{ref-Borthwick2020}
+Borthwick R, de Flamingh A, Hesselbarth MHK, et al (2020) Alternative
+quantifications of landscape complementation to model gene flow in
+banded longhorn beetles {[}typocerus v. Velutinus (olivier){]}.
+Frontiers in Genetics 11:307.
+\url{https://doi.org/10.3389/fgene.2020.00307}
 
@@
 
+\bibitem[\citeproctext]{ref-Cushman2010}
+Cushman SA, Gutzweiler K, Evans JS, McGarigal K (2010) The gradient
+paradigm: A conceptual and analytical framework for landscape ecology.
+In: Cushman SA, Huettmann F (eds) Spatial complexity informatics and
+wildlife conservation. Springer International Publishing, Basel, CH, pp
+83--108
+
 \bibitem[\citeproctext]{ref-Dale1999}
@@
 
-\bibitem[\citeproctext]{ref-Etherington2015}
-Etherington TR, Holland EP, O'Sullivan D (2015) NLMpy: A python software
-package for the creation of neutral landscape models within a general
-numerical framework. Methods in Ecology and Evolution 6:164--168.
-\url{https://doi.org/10.1111/2041-210X.12308}
+\bibitem[\citeproctext]{ref-Etherington2022a}
+Etherington TR, Morgan FJ, O'Sullivan D (2022) Binary space partitioning
+generates hierarchical and rectilinear neutral landscape models suitable
+for human-dominated landscapes. Landscape Ecology 37:1761--1769.
+\url{https://doi.org/10.1007/s10980-022-01452-6}
+
+\bibitem[\citeproctext]{ref-Etherington2024}
+Etherington TR, O'Sullivan D, Perry GLW, et al (2024) A least-cost
+network neutral landscape model of human sites and routes. Landscape
+Ecology 39:52. \url{https://doi.org/10.1007/s10980-024-01836-w}
 
@@
 
+\bibitem[\citeproctext]{ref-Frazier2019}
+Frazier AE (2019) Emerging trajectories for spatial pattern analysis in
+landscape ecology. Landscape Ecology 34:2073--2082.
+\url{https://doi.org/10.1007/s10980-019-00880-1}
+
+\bibitem[\citeproctext]{ref-Frazier2017}
+Frazier AE, Kedron P (2017) Landscape metrics: Past progress and future
+directions. Current Landscape Ecology Reports 63--72.
+\url{https://doi.org/10.1007/s40823-017-0026-0}
+
+\bibitem[\citeproctext]{ref-Gao2019}
+Gao P, Li Z (2019) Computation of the boltzmann entropy of a landscape:
+A review and a generalization. Landscape Ecology 34:2183--2196.
+\url{https://doi.org/10.1007/s10980-019-00814-x}
+
 \bibitem[\citeproctext]{ref-Gao2017}
@@
 \url{https://doi.org/10.1007/s10980-017-0515-x}
+
+\bibitem[\citeproctext]{ref-Gardner1987}
+Gardner RH, Milne BT, Turnei MG, O'Neill RV (1987) Neutral models for
+the analysis of broad-scale landscape pattern. Landscape Ecology
+1:19--28. \url{https://doi.org/10.1007/BF02275262}
 
@@
 
+\bibitem[\citeproctext]{ref-Hesselbarth2019}
+Hesselbarth MHK, Sciaini M, With KA, et al (2019) Landscapemetrics: An
+open-source r tool to calculate landscape metrics. Ecography
+42:1648--1657. \url{https://doi.org/10.1111/ecog.04617}
+
 \bibitem[\citeproctext]{ref-Huckeba2023}
@@
 38:645--657. \url{https://doi.org/10.1007/s10980-022-01514-9}
+
+\bibitem[\citeproctext]{ref-Inkoom2017}
+Inkoom JN, Frank S, Greve K, Fürst C (2017) Designing neutral landscapes
+for data scarce regions in west africa. Ecological Informatics 42:1--13.
+\url{https://doi.org/10.1016/j.ecoinf.2017.08.003}
 
@@
 
+\bibitem[\citeproctext]{ref-Justeau-Allaire2022}
+Justeau-Allaire D, Blanchard G, Ibanez T, et al (2022) Fragmented
+landscape generator (flsgen): A neutral landscape generator with control
+of landscape structure and fragmentation indices. Methods in Ecology and
+Evolution 13:1412--1420. \url{https://doi.org/10.1111/2041-210X.13859}
+
+\bibitem[\citeproctext]{ref-Kedron2018}
+Kedron PJ, Frazier AE, Ovando-Montejo GA, Wang J (2018) Surface metrics
+for landscape ecology: A comparison of landscape models across
+ecoregions and scales. Landscape Ecology 33:1489--1504.
+\url{https://doi.org/10.1007/s10980-018-0685-1}
+
+\bibitem[\citeproctext]{ref-Langhammer2019}
+Langhammer M, Thober J, Lange M, et al (2019) Agricultural landscape
+generators for simulation models: A review of existing solutions and an
+outline of future directions. Ecological Modelling 393:135--151.
+\url{https://doi.org/10.1016/j.ecolmodel.2018.12.010}
+
+\bibitem[\citeproctext]{ref-Lausch2015}
+Lausch A, Blaschke T, Haase D, et al (2015) Understanding and
+quantifying landscape structure - a review on relevant process
+characteristics, data models and landscape metrics. Ecological Modelling
+295:31--41. \url{https://doi.org/10.1016/j.ecolmodel.2014.08.018}
+
+\bibitem[\citeproctext]{ref-Lechner2016}
+Lechner AM, Rhodes JR (2016) Recent progress on spatial and thematic
+resolution in landscape ecology. Current Landscape Ecology Reports
+1:98--105. \url{https://doi.org/10.1007/s40823-016-0011-z}
+
 \bibitem[\citeproctext]{ref-Li2004}
@@
 
+\bibitem[\citeproctext]{ref-Lovelace2019}
+Lovelace R, Nowosad J, Münchow J (2019) Geocomputation with r, 1st edn.
+Chapman \& Hall/CRC, Boca Raton, USA
+
 \bibitem[\citeproctext]{ref-Maceachren1985}
@@
 67:53--67. \url{https://doi.org/10.1080/04353684.1985.11879515}
+
+\bibitem[\citeproctext]{ref-McGarigal2009}
+McGarigal K, Tagil S, Cushman SA (2009) Surface metrics: An alternative
+to patch metrics for the quantification of landscape structure.
+Landscape Ecology 24:433--450.
+\url{https://doi.org/10.1007/s10980-009-9327-y}
 
@@
 
+\bibitem[\citeproctext]{ref-ONeill1992}
+O'Neill RV, Gardner RH, Turner MG (1992) A hierarchical neutral model
+for landscape analysis. Landscape Ecology 7:55--61.
+\url{https://doi.org/10.1007/BF02573957}
+
+\bibitem[\citeproctext]{ref-Palmer1992}
+Palmer MW (1992) The coexistence of species in fractal landscapes. The
+American Naturalist 139:375--397. \url{https://doi.org/10.1086/285332}
+
 \bibitem[\citeproctext]{ref-Pascual2005}
@@
 Evolution 12: \url{https://doi.org/10.24908/iee.2019.12.2.e}
+
+\bibitem[\citeproctext]{ref-Powers2019}
+Powers SM, Hampton SE (2019) Open science, reproducibility, and
+transparency in ecology. Ecological Applications 29:e01822.
+\url{https://doi.org/10.1002/eap.1822}
 
@@
 
+\bibitem[\citeproctext]{ref-Salecker2019}
+Salecker J, Dislich C, Wiegand K, et al (2019) EFForTS-LGraf: A
+landscape generator for creating smallholder-driven land-use mosaics.
+PLOS ONE 14:e0222949. \url{https://doi.org/10.1371/journal.pone.0222949}
+
 \bibitem[\citeproctext]{ref-Schroder2006}
@@
 \url{https://doi.org/10.1016/j.ecolmodel.2006.05.036}
-
-\bibitem[\citeproctext]{ref-Sciaini2018}
-Sciaini M, Fritsch M, Scherer C, Simpkins CE (2018) NLMR and
-landscapetools: An integrated environment for simulating and modifying
-neutral landscape models in r. Methods in Ecology and Evolution
-9:2240--2248. \url{https://doi.org/10.1111/2041-210X.13076}
 
@@
 
-\bibitem[\citeproctext]{ref-vonKrogh2006}
-von Krogh G, von Hippel E (2006) The promise of research on open source
-software. Management Science 52:975--983.
-\url{https://doi.org/10.1287/mnsc.1060.0560}
+\bibitem[\citeproctext]{ref-VanStrien2016}
+Van Strien MJ, Slager CTJ, De Vries B, Grêt-Regamey A (2016) An improved
+neutral landscape model for recreating real landscapes and generating
+landscape series for spatial ecological simulations. Ecology and
+Evolution 6:3808--3821. \url{https://doi.org/10.1002/ece3.2145}
 
@@
 30:51--65. \url{https://doi.org/10.1007/s10980-014-0105-0}
+
+\bibitem[\citeproctext]{ref-Wade2003}
+Wade TG, Wickham JD, Nash MS, et al (2003) A comparison of vector and
+raster GIS methods for calculating landscape metrics used in
+environmental assessments. Photogrammetric Engineering \& Remote Sensing
+69:1399--1405. \url{https://doi.org/10.14358/PERS.69.12.1399}
 

### 2024-07-01 Manuscript.tex
Rename files

@@
 
-\subsection{Surface metrics}\label{surface-metrics}
-
-Surface metrics are based on the gradient surface model using raster
-data and continuous values (McGarigal et al. 2009; Cushman et al. 2010)
-and were adapted from microscopy and molecular physics (McGarigal et al.
-2009; Kedron et al. 2018). The gradient surface model can increase the
-resemblance of the data model and the natural world because it allows to
-include more heterogeneity within each grid cell (Cushman et al. 2010;
-Kedron et al. 2018). Surface metrics are able to quantify various
-characteristics, such as roughness, skewness and kurtosis, total and
-relative amplitudes, curvatures of local peaks, or surface bearing area
-ratios (Cushman et al. 2010). Many surface metrics have analogous
-landscape metrics (Cushman et al. 2010), however, using surface metrics
-allows to reveal different or additional patterns and potentially
-pattern-process links (McGarigal et al. 2009; Borthwick et al. 2020).
-Yet, software to calculate these metrics is still rare and further
-research is needed into the specific pattern-process links and
-ecological meaningful interpretations (Kedron et al. 2018; Frazier
-2019).
-
 \subsection{Entropy}\label{entropy}
@@
 
+\subsection{Operations on spatial
+patterns}\label{operations-on-spatial-patterns}
+
+Spatial patterns can be analyzed through a range of computational
+operations, such as comparing, searching, or grouping (Turner 1990;
+Wickham and Norton 1994; Jasiewicz et al. 2015; Remmel 2020; Nowosad
+2021). These operations are based on spatial signatures which are
+multi-numerical representations of landscape pattern, and dissimilarity
+measures which are functions that quantify differences between the
+signatures.
+
+Spatial patterns can be compared between different areas, or for the
+same area at two different moments in time. Comparing spatial patterns
+is often used to analyze landscape dynamics, e.g., to detect changes in
+landscape structure over time (Duncan and Boruff 2023). Furthermore,
+pattern-based search is a one-to-many operation in which the spatial
+signature of a query area is compared to the signatures of all other
+areas. This allows to find areas with similar patterns compared to the
+query area, e.g., areas with similar environmental conditions (Jasiewicz
+et al. 2014). Finally, the grouping of spatial patterns is a
+many-to-many operation in which spatial signatures are calculated for
+all areas and clustered based on their similarity (Nowosad and Stepinski
+2021).
+
 \subsection{Landscape mosaic}\label{landscape-mosaic}
+
+\subsection{Surface metrics}\label{surface-metrics}
+
+Surface metrics are based on the gradient surface model using raster
+data and continuous values (McGarigal et al. 2009; Cushman et al. 2010)
+and were adapted from microscopy and molecular physics (McGarigal et al.
+2009; Kedron et al. 2018). The gradient surface model can increase the
+resemblance of the data model and the natural world because it allows to
+include more heterogeneity within each grid cell (Cushman et al. 2010;
+Kedron et al. 2018). Surface metrics are able to quantify various
+characteristics, such as roughness, skewness and kurtosis, total and
+relative amplitudes, curvatures of local peaks, or surface bearing area
+ratios (Cushman et al. 2010). Many surface metrics have analogous
+landscape metrics (Cushman et al. 2010), however, using surface metrics
+allows to reveal different or additional patterns and potentially
+pattern-process links (McGarigal et al. 2009; Borthwick et al. 2020).
+Yet, software to calculate these metrics is still rare and further
+research is needed into the specific pattern-process links and
+ecological meaningful interpretations (Kedron et al. 2018; Frazier
+2019).
 
@@
 
-\subsection{Operations on spatial
-patterns}\label{operations-on-spatial-patterns}
-
-Spatial patterns can be analyzed through a range of computational
-operations, such as comparing, searching, or grouping (Turner 1990;
-Wickham and Norton 1994; Jasiewicz et al. 2015; Remmel 2020; Nowosad
-2021). These operations are based on spatial signatures which are
-multi-numerical representations of landscape pattern, and dissimilarity
-measures which are functions that quantify differences between the
-signatures.
-
-Spatial patterns can be compared between different areas, or for the
-same area at two different moments in time. Comparing spatial patterns
-is often used to analyze landscape dynamics, e.g., to detect changes in
-landscape structure over time (Duncan and Boruff 2023). Furthermore,
-pattern-based search is a one-to-many operation in which the spatial
-signature of a query area is compared to the signatures of all other
-areas. This allows to find areas with similar patterns compared to the
-query area, e.g., areas with similar environmental conditions (Jasiewicz
-et al. 2014). Finally, the grouping of spatial patterns is a
-many-to-many operation in which spatial signatures are calculated for
-all areas and clustered based on their similarity (Nowosad and Stepinski
-2021).
-
 \section{Simulation models}\label{simulation-models}
@@
 
-\subsection{Neutral landscapes models}\label{neutral-landscapes-models}
+\subsection{Landscapes generators}\label{landscapes-generators}
 
-Neutral landscapes models simulate landscape patterns without assuming
-any underlying abiotic or biotic processes. Thus, they are often used as
-null hypothesis, baselines, or scenarios for comparisons with real
-landscape patterns, while controlling certain aspects of the landscape
-(Li et al. 2004; Wang and Malanson 2008). The first neutral models
-simply assigned land cover types randomly to cells in the landscape and
-were based on percolation theory (Gardner et al. 1987). Further
-developments included hierarchical models that considered different
-spatial scales while assigning land cover types (O'Neill et al. 1992).
-While these first two classes can only simulate landscape with discrete
-land cover classes, fractal neutral models can also simulate continuous
-land cover maps by utilizing e.g., Brownian motion (Palmer 1992).
-Borrowing from computer graphics, more recent neutral landscape models
-make use of spectral synthesis (e.g., Perlin noise; Etherington 2022) or
-Binary space partitioning (Etherington et al. 2022) hierarchical neutral
-landscapes. In order to simulate landscapes dominated by human
-activities approaches based on least-cost network movement exist
-(Etherington et al. 2024). Neutral landscapes models can also be used to
-simulated landscapes similar to real landscapes (Inkoom et al. 2017), or
-based on target values of landscape metrics (Van Strien et al. 2016;
-Justeau-Allaire et al. 2022).
+Landscape generators are typically used as null hypotheses, baselines,
+or scenarios in landscape ecology analyses because they allow to control
+certain aspects of the landscape (Li et al. 2004; Wang and Malanson
+2008) and are generally simpler as ecological simulations models (Hess
+et al. 2020). Landscape generators can be classified into two major
+categories, namely pattern-based and process-based approaches (Pe'er et
+al. 2013).
 
-\subsection{Landscape Generator}\label{landscape-generator}
+Pattern-based landscape generators simulate landscape patterns without
+assuming any underlying abiotic or biotic processes (i.e., neutral
+landscape model; Gardner et al. 1987). The earliest neutral models are
+based on percolation theory and assign LULC types randomly to cells in
+the landscape(Gardner et al. 1987) or include hierarchical models that
+consider different spatial scales while assigning cell values (O'Neill
+et al. 1992). Landscapes characterized by continuous values can be
+simulated by fractal models, such as Brownian motion (Palmer 1992).
+Borrowing from computer graphics and being able to simulate hierarchical
+landscapes, more recent neutral landscape models make use of spectral
+synthesis (e.g., Perlin noise; Etherington 2022) or Binary space
+partitioning (Etherington et al. 2022). Neutral landscape models are
+also able to simulate landscapes dominated by anthropogenic activities,
+e.g., based on least-cost networks (Etherington et al. 2024). In order
+to ensure realistic landscapes, several approaches exist that use real
+landscapes (Inkoom et al. 2017) or are based on target values
+optimization (Van Strien et al. 2016; Justeau-Allaire et al. 2022) to
+simulate neutral landscapes.
 
-Processes-based landscape generators are related to neutral landscapes
-models because they can simulate virtual landscapes. However, in
-contrast to neutral landscape models, processes-based landscape
-generators are able to simulate structurally more realistic landscapes
-by including pattern creating processes (Langhammer et al. 2019;
+Contrastingly, processes-based landscape generators explicitly include
+hypothesized pattern creating processes (Langhammer et al. 2019;
 Salecker et al. 2019).
@@
 
+\bibitem[\citeproctext]{ref-Hess2020}
+Hess B, Dreber N, Liu Y, et al (2020) PioLaG: A piosphere landscape
+generator for savanna rangeland modelling. Landscape Ecology
+35:2061--2082. \url{https://doi.org/10.1007/s10980-020-01066-w}
+
 \bibitem[\citeproctext]{ref-Hesselbarth2019}
@@
 \url{https://doi.org/10.1371/journal.pcbi.0010018}
+
+\bibitem[\citeproctext]{ref-Peer2013}
+Pe'er G, Zurita GA, Schober L, et al (2013) Simple process-based
+simulators for generating spatial patterns of habitat loss and
+fragmentation: A review and introduction to the g-RaFFe model. PLoS ONE
+8:e64968. \url{https://doi.org/10.1371/journal.pone.0064968}
 

### 2024-07-01 Sections/01-introduction.qmd
Rename files

---
bibliography: CLER.bib
---

<!-- Person responsible: Hesselbarth -->

# Introduction

Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].

Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012; but see @Wiersma2022].
Because landscape ecology is a cross-disciplinary field including [e.g., social sciences, geography, or ecology and evolution\; @Wiens1997; @With2019], the general increase of data availability [@Chi2016; @Jaric2020; @Wuest2020; @Nathan2022], or the studied complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].

For all computational methods the natural world must be translated into representative data models.
The spatial vector data model is based on points, lines, and polygons features that are defined by their geometry and additional attributes related to them.
Contrastingly, the raster data model is based on evenly-spaced grid cells and discrete or continuous values related to each cell [@Lovelace2019; @With2019].
The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003; @Lechner2016; @Nedd2021].

Here, we aim to introduce the latest developments of computational methods in landscape ecology.
However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
For a more comprehensive, general introductions to landscape ecology, please see @Turner2015, @Gergel2017, or @With2019.
First, we will introduce recent computational developments of the most important topics of landscape ecology.
This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations.
Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
However, due to its many advantages [@Powers2019], we will focus on open-source software , and specifically scripting languages, i.e., R, Python, and Julia.
Last, we will discuss potential future needs developments of computational methods in landscape ecology.

### 2024-07-01 Sections/05-simulations.qmd
Rename files

--- 2024-06-24 Sections/simulations.qmd
@@
 
-## Neutral landscapes models
-Neutral landscapes models simulate landscape patterns without assuming any underlying abiotic or biotic processes.
-Thus, they are often used as null hypothesis, baselines, or scenarios for comparisons with real landscape patterns, while controlling certain aspects of the landscape [@Li2004; @Wang2008].
-The first neutral models simply assigned land cover types randomly to cells in the landscape and were based on percolation theory [@Gardner1987].
-Further developments included hierarchical models that considered different spatial scales while assigning land cover types [@ONeill1992].
-While these first two classes can only simulate landscape with discrete land cover classes, fractal neutral models can also simulate continuous land cover maps by utilizing e.g., Brownian motion [@Palmer1992].
-Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or Binary space partitioning [@Etherington2022a] hierarchical neutral landscapes.
-In order to simulate landscapes dominated by human activities approaches based on least-cost network movement exist [@Etherington2024].
-Neutral landscapes models can also be used to simulated landscapes similar to real landscapes [@Inkoom2017], or based on target values of landscape metrics [@VanStrien2016; @Justeau-Allaire2022].
+## Landscapes generators
+
+Landscape generators are typically used as null hypotheses, baselines, or scenarios in landscape ecology analyses because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler as ecological simulations models [@Hess2020].
+Landscape generators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013].
+
+Pattern-based landscape generators simulate landscape patterns without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
+The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or include hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
+Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
+Borrowing from computer graphics and being able to simulate hierarchical landscapes, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or Binary space partitioning [@Etherington2022a].
+Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities, e.g., based on least-cost networks [@Etherington2024].
+In order to ensure realistic landscapes, several approaches exist that use real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022] to simulate neutral landscapes.
 <!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], or Julia (https://github.com/EcoJulia/NeutralLandscapes.jl).
 -->
+<!-- QRULE -->
 
-## Landscape Generator
-Processes-based landscape generators are related to neutral landscapes models because they can simulate virtual landscapes.
-However, in contrast to neutral landscape models, processes-based landscape generators are able to simulate structurally more realistic landscapes by including pattern creating processes [@Langhammer2019; @Salecker2019].
+Contrastingly, processes-based landscape generators explicitly include hypothesized pattern creating processes [@Langhammer2019; @Salecker2019].
+<!-- are able to simulate structurally more realistic landscapes by including -->
 
 <!--
-Cellular automaton
+Cellular automata
 [G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
-[GradientLand Software](https://doi.org/10.1016/j.ecoinf.2014.12.001)
 [PioLaG](https://doi.org/10.1007/s10980-020-01066-w)

### 2024-07-01 Sections/07-spatial-planning.qmd
Rename files

--- 2024-03-28 01-Manuscript/Manuscript.qmd
@@
 ---
-title: "Computational methods in landscape ecology"
-author:
-  - name: Maximilian H.K. Hesselbarth
-    orcid: 0000-0003-1125-9918
-    email: mhk.hesselbarth@gmail.com
-    affiliations:
-      - name: International Institute for Applied Systems Analysis (IIASA) 
-        city: Laxenburg
 bibliography: CLER.bib
-format:
-  pdf
 ---
 
+<!-- Person responsible: Jung -->
+
+# Spatial planning
+
 Landscape ecology is fun [@Turner1989].
+There are many things to explore [@With2019].

### 2024-07-02 Sections/05-simulations.qmd
Update simulations section

@@
 
-## Landscapes generators
+## Landscapes simulators
 
-Landscape generators are typically used as null hypotheses, baselines, or scenarios in landscape ecology analyses because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler as ecological simulations models [@Hess2020].
-Landscape generators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013].
+Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
+Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; Langhammer2019].
 
-Pattern-based landscape generators simulate landscape patterns without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
-The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or include hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
+Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
+The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
 Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
-Borrowing from computer graphics and being able to simulate hierarchical landscapes, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or Binary space partitioning [@Etherington2022a].
-Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities, e.g., based on least-cost networks [@Etherington2024].
-In order to ensure realistic landscapes, several approaches exist that use real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022] to simulate neutral landscapes.
-<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], or Julia (https://github.com/EcoJulia/NeutralLandscapes.jl).
--->
-<!-- QRULE -->
+Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or binary space partitioning [@Etherington2022a].
+Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024].
+In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022].
+<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->
 
-Contrastingly, processes-based landscape generators explicitly include hypothesized pattern creating processes [@Langhammer2019; @Salecker2019].
-<!-- are able to simulate structurally more realistic landscapes by including -->
+Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019; @Salecker2019].
+Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998] or deforestation [@Soares-Filho2002].
+More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006], deforestation based on road and agricultural fields access [@Peer2013], agricultural areas dominated by smallholders [@Salecker2019], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020].
 
-<!--
-Cellular automata
-[G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
-[PioLaG](https://doi.org/10.1007/s10980-020-01066-w)
--->
+## Individual-based models
 
-## IBMs /Coupled lattice
+Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014].
+In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018; @Dislich2018; @Gonzalez-Redin2019], but also disturbances [e.g., @Seidl2017a; @Spies2017; @Kamimura2019] or connectivity [e.g., @Day2020].
+In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018] or reusable building block [@Berger2024].
+Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016], e.g., linking an individual-based and ecosystem model [@Polhill2011].
 
@@
 <!--
-[RangeShifter](https://doi.org/10.1111/ecog.05687)
+[Digital twins](https://doi.org/10.1016/j.tree.2023.04.010) 
+[Rural ecological landscape control](https://doi.org/10.1186/s12302-024-00888-8)
+[Urban Expansion and Vegetation Coverage](https://doi.org/10.3389/fpls.2022.840471)
+[Digital Twins in agriculture](https://doi.org/10.1016/j.cosust.2022.101252)
+[Future landscape visualization](https://doi.org/10.1093/jcde/qwac032)
+[Digital Twin of the Agricultural Landscape](https://doi.org/10.14627/537690019)
+-->
+
+<!--
 [Marine SDMS](https://doi.org/10.1016/j.ecolmodel.2019.108837)
@@
 [Simulations in landscape genetics](https://doi.org/10.1111/j.1365-294X.2010.04678.x)
-[HexSim modeling environment](https://doi.org/10.1007/s10980-017-0605-9)
-[Digital twins?](https://doi.org/10.1016/j.tree.2023.04.010)
 -->

### 2024-07-02 Manuscript.qmd
Rename files
render MS

--- 2024-05-28 Manuscript.qmd
@@
 papersize: a4paper
-fontsize: 12pt
+mainfont: "Garamond"
+fontsize: 11pt
 geometry: margin=25mm
@@
 
-{{< include Sections/introduction.qmd >}}
-{{< include Sections/simulations.qmd >}}
+{{< include Sections/01-introduction.qmd >}}
+{{< include Sections/02-spatial-patterns.qmd >}}
+{{< include Sections/05-simulations.qmd >}}
 
@@
 
-### References
+### References {.unnumbered}
 
+\begingroup
+\raggedright
 ::: {#refs}
 :::
+\endgroup

### 2024-07-02 Manuscript.tex
render MS

@@
 \documentclass[
-  12pt,
+  11pt,
   a4paperpaper,
@@
 
-\subsection{Landscapes generators}\label{landscapes-generators}
-
-Landscape generators are typically used as null hypotheses, baselines,
-or scenarios in landscape ecology analyses because they allow to control
-certain aspects of the landscape (Li et al. 2004; Wang and Malanson
-2008) and are generally simpler as ecological simulations models (Hess
-et al. 2020). Landscape generators can be classified into two major
-categories, namely pattern-based and process-based approaches (Pe'er et
-al. 2013).
-
-Pattern-based landscape generators simulate landscape patterns without
-assuming any underlying abiotic or biotic processes (i.e., neutral
-landscape model; Gardner et al. 1987). The earliest neutral models are
-based on percolation theory and assign LULC types randomly to cells in
-the landscape(Gardner et al. 1987) or include hierarchical models that
-consider different spatial scales while assigning cell values (O'Neill
-et al. 1992). Landscapes characterized by continuous values can be
-simulated by fractal models, such as Brownian motion (Palmer 1992).
-Borrowing from computer graphics and being able to simulate hierarchical
-landscapes, more recent neutral landscape models make use of spectral
-synthesis (e.g., Perlin noise; Etherington 2022) or Binary space
-partitioning (Etherington et al. 2022). Neutral landscape models are
-also able to simulate landscapes dominated by anthropogenic activities,
-e.g., based on least-cost networks (Etherington et al. 2024). In order
-to ensure realistic landscapes, several approaches exist that use real
-landscapes (Inkoom et al. 2017) or are based on target values
-optimization (Van Strien et al. 2016; Justeau-Allaire et al. 2022) to
-simulate neutral landscapes.
-
-Contrastingly, processes-based landscape generators explicitly include
-hypothesized pattern creating processes (Langhammer et al. 2019;
-Salecker et al. 2019).
-
-\subsection{IBMs /Coupled lattice}\label{ibms-coupled-lattice}
+\subsection{Landscapes simulators}\label{landscapes-simulators}
+
+Landscape simulators are typically used as null hypotheses, baselines,
+or scenarios because they allow to control certain aspects of the
+landscape (Li et al. 2004; Wang and Malanson 2008) and are generally
+simpler than ecological simulations models (Hess et al. 2020). Landscape
+simulators can be classified into two major categories, namely
+pattern-based and process-based approaches {[}Pe'er et al. (2013);
+Langhammer2019{]}.
+
+Pattern-based approaches simulate landscapes without assuming any
+underlying abiotic or biotic processes (i.e., neutral landscape model;
+Gardner et al. 1987). The earliest neutral models are based on
+percolation theory and assign LULC types randomly to cells in the
+landscape(Gardner et al. 1987) or hierarchical models that consider
+different spatial scales while assigning cell values (O'Neill et al.
+1992). Landscapes characterized by continuous values can be simulated by
+fractal models, such as Brownian motion (Palmer 1992). Borrowing from
+computer graphics, more recent neutral landscape models make use of
+spectral synthesis (e.g., Perlin noise; Etherington 2022) or binary
+space partitioning (Etherington et al. 2022). Neutral landscape models
+are also able to simulate landscapes dominated by anthropogenic
+activities using based on least-cost networks (Etherington et al. 2024).
+In order to ensure realistic neutral landscapes several approaches exist
+that use comparisons with real landscapes (Inkoom et al. 2017) or are
+based on target values optimization (Van Strien et al. 2016;
+Justeau-Allaire et al. 2022).
+
+Contrastingly, processes-based approaches explicitly include abiotic or
+biotic pattern-forming processes (i.e., landscape generators; Langhammer
+et al. 2019; Salecker et al. 2019). Several earlier landscape generators
+are based on cellular automata models and are able to simulate urban
+growth (Clarke and Gaydos 1998) or deforestation (Soares-Filho et al.
+2002). More recent landscape generators allow to simulate patchy
+landscapes based on Gibbs processes (Gaucherel et al. 2006),
+deforestation based on road and agricultural fields access (Pe'er et al.
+2013), agricultural areas dominated by smallholders (Salecker et al.
+2019), or vegetation surround watering points in semi-arid savanna
+rangelands (combining pattern- and process-bases approaches; Hess et al.
+2020).
+
+\subsection{Individual-based models}\label{individual-based-models}
+
+Individual-based models (or agent-based models) simulate discrete
+entities that are described by attributes and behaviors and patterns
+emerge from bottom-up interactions of individuals with each other and
+their environment (DeAngelis and Grimm 2014). In landscape ecology
+individual-based models are increasingly used to model especially
+social-ecological systems (e.g, Carauta et al. 2018; Dislich et al.
+2018; Gonzalez-Redin et al. 2019), but also disturbances (e.g., Seidl
+and Rammer 2017; Spies et al. 2017; Kamimura et al. 2019) or
+connectivity (e.g., Day et al. 2020). In order to facilitate future
+model developments, recent progress includes adaptable modeling
+frameworks (Schumaker and Brookes 2018) or reusable building block
+(Berger et al. 2024). Furthermore, hybrid approaches that couple or even
+integrated different model types with individual-based models could
+benefit from rich development history or facilitate corresponding
+strengths (O'Sullivan et al. 2016), e.g., linking an individual-based
+and ecosystem model (Polhill et al. 2011).
 
@@
 
-\subsubsection{References}\label{references}
+\subsubsection*{References}\label{references}
+\addcontentsline{toc}{subsubsection}{References}
+
+\begingroup
+\raggedright
 
@@
 
+\bibitem[\citeproctext]{ref-Berger2024}
+Berger U, Bell A, Barton CM, et al (2024) Towards reusable building
+blocks for agent-based modelling and theory development. Environmental
+Modelling \& Software 175:106003.
+\url{https://doi.org/10.1016/j.envsoft.2024.106003}
+
 \bibitem[\citeproctext]{ref-Borthwick2020}
@@
 
+\bibitem[\citeproctext]{ref-Carauta2018}
+Carauta M, Latynskiy E, Mössinger J, et al (2018) Can preferential
+credit programs speed up the adoption of low-carbon agricultural systems
+in mato grosso, brazil? Results from bioeconomic microsimulation.
+Regional Environmental Change 18:117--128.
+\url{https://doi.org/10.1007/s10113-017-1104-x}
+
 \bibitem[\citeproctext]{ref-Chi2016}
@@
 Springer Berlin Heidelberg, Berlin, Heidelberg, pp 221--230
+
+\bibitem[\citeproctext]{ref-Clarke1998}
+Clarke KC, Gaydos LJ (1998) Loose-coupling a cellular automaton model
+and GIS: Long-term urban growth prediction for san francisco and
+washington/baltimore. International Journal of Geographical Information
+Science 12:699--714. \url{https://doi.org/10.1080/136588198241617}
 
@@
 
+\bibitem[\citeproctext]{ref-Day2020}
+Day CC, Zollner PA, Gilbert JH, McCann NP (2020) Individual-based
+modeling highlights the importance of mortality and landscape structure
+in measures of functional connectivity. Landscape Ecology 35:2191--2208.
+\url{https://doi.org/10.1007/s10980-020-01095-5}
+
+\bibitem[\citeproctext]{ref-DeAngelis2014}
+DeAngelis DL, Grimm V (2014) Individual-based models in ecology after
+four decades. F1000Prime Reports 6:1--6.
+\url{https://doi.org/10.12703/P6-39}
+
+\bibitem[\citeproctext]{ref-Dislich2018}
+Dislich C, Hettig E, Salecker J, et al (2018) Land-use change in oil
+palm dominated tropical landscapes - an agent-based model to explore
+ecological and socio-economic trade-offs. PLoS ONE 13:e0190506.
+\url{https://doi.org/10.1371/journal.pone.0190506}
+
 \bibitem[\citeproctext]{ref-Duncan2023}
@@
 
+\bibitem[\citeproctext]{ref-Gaucherel2006}
+Gaucherel C, Fleury D, Auclair D, Dreyfus P (2006) Neutral models for
+patchy landscapes. Ecological Modelling 197:159--170.
+\url{https://doi.org/10.1016/j.ecolmodel.2006.02.044}
+
 \bibitem[\citeproctext]{ref-Gergel2017}
@@
 ecology}. Springer New York, New York, NY
+
+\bibitem[\citeproctext]{ref-Gonzalez-Redin2019}
+Gonzalez-Redin J, Gordon IJ, Hill R, et al (2019) Exploring sustainable
+land use in forested tropical social-ecological systems: A case-study in
+the wet tropics. Journal of Environmental Management 231:940--952.
+\url{https://doi.org/10.1016/j.jenvman.2018.10.079}
 
@@
 
+\bibitem[\citeproctext]{ref-Kamimura2019}
+Kamimura K, Gardiner B, Dupont S, Finnigan J (2019) Agent-based
+modelling of wind damage processes and patterns in forests. Agricultural
+and Forest Meteorology 268:279--288.
+\url{https://doi.org/10.1016/j.agrformet.2019.01.020}
+
 \bibitem[\citeproctext]{ref-Kedron2018}
@@
 
+\bibitem[\citeproctext]{ref-OSullivan2016}
+O'Sullivan D, Evans T, Manson S, et al (2016) Strategic directions for
+agent-based modeling: Avoiding the YAAWN syndrome. Journal of Land Use
+Science 11:177--187. \url{https://doi.org/10.1080/1747423X.2015.1030463}
+
 \bibitem[\citeproctext]{ref-Palmer1992}
@@
 
+\bibitem[\citeproctext]{ref-Polhill2011}
+Polhill JG, Gimona A, Aspinall RJ (2011) Agent-based modelling of land
+use effects on ecosystem processes and services. Journal of Land Use
+Science 6:75--81. \url{https://doi.org/10.1080/1747423X.2011.558603}
+
 \bibitem[\citeproctext]{ref-Powers2019}
@@
 
+\bibitem[\citeproctext]{ref-Schumaker2018}
+Schumaker NH, Brookes A (2018) HexSim: A modeling environment for
+ecology and conservation. Landscape Ecology 33:197--211.
+\url{https://doi.org/10.1007/s10980-017-0605-9}
+
+\bibitem[\citeproctext]{ref-Seidl2017a}
+Seidl R, Rammer W (2017) Climate change amplifies the interactions
+between wind and bark beetle disturbances in forest landscapes.
+Landscape Ecology 32:1485--1498.
+\url{https://doi.org/10.1007/s10980-016-0396-4}
+
 \bibitem[\citeproctext]{ref-Shannon1948}
@@
 \url{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}
+
+\bibitem[\citeproctext]{ref-Soares-Filho2002}
+Soares-Filho BS, Coutinho Cerqueira G, Lopes Pennachin C (2002)
+Dinamica---a stochastic cellular automata model designed to simulate the
+landscape dynamics in an amazonian colonization frontier. Ecological
+Modelling 154:217--235.
+\url{https://doi.org/10.1016/S0304-3800(02)00059-5}
+
+\bibitem[\citeproctext]{ref-Spies2017}
+Spies TA, White E, Ager A, et al (2017) Using an agent-based model to
+examine forest management outcomes in a fire-prone landscape in oregon,
+USA. Ecology and Society 22:art25.
+\url{https://doi.org/10.5751/ES-08841-220125}
 
@@
 
+\endgroup
+
 \end{document}

### 2024-07-11 Manuscript.qmd
Update MS and software

@@
 papersize: a4paper
-mainfont: "Garamond"
-fontsize: 11pt
+mainfont: "Verdana"
+fontsize: 10pt
 geometry: margin=25mm
@@
 {{< include Sections/05-simulations.qmd >}}
+{{< include Sections/08-software.qmd >}}
 

### 2024-07-11 Manuscript.tex
Update MS and software

@@
 \documentclass[
-  11pt,
+  10pt,
   a4paperpaper,
@@
     % xetex/luatex font selection
-  \setmainfont[]{Garamond}
+    \setmainfont[]{Verdana}
 \fi
@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \makeatother
+
 \ifLuaTeX
@@
 
+\section{Software and programming
+languages}\label{software-and-programming-languages}
+
+``Open science'' involves full transparency and data access at all
+stages of the research processes and the integrative and
+multi-disciplinary field of ecology can massively benefit from this
+concept (Hampton et al. 2015). One key aspect of achieving open science
+is to use open source software, which ensures transparency and
+consequently increasing the (public) trust into the analyses (Powers and
+Hampton 2019). Related to this, script-based workflows have the
+advantage over workflows involving graphical user interfaces that the
+former are shareable and reproducible (Hampton et al. 2015; Powers and
+Hampton 2019). Thus, we will focus the following section on script-based
+open-source software suitable for landscape ecology. Given the
+increasing availability of data and complexity of methods, future
+progress within the field of landscape ecology will depend on the
+software and analysis tools available (Costanza et al. 2019)
+
+The open-source and script-based R programming languages (R Core Team
+2024) was first released in 1995 (Smith 2016) with a focus on
+statistical computing and data visualization. However, due to its
+extensions (also called packages), the language is popular in many
+academic fields including ecology (Lai et al. 2019) and it is suitable
+for spatial data analysis or landscape ecology (e.g., Bivand 2020;
+Hesselbarth et al. 2021).
+
 \newpage{}
@@
 
+\bibitem[\citeproctext]{ref-Bivand2020}
+Bivand RS (2020) Progress in the r ecosystem for representing and
+handling spatial data. Journal of Geographical Systems.
+\url{https://doi.org/10.1007/s10109-020-00336-0}
+
 \bibitem[\citeproctext]{ref-Borthwick2020}
@@
 
+\bibitem[\citeproctext]{ref-Hampton2015}
+Hampton SE, Anderson SS, Bagby SC, et al (2015) The tao of open science
+for ecology. Ecosphere 6:1--13.
+\url{https://doi.org/10.1890/ES14-00402.1}
+
 \bibitem[\citeproctext]{ref-Hess2020}
@@
 35:2061--2082. \url{https://doi.org/10.1007/s10980-020-01066-w}
+
+\bibitem[\citeproctext]{ref-Hesselbarth2021}
+Hesselbarth MHK, Nowosad J, Signer J, Graham LJ (2021) Open-source tools
+in r for landscape ecology. Current Landscape Ecology Reports 6:97--111.
+\url{https://doi.org/10.1007/s40823-021-00067-y}
 
@@
 
+\bibitem[\citeproctext]{ref-Lai2019}
+Lai J, Lortie CJ, Muenchen RA, et al (2019) Evaluating the popularity of
+r in ecology. Ecosphere 10: \url{https://doi.org/10.1002/ecs2.2567}
+
 \bibitem[\citeproctext]{ref-Langhammer2019}
@@
 
+\bibitem[\citeproctext]{ref-RCoreTeam2024}
+R Core Team (2024) \href{https://www.R-project.org/}{R: A language and
+environment for statistical computing}. R Foundation for Statistical
+Computing, Vienna, Austria
+
 \bibitem[\citeproctext]{ref-Rao1982}
@@
 \url{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}
+
+\bibitem[\citeproctext]{ref-Smith2016}
+Smith D (2016) Over 16 years of r project history
 

### 2024-07-11 Sections/03-connectivity.qmd
Update MS and software

---
bibliography: CLER.bib
---

<!-- Person responsible: -->

# Connectivity / Movement / Dispersal

<!-- 
https://link.springer.com/article/10.1007/s10980-023-01741-8
https://link.springer.com/article/10.1007/s10980-022-01504-x?fromPaywallRec=false
https://link.springer.com/article/10.1007/s10980-019-00935-3?fromPaywallRec=false
https://link.springer.com/article/10.1038/s41598-022-20370-w?fromPaywallRec=true
-->

### 2024-07-11 Sections/08-software.qmd
Update MS and software

---
bibliography: CLER.bib
---

<!-- Person responsible: Hesselbarth, Nowosad, Bosch, Jung, Schouten  -->

# Software and programming languages

<!-- "The future applications of landscape pattern metrics will almost certainly depend on the software and analysis tools available" -- @costanzaDescribingAnalyzingLandscape2019 -->

"Open science" involves full transparency and data access at all stages of the research processes and the integrative and multi-disciplinary field of ecology can massively benefit from this concept [@Hampton2015].
One key aspect of achieving open science is to use open source software, which ensures transparency and consequently increasing the (public) trust into the analyses [@Powers2019].
Related to this, script-based workflows have the advantage over workflows involving graphical user interfaces that the former are shareable and reproducible [@Hampton2015; @Powers2019].
Thus, we will focus the following section on script-based open-source software suitable for landscape ecology.
Given the increasing availability of data and complexity of methods, future progress within the field of landscape ecology will depend on the software and analysis tools available [@Costanza2019]

<!-- R: Hesselbarth -->
The open-source and script-based R programming languages [@RCoreTeam2024] was first released in 1995 [@Smith2016] with a focus on statistical computing and data visualization.
However, due to its extensions (also called packages), the language is popular in many academic fields including ecology [@Lai2019] and it is suitable for spatial data analysis or landscape ecology [e.g., @Bivand2020; @Hesselbarth2021].

<!-- 
landscapemetrics
geodiv
vectormetrics (also: VecLI -- they mention open source in their paper a few times, but the software seems to be closed source -- https://urbancomp.net/archives/vecliv3beta)
bespatial
LandComp
motif
?GLCMTextures
?rasterdiv
?terra 
-->

<!-- Python: Bosch -->

<!--Julia: Schouten -->

<!--Others: -->

<!-- 
GuidosToolbox
?GRASS GIS
?QGIS: LecoS plugin 
-->

### 2024-07-12 Sections/04-sampling.qmd
Moving potential literature to github issue

---
bibliography: CLER.bib
---

<!-- Person responsible: Bosch -->

# Sampling in landscape ecology

### 2024-07-12 Sections/05-simulations.qmd
Moving potential literature to github issue

@@
 ## Digital twins
-
-<!--
-[Digital twins](https://doi.org/10.1016/j.tree.2023.04.010) 
-[Rural ecological landscape control](https://doi.org/10.1186/s12302-024-00888-8)
-[Urban Expansion and Vegetation Coverage](https://doi.org/10.3389/fpls.2022.840471)
-[Digital Twins in agriculture](https://doi.org/10.1016/j.cosust.2022.101252)
-[Future landscape visualization](https://doi.org/10.1093/jcde/qwac032)
-[Digital Twin of the Agricultural Landscape](https://doi.org/10.14627/537690019)
--->
-
-<!--
-[Marine SDMS](https://doi.org/10.1016/j.ecolmodel.2019.108837)
-[SDM past present future](https://doi.org/10.1111/jbi.14617)
-[Simulations in landscape genetics](https://doi.org/10.1111/j.1365-294X.2010.04678.x)
--->

### 2024-07-12 Manuscript.qmd
Update digital twins section

@@
 papersize: a4paper
-mainfont: "Verdana"
 fontsize: 10pt

### 2024-07-12 Manuscript.tex
Moving potential literature to github issue
Update digital twins section

@@
     % xetex/luatex font selection
-    \setmainfont[]{Verdana}
 \fi
@@
 % Make \paragraph and \subparagraph free-standing
-\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-\makeatother
 
@@
 \makeatother
-
 \ifLuaTeX
@@
 
+Digital Twins include three connected elements, namely a physical object
+or system, its digital representation, and a data exchange between the
+former two (Jones et al. 2020; De Koning et al. 2023) and there is an
+increasing interest and use of them in both the industry and academia
+(Jones et al. 2020). The need for timely and evidence-based decision
+making in combination with increasing data availability makes them a
+powerful tool for ecology (De Koning et al. 2023). Digital twins are
+increasingly applied in agricultural landscapes research, including
+topics such as crop and land cultivation, livestock farming, controlled
+environment farming, or fertilization management (Cesco et al. 2023;
+Purcell et al. 2023). Further applications include planning of rural
+ecological landscapes (Tan and Cheng 2024), or exploration of
+relationships between urban expansion and vegetation coverage (Zhao et
+al. 2022). The use of general data processing, statistical and
+mechanistic models, or artificial intelligence allows them to constantly
+update the digital representation and analyze, visualize, or predict the
+physical counterpart. Thus, Digital Twins can provide real-time state of
+nature measurements, early detection of conservation trends,
+relationships of ecosystem trends and environmental conditions, impact
+assessments of interventions, or identify uncertainties and information
+gaps (De Koning et al. 2023).
+
 \section{Software and programming
@@
 
+\bibitem[\citeproctext]{ref-Cesco2023}
+Cesco S, Sambo P, Borin M, et al (2023) Smart agriculture and digital
+twins: Applications and challenges in a vision of sustainability.
+European Journal of Agronomy 146:126809.
+\url{https://doi.org/10.1016/j.eja.2023.126809}
+
 \bibitem[\citeproctext]{ref-Chi2016}
@@
 
+\bibitem[\citeproctext]{ref-DeKoning2023}
+De Koning K, Broekhuijsen J, Kühn I, et al (2023) Digital twins: Dynamic
+model-data fusion for ecology. Trends in Ecology \& Evolution
+38:916--926. \url{https://doi.org/10.1016/j.tree.2023.04.010}
+
 \bibitem[\citeproctext]{ref-DeAngelis2014}
@@
 
+\bibitem[\citeproctext]{ref-Jones2020}
+Jones D, Snider C, Nassehi A, et al (2020) Characterising the digital
+twin: A systematic literature review. CIRP Journal of Manufacturing
+Science and Technology 29:36--52.
+\url{https://doi.org/10.1016/j.cirpj.2020.02.002}
+
 \bibitem[\citeproctext]{ref-Justeau-Allaire2022}
@@
 \url{https://doi.org/10.1371/journal.pcbi.1002802}
+
+\bibitem[\citeproctext]{ref-Purcell2023}
+Purcell W, Neubauer T, Mallinger K (2023) Digital twins in agriculture:
+Challenges and opportunities for environmental sustainability. Current
+Opinion in Environmental Sustainability 61:101252.
+\url{https://doi.org/10.1016/j.cosust.2022.101252}
 
@@
 
+\bibitem[\citeproctext]{ref-Tan2024}
+Tan F, Cheng Y (2024) A digital twin framework for innovating rural
+ecological landscape control. Environmental Sciences Europe 36:59.
+\url{https://doi.org/10.1186/s12302-024-00888-8}
+
 \bibitem[\citeproctext]{ref-Turner1989}
@@
 
+\bibitem[\citeproctext]{ref-Zhao2022}
+Zhao D, Li X, Wang X, et al (2022) Applying digital twins to research
+the relationship between urban expansion and vegetation coverage: A case
+study of natural preserve. Frontiers in Plant Science 13:840471.
+\url{https://doi.org/10.3389/fpls.2022.840471}
+
 \bibitem[\citeproctext]{ref-Zhao2019}

### 2024-07-12 Sections/05-simulations.qmd
Update digital twins section

@@
 ## Digital twins
+
+Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020; @DeKoning2023] and there is an increasing interest and use of them in both the industry and academia [@Jones2020].
+The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@DeKoning2023].
+Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023; @Purcell2023].
+Further applications include planning of rural ecological landscapes [@Tan2024], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022].
+The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
+Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@DeKoning2023].

### 2024-08-19 Manuscript.qmd
adds my second affiliation

@@
       - name: Adam Mickiewicz University, Poznan, Poland
+      - name: University of Münster, Münster, Germany
   - name: Marti Bosch

### 2024-08-28 Sections/02-spatial-patterns.qmd
Rename files
Moving potential literature to github issue

--- 2024-06-24 Sections/spatial-patterns.qmd
@@
 <!-- jn: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
-<!-- MH: I think we decided to not reference software here but in a seperate section -->
+<!-- MH: I think we decided to not reference software here but in a separate section -->
+<!-- MH: Alternatively, if we move the software section to the beginning, we could reference software here -->
 
@@
 However, spatial patterns are not only a result of processes, but could also be a driver of them.
-Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989].
+Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989Landscape].
 
@@
 <!-- https://fragstats.org/ -->
-
-## Surface metrics
-<!-- Bosch, Hesselbarth -->
-
-Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009; @Cushman2010] and were adapted from microscopy and molecular physics [@McGarigal2009; @Kedron2018].
-The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010; @Kedron2018].
-Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010].
-Many surface metrics have analogous landscape metrics [@Cushman2010], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009; @Borthwick2020].
-Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018; @Frazier2019].
-<!-- MH: What about Spectral and Wavelet Analysis?
-
-<!-- https://www.digitalsurf.com/free-trial/ -->
-<!-- https://github.com/bioXgeo/geodiv -->
 
@@
 
+## Operations on spatial patterns
+
+Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990; @Wickham1994; @Jasiewicz2015; @Remmel2020; @Nowosad2021a].
+These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between the signatures.
+
+Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
+Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023].
+Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
+This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014].
+Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021b].
+
+<!-- https://github.com/Nowosad/motif/ -->
+<!-- https://jakubnowosad.com/spquery/ -->
+<!-- FOSS4G Europe pap? -->
+
 ## Landscape mosaic
 <!-- Bosch -->
+
+<!-- https://doi.org/10.1371/journal.pone.0304215 -->
+Despite the extensive use of the landscape mosaic model and the landscape metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004Use; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
+<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
+<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
+<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation.
+Let us go through them in a different order: -->
+In line with Dramstad [@Dramstad2009], we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation.
+
+### Sensitivity to spatial scale and thematic resolution
+
+<!-- (i) scale: https://doi.org/10.1007/s10980-014-0066-3 "Behavior of class-level landscape metrics across gradients of class aggregation and area"-->
+<!-- "identifying the characteristic scales of relevant patterns and processes is a critical first step in designing a successful research project" [@Wu2007, pp. 119]. -->
+Identifying the characteristic scale at which the organisms or processes of interest operate is a critical first step towards establishing sound relationships between pattern and process [@Levin1992; @Wu2007; @Cushman2008Landscape].
+To that end, a key requirement it to understand the sensitivity of landscape metrics to spatial scale in terms of both resolution (grain) and extent.
+
+A key concept to consider regarding grain size is the minimum mapping unit (MMU), namely the smallest area entity that can be mapped and analyzed as a discrete patch.
+<!-- Accordingly, a grain size several times smaller than the smallest patch in order to grasp (even schematically) its shape. -->
+Since the input maps should represent not only the location of patches but also their shape, a grain size several times smaller than the smallest patch (i.e., the MMU) in order to grasp (even schematically) its shape.
+Even though the MMU is often assumed as a natural characteristic of the dataset, it has an unsurprising impact on the computed landscape metrics, especially in terms of the number of patches and their size distribution [@Saura2002].
+<!-- On the other hand, landscape metrics such as the landscape division index, largest patch index and area-weighted mean shape indices show more predictable responses to changing the MMU. -->
+
+<!-- Since the applications of landscape metrics are manifold, the choices of grain and especially extent vary largely depending on the context of the study. -->
+<!-- The growing availability of global remote sensing products at homogenized resolution has resulted in a myriad of studies sharing the same resolution, e.g., based on Landsat scenes at 30 m.
+At the same time, the choice of spatial extent varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient.
+-->
+While the choice of grain is often constrained by the available data, e.g., Landsat scenes at 30 m resolution, the selection of spatial extent and the way that it is sampled varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient.
+At the same time, the choice of spatial extent and sampling strategy is often subject to many arbitrary criteria that can largely influence the ultimate results.
+Such an issue has long been recognized in physical geography as the modifiable areal unit problem (MAUP) [@Openshaw1979; @Openshaw1984].
+<!-- The MAUP has two interrelated components, namely the scale problem and the zoning problem -->
+<!-- A first approach to dealing with the MAUP is to identify individual entities that are ecologically meaninful and not modifiable [@Jelinski1996].
+In the landscape matrix model, such entities could be target patches of interest.
+However, it is not always possible to identify basic entities, especially since ecological processes of interest typically involve the integration of spatially distributed data.
+-->
+Approaches to dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation and the analysis of the sensitivity of the zoning configuration [@Jelinski1996; @Duque2018; @Fletcher2023].
+<!-- TODO: use a moving window measurement device to "postpone" the MAUP [@Riitters2019] -->
+
+TODO: section about approaches to sample landscapes to compute metrics and available software tools, e.g., zonal analysis with polygons (e.g., administrative limits), grids, transects or buffers along a landscape gradient, moving windows [@Bosch2019; @Hesselbarth2019; @Huais2024]?
+
+<!-- While selecting the adequate scale should be based on the phenomenon of interest, researchers are often constrained by the availability of data. -->
+<!-- Therefore, a thorough understanding of these relationships is essential in order to compare spatial patterns measured at different scales. -->
+<!-- Several types of responses of landscape metrics to changing scale have been identified in computational experiments by changing both the spatial resolution (grain) and extent. -->
+The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales.
+For instance, some metrics related to the number and density of patches and edges show consistent linear and power-law scaling relationships to changing grain across a wide diversity of simulated and real landscapes, whereas other metrics exhibit less predictable responses [@Turner1989Effects; @Saura2001; @Wu2002Empirical; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
+<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->
+<!-- also multi-scale is key to habitat assessment, see "Multi-scale habitat selection modeling: a review and outlook" -->
+<!-- Predictable scaling relationships can be exploited to estimate spatial patterns at higher resolutions -->
+Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004].
+For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches, resulting in unreliable landscape metric computations.
+In order to address scale issues and advance the science of scaling in landscape ecology, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
+As suggested by Frazier [@Frazier2023], this can be done following three key steps, namely by considering scope guidelines when designing experiments, explicitly reporting grain and extent so that scope can be calculated allowing quantitative comparability with other studes and undertaking comparisons and replications based on scope.
+The latter is showcased in a proof of concept example experiment, where landscapes from the same ecoregion are sampled using different grain and extent and then the distributions of the computed landscape metrics are visualized through ridgeline plots colored by scope [@Frazier2023].
+
+TODO: section about landscape metric rescaling [@Saura2007; @Arganaraz2014; @Frazier2014; @Frazier2016; @Frazier2021]?
+
+<!-- Overall, robust metrics seem more desirable because (i) they improve comparability between studies and (ii) the relationships between pattern and process occur at multiple scales -->
+<!-- which highlights the importance of considering multiple scales when attempting to measure landscape patterns. -->
+
+TODO: mention hierarchichal patch dynamics [@Wu1999]?
+
+<!-- (ii) choice of LULC scheme/thematic resolution and image processing/classification procedure: https://doi.org/10.1016/j.foreco.2009.12.016 https://doi.org/10.1016/j.isprsjprs.2007.03.004 https://doi.org/10.1007/s10021-005-0119-1 https://doi.org/10.1007/s10980-006-9035-9 https://doi.org/10.1007/s10980-022-01546-1 -->
+<!-- Scale and heterogeneity are inherently linked (Wu 2007), so it is not surprising that the results indicate downscaling accuracy varies with heterogeneity. -->
+TODO: sensitivity to thematic resolution [@Huang2006; @Bailey2007; @Buyantuyev2007; @Castilla2009].
+
+### Redundancy
+<!-- (iii) selecting metrics -->
+
+TODO
+
+- Factor analysis of Riitters et al.
+[@Riitters1995] found six main components: average patch compaction, image texture, average patch shape, patch perimeter-area scaling, number of attribute classes and large-patch density-area scaling.
+
+- Cushman et al.
+[@Cushman2008Parsimony] not only isolate the components but also assess their strength (proportion of variance explained), universality (how present they are accross landscapes) and consistency (how stable they are accross landscapes).
+At the class level, they find seven "highly universal and consistent structure components" (page 700), namely edge contrast, patch shape complexity, aggregation, nearest neighbor distance, patch dispersion, large patch dominance and neighborhood similarity.
+At the landscape level, five components are identified, namely contagion/diversity (similar to image texture in Riitters et al.
+[@Riitters1995], large patch dominance, interspersion/juxtaposition, edge contrast, patch shape variability, proximity and nearest neighbor distance.
+Overall, components are more consistent and universal at the landscape level than at the class level.
+
+- Uuemaa et al.
+[@Uuemaa2011] find four main components: dominance (also identified by Riitters et al.
+[@Riitters1995] and Cushman et al.
+[@Cushman2008Parsimony]), contrast, shape complexity and composition.
+
+- Amount (composition) and adjacency (configuration) are the fundamental elements of pattern [@Riitters2019]
+
+- Entropy [@Nowosad2019a]
+
+- Spatial signatures [@Nowosad2021a]
+
+### Interpretability
+<!-- (iv) interpretatability -->
+
+TODO
+- do landscape metrics measure fragmentation?
+- normalized metrics [@Wang2011]
+- [@Frazier2021]
+
+<!-- CLOSING THOUGHTS AND BRIDGING WITH THE NEXT SECTION -->
+Responses:
+- incorporating aspects of ecological function into landscape metrics (e.g., core area, least-cost distances) and moving towards graph-based and circuit theory approaches [@Kupfer2012].
+- point-data analysis [@Gustafson1998] and gradient surface model [@Lausch2015] require fewer assumptions regarding LULC categorization schemes or spatial scale
+
+<!-- Note that all approaches (landscape metrics, graph and surface - TODO: what about point?) suffer from correlation and redundancy issues, scale dependency and target problem [@Frazier2017]. -->
+
+## Surface metrics
+<!-- Bosch, Hesselbarth -->
+
+Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009; @Cushman2010] and were adapted from microscopy and molecular physics [@McGarigal2009; @Kedron2018].
+The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010; @Kedron2018].
+Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010].
+Many surface metrics have analogous landscape metrics [@Cushman2010], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009; @Borthwick2020].
+Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018; @Frazier2019].
+<!-- MH: What about Spectral and Wavelet Analysis?
+
+<!-- https://www.digitalsurf.com/free-trial/ -->
+<!-- https://github.com/bioXgeo/geodiv -->
 
@@
 <!-- https://github.com/r-spatialecology/vectormetrics -->
-
-## Operations on spatial patterns
-
-Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990; @Wickham1994; @Jasiewicz2015; @Remmel2020; @Nowosad2021a].
-These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between the signatures.
-
-Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
-Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023].
-Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
-This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014].
-Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021b].
-
-<!-- https://github.com/Nowosad/motif/ -->
-<!-- https://jakubnowosad.com/spquery/ -->
-<!-- FOSS4G Europe pap? -->

### 2024-08-28 Manuscript.tex
adds my second affiliation
Update bib

@@
               }
+          \affil{%
+                  University of Münster, Münster, Germany
+              }
         \author{Marti Bosch}
@@
 \subsection{Landscape mosaic}\label{landscape-mosaic}
+
+Despite the extensive use of the landscape mosaic model and the
+landscape metrics based on it, major limitations have been identified
+regarding both its ability to quantify the spatial structure of
+landscapes and their relationship to ecological processes (Gustafson
+1998; Li and Wu 2004; Dramstad 2009; Kupfer 2012; Lausch et al. 2015;
+Frazier and Kedron 2017). In line with Dramstad (Dramstad 2009), we
+review four major issues of landscape metrics, namely their sensitivity
+to both spatial scale and thematic resolution, redundancy and
+interpretation.
+
+\subsubsection{Sensitivity to spatial scale and thematic
+resolution}\label{sensitivity-to-spatial-scale-and-thematic-resolution}
+
+Identifying the characteristic scale at which the organisms or processes
+of interest operate is a critical first step towards establishing sound
+relationships between pattern and process (Levin 1992; Wu 2007; Cushman
+and McGarigal 2008). To that end, a key requirement it to understand the
+sensitivity of landscape metrics to spatial scale in terms of both
+resolution (grain) and extent.
+
+A key concept to consider regarding grain size is the minimum mapping
+unit (MMU), namely the smallest area entity that can be mapped and
+analyzed as a discrete patch. Since the input maps should represent not
+only the location of patches but also their shape, a grain size several
+times smaller than the smallest patch (i.e., the MMU) in order to grasp
+(even schematically) its shape. Even though the MMU is often assumed as
+a natural characteristic of the dataset, it has an unsurprising impact
+on the computed landscape metrics, especially in terms of the number of
+patches and their size distribution (Saura 2002).
+
+While the choice of grain is often constrained by the available data,
+e.g., Landsat scenes at 30 m resolution, the selection of spatial extent
+and the way that it is sampled varies largely depending on the
+phenomenon of interest, e.g., from large transects of forests in natural
+parks to small buffers along the urban-rural gradient. At the same time,
+the choice of spatial extent and sampling strategy is often subject to
+many arbitrary criteria that can largely influence the ultimate results.
+Such an issue has long been recognized in physical geography as the
+modifiable areal unit problem (MAUP) (Openshaw 1979, 1984). Approaches
+to dealing with the MAUP in the context of landscape metrics include the
+identification of individual entities that are ecologically meaningful
+and not modifiable, the evaluation of ``optimal'' zoning systems that
+maximize interzonal variations and minimize intrazonal variation and the
+analysis of the sensitivity of the zoning configuration (Jelinski and Wu
+1996; Duque et al. 2018; Fletcher Jr et al. 2023).
+
+TODO: section about approaches to sample landscapes to compute metrics
+and available software tools, e.g., zonal analysis with polygons (e.g.,
+administrative limits), grids, transects or buffers along a landscape
+gradient, moving windows (Bosch 2019; Hesselbarth et al. 2019; Huais
+2024)?
+
+The scale effects of landscape metrics can be assessed by computing
+scalograms, i.e., the response curves of metrics to changing scales. For
+instance, some metrics related to the number and density of patches and
+edges show consistent linear and power-law scaling relationships to
+changing grain across a wide diversity of simulated and real landscapes,
+whereas other metrics exhibit less predictable responses (Turner et al.
+1989; Saura and Martínez-Millán 2001; Wu et al. 2002; Wu 2004; Saura and
+Castro 2007; Šı́mová and Gdulová 2012; Argañaraz and Entraigas 2014;
+Frazier et al. 2021). Nonetheless, the effects of changing the sampling
+extent can be less predictable than those of changing grain size (Wu
+2004). For instance, in landscapes with small extent to grain ratio,
+boundary effects are likely to dominate and truncate patches, resulting
+in unreliable landscape metric computations. In order to address scale
+issues and advance the science of scaling in landscape ecology, studies
+should explicitly consider the concept of scope, namely the ratio of
+extent to resolution (Schneider 2001). As suggested by Frazier (Frazier
+2023), this can be done following three key steps, namely by considering
+scope guidelines when designing experiments, explicitly reporting grain
+and extent so that scope can be calculated allowing quantitative
+comparability with other studes and undertaking comparisons and
+replications based on scope. The latter is showcased in a proof of
+concept example experiment, where landscapes from the same ecoregion are
+sampled using different grain and extent and then the distributions of
+the computed landscape metrics are visualized through ridgeline plots
+colored by scope (Frazier 2023).
+
+TODO: section about landscape metric rescaling (Saura and Castro 2007;
+Argañaraz and Entraigas 2014; Frazier 2014, 2016; Frazier et al. 2021)?
+
+TODO: mention hierarchichal patch dynamics (Wu 1999)?
+
+TODO: sensitivity to thematic resolution (Huang et al. 2006; Bailey et
+al. 2007; Buyantuyev and Wu 2007; Castilla et al. 2009).
+
+\subsubsection{Redundancy}\label{redundancy}
+
+TODO
+
+\begin{itemize}
+\item
+  Factor analysis of Riitters et al. (Riitters et al. 1995) found six
+  main components: average patch compaction, image texture, average
+  patch shape, patch perimeter-area scaling, number of attribute classes
+  and large-patch density-area scaling.
+\item
+  Cushman et al. (Cushman et al. 2008) not only isolate the components
+  but also assess their strength (proportion of variance explained),
+  universality (how present they are accross landscapes) and consistency
+  (how stable they are accross landscapes). At the class level, they
+  find seven ``highly universal and consistent structure components''
+  (page 700), namely edge contrast, patch shape complexity, aggregation,
+  nearest neighbor distance, patch dispersion, large patch dominance and
+  neighborhood similarity. At the landscape level, five components are
+  identified, namely contagion/diversity (similar to image texture in
+  Riitters et al. (Riitters et al. 1995), large patch dominance,
+  interspersion/juxtaposition, edge contrast, patch shape variability,
+  proximity and nearest neighbor distance. Overall, components are more
+  consistent and universal at the landscape level than at the class
+  level.
+\item
+  Uuemaa et al. (Uuemaa et al. 2011) find four main components:
+  dominance (also identified by Riitters et al. (Riitters et al. 1995)
+  and Cushman et al. (Cushman et al. 2008)), contrast, shape complexity
+  and composition.
+\item
+  Amount (composition) and adjacency (configuration) are the fundamental
+  elements of pattern (Riitters 2019)
+\item
+  Entropy (Nowosad and Stepinski 2019)
+\item
+  Spatial signatures (Nowosad 2021)
+\end{itemize}
+
+\subsubsection{Interpretability}\label{interpretability}
+
+TODO - do landscape metrics measure fragmentation? - normalized metrics
+(Wang and Cumming 2011) - (Frazier et al. 2021)
+
+Responses: - incorporating aspects of ecological function into landscape
+metrics (e.g., core area, least-cost distances) and moving towards
+graph-based and circuit theory approaches (Kupfer 2012). - point-data
+analysis (Gustafson 1998) and gradient surface model (Lausch et al.
+2015) require fewer assumptions regarding LULC categorization schemes or
+spatial scale
 
@@
 
+\bibitem[\citeproctext]{ref-Arganaraz2014}
+Argañaraz JP, Entraigas I (2014) Scaling functions evaluation for
+estimation of landscape metrics at higher resolutions. Ecological
+informatics 22:1--12
+
+\bibitem[\citeproctext]{ref-Bailey2007}
+Bailey D, Billeter R, Aviron S, et al (2007) The influence of thematic
+resolution on metric selection for biodiversity monitoring in
+agricultural landscapes. Landscape Ecology 22:461--473
+
 \bibitem[\citeproctext]{ref-Basaraner2017}
@@
 
+\bibitem[\citeproctext]{ref-Bosch2019}
+Bosch M (2019) PyLandStats: An open-source pythonic library to compute
+landscape metrics. PLoS One 14:e0225734
+
+\bibitem[\citeproctext]{ref-Buyantuyev2007}
+Buyantuyev A, Wu J (2007) Effects of thematic resolution on landscape
+pattern analysis. Landscape Ecology 22:7--13
+
 \bibitem[\citeproctext]{ref-Carauta2018}
@@
 
+\bibitem[\citeproctext]{ref-Castilla2009}
+Castilla G, Larkin K, Linke J, Hay GJ (2009) The impact of thematic
+resolution on the patch-mosaic model of natural landscapes. Landscape
+Ecology 24:15--23
+
 \bibitem[\citeproctext]{ref-Cesco2023}
@@
 
+\bibitem[\citeproctext]{ref-Cushman2008Landscape}
+Cushman SA, McGarigal K (2008) Landscape metrics, scales of resolution.
+In: Designing green landscapes. Springer, pp 33--51
+
+\bibitem[\citeproctext]{ref-Cushman2008Parsimony}
+Cushman SA, McGarigal K, Neel MC (2008) Parsimony in landscape metrics:
+Strength, universality, and consistency. Ecological indicators
+8:691--703
+
 \bibitem[\citeproctext]{ref-Dale1999}
@@
 
+\bibitem[\citeproctext]{ref-Dramstad2009}
+Dramstad WE (2009) Spatial metrics--useful indicators for society or
+mainly fun tools for landscape ecologists? Norsk Geografisk
+Tidsskrift-Norwegian Journal of Geography 63:246--254
+
 \bibitem[\citeproctext]{ref-Duncan2023}
@@
 
+\bibitem[\citeproctext]{ref-Duque2018}
+Duque JC, Laniado H, Polo A (2018) S-maup: Statistical test to measure
+the sensitivity to the modifiable areal unit problem. PloS one
+13:e0207377
+
 \bibitem[\citeproctext]{ref-Etherington2022}
@@
 
+\bibitem[\citeproctext]{ref-Fletcher2023}
+Fletcher Jr RJ, Betts MG, Damschen EI, et al (2023) Addressing the
+problem of scale that emerges with habitat fragmentation. Global Ecology
+and Biogeography 32:828--841
+
 \bibitem[\citeproctext]{ref-Forman1995}
@@
 
+\bibitem[\citeproctext]{ref-Frazier2014}
+Frazier AE (2014) A new data aggregation technique to improve landscape
+metric downscaling. Landscape ecology 29:1261--1276
+
+\bibitem[\citeproctext]{ref-Frazier2016}
+Frazier AE (2016) Surface metrics: Scaling relationships and downscaling
+behavior. Landscape ecology 31:351--363
+
+\bibitem[\citeproctext]{ref-Frazier2023}
+Frazier AE (2023) Scope and its role in advancing a science of scaling
+in landscape ecology. Landscape Ecology 38:637--643
+
 \bibitem[\citeproctext]{ref-Frazier2017}
@@
 \url{https://doi.org/10.1007/s40823-017-0026-0}
+
+\bibitem[\citeproctext]{ref-Frazier2021}
+Frazier AE, Kedron P, Ovando-Montejo GA, Zhao Y (2021) Scaling spatial
+pattern metrics: Impacts of composition and configuration on downscaling
+accuracy. Landscape Ecology 1--16
 
@@
 
+\bibitem[\citeproctext]{ref-Gustafson1998}
+Gustafson EJ (1998) Quantifying landscape spatial pattern: What is the
+state of the art? Ecosystems 1:143--156
+
 \bibitem[\citeproctext]{ref-Hampton2015}
@@
 42:1648--1657. \url{https://doi.org/10.1111/ecog.04617}
+
+\bibitem[\citeproctext]{ref-Huais2024}
+Huais PY (2024) Multilandr: An r package for multi-scale landscape
+analysis. Landscape Ecology 39:140
+
+\bibitem[\citeproctext]{ref-Huang2006}
+Huang C, Geiger E, Kupfer J (2006) Sensitivity of landscape metrics to
+classification scheme. International Journal of Remote Sensing
+27:2927--2948
 
@@
 
+\bibitem[\citeproctext]{ref-Jelinski1996}
+Jelinski DE, Wu J (1996) The modifiable areal unit problem and
+implications for landscape ecology. Landscape ecology 11:129--140
+
 \bibitem[\citeproctext]{ref-Jones2020}
@@
 
+\bibitem[\citeproctext]{ref-Kupfer2012}
+Kupfer JA (2012) Landscape ecology and biogeography: Rethinking
+landscape metrics in a post-FRAGSTATS landscape. Progress in physical
+geography 36:400--420
+
 \bibitem[\citeproctext]{ref-Lai2019}
@@
 
+\bibitem[\citeproctext]{ref-Levin1992}
+Levin SA (1992) The problem of pattern and scale in ecology: The robert
+h. MacArthur award lecture. Ecology 73:1943--1967
+
+\bibitem[\citeproctext]{ref-Li2004Use}
+Li H, Wu J (2004) Use and misuse of landscape indices. Landscape ecology
+19:389--399
+
 \bibitem[\citeproctext]{ref-Li2004}
@@
 
+\bibitem[\citeproctext]{ref-Openshaw1979}
+Openshaw S (1979) A million or so correlated coefficients: Three
+experiment on the modifiable areal unit problem. Statistical
+applications in the spatial sciences
+
+\bibitem[\citeproctext]{ref-Openshaw1984}
+Openshaw S (1984) Ecological fallacies and the analysis of areal census
+data. Environment and planning A 16:17--31
+
 \bibitem[\citeproctext]{ref-Palmer1992}
@@
 
+\bibitem[\citeproctext]{ref-Riitters2019}
+Riitters K (2019) Pattern metrics for a transdisciplinary landscape
+ecology. Landscape Ecology 34:2057--2063
+
+\bibitem[\citeproctext]{ref-Riitters1995}
+Riitters KH, O'neill R, Hunsaker C, et al (1995) A factor analysis of
+landscape pattern and structure metrics. Landscape ecology 10:23--39
+
 \bibitem[\citeproctext]{ref-Risser1984}
@@
 
+\bibitem[\citeproctext]{ref-Saura2002}
+Saura S (2002) Effects of minimum mapping unit on land cover data
+spatial configuration and composition. International Journal of Remote
+Sensing 23:4853--4880
+
+\bibitem[\citeproctext]{ref-Saura2007}
+Saura S, Castro S (2007) Scaling functions for landscape pattern metrics
+derived from remotely sensed data: Are their subpixel estimates really
+accurate? ISPRS Journal of Photogrammetry and Remote sensing 62:201--216
+
+\bibitem[\citeproctext]{ref-Saura2001}
+Saura S, Martínez-Millán J (2001) Sensitivity of landscape pattern
+metrics to map spatial extent. Photogrammetric Engineering \& Remote
+Sensing 67:1027--1036
+
+\bibitem[\citeproctext]{ref-Schneider2001}
+Schneider DC (2001) The rise of the concept of scale in ecology: The
+concept of scale is evolving from verbal expression to quantitative
+expression. BioScience 51:545--553
+
 \bibitem[\citeproctext]{ref-Schroder2006}
@@
 \url{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}
+
+\bibitem[\citeproctext]{ref-Simova2012}
+Šı́mová P, Gdulová K (2012) Landscape indices behavior: A review of scale
+effects. Applied geography 34:385--394
 
@@
 
-\bibitem[\citeproctext]{ref-Turner1989}
+\bibitem[\citeproctext]{ref-Turner1989Landscape}
 Turner MG (1989) Landscape ecology: The effect of pattern on process.
@@
 Pattern and process, 2nd edition. Springer, New York
+
+\bibitem[\citeproctext]{ref-Turner1989Effects}
+Turner MG, O'Neill RV, Gardner RH, Milne BT (1989) Effects of changing
+spatial scale on the analysis of landscape pattern. Landscape ecology
+3:153--162
+
+\bibitem[\citeproctext]{ref-Uuemaa2011}
+Uuemaa E, Roosaare J, Oja T, Mander Ü (2011) Analysing the spatial
+structure of the estonian landscapes: Which landscape metrics are the
+most suitable for comparing different landscapes? Estonian Journal of
+Ecology 60:70
 
@@
 
+\bibitem[\citeproctext]{ref-Wang2011}
+Wang X, Cumming SG (2011) Measuring landscape configuration with
+normalized metrics. Landscape Ecology 26:723--736
+
 \bibitem[\citeproctext]{ref-Wickham1994}
@@
 
-\bibitem[\citeproctext]{ref-Wu2002}
+\bibitem[\citeproctext]{ref-Wu1999}
+Wu J (1999) Hierarchy and scaling: Extrapolating information along a
+scaling ladder. Canadian journal of remote sensing 25:367--380
+
+\bibitem[\citeproctext]{ref-Wu2004}
+Wu J (2004) Effects of changing scale on landscape pattern analysis:
+Scaling relations. Landscape ecology 19:125--138
+
+\bibitem[\citeproctext]{ref-Wu2007}
+Wu J (2007) Scale and scaling: A cross-disciplinary perspective. In: Key
+topics in landscape ecology. Cambridge University Press, pp 115--142
+
+\bibitem[\citeproctext]{ref-Wu2002Key}
 Wu J, Hobbs R (2002) Key issues and research priorities in landscape
@@
 \url{https://doi.org/10.1023/A:1020561630963}
+
+\bibitem[\citeproctext]{ref-Wu2002Empirical}
+Wu J, Shen W, Sun W, Tueller PT (2002) Empirical patterns of the effects
+of changing scale on landscape metrics. Landscape ecology 17:761--782
 

### 2024-08-28 Sections/01-introduction.qmd
Update bib

@@
 Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
-Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
-Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].
+Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989Landscape; @Turner2005].
+Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002Key; @Wu2013].
 

### 2024-08-29 Manuscript.tex
Updating software section and tiny cosmetic changes

@@
 simulators can be classified into two major categories, namely
-pattern-based and process-based approaches {[}Pe'er et al. (2013);
-Langhammer2019{]}.
+pattern-based and process-based approaches (Pe'er et al. 2013;
+Langhammer et al. 2019).
 
@@
 ``Open science'' involves full transparency and data access at all
-stages of the research processes and the integrative and
-multi-disciplinary field of ecology can massively benefit from this
-concept (Hampton et al. 2015). One key aspect of achieving open science
-is to use open source software, which ensures transparency and
-consequently increasing the (public) trust into the analyses (Powers and
-Hampton 2019). Related to this, script-based workflows have the
-advantage over workflows involving graphical user interfaces that the
-former are shareable and reproducible (Hampton et al. 2015; Powers and
-Hampton 2019). Thus, we will focus the following section on script-based
-open-source software suitable for landscape ecology. Given the
-increasing availability of data and complexity of methods, future
-progress within the field of landscape ecology will depend on the
-software and analysis tools available (Costanza et al. 2019)
-
-The open-source and script-based R programming languages (R Core Team
-2024) was first released in 1995 (Smith 2016) with a focus on
-statistical computing and data visualization. However, due to its
-extensions (also called packages), the language is popular in many
-academic fields including ecology (Lai et al. 2019) and it is suitable
-for spatial data analysis or landscape ecology (e.g., Bivand 2020;
-Hesselbarth et al. 2021).
+stages of the research processes and integrative and multi-disciplinary
+fields such as ecology can massively benefit from this concept (Hampton
+et al. 2015). Using open source software is one key aspect of open
+science because it ensures transparency and consequently increases trust
+in the analyses (Powers and Hampton 2019). Given the increasing data
+availability and method complexity, future progress within the research
+field will highly depend on available software and analysis tools
+(Costanza et al. 2019). In this context, script-based workflows have an
+advantage over workflows involving graphical user interfaces because the
+former are more shareable and reproducible (Hampton et al. 2015; Powers
+and Hampton 2019). Thus, we will focus the following section on
+script-based open-source software suitable for landscape ecology.
+
+The open-source, script-based R programming languages (R Core Team 2024)
+was first released in 1995 (Smith 2016) with a focus on statistical
+computing and data visualization. However, due to its extensions (also
+called packages), the language is popular in many academic fields,
+including ecology (Lai et al. 2019; Lortie et al. 2020), and it is
+suitable for spatial data analysis and landscape ecology (e.g., Bivand
+2020; Hesselbarth et al. 2021). Fundamental packages for spatial data
+analysis include \emph{sf} for simple-features vector data (Pebesma
+2018), \emph{terra} for raster data (Hijmans 2024), \emph{stars} for
+spatiotemporal arrays, raster, and vector data cubes (Pebesma and Bivand
+2023), and \emph{spatstat} for point pattern data (Baddeley et al.
+2015). Several packages allow to create maps, such as \emph{tmap}
+(Tennekes 2018) or \emph{ggplot2} (Wickham 2016). For a comprehensive
+overview of packages related to geographical spatial data, please see
+the CRAN Task View (Bivand and Nowosad 2024).
+
+R packages developed specifically for spatial pattern analysis include,
+besides others, \emph{landscapemetrics} (Hesselbarth et al. 2019),
+\emph{vectormetrics} {[}still being developed; Matuszek et al.
+(2024){]}, \emph{motif} (Nowosad 2021), \emph{rasterdiv} (Rocchini et
+al. 2021), or \emph{geodiv} (Smith et al. 2021). Packages with a focus
+on scale include e.g., \emph{multilandr} (Huais 2024) or
+\emph{grainchanger} (Graham 2024). Connectivity analysis can be
+conducted using packages including \emph{lconnect} (Mestre and Silva
+2023), or \emph{grainscape} (Chubaty et al. 2020). Furthermore, neutral
+landscapes can be simulated using \emph{NLMR} (Sciaini et al. 2018), and
+the \emph{graph4lg} (Savary et al. 2021) provides functions for
+landscape genetics.
 
@@
 
+\bibitem[\citeproctext]{ref-Baddeley2015}
+Baddeley A, Rubak E, Turner R (2015) Spatial point patterns: Methodology
+and applications with r. Chapman \& Hall/CRC, London, UK
+
 \bibitem[\citeproctext]{ref-Bailey2007}
@@
 
+\bibitem[\citeproctext]{ref-Bivand2024}
+Bivand RS, Nowosad J (2024) CRAN task view: Analysis of spatial data.
+{https://CRAN.R-project.org/view=Spatial}
+
 \bibitem[\citeproctext]{ref-Borthwick2020}
@@
 
+\bibitem[\citeproctext]{ref-Chubaty2020a}
+Chubaty AM, Galpern P, Doctolero SC (2020) The r toolbox grainscape for
+modelling and visualizing landscape connectivity using spatially
+explicit networks. Methods in Ecology and Evolution 11:591--595.
+\url{https://doi.org/10.1111/2041-210X.13350}
+
 \bibitem[\citeproctext]{ref-Claramunt2012}
@@
 
+\bibitem[\citeproctext]{ref-Graham2024}
+Graham L (2024) Grainchanger: Moving-window and direct data aggregation.
+{https://ropensci.github.io/grainchanger/}
+
 \bibitem[\citeproctext]{ref-Grimm1996}
@@
 42:1648--1657. \url{https://doi.org/10.1111/ecog.04617}
+
+\bibitem[\citeproctext]{ref-Hijmans2024}
+Hijmans RJ (2024) Terra: Spatial data analysis. R package version
+1.7-78. {\(<\)}{https://CRAN.R-project.org/package=terra\(>\)}
 
@@
 
-\bibitem[\citeproctext]{ref-Li2004Use}
+\bibitem[\citeproctext]{ref-Li2004b}
 Li H, Wu J (2004) Use and misuse of landscape indices. Landscape ecology
@@
 
-\bibitem[\citeproctext]{ref-Li2004}
+\bibitem[\citeproctext]{ref-Li2004a}
 Li X, He HS, Wang X, et al (2004) Evaluating the effectiveness of
@@
 
+\bibitem[\citeproctext]{ref-Lortie2020}
+Lortie CJ, Braun J, Filazzola A, Miguel F (2020) A checklist for
+choosing between r packages in ecology and evolution. Ecology and
+Evolution 10:1098--1105. \url{https://doi.org/10.1002/ece3.5970}
+
 \bibitem[\citeproctext]{ref-Lovelace2019}
@@
 67:53--67. \url{https://doi.org/10.1080/04353684.1985.11879515}
+
+\bibitem[\citeproctext]{ref-Matuszek2024}
+Matuszek T, Nowosad J, Sciaini M, et al (2024) Vectormetrics: Landscape
+metrics for categorical map patterns in vector data.
+{https://github.com/r-spatialecology/vectormetrics}
 
@@
 
+\bibitem[\citeproctext]{ref-Mestre2023}
+Mestre F, Silva B (2023) Lconnect r package: A versatile tool for
+evaluating landscape connectivity and prioritizing habitat patches in
+conservation research. Ecological Modelling 484:110489.
+\url{https://doi.org/10.1016/j.ecolmodel.2023.110489}
+
 \bibitem[\citeproctext]{ref-Nathan2022}
@@
 
+\bibitem[\citeproctext]{ref-Pebesma2018}
+Pebesma E (2018) Simple features for r: Standardized support for spatial
+vector data. The R Journal 10:439.
+\url{https://doi.org/10.32614/RJ-2018-009}
+
+\bibitem[\citeproctext]{ref-Pebesma2023}
+Pebesma E, Bivand R (2023)
+\href{https://doi.org/10.1201/9780429459016}{Spatial data science: With
+applications in r}, 1st edn. {Chapman and Hall/CRC}, New York
+
 \bibitem[\citeproctext]{ref-Peck2004}
@@
 
+\bibitem[\citeproctext]{ref-Rocchini2021}
+Rocchini D, Thouverai E, Marcantonio M, et al (2021) Rasterdiv---an
+information theory tailored r package for measuring ecosystem
+heterogeneity from space: To the origin and back. Methods in Ecology and
+Evolution 12:1093--1102. \url{https://doi.org/10.1111/2041-210X.13583}
+
 \bibitem[\citeproctext]{ref-Salecker2019}
@@
 Sensing 67:1027--1036
+
+\bibitem[\citeproctext]{ref-Savary2021}
+Savary P, Foltête J-C, Moal H, et al (2021) graph4lg: A package for
+constructing and analysing graphs for landscape genetics in r. Methods
+in Ecology and Evolution 12:539--547.
+\url{https://doi.org/10.1111/2041-210X.13530}
 
@@
 
+\bibitem[\citeproctext]{ref-Sciaini2018}
+Sciaini M, Fritsch M, Scherer C, Simpkins CE (2018) NLMR and
+landscapetools: An integrated environment for simulating and modifying
+neutral landscape models in r. Methods in Ecology and Evolution
+9:2240--2248. \url{https://doi.org/10.1111/2041-210X.13076}
+
 \bibitem[\citeproctext]{ref-Seidl2017a}
@@
 effects. Applied geography 34:385--394
+
+\bibitem[\citeproctext]{ref-Smith2021b}
+Smith AC, Dahlin KM, Record S, et al (2021) The {\textsc{geodiv r}}
+package: Tools for calculating gradient surface metrics. Methods in
+Ecology and Evolution 12:2094--2100.
+\url{https://doi.org/10.1111/2041-210X.13677}
 
@@
 
-\bibitem[\citeproctext]{ref-Turner1989Landscape}
+\bibitem[\citeproctext]{ref-Tennekes2018}
+Tennekes M (2018) Tmap: Thematic maps in r. Journal of Statistical
+Software 84: \url{https://doi.org/10.18637/jss.v084.i06}
+
+\bibitem[\citeproctext]{ref-Turner1989a}
 Turner MG (1989) Landscape ecology: The effect of pattern on process.
@@
 
-\bibitem[\citeproctext]{ref-Turner1989Effects}
+\bibitem[\citeproctext]{ref-Turner1989b}
 Turner MG, O'Neill RV, Gardner RH, Milne BT (1989) Effects of changing
@@
 
+\bibitem[\citeproctext]{ref-Wickham2016}
+Wickham H (2016) ggplot2: Elegant graphics for data analysis. Springer
+New York, New York, USA
+
 \bibitem[\citeproctext]{ref-Wickham1994}
@@
 
-\bibitem[\citeproctext]{ref-Wu2002Key}
+\bibitem[\citeproctext]{ref-Wu2002b}
 Wu J, Hobbs R (2002) Key issues and research priorities in landscape
@@
 
-\bibitem[\citeproctext]{ref-Wu2002Empirical}
+\bibitem[\citeproctext]{ref-Wu2002a}
 Wu J, Shen W, Sun W, Tueller PT (2002) Empirical patterns of the effects

### 2024-08-29 Sections/01-introduction.qmd
Updating software section and tiny cosmetic changes

--- 2024-07-01 Sections/01-introduction.qmd
@@
 Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
-Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
-Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].
+Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989a; @Turner2005].
+Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002b; @Wu2013].
 

### 2024-08-29 Sections/02-spatial-patterns.qmd
Updating software section and tiny cosmetic changes

@@
 
-<!-- Spatial patterns are a fundamental concept not only in landscape ecology, but also in many other fields, such as geography, biology, urban planning, and even medical studies. -->
 Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996].
@@
 However, spatial patterns are not only a result of processes, but could also be a driver of them.
-Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989Landscape].
+Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989a].
 
@@
 <!-- https://doi.org/10.1371/journal.pone.0304215 -->
-Despite the extensive use of the landscape mosaic model and the landscape metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004Use; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
+Despite the extensive use of the landscape mosaic model and the landscape metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004b; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
 <!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
@@
 The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales.
-For instance, some metrics related to the number and density of patches and edges show consistent linear and power-law scaling relationships to changing grain across a wide diversity of simulated and real landscapes, whereas other metrics exhibit less predictable responses [@Turner1989Effects; @Saura2001; @Wu2002Empirical; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
+For instance, some metrics related to the number and density of patches and edges show consistent linear and power-law scaling relationships to changing grain across a wide diversity of simulated and real landscapes, whereas other metrics exhibit less predictable responses [@Turner1989b; @Saura2001; @Wu2002a; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
 <!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->

### 2024-08-29 Sections/05-simulations.qmd
Updating software section and tiny cosmetic changes

@@
 
-Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
-Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; Langhammer2019].
+Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004a; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
+Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; @Langhammer2019].
 

### 2024-08-29 Sections/08-software.qmd
Updating software section and tiny cosmetic changes

@@
 
-<!-- "The future applications of landscape pattern metrics will almost certainly depend on the software and analysis tools available" -- @costanzaDescribingAnalyzingLandscape2019 -->
-
-"Open science" involves full transparency and data access at all stages of the research processes and the integrative and multi-disciplinary field of ecology can massively benefit from this concept [@Hampton2015].
-One key aspect of achieving open science is to use open source software, which ensures transparency and consequently increasing the (public) trust into the analyses [@Powers2019].
-Related to this, script-based workflows have the advantage over workflows involving graphical user interfaces that the former are shareable and reproducible [@Hampton2015; @Powers2019].
+"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields such as ecology can massively benefit from this concept [@Hampton2015].
+Using open source software is one key aspect of open science because it ensures transparency and consequently increases trust in the analyses [@Powers2019].
+Given the increasing data availability and method complexity, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019].
+In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015; @Powers2019].
 Thus, we will focus the following section on script-based open-source software suitable for landscape ecology.
-Given the increasing availability of data and complexity of methods, future progress within the field of landscape ecology will depend on the software and analysis tools available [@Costanza2019]
 
 <!-- R: Hesselbarth -->
-The open-source and script-based R programming languages [@RCoreTeam2024] was first released in 1995 [@Smith2016] with a focus on statistical computing and data visualization.
-However, due to its extensions (also called packages), the language is popular in many academic fields including ecology [@Lai2019] and it is suitable for spatial data analysis or landscape ecology [e.g., @Bivand2020; @Hesselbarth2021].
+The open-source, script-based R programming languages [@RCoreTeam2024] was first released in 1995 [@Smith2016] with a focus on statistical computing and data visualization.
+However, due to its extensions (also called packages), the language is popular in many academic fields, including ecology [@Lai2019; @Lortie2020], and it is suitable for spatial data analysis and landscape ecology [e.g., @Bivand2020; @Hesselbarth2021].
+Fundamental packages for spatial data analysis include *sf* for simple-features vector data [@Pebesma2018], *terra* for raster data [@Hijmans2024], *stars* for spatiotemporal arrays, raster, and vector data cubes [@Pebesma2023], and *spatstat* for point pattern data [@Baddeley2015].
+Several packages allow to create maps, such as *tmap* [@Tennekes2018] or *ggplot2* [@Wickham2016]. 
+For a comprehensive overview of packages related to geographical spatial data, please see the CRAN Task View [@Bivand2024].
 
-<!-- 
-landscapemetrics
-geodiv
-vectormetrics (also: VecLI -- they mention open source in their paper a few times, but the software seems to be closed source -- https://urbancomp.net/archives/vecliv3beta)
-bespatial
-LandComp
-motif
-?GLCMTextures
-?rasterdiv
-?terra 
--->
+R packages developed specifically for spatial pattern analysis include, besides others, *landscapemetrics* [@Hesselbarth2019], *vectormetrics* [still being developed; @Matuszek2024], *motif* [@Nowosad2021a], *rasterdiv* [@Rocchini2021], or *geodiv* [@Smith2021b].
+<!-- BESSpatial: https://github.com/Nowosad/bespatial/ -->
+Packages with a focus on scale include e.g., *multilandr* [@Huais2024] or *grainchanger* [@Graham2024].
+Connectivity analysis can be conducted using packages including *lconnect* [@Mestre2023], or *grainscape* [@Chubaty2020a].
+Furthermore, neutral landscapes can be simulated using *NLMR* [@Sciaini2018], and the *graph4lg* [@Savary2021] provides functions for landscape genetics.
 
@@
 <!--Julia: Schouten -->
-
-<!--Others: -->
-
-<!-- 
-GuidosToolbox
-?GRASS GIS
-?QGIS: LecoS plugin 
--->

### 2024-08-29 Sections/02-spatial-patterns.qmd
Adding comments to spatial patterns section

@@
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961] and Gibbs entropies <!-- REF -->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
+The Renyi [@Renyi1961] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
 The Rao quadratic entropy [@Rao1982] has also been applied recently [@Rocchini2017], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
@@
 
+<!-- MH: Is the landscape mosaic after e.g., Vogt et al.
+10.1371/journal.pone.0304215 the same as the patch mosaic model and landscape metrics?
+I think there is some confusion here.
+We already discuss metrics above.
+I think it would be more valuable to discuss the LLM and the most recent additions after Vogt et al.
+here?
+-->
+
 <!-- https://doi.org/10.1371/journal.pone.0304215 -->
-Despite the extensive use of the landscape mosaic model and the landscape metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004b; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
+Despite the extensive use of the landscape mosaic model and metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004b; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
 <!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
@@
 Let us go through them in a different order: -->
-In line with Dramstad [@Dramstad2009], we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation.
+Here, we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation [@Dramstad2009].
 
@@
 
+<!-- MH: Suggestion is to focus much more on computational issues here.
+Currently, this is more about landscape metrics and their behavior in general.
+I like the point about growing availability of remote sensing products.
+Maybe we can focus a bit more on that and cut a lot of the more general, theoretical behavior parts?
+-->
+
 <!-- (i) scale: https://doi.org/10.1007/s10980-014-0066-3 "Behavior of class-level landscape metrics across gradients of class aggregation and area"-->
 <!-- "identifying the characteristic scales of relevant patterns and processes is a critical first step in designing a successful research project" [@Wu2007, pp. 119]. -->
-Identifying the characteristic scale at which the organisms or processes of interest operate is a critical first step towards establishing sound relationships between pattern and process [@Levin1992; @Wu2007; @Cushman2008Landscape].
-To that end, a key requirement it to understand the sensitivity of landscape metrics to spatial scale in terms of both resolution (grain) and extent.
+Identifying scales at which the ecological processes of interest operate is a critical first step towards establishing relationships between pattern and process [@Levin1992; @Wu2007; @Cushman2008Landscape].
+To that end, a key requirement is to understand the sensitivity of landscape metrics to spatial scale in terms of both grain and extent.
 
@@
 <!-- Accordingly, a grain size several times smaller than the smallest patch in order to grasp (even schematically) its shape. -->
-Since the input maps should represent not only the location of patches but also their shape, a grain size several times smaller than the smallest patch (i.e., the MMU) in order to grasp (even schematically) its shape.
-Even though the MMU is often assumed as a natural characteristic of the dataset, it has an unsurprising impact on the computed landscape metrics, especially in terms of the number of patches and their size distribution [@Saura2002].
+Since the input maps should represent not only the location of patches but also their shape, the grain size should be several times smaller than the smallest patch (i.e., the MMU) in order to quantify shapes.
+Even though the MMU is often assumed as a natural characteristic of the dataset, it has an impact on the computed landscape metrics, especially on the number of patches and their size distribution [@Saura2002].
 <!-- On the other hand, landscape metrics such as the landscape division index, largest patch index and area-weighted mean shape indices show more predictable responses to changing the MMU. -->
@@
 -->
-While the choice of grain is often constrained by the available data, e.g., Landsat scenes at 30 m resolution, the selection of spatial extent and the way that it is sampled varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient.
-At the same time, the choice of spatial extent and sampling strategy is often subject to many arbitrary criteria that can largely influence the ultimate results.
-Such an issue has long been recognized in physical geography as the modifiable areal unit problem (MAUP) [@Openshaw1979; @Openshaw1984].
+While the choice of grain is often constrained by the available data, e.g., Landsat scenes at 30 m resolution, the selection of spatial extent varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient.
+<!-- REF -->
+The spatial extent and sampling strategy is often subject to many arbitrary criteria that can largely influence results.
+This issue has long been recognized in physical geography as the modifiable areal unit problem [MAUP; @Openshaw1979; @Openshaw1984].
 <!-- The MAUP has two interrelated components, namely the scale problem and the zoning problem -->
@@
 -->
-Approaches to dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation and the analysis of the sensitivity of the zoning configuration [@Jelinski1996; @Duque2018; @Fletcher2023].
+Dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation, and the analysis of the sensitivity of the zoning configuration [@Jelinski1996; @Duque2018; @Fletcher2023].
 <!-- TODO: use a moving window measurement device to "postpone" the MAUP [@Riitters2019] -->
 
-TODO: section about approaches to sample landscapes to compute metrics and available software tools, e.g., zonal analysis with polygons (e.g., administrative limits), grids, transects or buffers along a landscape gradient, moving windows [@Bosch2019; @Hesselbarth2019; @Huais2024]?
+<!-- MH: Can we move this to the sampling section maybe? -->
+<!-- TODO: section about approaches to sample landscapes to compute metrics and available software tools, e.g., zonal analysis with polygons (e.g., administrative limits), grids, transects or buffers along a landscape gradient, moving windows [@Bosch2019; @Hesselbarth2019; @Huais2024]?
+-->
 
@@
 The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales. 
-For instance, some metrics related to the number and density of patches and edges show consistent linear and power-law scaling relationships to changing grain across a wide diversity of simulated and real landscapes, whereas other metrics exhibit less predictable responses [@Turner1989b; @Saura2001; @Wu2002a; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
+Some metrics related to the number of patches, edges show consistent linear and power-law scaling relationships to changing grain, while other metrics exhibit less predictable responses [@Turner1989b; @Saura2001; @Wu2002a; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
 <!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->
@@
 Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004].
-For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches, resulting in unreliable landscape metric computations.
-In order to address scale issues and advance the science of scaling in landscape ecology, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
-As suggested by Frazier [@Frazier2023], this can be done following three key steps, namely by considering scope guidelines when designing experiments, explicitly reporting grain and extent so that scope can be calculated allowing quantitative comparability with other studes and undertaking comparisons and replications based on scope.
-The latter is showcased in a proof of concept example experiment, where landscapes from the same ecoregion are sampled using different grain and extent and then the distributions of the computed landscape metrics are visualized through ridgeline plots colored by scope [@Frazier2023].
+For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches resulting in unreliable metric computations.
+In order to address scale issues, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
+This involves three fundamental steps, i) considering scope guidelines when designing experiments, ii) explicitly reporting grain and extent allowing quantitative scope comparisons with other studies, iii) and comparisons and replications based on scope [@Frazier2023].
 
-TODO: section about landscape metric rescaling [@Saura2007; @Arganaraz2014; @Frazier2014; @Frazier2016; @Frazier2021]?
+<!-- TODO: section about landscape metric rescaling [@Saura2007; @Arganaraz2014; @Frazier2014; @Frazier2016; @Frazier2021]? -->
 
@@
 
-TODO: mention hierarchichal patch dynamics [@Wu1999]?
+<!-- TODO: mention hierarchichal patch dynamics [@Wu1999]? -->
 
@@
 <!-- Scale and heterogeneity are inherently linked (Wu 2007), so it is not surprising that the results indicate downscaling accuracy varies with heterogeneity. -->
-TODO: sensitivity to thematic resolution [@Huang2006; @Bailey2007; @Buyantuyev2007; @Castilla2009].
+<!-- TODO: sensitivity to thematic resolution [@Huang2006; @Bailey2007; @Buyantuyev2007; @Castilla2009]. -->
 
@@
 TODO
+
+<!-- MH: I think most of this is either not relevant here because not computational focused or already covered above -->
 
@@
 However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019].
-<!-- Most LULC maps are provided as raster as they often related to underlying remote sensing products. -->
-For example, several early landscape ecology studies were based on polygonal vector data that was converted to raster data. <!-- REF -->
+For example, many LULC maps are provided as raster as they often related to underlying remote sensing products.
 Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.

### 2024-08-29 Sections/03-connectivity.qmd
Starting to work on connectivity

@@
 
-<!-- Person responsible: -->
+<!-- Person responsible: Hesselbarth, ??? -->
 
-# Connectivity / Movement / Dispersal
+# Connectivity
 
-<!-- 
-https://link.springer.com/article/10.1007/s10980-023-01741-8
-https://link.springer.com/article/10.1007/s10980-022-01504-x?fromPaywallRec=false
-https://link.springer.com/article/10.1007/s10980-019-00935-3?fromPaywallRec=false
-https://link.springer.com/article/10.1038/s41598-022-20370-w?fromPaywallRec=true
--->
+Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993; @Taylor2006].
+While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006].
+Connectivity underpins ecosystem functionality and maintains biodiversity, and plays a large role in many conservation actions [@CorreaAyram2016; @Hilty2020].
+Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018; @Keeley2021].
+
+## Landscape metrics
+
+- Many traditional lsm measure connectivity, e.g, ENN
+- Recently, also focus in within-patch connectivity (https://doi.org/10.1007/s10980-019-00881-0)
+
+## Resistance surfaces
+
+https://doi.org/10.1111/2041-210X.12750
+https://doi.org/10.1007/s10980-019-00789-9
+https://doi.org/10.1007/s10980-022-01504-x
+https://doi.org/10.1007/s10980-022-01469-x
+
+## Circuit theory-based
+
+https://circuitscape.org
+https://doi.org/10.1111/2041-210X.12689
+https://doi.org/10.1016/j.ecolind.2016.03.032
+
+##  Least-cost paths (LCPs)
+
+https://doi.org/10.1016/S0169-2046(02)00242-6
+https://doi.org/10.1007/s40823-016-0006-9
+
+## Graph theory
+
+https://doi.org/10.1007/s10980-020-00998-7
+https://doi.org/10.1007/s10980-020-01164-9
+
+## Various
+
+https://doi.org/10.1890/14-1690.1
+https://doi.org/10.1016/j.ecolind.2016.03.032
+https://doi.org/10.1007/s10980-019-00935-3
+https://doi.org/10.1126/science.abg1780

### 2024-08-29 Manuscript.qmd
Update Bib and manuscript

@@
 {{< include Sections/02-spatial-patterns.qmd >}}
+{{< include Sections/03-connectivity.qmd >}}
 {{< include Sections/05-simulations.qmd >}}

### 2024-08-29 Manuscript.tex
Update Bib and manuscript

@@
 
-Despite the extensive use of the landscape mosaic model and the
-landscape metrics based on it, major limitations have been identified
-regarding both its ability to quantify the spatial structure of
-landscapes and their relationship to ecological processes (Gustafson
-1998; Li and Wu 2004; Dramstad 2009; Kupfer 2012; Lausch et al. 2015;
-Frazier and Kedron 2017). In line with Dramstad (Dramstad 2009), we
-review four major issues of landscape metrics, namely their sensitivity
-to both spatial scale and thematic resolution, redundancy and
-interpretation.
+Despite the extensive use of the landscape mosaic model and metrics
+based on it, major limitations have been identified regarding both its
+ability to quantify the spatial structure of landscapes and their
+relationship to ecological processes (Gustafson 1998; Li and Wu 2004;
+Dramstad 2009; Kupfer 2012; Lausch et al. 2015; Frazier and Kedron
+2017). Here, we review four major issues of landscape metrics, namely
+their sensitivity to both spatial scale and thematic resolution,
+redundancy and interpretation (Dramstad 2009).
 
@@
 
-Identifying the characteristic scale at which the organisms or processes
-of interest operate is a critical first step towards establishing sound
-relationships between pattern and process (Levin 1992; Wu 2007; Cushman
-and McGarigal 2008). To that end, a key requirement it to understand the
-sensitivity of landscape metrics to spatial scale in terms of both
-resolution (grain) and extent.
+Identifying scales at which the ecological processes of interest operate
+is a critical first step towards establishing relationships between
+pattern and process (Levin 1992; Wu 2007; Cushman and McGarigal 2008).
+To that end, a key requirement is to understand the sensitivity of
+landscape metrics to spatial scale in terms of both grain and extent.
 
@@
 analyzed as a discrete patch. Since the input maps should represent not
-only the location of patches but also their shape, a grain size several
-times smaller than the smallest patch (i.e., the MMU) in order to grasp
-(even schematically) its shape. Even though the MMU is often assumed as
-a natural characteristic of the dataset, it has an unsurprising impact
-on the computed landscape metrics, especially in terms of the number of
-patches and their size distribution (Saura 2002).
+only the location of patches but also their shape, the grain size should
+be several times smaller than the smallest patch (i.e., the MMU) in
+order to quantify shapes. Even though the MMU is often assumed as a
+natural characteristic of the dataset, it has an impact on the computed
+landscape metrics, especially on the number of patches and their size
+distribution (Saura 2002).
 
@@
 e.g., Landsat scenes at 30 m resolution, the selection of spatial extent
-and the way that it is sampled varies largely depending on the
-phenomenon of interest, e.g., from large transects of forests in natural
-parks to small buffers along the urban-rural gradient. At the same time,
-the choice of spatial extent and sampling strategy is often subject to
-many arbitrary criteria that can largely influence the ultimate results.
-Such an issue has long been recognized in physical geography as the
-modifiable areal unit problem (MAUP) (Openshaw 1979, 1984). Approaches
-to dealing with the MAUP in the context of landscape metrics include the
-identification of individual entities that are ecologically meaningful
-and not modifiable, the evaluation of ``optimal'' zoning systems that
-maximize interzonal variations and minimize intrazonal variation and the
-analysis of the sensitivity of the zoning configuration (Jelinski and Wu
-1996; Duque et al. 2018; Fletcher Jr et al. 2023).
-
-TODO: section about approaches to sample landscapes to compute metrics
-and available software tools, e.g., zonal analysis with polygons (e.g.,
-administrative limits), grids, transects or buffers along a landscape
-gradient, moving windows (Bosch 2019; Hesselbarth et al. 2019; Huais
-2024)?
+varies largely depending on the phenomenon of interest, e.g., from large
+transects of forests in natural parks to small buffers along the
+urban-rural gradient. The spatial extent and sampling strategy is often
+subject to many arbitrary criteria that can largely influence results.
+This issue has long been recognized in physical geography as the
+modifiable areal unit problem {[}MAUP; Openshaw (1979); Openshaw
+(1984){]}. Dealing with the MAUP in the context of landscape metrics
+include the identification of individual entities that are ecologically
+meaningful and not modifiable, the evaluation of ``optimal'' zoning
+systems that maximize interzonal variations and minimize intrazonal
+variation, and the analysis of the sensitivity of the zoning
+configuration (Jelinski and Wu 1996; Duque et al. 2018; Fletcher Jr et
+al. 2023).
 
 The scale effects of landscape metrics can be assessed by computing
-scalograms, i.e., the response curves of metrics to changing scales. For
-instance, some metrics related to the number and density of patches and
-edges show consistent linear and power-law scaling relationships to
-changing grain across a wide diversity of simulated and real landscapes,
-whereas other metrics exhibit less predictable responses (Turner et al.
-1989; Saura and Martínez-Millán 2001; Wu et al. 2002; Wu 2004; Saura and
+scalograms, i.e., the response curves of metrics to changing scales.
+Some metrics related to the number of patches, edges show consistent
+linear and power-law scaling relationships to changing grain, while
+other metrics exhibit less predictable responses (Turner et al. 1989;
+Saura and Martínez-Millán 2001; Wu et al. 2002; Wu 2004; Saura and
 Castro 2007; Šı́mová and Gdulová 2012; Argañaraz and Entraigas 2014;
@@
 2004). For instance, in landscapes with small extent to grain ratio,
-boundary effects are likely to dominate and truncate patches, resulting
-in unreliable landscape metric computations. In order to address scale
-issues and advance the science of scaling in landscape ecology, studies
-should explicitly consider the concept of scope, namely the ratio of
-extent to resolution (Schneider 2001). As suggested by Frazier (Frazier
-2023), this can be done following three key steps, namely by considering
-scope guidelines when designing experiments, explicitly reporting grain
-and extent so that scope can be calculated allowing quantitative
-comparability with other studes and undertaking comparisons and
-replications based on scope. The latter is showcased in a proof of
-concept example experiment, where landscapes from the same ecoregion are
-sampled using different grain and extent and then the distributions of
-the computed landscape metrics are visualized through ridgeline plots
-colored by scope (Frazier 2023).
-
-TODO: section about landscape metric rescaling (Saura and Castro 2007;
-Argañaraz and Entraigas 2014; Frazier 2014, 2016; Frazier et al. 2021)?
-
-TODO: mention hierarchichal patch dynamics (Wu 1999)?
-
-TODO: sensitivity to thematic resolution (Huang et al. 2006; Bailey et
-al. 2007; Buyantuyev and Wu 2007; Castilla et al. 2009).
+boundary effects are likely to dominate and truncate patches resulting
+in unreliable metric computations. In order to address scale issues,
+studies should explicitly consider the concept of scope, namely the
+ratio of extent to resolution (Schneider 2001). This involves three
+fundamental steps, i) considering scope guidelines when designing
+experiments, ii) explicitly reporting grain and extent allowing
+quantitative scope comparisons with other studies, iii) and comparisons
+and replications based on scope (Frazier 2023).
 
@@
 software availability, or the familiarity with the approach (Costanza et
-al. 2019). For example, several early landscape ecology studies were
-based on polygonal vector data that was converted to raster data. Yet,
-nowadays, vector data is increasingly available, and most geospatial
-software supports it.
+al. 2019). For example, many LULC maps are provided as raster as they
+often related to underlying remote sensing products. Yet, nowadays,
+vector data is increasingly available, and most geospatial software
+supports it.
 
@@
 compared to raster-based metrics.
+
+\section{Connectivity}\label{connectivity}
+
+Connectivity describes how movement among resources patches is
+facilitated or hindered by landscape features such as types, amounts,
+and arrangement of LULC (Taylor et al. 1993, 2006). While structural
+connectivity focuses on physical and spatial relationships among
+resource patches, functional connectivity also includes the movement or
+flow of individuals and populations through the landscape (Taylor et al.
+2006). Connectivity underpins ecosystem functionality and maintains
+biodiversity, and plays a large role in many conservation actions
+(Correa Ayram et al. 2016; Hilty et al. 2020). Thus, in the past
+multiple approaches how to measure connectivity were developed (Simpkins
+et al. 2018; Keeley et al. 2021).
+
+\subsection{Landscape metrics}\label{landscape-metrics-1}
+
+\begin{itemize}
+\tightlist
+\item
+  Many traditional lsm measure connectivity, e.g, ENN
+\item
+  Recently, also focus in within-patch connectivity
+  (https://doi.org/10.1007/s10980-019-00881-0)
+\end{itemize}
+
+\subsection{Resistance surfaces}\label{resistance-surfaces}
+
+https://doi.org/10.1111/2041-210X.12750
+https://doi.org/10.1007/s10980-019-00789-9
+https://doi.org/10.1007/s10980-022-01504-x
+https://doi.org/10.1007/s10980-022-01469-x
+
+\subsection{Circuit theory-based}\label{circuit-theory-based}
+
+https://circuitscape.org https://doi.org/10.1111/2041-210X.12689
+https://doi.org/10.1016/j.ecolind.2016.03.032
+
+\subsection{Least-cost paths (LCPs)}\label{least-cost-paths-lcps}
+
+https://doi.org/10.1016/S0169-2046(02)00242-6
+https://doi.org/10.1007/s40823-016-0006-9
+
+\subsection{Graph theory}\label{graph-theory}
+
+https://doi.org/10.1007/s10980-020-00998-7
+https://doi.org/10.1007/s10980-020-01164-9
+
+\subsection{Various}\label{various}
+
+https://doi.org/10.1890/14-1690.1
+https://doi.org/10.1016/j.ecolind.2016.03.032
+https://doi.org/10.1007/s10980-019-00935-3
+https://doi.org/10.1126/science.abg1780
 
@@
 overview of packages related to geographical spatial data, please see
-the CRAN Task View (Bivand and Nowosad 2024).
-
-R packages developed specifically for spatial pattern analysis include,
-besides others, \emph{landscapemetrics} (Hesselbarth et al. 2019),
-\emph{vectormetrics} {[}still being developed; Matuszek et al.
-(2024){]}, \emph{motif} (Nowosad 2021), \emph{rasterdiv} (Rocchini et
-al. 2021), or \emph{geodiv} (Smith et al. 2021). Packages with a focus
-on scale include e.g., \emph{multilandr} (Huais 2024) or
-\emph{grainchanger} (Graham 2024). Connectivity analysis can be
-conducted using packages including \emph{lconnect} (Mestre and Silva
-2023), or \emph{grainscape} (Chubaty et al. 2020). Furthermore, neutral
-landscapes can be simulated using \emph{NLMR} (Sciaini et al. 2018), and
-the \emph{graph4lg} (Savary et al. 2021) provides functions for
-landscape genetics.
+the CRAN Task View (Bivand and Nowosad 2024). R packages developed
+specifically for spatial pattern analysis include, besides others,
+\emph{landscapemetrics} (Hesselbarth et al. 2019), \emph{vectormetrics}
+{[}still being developed; Matuszek et al. (2024){]}, \emph{motif}
+(Nowosad 2021), \emph{rasterdiv} (Rocchini et al. 2021), or
+\emph{geodiv} (Smith et al. 2021). Packages with a focus on scale
+include e.g., \emph{multilandr} (Huais 2024) or \emph{grainchanger}
+(Graham 2024). Connectivity analysis can be conducted using packages
+including \emph{lconnect} (Mestre and Silva 2023), or \emph{grainscape}
+(Chubaty et al. 2020). Furthermore, neutral landscapes can be simulated
+using \emph{NLMR} (Sciaini et al. 2018), and the \emph{graph4lg} (Savary
+et al. 2021) provides functions for landscape genetics.
 
@@
 
-\bibitem[\citeproctext]{ref-Bailey2007}
-Bailey D, Billeter R, Aviron S, et al (2007) The influence of thematic
-resolution on metric selection for biodiversity monitoring in
-agricultural landscapes. Landscape Ecology 22:461--473
-
 \bibitem[\citeproctext]{ref-Basaraner2017}
@@
 
-\bibitem[\citeproctext]{ref-Bosch2019}
-Bosch M (2019) PyLandStats: An open-source pythonic library to compute
-landscape metrics. PLoS One 14:e0225734
-
-\bibitem[\citeproctext]{ref-Buyantuyev2007}
-Buyantuyev A, Wu J (2007) Effects of thematic resolution on landscape
-pattern analysis. Landscape Ecology 22:7--13
-
 \bibitem[\citeproctext]{ref-Carauta2018}
@@
 
-\bibitem[\citeproctext]{ref-Castilla2009}
-Castilla G, Larkin K, Linke J, Hay GJ (2009) The impact of thematic
-resolution on the patch-mosaic model of natural landscapes. Landscape
-Ecology 24:15--23
-
 \bibitem[\citeproctext]{ref-Cesco2023}
@@
 Science 12:699--714. \url{https://doi.org/10.1080/136588198241617}
+
+\bibitem[\citeproctext]{ref-CorreaAyram2016}
+Correa Ayram CA, Mendoza ME, Etter A, Salicrup DRP (2016) Habitat
+connectivity in biodiversity conservation: A review of recent studies
+and applications. Progress in Physical Geography: Earth and Environment
+40:7--37. \url{https://doi.org/10.1177/0309133315598713}
 
@@
 
-\bibitem[\citeproctext]{ref-Frazier2014}
-Frazier AE (2014) A new data aggregation technique to improve landscape
-metric downscaling. Landscape ecology 29:1261--1276
-
-\bibitem[\citeproctext]{ref-Frazier2016}
-Frazier AE (2016) Surface metrics: Scaling relationships and downscaling
-behavior. Landscape ecology 31:351--363
-
 \bibitem[\citeproctext]{ref-Frazier2023}
@@
 
+\bibitem[\citeproctext]{ref-Hilty2020}
+Hilty J, Worboys GL, Keeley A, et al (2020)
+\href{https://doi.org/10.2305/IUCN.CH.2020.PAG.30.en}{Guidelines for
+conserving connectivity through ecological networks and corridors}.
+IUCN, International Union for Conservation of Nature
+
 \bibitem[\citeproctext]{ref-Huais2024}
@@
 analysis. Landscape Ecology 39:140
-
-\bibitem[\citeproctext]{ref-Huang2006}
-Huang C, Geiger E, Kupfer J (2006) Sensitivity of landscape metrics to
-classification scheme. International Journal of Remote Sensing
-27:2927--2948
 
@@
 
+\bibitem[\citeproctext]{ref-Keeley2021}
+Keeley ATH, Beier P, Jenness JS (2021) Connectivity metrics for
+conservation planning and monitoring. Biological Conservation
+255:109008. \url{https://doi.org/10.1016/j.biocon.2021.109008}
+
 \bibitem[\citeproctext]{ref-Kupfer2012}
@@
 \url{https://doi.org/10.1002/j.1538-7305.1948.tb01338.x}
+
+\bibitem[\citeproctext]{ref-Simpkins2018}
+Simpkins CE, Dennis TE, Etherington TR, Perry GLW (2018) Assessing the
+performance of common landscape connectivity metrics using a virtual
+ecologist approach. Ecological Modelling 367:13--23.
+\url{https://doi.org/10.1016/j.ecolmodel.2017.11.001}
 
@@
 
+\bibitem[\citeproctext]{ref-Taylor1993}
+Taylor PD, Fahrig L, Henein K, Merriam G (1993) Connectivity is a vital
+element of landscape structure. Oikos 68:571.
+\url{https://doi.org/10.2307/3544927}
+
+\bibitem[\citeproctext]{ref-Taylor2006}
+Taylor PD, Fahrig L, With KA (2006)
+\href{https://doi.org/10.1017/CBO9780511754821}{Landscape connectivity:
+A return to the basics}. In: Crooks KR, Sanjayan M (eds) Connectivity
+conservation. Cambridge University Press, Cambridge, pp 29--43
+
 \bibitem[\citeproctext]{ref-Tennekes2018}
@@
 
-\bibitem[\citeproctext]{ref-Wu1999}
-Wu J (1999) Hierarchy and scaling: Extrapolating information along a
-scaling ladder. Canadian journal of remote sensing 25:367--380
-
 \bibitem[\citeproctext]{ref-Wu2004}

### 2024-08-30 Manuscript.qmd
Process to connectivity section

@@
       - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
-  - name: Rafael Schouten
-    orcid: 0000-0002-8380-0884
-    email: rafael.schouten@sund.ku.dk
-    affiliations:
-      - name: Globe Institute, University of Copenhagen, Copenhagen, Denmark
 format:

### 2024-08-30 Manuscript.tex
Process to connectivity section

@@
   pdftitle={Computational methods in landscape ecology},
-  pdfauthor={Maximilian H.K. Hesselbarth; Jakub Nowosad; Marti Bosch; Martin Jung; Rafael Schouten},
+  pdfauthor={Maximilian H.K. Hesselbarth; Jakub Nowosad; Marti Bosch; Martin Jung},
   pdfkeywords={keyword1, keyword2},
@@
                   Laxenburg, Austria
-              }
-        \author{Rafael Schouten}
-            \affil{%
-                  Globe Institute, University of Copenhagen, Copenhagen,
-                  Denmark
               }
@@
 2006). Connectivity underpins ecosystem functionality and maintains
-biodiversity, and plays a large role in many conservation actions
-(Correa Ayram et al. 2016; Hilty et al. 2020). Thus, in the past
-multiple approaches how to measure connectivity were developed (Simpkins
-et al. 2018; Keeley et al. 2021).
+biodiversity and populations, and plays a large role in many
+conservation actions (Correa Ayram et al. 2016; Hilty et al. 2020).
+Thus, in the past multiple approaches how to measure connectivity were
+developed (Simpkins et al. 2018; Keeley et al. 2021).
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Many traditional lsm measure connectivity, e.g, ENN
-\item
-  Recently, also focus in within-patch connectivity
-  (https://doi.org/10.1007/s10980-019-00881-0)
-\end{itemize}
+There are several landscape metrics that are traditionally used to
+measure landscape connectivity. For example, the habitat amount within a
+buffer zone of focal patches, the nearest-neighbor distance, or the
+connectance index (Moilanen and Nieminen 2002; Prugh 2009; Spanowicz and
+Jaeger 2019). However, these metrics are frequently criticised for being
+to simple and especially the link to functional connectivity is often
+ambiguous. Thus, more functionality needs to be included using
+additional ecological meaningful metrics, such as the core area (e.g.,
+Kupfer 2012). More recently there is also an increasing focus in
+within-patch connectivity using metrics such as the effective mesh size
+in addition to the usual measure between-patch connectivity (Spanowicz
+and Jaeger 2019).
 
@@
 
-https://doi.org/10.1111/2041-210X.12750
-https://doi.org/10.1007/s10980-019-00789-9
-https://doi.org/10.1007/s10980-022-01504-x
-https://doi.org/10.1007/s10980-022-01469-x
-
-\subsection{Circuit theory-based}\label{circuit-theory-based}
-
-https://circuitscape.org https://doi.org/10.1111/2041-210X.12689
-https://doi.org/10.1016/j.ecolind.2016.03.032
-
-\subsection{Least-cost paths (LCPs)}\label{least-cost-paths-lcps}
-
-https://doi.org/10.1016/S0169-2046(02)00242-6
-https://doi.org/10.1007/s40823-016-0006-9
+Resistance surfaces are one of the most popular approaches to model
+connectivity and developed from simple isolation-by-distance approaches.
+The approach uses environmental layers and biodiversity data or expert
+opinions to estimate the cost and of movement through landscapes (Zeller
+et al. 2012). The latest developments of resistance surfaces have
+recently been reviewed (Dutta et al. 2022). The resistance surface can
+be used to model movement based on different approaches such as
+leas-cost paths (Adriaensen et al. 2003; Etherington 2016) or circuit
+theory (McRae et al. 2008). However, resistance surfaces are not free of
+critique including missing spatial-temporal variability or context
+dependency (Unnithan Kumar et al. 2022b; Unnithan Kumar and Cushman
+2022). Thus, recent developments include the combination with more
+processes-based modelling approaches (Schumaker and Brookes 2018; Bocedi
+et al. 2021; Unnithan Kumar et al. 2022a), and modelling of dynamic
+landscape connectivity (Zeller et al. 2020).
 
@@
 
-https://doi.org/10.1007/s10980-020-00998-7
-https://doi.org/10.1007/s10980-020-01164-9
-
 \subsection{Various}\label{various}
-
-https://doi.org/10.1890/14-1690.1
-https://doi.org/10.1016/j.ecolind.2016.03.032
-https://doi.org/10.1007/s10980-019-00935-3
-https://doi.org/10.1126/science.abg1780
 
@@
 \begin{CSLReferences}{1}{1}
+\bibitem[\citeproctext]{ref-Adriaensen2003}
+Adriaensen F, Chardon JP, De Blust G, et al (2003) The application of
+{``least-cost''} modelling as a functional landscape model. Landscape
+and Urban Planning 64:233--247.
+\url{https://doi.org/10.1016/S0169-2046(02)00242-6}
+
 \bibitem[\citeproctext]{ref-Angel2010}
@@
 
+\bibitem[\citeproctext]{ref-Bocedi2021}
+Bocedi G, Palmer SCF, Malchow A-K, et al (2021) RangeShifter 2.0: An
+extended and enhanced platform for modelling spatial eco-evolutionary
+dynamics and species' responses to environmental changes. Ecography
+44:1453--1462. \url{https://doi.org/10.1111/ecog.05687}
+
 \bibitem[\citeproctext]{ref-Borthwick2020}
@@
 
+\bibitem[\citeproctext]{ref-Dutta2022}
+Dutta T, Sharma S, Meyer NFV, et al (2022) An overview of computational
+tools for preparing, constructing and using resistance surfaces in
+connectivity research. Landscape Ecology 37:2195--2224.
+\url{https://doi.org/10.1007/s10980-022-01469-x}
+
+\bibitem[\citeproctext]{ref-Etherington2016}
+Etherington TR (2016) Least-cost modelling and landscape ecology:
+Concepts, applications, and opportunities. Current Landscape Ecology
+Reports 1:40--53. \url{https://doi.org/10.1007/s40823-016-0006-9}
+
 \bibitem[\citeproctext]{ref-Etherington2022}
@@
 
+\bibitem[\citeproctext]{ref-McRae2008}
+McRae B, Dickson BG, Keitt TH, Shah VB (2008) Using circuit theory to
+model connectivity in ecology, evolution, and conservation. Ecology
+89:2712--2724. \url{https://doi.org/10.1890/07-1861.1}
+
 \bibitem[\citeproctext]{ref-Mestre2023}
@@
 
+\bibitem[\citeproctext]{ref-Moilanen2002}
+Moilanen A, Nieminen M (2002) Simple connectivity measures in spatial
+ecology. Ecology 83:1131--1145.
+\url{https://doi.org/10.1890/0012-9658(2002)083\%5B1131:SCMISE\%5D2.0.CO;2}
+
 \bibitem[\citeproctext]{ref-Nathan2022}
@@
 \url{https://doi.org/10.1371/journal.pcbi.1002802}
+
+\bibitem[\citeproctext]{ref-Prugh2009}
+Prugh LR (2009) An evaluation of patch connectivity measures. Ecological
+Applications 19:1300--1310. \url{https://doi.org/10.1890/08-1524.1}
 
@@
 
+\bibitem[\citeproctext]{ref-Spanowicz2019}
+Spanowicz AG, Jaeger JAG (2019) Measuring landscape connectivity: On the
+importance of within-patch connectivity. Landscape Ecology
+34:2261--2278. \url{https://doi.org/10.1007/s10980-019-00881-0}
+
 \bibitem[\citeproctext]{ref-Spies2017}
@@
 
+\bibitem[\citeproctext]{ref-UnnithanKumar2022a}
+Unnithan Kumar S, Cushman SA (2022) Connectivity modelling in
+conservation science: A comparative evaluation. Scientific Reports
+12:16680. \url{https://doi.org/10.1038/s41598-022-20370-w}
+
+\bibitem[\citeproctext]{ref-UnnithanKumar2022b}
+Unnithan Kumar S, Kaszta Z, Cushman SA (2022a) Pathwalker: A new
+individual-based movement model for conservation science and
+connectivity modelling. ISPRS International Journal of Geo-Information
+11:329. \url{https://doi.org/10.3390/ijgi11060329}
+
+\bibitem[\citeproctext]{ref-UnnithanKumar2022}
+Unnithan Kumar S, Turnbull J, Hartman Davies O, et al (2022b) Moving
+beyond landscape resistance: Considerations for the future of
+connectivity modelling and conservation science. Landscape Ecology
+37:2465--2480. \url{https://doi.org/10.1007/s10980-022-01504-x}
+
 \bibitem[\citeproctext]{ref-Uuemaa2011}
@@
 
+\bibitem[\citeproctext]{ref-Zeller2012}
+Zeller KA, McGarigal K, Whiteley AR (2012) Estimating landscape
+resistance to movement: A review. Landscape Ecology 27:777--797.
+\url{https://doi.org/10.1007/s10980-012-9737-0}
+
+\bibitem[\citeproctext]{ref-Zeller2020}
+Zeller K, Lewison R, Fletcher R, et al (2020) Understanding the
+importance of dynamic landscape connectivity. Land 9:303.
+\url{https://doi.org/10.3390/land9090303}
+
 \bibitem[\citeproctext]{ref-Zhang2020}

### 2024-08-30 Sections/03-connectivity.qmd
Process to connectivity section

@@
 While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006].
-Connectivity underpins ecosystem functionality and maintains biodiversity, and plays a large role in many conservation actions [@CorreaAyram2016; @Hilty2020].
+Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016; @Hilty2020].
 Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018; @Keeley2021].
+
+<!-- 
+Connectivity modelling in conservation science: https://doi.org/10.1038/s41598-022-20370-w
+-->
 
@@
 
-- Many traditional lsm measure connectivity, e.g, ENN
-- Recently, also focus in within-patch connectivity (https://doi.org/10.1007/s10980-019-00881-0)
+There are several landscape metrics that are traditionally used to measure landscape connectivity.
+For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002; @Prugh2009; @Spanowicz2019].
+However, these metrics are frequently criticised for being to simple and especially the link to functional connectivity is often ambiguous. 
+Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012].
+More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019].
 
@@
 
-https://doi.org/10.1111/2041-210X.12750
-https://doi.org/10.1007/s10980-019-00789-9
-https://doi.org/10.1007/s10980-022-01504-x
-https://doi.org/10.1007/s10980-022-01469-x
+Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
+The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012].
+The latest developments of resistance surfaces have recently been reviewed [@Dutta2022].
+The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003; @Etherington2016] or circuit theory [@McRae2008].
+However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022; @UnnithanKumar2022a].
+Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018; @Bocedi2021; @UnnithanKumar2022b], and modelling of dynamic landscape connectivity [@Zeller2020].
+<!-- MH: Not complete yet -->
 
-## Circuit theory-based
-
-https://circuitscape.org
+<!--
+lscorridors: https://doi.org/10.1111/2041-210X.12750
+Complementary landscape genetics and habitat modelling approaches: https://doi.org/10.1007/s10980-019-00789-9
 https://doi.org/10.1111/2041-210X.12689
 https://doi.org/10.1016/j.ecolind.2016.03.032
-
-##  Least-cost paths (LCPs)
-
-https://doi.org/10.1016/S0169-2046(02)00242-6
-https://doi.org/10.1007/s40823-016-0006-9
+-->
 
@@
 
+<!--
 https://doi.org/10.1007/s10980-020-00998-7
 https://doi.org/10.1007/s10980-020-01164-9
+-->
 
@@
 
+<!--
 https://doi.org/10.1890/14-1690.1
@@
 https://doi.org/10.1126/science.abg1780
+-->

### 2024-08-30 Sections/05-simulations.qmd
Process to connectivity section

@@
 
-<!-- Person responsible: Hesselbarth, Schouten -->
+<!-- Person responsible: Hesselbarth -->
 
@@
 In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022].
-<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->
+<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [ c], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->
 

### 2024-08-30 Manuscript.tex
Adding some references

@@
 processes-based modelling approaches (Schumaker and Brookes 2018; Bocedi
-et al. 2021; Unnithan Kumar et al. 2022a), and modelling of dynamic
-landscape connectivity (Zeller et al. 2020).
+et al. 2021; Unnithan Kumar et al. 2022a), modelling of dynamic
+landscape connectivity (Zeller et al. 2020), inclusion of stochasticity
+and spatial context (Ribeiro et al. 2017), using genetic optimization
+algorithms (Peterman 2018), or computational improvements (Leonard et
+al. 2017).
 
@@
 
-\subsection{Various}\label{various}
+Graph theory describes a landscape as a set of nodes representing
+habitat patches and edges as potential links between the nodes. Thus,
+graph theory is a powerful tool to model landscape connectivity,
+especially in a metapopulation context (Urban and Keitt 2001).
+
+\subsection{Other approaches}\label{other-approaches}
 
@@
 
+\bibitem[\citeproctext]{ref-Leonard2017}
+Leonard PB, Duffy EB, Baldwin RF, et al (2017) Glow: Software for
+modelling circuit theory-based connectivity at any scale. Methods in
+Ecology and Evolution 8:519--526.
+\url{https://doi.org/10.1111/2041-210X.12689}
+
 \bibitem[\citeproctext]{ref-Levin1992}
@@
 
+\bibitem[\citeproctext]{ref-Peterman2018}
+Peterman WE (2018) ResistanceGA: An r package for the optimization of
+resistance surfaces using genetic algorithms. Methods in Ecology and
+Evolution 9:1638--1647. \url{https://doi.org/10.1111/2041-210X.12984}
+
 \bibitem[\citeproctext]{ref-Petrovskii2012}
@@
 University of California Press, pp 547--562
+
+\bibitem[\citeproctext]{ref-Ribeiro2017}
+Ribeiro JW, Silveira Dos Santos J, Dodonov P, et al (2017) LandScape
+corridors ( {\textsc{lscorridors}} ): A new software package for
+modelling ecological corridors based on landscape patterns and species
+requirements. Methods in Ecology and Evolution 8:1425--1432.
+\url{https://doi.org/10.1111/2041-210X.12750}
 
@@
 
+\bibitem[\citeproctext]{ref-Urban2001}
+Urban D, Keitt T (2001) Landscape connectivity: A graph-theoretic
+perspective. Ecology 82:1205--1218.
+\url{https://doi.org/10.1890/0012-9658(2001)082\%5B1205:LCAGTP\%5D2.0.CO;2}
+
 \bibitem[\citeproctext]{ref-Uuemaa2011}

### 2024-08-30 Sections/03-connectivity.qmd
Adding some references

@@
 <!-- 
-Connectivity modelling in conservation science: https://doi.org/10.1038/s41598-022-20370-w
+Connectivity modelling in conservation science https://doi.org/10.1038/s41598-022-20370-w
 -->
@@
 However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022; @UnnithanKumar2022a].
-Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018; @Bocedi2021; @UnnithanKumar2022b], and modelling of dynamic landscape connectivity [@Zeller2020].
-<!-- MH: Not complete yet -->
+Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018; @Bocedi2021; @UnnithanKumar2022b], modelling of dynamic landscape connectivity [@Zeller2020], inclusion of stochasticity and spatial context [@Ribeiro2017], using genetic optimization algorithms [@Peterman2018], or computational improvements [@Leonard2017].
 
-<!--
-lscorridors: https://doi.org/10.1111/2041-210X.12750
-Complementary landscape genetics and habitat modelling approaches: https://doi.org/10.1007/s10980-019-00789-9
-https://doi.org/10.1111/2041-210X.12689
-https://doi.org/10.1016/j.ecolind.2016.03.032
--->
+<!-- What about multi-species approaches? -->
 
@@
 
+Graph theory describes a landscape as a set of nodes representing habitat patches and edges as potential links between the nodes.
+Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001].
+
 <!--
-https://doi.org/10.1007/s10980-020-00998-7
-https://doi.org/10.1007/s10980-020-01164-9
+Coupling landscape graph modeling and biological data https://doi.org/10.1007/s10980-020-00998-7
+Temporal dynamics of landscape connectivity in ever-changing landscapes: a literature review https://doi.org/10.1007/s10980-021-01277-9
+Some methodological considerations to improve landscape planning https://doi.org/10.1016/j.ecolind.2017.05.052
+Comparison of landscape graph modelling methods https://doi.org/10.1007/s10980-020-01164-9
+
 -->
 
-## Various
+## Other approaches
 
 <!--
-https://doi.org/10.1890/14-1690.1
-https://doi.org/10.1016/j.ecolind.2016.03.032
-https://doi.org/10.1007/s10980-019-00935-3
-https://doi.org/10.1126/science.abg1780
+A stochastic movement simulator improves estimates of landscape connectivity https://doi.org/10.1890/14-1690.1
+Landscape connectivity modeling from the perspective of animal dispersal  https://doi.org/10.1007/s10980-019-00935-3
+Big-data approaches https://doi.org/10.1126/science.abg1780
 -->

### 2024-09-02 Manuscript.qmd

@@
 linestretch: 1.5
+number-sections: true
 bibliography: CLER.bib
@@
 {{< include Sections/05-simulations.qmd >}}
+{{< include Sections/07-spatial-planning.qmd >}}
 {{< include Sections/08-software.qmd >}}

### 2024-09-02 Manuscript.tex

@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
-\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
 
-\section{Spatial patterns}\label{spatial-patterns}
+\section{Spatial patterns}\label{sec-spatialpattern}
 
@@
 
-\section{Connectivity}\label{connectivity}
+\section{Connectivity}\label{sec-connectivity}
 
@@
 
-\section{Simulation models}\label{simulation-models}
+\section{Simulation models}\label{sec-simulation}
 
@@
 gaps (De Koning et al. 2023).
+
+\section{Spatial conservation planning}\label{sec-spatialplanning}
+
+Spatial planning broadly considers the use of decision theory to
+identify and allocate areas to specific purposes, such as reserve
+selection through spatial conservation prioritization approaches
+(Margules and Pressey 2000; Kukkala and Moilanen 2013; Jung et al.
+2024). In the context of landscape-ecology it provides the critical link
+between the measuring of patterns and processes (see section
+Section~\ref{sec-spatialpattern}), and outcomes related to actual
+decision making by distinguishing between structure, function and scales
+(Hersperger et al. 2021). Although - as integrative approach - most
+spatial planning approaches can benefit from other computational
+advances in landscape ecology, such as for example through using
+simulation model outputs (see section Section~\ref{sec-simulation}) as
+input features to the planning (see also Daigle et al. (2020)), there
+have also been several recent frontiers in directly considering
+landscape patterns directly in spatial planning.
+
+Spatial patterns in terms of shape and patterns have always been of
+interest to spatial planning. Generally, such patterns influence
+planning outcomes because of their spatial \emph{compactness},
+\emph{contiguity} and \emph{connectedness}. Aggregation can be important
+because of ecological reasoning or practical feasibility, for example by
+ensuring that any selected areas satisfy certain minimum patch size
+constraints (Smith et al. 2010; Justeau-Allaire et al. 2021). Novel
+computational approaches make use of graph theory to identify those
+areas in a landscape more likely to be of values as compact core habitat
+and reducing boundary exposure (Weerasena et al. 2023). Similarly
+contiguity and connectivity are often seen as important to ensure that
+landscapes and patches of specific characteristics are connected, such
+as for example river networks (Beger et al. 2010). In particular the
+direct consideration of connectivity in spatial conservation planning
+has received much recent attention and computational methods are
+increasingly developed (Hanson et al. 2019a, 2022; Daigle et al. 2020;
+Beger et al. 2022).
+
+Particular noteworthy here are two advances in incorporating landscape
+indices in spatial planning. One is the use of graph theory in linear
+programming to guide conservation and restoration efforts
+(Justeau-Allaire et al. 2021; Hamonic et al. 2023). For example, in a
+New Caledonia case study Justeau-Allaire et al. (2021) showed how theory
+and indicators from landscape ecology to determine where restoration
+efforts would best be located to maximize broader landscape connectivity
+(Justeau-Allaire et al. 2021).
+
+A second main advance is the increasing use of deep reinforcement
+learning (RL), a machine learning approach based not on patterns but
+agents and pathway, in spatial planning. RL can be used to identify
+cost-efficient solutions to area-based conservation planning (Silvestro
+et al. 2022) or typical connectivity problems (Equihua et al. 2024),
+where the objective is to identify best possible improvements to
+landscape connectivity indices such as the Integral Index of
+Connectivity. Advantages of RL are that they are able to easier
+incorporate linear and non-linear functions and spatial-temporal
+processes in the planning and are generally more readibly scalable to
+larger spatial extent (Equihua et al. 2024), which however comes with
+the downside of decreasing interpretability.
+
+\subsection{Open-source software in spatial conservation
+planning}\label{sec-spatialplanning-software}
+
+Multiple open-source software solutions have been developed to support
+further integration of landscape ecology theory and indicators into
+spatial planning. Most of these software are based on linear or
+constrained programming approaches, which however require a solver to
+create any outputs. Unfortunately there is still a gap between
+state-of-the art open-source and proprietary solvers in terms of
+computational efficiency (Hanson et al. 2019b). Noteworthy software is
+the \emph{prioritizr} R-package (Hanson et al. 2024) which supports a
+range of compactness and contiguity constraints and penalties, and
+generally allows for the flexibly creation of planning problems in
+multiple configurations. Similarly, the \emph{coco} package makes use of
+connectivity constraints to identify potential outcomes that minimize
+further fragmentation of landscapes (Mantgem et al. 2023). The new
+\emph{restoptr} R-package makes use of constrained programming for those
+areas to be restored in a region that would most effectively contribute
+to greater landscape connectivity (Justeau-Allaire et al. 2023).
+Innovatively, it uses landscape ecology indices not as constraints, but
+as specific targets (e.g.~where one could best maximize the effective
+mesh size in a region). The CAPTAIN software suite enables the
+application of reinforcement learning for area-based conservation
+efforts (Silvestro et al. 2022). Overall, most of these software however
+still require substantial prior knowledge on (spatial) optimization and
+statistics, which can be an entry barrier to more widespread uptake.
 
@@
 
+\bibitem[\citeproctext]{ref-beger2010incorporating}
+Beger M, Linke S, Watts M, et al (2010) Incorporating asymmetric
+connectivity into spatial decision making for conservation. Conservation
+Letters 3:359--368
+
+\bibitem[\citeproctext]{ref-beger2022demystifying}
+Beger M, Metaxas A, Balbar AC, et al (2022) Demystifying ecological
+connectivity for actionable spatial conservation planning. Trends in
+ecology \& evolution 37:1079--1091
+
 \bibitem[\citeproctext]{ref-Berger2024}
@@
 
+\bibitem[\citeproctext]{ref-daigle2020operationalizing}
+Daigle RM, Metaxas A, Balbar AC, et al (2020) Operationalizing
+ecological connectivity in spatial conservation planning with marxan
+connect. Methods in Ecology and Evolution 11:570--579
+
 \bibitem[\citeproctext]{ref-Dale1999}
@@
 
+\bibitem[\citeproctext]{ref-equihua2024connectivity}
+Equihua J, Beckmann M, Seppelt R (2024) Connectivity conservation
+planning through deep reinforcement learning. Methods in Ecology and
+Evolution 15:779--790
+
 \bibitem[\citeproctext]{ref-Etherington2016}
@@
 
+\bibitem[\citeproctext]{ref-hamonic2023cumulative}
+Hamonic F, Couëtoux B, Vaxès Y, Albert CH (2023) Cumulative effects on
+habitat networks: How greedy should we be? Biological Conservation
+282:110066
+
 \bibitem[\citeproctext]{ref-Hampton2015}
@@
 \url{https://doi.org/10.1890/ES14-00402.1}
+
+\bibitem[\citeproctext]{ref-hanson2019conventional}
+Hanson JO, Fuller RA, Rhodes JR (2019a) Conventional methods for
+enhancing connectivity in conservation planning do not always maintain
+gene flow. Journal of Applied Ecology 56:913--922
+
+\bibitem[\citeproctext]{ref-hanson2024prioritizr}
+Hanson JO, Schuster R, Morrell N, et al (2024)
+\href{https://CRAN.R-project.org/package=prioritizr}{Prioritizr:
+Systematic conservation prioritization in r}
+
+\bibitem[\citeproctext]{ref-hanson2019optimality}
+Hanson JO, Schuster R, Strimas-Mackey M, Bennett JR (2019b) Optimality
+in prioritizing conservation projects. Methods in Ecology and Evolution
+10:1655--1663
+
+\bibitem[\citeproctext]{ref-hanson2022comparison}
+Hanson JO, Vincent J, Schuster R, et al (2022) A comparison of
+approaches for including connectivity in systematic conservation
+planning. Journal of Applied Ecology 59:2507--2519
+
+\bibitem[\citeproctext]{ref-hersperger2021landscape}
+Hersperger AM, Grădinaru SR, Pierri Daunt AB, et al (2021) Landscape
+ecological concepts in planning: Review of recent developments.
+Landscape ecology 1--17
 
@@
 
+\bibitem[\citeproctext]{ref-jung2024assessment}
+Jung M, Alagador D, Chapman M, et al (2024) An assessment of the state
+of conservation planning in europe. Philosophical Transactions of the
+Royal Society B 379:20230015
+
 \bibitem[\citeproctext]{ref-Justeau-Allaire2022}
@@
 
+\bibitem[\citeproctext]{ref-justeau2023restoptr}
+Justeau-Allaire D, Hanson JO, Lannuzel G, et al (2023) Restoptr: An r
+package for ecological restoration planning. Restoration Ecology
+31:e13910
+
+\bibitem[\citeproctext]{ref-justeau2021constrained}
+Justeau-Allaire D, Vieilledent G, Rinck N, et al (2021) Constrained
+optimization of landscape indices in conservation planning to support
+ecological restoration in new caledonia. Journal of Applied Ecology
+58:744--754
+
 \bibitem[\citeproctext]{ref-Kamimura2019}
@@
 
+\bibitem[\citeproctext]{ref-kukkala2013core}
+Kukkala AS, Moilanen A (2013) Core concepts of spatial prioritisation in
+systematic conservation planning. Biological Reviews 88:443--464
+
 \bibitem[\citeproctext]{ref-Kupfer2012}
@@
 67:53--67. \url{https://doi.org/10.1080/04353684.1985.11879515}
+
+\bibitem[\citeproctext]{ref-van2023coco}
+Mantgem ES van, Hillebrand J, Rose L, Klau GW (2023) Coco: Conservation
+design for optimal ecological connectivity. Frontiers in Ecology and
+Evolution 11:1149571
+
+\bibitem[\citeproctext]{ref-margules2000systematic}
+Margules CR, Pressey RL (2000) Systematic conservation planning. Nature
+405:243--253
 
@@
 
+\bibitem[\citeproctext]{ref-silvestro2022improving}
+Silvestro D, Goria S, Sterner T, Antonelli A (2022) Improving
+biodiversity protection through artificial intelligence. Nature
+sustainability 5:415--424
+
 \bibitem[\citeproctext]{ref-Simpkins2018}
@@
 Smith D (2016) Over 16 years of r project history
+
+\bibitem[\citeproctext]{ref-smith2010approach}
+Smith RJ, Di Minin E, Linke S, et al (2010) An approach for ensuring
+minimum protected area size in systematic conservation planning.
+Biological Conservation 143:2525--2531
 
@@
 
+\bibitem[\citeproctext]{ref-weerasena2023sequential}
+Weerasena L, Shier D, Tonkyn D, et al (2023) A sequential approach to
+reserve design with compactness and contiguity considerations.
+Ecological Modelling 478:110281
+
 \bibitem[\citeproctext]{ref-Wickham2016}

### 2024-09-02 Sections/07-spatial-planning.qmd
Updating software section and tiny cosmetic changes

@@
 
-# Spatial planning
+# Spatial conservation planning {#sec-spatialplanning}
 
-Landscape ecology is fun [@Turner1989].
-There are many things to explore [@With2019].
+Spatial planning broadly considers the use of decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization approaches [@margules2000systematic; @kukkala2013core; @jung2024assessment].
+In the context of landscape-ecology it provides the critical link between the measuring of patterns and processes (see section @sec-spatialpattern), and outcomes related to actual decision making by distinguishing between structure, function and scales [@hersperger2021landscape].
+Although - as integrative approach - most spatial planning approaches can benefit from other computational advances in landscape ecology, such as for example through using simulation model outputs (see section @sec-simulation) as input features to the planning (see also @daigle2020operationalizing), there have also been several recent frontiers in directly considering landscape patterns directly in spatial planning.
+
+Spatial patterns in terms of shape and patterns have always been of interest to spatial planning.
+Generally, such patterns influence planning outcomes because of their spatial *compactness*, *contiguity* and *connectedness*.
+Aggregation can be important because of ecological reasoning or practical feasibility, for example by ensuring that any selected areas satisfy certain minimum patch size constraints [@smith2010approach;@justeau2021constrained].
+Novel computational approaches make use of graph theory to identify those areas in a landscape more likely to be of values as compact core habitat and reducing boundary exposure [@weerasena2023sequential].
+Similarly contiguity and connectivity are often seen as important to ensure that landscapes and patches of specific characteristics are connected, such as for example river networks [@beger2010incorporating].
+In particular the direct consideration of connectivity in spatial conservation planning has received much recent attention and computational methods are increasingly developed [@hanson2019conventional; @daigle2020operationalizing; @hanson2022comparison; @beger2022demystifying].
+
+Particular noteworthy here are two advances in incorporating landscape indices in spatial planning.
+One is the use of graph theory in linear programming to guide conservation and restoration efforts [@justeau2021constrained; @hamonic2023cumulative].
+For example, in a New Caledonia case study @justeau2021constrained showed how theory and indicators from landscape ecology to determine where restoration efforts would best be located to maximize broader landscape connectivity.
+The direct consideration of graph theoretic approaches in spatial networks has the potential to be provide more cost-effective and precise solutions to the design of reserve networks [@hamonic2023cumulative].
+
+A second main advance is the increasing use of deep reinforcement learning (RL), a machine learning approach based not on patterns but agents and pathway, in spatial planning.
+RL can be used to identify cost-efficient solutions to area-based conservation planning [@silvestro2022improving] or typical connectivity problems [@equihua2024connectivity], where the objective is to identify best possible improvements to landscape connectivity indices such as the Integral Index of Connectivity.
+Advantages of RL are that they are able to easier incorporate linear and non-linear functions and spatial-temporal processes in the planning and are generally more readibly scalable to larger spatial extent [@equihua2024connectivity], which however comes with the downside of decreasing interpretability.
+
+## Open-source software in spatial conservation planning {#sec-spatialplanning-software}
+
+Multiple open-source software solutions have been developed to support further integration of landscape ecology theory and indicators into spatial planning.
+Most of these software are based on linear or constrained programming approaches, which however require a solver to create any outputs.
+Unfortunately there is still a gap between state-of-the art open-source and proprietary solvers in terms of computational efficiency [@hanson2019optimality].
+
+Noteworthy software is the *prioritizr* R-package [@hanson2024prioritizr] which supports a range of compactness and contiguity constraints and penalties, and generally allows for the flexibly creation of planning problems in multiple configurations.
+Similarly, the *coco* package makes use of connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of landscapes [@van2023coco].
+The new *restoptr* R-package makes use of constrained programming for those areas to be restored in a region that would most effectively contribute to greater landscape connectivity [@justeau2023restoptr].
+Innovatively, it uses landscape ecology indices not as constraints, but as specific targets (e.g.
+where one could best maximize the effective mesh size in a region).
+The *CAPTAIN* software suite enables the application of reinforcement learning for area-based conservation efforts [@silvestro2022improving].
+Overall, most of these software however still require substantial prior knowledge on (spatial) optimization and statistics, which can be an entry barrier to more widespread uptake.

### 2024-09-02 Manuscript.qmd
Update meta information and BIB

--- 2024-08-30 Manuscript.qmd
@@
 title: "Computational methods in landscape ecology"
+
 keywords: keyword1, keyword2
+
 author:
@@
     email: mhk.hesselbarth@gmail.com
-    affiliations:
-      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
+    affil-id: 1
     attributes:
@@
     email: nowosad.jakub@gmail.com
-    affiliations:
-      - name: Adam Mickiewicz University, Poznan, Poland
-      - name: University of Münster, Münster, Germany
+    affil-id: 2,3
   - name: Marti Bosch
@@
     email: marti.bosch@epfl.ch
-    affiliations:
-      - name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
+    affil-id: 4
   - name: Martin Jung
@@
     email: jung@iiasa.ac.at
+    affil-id: 1
+
 affiliations: 
-      - name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
+  - id: 1
+    name: International Institute for Applied Systems Analysis, Biodiversity, Ecology, and Conservation Group, Laxenburg, Austria
+  - id: 2
+    name: Adam Mickiewicz University, Poznan, Poland
+  - id: 3
+    name: University of Münster, Münster, Germany
+  - id: 4
+    name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
+
 format:
@@
         \renewcommand\Affilfont{\small}
+
 papersize: a4paper
@@
 linestretch: 1.5
+number-sections: true
+
 bibliography: CLER.bib
@@
 {{< include Sections/03-connectivity.qmd >}}
+{{< include Sections/04-sampling.qmd >}}
 {{< include Sections/05-simulations.qmd >}}
+{{< include Sections/06-spatial-planning.qmd >}}
+{{< include Sections/07-genetics.qmd >}}
 {{< include Sections/08-software.qmd >}}
+{{< include Sections/09-discussion.qmd >}}
 

### 2024-09-02 Sections/01-introduction.qmd
Update references

--- 2024-07-01 Sections/01-introduction.qmd
@@
 
-Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986; @Forman1995] - with empathize on heterogeneity of at least one factor of interest [@Turner2015].
-Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984; @Turner1989; @Turner2005].
-Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002; @Wu2013].
+Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology; @Forman1995LandMosaics] - with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
+Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984LandscapeEcology; @Turner1989LandscapeEcology; @Turner2005LandscapeEcology].
+Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].
 
 Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
-Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019].
-Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012; but see @Wiersma2022].
-Because landscape ecology is a cross-disciplinary field including [e.g., social sciences, geography, or ecology and evolution\; @Wiens1997; @With2019], the general increase of data availability [@Chi2016; @Jaric2020; @Wuest2020; @Nathan2022], or the studied complex landscape systems [@Newman2019], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
-Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012; @Wilson2014].
+Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019DatabasedSynthesisdriven].
+Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
+Because landscape ecology is a cross-disciplinary field including [e.g., social sciences, geography, or ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the studied complex landscape systems [@Newman2019ScalingComplexity], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
+Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012TenSimple; @Wilson2014BestPractices].
 
@@
 The spatial vector data model is based on points, lines, and polygons features that are defined by their geometry and additional attributes related to them.
-Contrastingly, the raster data model is based on evenly-spaced grid cells and discrete or continuous values related to each cell [@Lovelace2019; @With2019].
-The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003; @Lechner2016; @Nedd2021].
+Contrastingly, the raster data model is based on evenly-spaced grid cells and discrete or continuous values related to each cell [@Lovelace2019Geocomputation; @With2019EssentialsLandscape].
+The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003ComparisonVector; @Lechner2016RecentProgress; @Nedd2021SynthesisLand].
 
@@
 However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
-For a more comprehensive, general introductions to landscape ecology, please see @Turner2015, @Gergel2017, or @With2019.
+For a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
 First, we will introduce recent computational developments of the most important topics of landscape ecology.
@@
 Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
-However, due to its many advantages [@Powers2019], we will focus on open-source software , and specifically scripting languages, i.e., R, Python, and Julia.
+However, due to its many advantages [@Powers2019OpenScience], we will focus on open-source software , and specifically scripting languages, i.e., R, Python, and Julia.
 Last, we will discuss potential future needs developments of computational methods in landscape ecology.

### 2024-09-02 Sections/02-spatial-patterns.qmd
Update references

@@
 
-# Spatial patterns
+# Spatial patterns {#sec-spatialpattern}
 
-Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996].
-Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003].
+Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999SpatialPattern], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996PatternorientedModelling].
+Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
 However, spatial patterns are not only a result of processes, but could also be a driver of them.
-Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989a].
+Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989LandscapeEcology].
 
@@
 
-Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017; @Costanza2019].
-The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015], however, they are also critiqued for not being able to realistically represent continuous features of the natural word [@Frazier2017].
-Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019].
+Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
+The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural word [@Frazier2017LandscapeMetrics].
+Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpen].
 Recently also several metrics describing landscape complexity were introduced, e.g., several entropy measures (see following sections).
@@
 It is mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
-Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015].
-However, studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix [@Zhao2019].
+Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
+However, studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix [@Zhao2019CalculatingSpatial].
 
-The Shannon diversity index [@Shannon1948] quantifies the richness and evenness of categories in the landscape omitting spatial configuration.
+The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape omitting spatial configuration.
 The index is based on information theory and is an adaptation of the Shannon entropy which measures uncertainty of a random variable.
-Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012].
+Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012SpatiotemporalForm].
 However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory can be adapted.
-@Nowosad2019a proposed to treat categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
+@Nowosad2019InformationTheory proposed to treat categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
 This approach allows us to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
@@
 
-Based on notions from thermodynamics initiated by the work of @Cushman2015 or @Vranken2015, there was a surge in the development of entropy-based metrics for landscape ecology in the last few years.
-In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019].
+Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape or @Vranken2015ReviewUse, there was a surge in the development of entropy-based metrics for landscape ecology in the last few years.
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
 <!-- REF -->
-@Cushman2016 proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
-This approach was generalized for calculations based on raster's surface model and point patterns [@Cushman2021a].
-Subsequently, @Gao2017 proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
-Recently, @Zhang2020 extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
+@Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
+This approach was generalized for calculations based on raster's surface model and point patterns [@Cushman2021EntropyLandscape].
+Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
+Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
 
@@
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
-The Rao quadratic entropy [@Rao1982] has also been applied recently [@Rocchini2017], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
+The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
+The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRao], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
 Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
-Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023].
+Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].
 
@@
 
-Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990; @Wickham1994; @Jasiewicz2015; @Remmel2020; @Nowosad2021a].
+Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990SpatialTemporal; @Wickham1994MappingAnalyzing; @Jasiewicz2015GeoPATToolbox; @Remmel2020DistributionsHyperlocal; @Nowosad2021MotifOpensource].
 These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between the signatures.
@@
 Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
-Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023].
+Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
 Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
-This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014].
-Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021b].
+This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
+Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].
 
@@
 <!-- https://doi.org/10.1371/journal.pone.0304215 -->
-Despite the extensive use of the landscape mosaic model and metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998; @Li2004b; @Dramstad2009; @Kupfer2012; @Lausch2015; @Frazier2017].
+Despite the extensive use of the landscape mosaic model and metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998QuantifyingLandscape; @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology; @Lausch2015UnderstandingQuantifying; @Frazier2017LandscapeMetrics].
 <!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
@@
 Let us go through them in a different order: -->
-Here, we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation [@Dramstad2009].
+Here, we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation [@Dramstad2009SpatialMetrics].
 
@@
 <!-- "identifying the characteristic scales of relevant patterns and processes is a critical first step in designing a successful research project" [@Wu2007, pp. 119]. -->
-Identifying scales at which the ecological processes of interest operate is a critical first step towards establishing relationships between pattern and process [@Levin1992; @Wu2007; @Cushman2008Landscape].
+Identifying scales at which the ecological processes of interest operate is a critical first step towards establishing relationships between pattern and process [@Levin1992ProblemPattern; @Wu2007ScaleScaling; @Cushman2008LandscapeMetrics].
 To that end, a key requirement is to understand the sensitivity of landscape metrics to spatial scale in terms of both grain and extent.
@@
 Since the input maps should represent not only the location of patches but also their shape, the grain size should be several times smaller than the smallest patch (i.e., the MMU) in order to quantify shapes.
-Even though the MMU is often assumed as a natural characteristic of the dataset, it has an impact on the computed landscape metrics, especially on the number of patches and their size distribution [@Saura2002].
+Even though the MMU is often assumed as a natural characteristic of the dataset, it has an impact on the computed landscape metrics, especially on the number of patches and their size distribution [@Saura2002EffectsMinimum].
 <!-- On the other hand, landscape metrics such as the landscape division index, largest patch index and area-weighted mean shape indices show more predictable responses to changing the MMU. -->
@@
 The spatial extent and sampling strategy is often subject to many arbitrary criteria that can largely influence results.
-This issue has long been recognized in physical geography as the modifiable areal unit problem [MAUP; @Openshaw1979; @Openshaw1984].
+This issue has long been recognized in physical geography as the modifiable areal unit problem [MAUP; @Openshaw1979MillionCorrelated; @Openshaw1984EcologicalFallacies].
 <!-- The MAUP has two interrelated components, namely the scale problem and the zoning problem -->
@@
 -->
-Dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation, and the analysis of the sensitivity of the zoning configuration [@Jelinski1996; @Duque2018; @Fletcher2023].
+Dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation, and the analysis of the sensitivity of the zoning configuration [@Jelinski1996ModifiableAreal; @Duque2018SmaupStatistical; @Fletcher2023AddressingProblem].
 <!-- TODO: use a moving window measurement device to "postpone" the MAUP [@Riitters2019] -->
@@
 The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales. 
-Some metrics related to the number of patches, edges show consistent linear and power-law scaling relationships to changing grain, while other metrics exhibit less predictable responses [@Turner1989b; @Saura2001; @Wu2002a; @Wu2004; @Saura2007; @Simova2012; @Arganaraz2014; @Frazier2021].
-<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023] -->
+Some metrics related to the number of patches, edges show consistent linear and power-law scaling relationships to changing grain, while other metrics exhibit less predictable responses [@Turner1989EffectsChanging; @Saura2001SensitivityLandscape; @Wu2002EmpiricalPatterns; @Wu2004EffectsChanging; @Saura2007ScalingFunctions; @Simova2012LandscapeIndices; @Arganaraz2014ScalingFunctions; @Frazier2023ScalingSpatial].
+<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023ScalingSpatial] -->
 <!-- also multi-scale is key to habitat assessment, see "Multi-scale habitat selection modeling: a review and outlook" -->
 <!-- Predictable scaling relationships can be exploited to estimate spatial patterns at higher resolutions -->
-Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004].
+Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004EffectsChanging].
 For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches resulting in unreliable metric computations.
-In order to address scale issues, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001].
-This involves three fundamental steps, i) considering scope guidelines when designing experiments, ii) explicitly reporting grain and extent allowing quantitative scope comparisons with other studies, iii) and comparisons and replications based on scope [@Frazier2023].
+In order to address scale issues, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001RiseConcept].
+This involves three fundamental steps, i) considering scope guidelines when designing experiments, ii) explicitly reporting grain and extent allowing quantitative scope comparisons with other studies, iii) and comparisons and replications based on scope [@Frazier2023ScalingSpatial].
 
@@
 
-TODO
+<!-- MH: I think most of this is either not relevant here because not computational focused or already covered above -->
 
-<!-- MH: I think most of this is either not relevant here because not computational focused or already covered above -->
+<!-- TODO
 
@@
 
+-->
+
 ### Interpretability
@@
 
-TODO
+<!-- TODO
 - do landscape metrics measure fragmentation?
@@
 
-<!-- CLOSING THOUGHTS AND BRIDGING WITH THE NEXT SECTION -->
+CLOSING THOUGHTS AND BRIDGING WITH THE NEXT SECTION 
 Responses:
@@
 
-<!-- Note that all approaches (landscape metrics, graph and surface - TODO: what about point?) suffer from correlation and redundancy issues, scale dependency and target problem [@Frazier2017]. -->
+Note that all approaches (landscape metrics, graph and surface - TODO: what about point?) suffer from correlation and redundancy issues, scale dependency and target problem [@Frazier2017].
+-->
 
@@
 
-Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009; @Cushman2010] and were adapted from microscopy and molecular physics [@McGarigal2009; @Kedron2018].
-The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010; @Kedron2018].
-Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010].
-Many surface metrics have analogous landscape metrics [@Cushman2010], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009; @Borthwick2020].
-Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018; @Frazier2019].
+Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
+The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
+Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
+Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
+Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
 <!-- MH: What about Spectral and Wavelet Analysis?
@@
 Many landscape ecology studies are based on the raster data model.
-However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019].
+However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
 For example, many LULC maps are provided as raster as they often related to underlying remote sensing products.
@@
 Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
-This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022].
+This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
 Yet, the vector data model has been widely used in other fields, such as geography or urban studies.
-For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017].
-Earlier approaches included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985].
-More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010]
+For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
+Earlier approaches included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
+More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness]
 

### 2024-09-02 Sections/03-connectivity.qmd
Update references

@@
 
-# Connectivity
+# Connectivity {#sec-connectivity}
 
-Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993; @Taylor2006].
-While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006].
-Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016; @Hilty2020].
-Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018; @Keeley2021].
+Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
+While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006LandscapeConnectivity].
+Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
+Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
 
@@
 There are several landscape metrics that are traditionally used to measure landscape connectivity.
-For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002; @Prugh2009; @Spanowicz2019].
-However, these metrics are frequently criticised for being to simple and especially the link to functional connectivity is often ambiguous. 
-Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012].
-More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019].
+For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002SimpleConnectivity; @Prugh2009EvaluationPatch; @Spanowicz2019MeasuringLandscape].
+However, these metrics are frequently criticized for being to simple and especially the link to functional connectivity is often ambiguous. 
+Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
+More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].
 
@@
 Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
-The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012].
-The latest developments of resistance surfaces have recently been reviewed [@Dutta2022].
-The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003; @Etherington2016] or circuit theory [@McRae2008].
-However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022; @UnnithanKumar2022a].
-Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018; @Bocedi2021; @UnnithanKumar2022b], modelling of dynamic landscape connectivity [@Zeller2020], inclusion of stochasticity and spatial context [@Ribeiro2017], using genetic optimization algorithms [@Peterman2018], or computational improvements [@Leonard2017].
+The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
+The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
+The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
+However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
+Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
 
@@
 Graph theory describes a landscape as a set of nodes representing habitat patches and edges as potential links between the nodes.
-Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001].
+Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001LandscapeConnectivity].
 

### 2024-09-02 Sections/05-simulations.qmd
Update references

@@
 
-# Simulation models
+# Simulation models {#sec-simulation}
 
-Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005; @Petrovskii2012].
-Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004].
-Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006; @Synes2016].
-In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004;@Synes2016].
+Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005ComputationalEcology; @Petrovskii2012ComputationalEcology].
+Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
+Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
+In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
 Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.
@@
 
-Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004a; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
-Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; @Langhammer2019].
+Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
+Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].
 
-Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
-The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
-Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
-Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or binary space partitioning [@Etherington2022a].
-Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024].
-In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022].
-<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [ c], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->
+Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
+The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
+Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
+Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
+Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
+In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@VanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
+<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->
 
-Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019; @Salecker2019].
-Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998] or deforestation [@Soares-Filho2002].
-More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006], deforestation based on road and agricultural fields access [@Peer2013], agricultural areas dominated by smallholders [@Salecker2019], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020].
+Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
+Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
+More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
 
@@
 
-Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014].
-In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018; @Dislich2018; @Gonzalez-Redin2019], but also disturbances [e.g., @Seidl2017a; @Spies2017; @Kamimura2019] or connectivity [e.g., @Day2020].
-In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018] or reusable building block [@Berger2024].
-Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016], e.g., linking an individual-based and ecosystem model [@Polhill2011].
+Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
+In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling] or connectivity [e.g., @Day2020IndividualbasedModeling].
+In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building block [@Berger2024ReusableBuilding].
+Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
 
@@
 
-Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020; @DeKoning2023] and there is an increasing interest and use of them in both the industry and academia [@Jones2020].
-The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@DeKoning2023].
-Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023; @Purcell2023].
-Further applications include planning of rural ecological landscapes [@Tan2024], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022].
+Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @DeKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
+The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@DeKoning2023DigitalTwins].
+Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
+Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
-Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@DeKoning2023].
+Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@DeKoning2023DigitalTwins].

### 2024-09-02 Sections/07-genetics.qmd
Update references

---
bibliography: CLER.bib
---

<!-- Person responsible: De Flamingh -->

# Landscape genetics

### 2024-09-02 Sections/08-software.qmd
Update Bib and manuscript
Process to connectivity section
Update references

--- 2024-07-11 Sections/08-software.qmd
@@
 
-<!-- Person responsible: Hesselbarth, Nowosad, Bosch, Jung, Schouten  -->
+<!-- Person responsible: Hesselbarth, Nowosad, Bosch, Jung  -->
 
@@
 
-<!-- "The future applications of landscape pattern metrics will almost certainly depend on the software and analysis tools available" -- @costanzaDescribingAnalyzingLandscape2019 -->
-
-"Open science" involves full transparency and data access at all stages of the research processes and the integrative and multi-disciplinary field of ecology can massively benefit from this concept [@Hampton2015].
-One key aspect of achieving open science is to use open source software, which ensures transparency and consequently increasing the (public) trust into the analyses [@Powers2019].
-Related to this, script-based workflows have the advantage over workflows involving graphical user interfaces that the former are shareable and reproducible [@Hampton2015; @Powers2019].
+"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields such as ecology can massively benefit from this concept [@Hampton2015TaoOpen].
+Using open source software is one key aspect of open science because it ensures transparency and consequently increases trust in the analyses [@Powers2019OpenScience].
+Given the increasing data availability and method complexity, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
+In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015TaoOpen; @Powers2019OpenScience].
 Thus, we will focus the following section on script-based open-source software suitable for landscape ecology.
-Given the increasing availability of data and complexity of methods, future progress within the field of landscape ecology will depend on the software and analysis tools available [@Costanza2019]
 
 <!-- R: Hesselbarth -->
-The open-source and script-based R programming languages [@RCoreTeam2024] was first released in 1995 [@Smith2016] with a focus on statistical computing and data visualization.
-However, due to its extensions (also called packages), the language is popular in many academic fields including ecology [@Lai2019] and it is suitable for spatial data analysis or landscape ecology [e.g., @Bivand2020; @Hesselbarth2021].
+The open-source, script-based R programming languages [@RCoreTeam2024LanguageEnvironment] was first released in 1995 [@Smith201616Years] with a focus on statistical computing and data visualization.
+However, due to its extensions (also called packages), the language is popular in many academic fields, including ecology [@Lai2019EvaluatingPopularity; @Lortie2020ChecklistChoosing], and it is suitable for spatial data analysis and landscape ecology [e.g., @Bivand2020ProgressEcosystem; @Hesselbarth2021OpensourceTools].
+Fundamental packages for spatial data analysis include *sf* for simple-features vector data [@Pebesma2018SimpleFeatures], *terra* for raster data [@Hijmans2024TerraSpatial], *stars* for spatiotemporal arrays, raster, and vector data cubes [@Pebesma2023SpatialData], and *spatstat* for point pattern data [@Baddeley2015SpatialPoint].
+Several packages allow to create maps, such as *tmap* [@Tennekes2018TmapThematic] or *ggplot2* [@Wickham2016Ggplot2Elegant]. 
+For a comprehensive overview of packages related to geographical spatial data, please see the CRAN Task View [@Bivand2024CRANTask].
+R packages developed specifically for spatial pattern analysis include, besides others, *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpen], *vectormetrics* [still being developed; @Matuszek2024VectormetricsLandscape], *motif* [@Nowosad2021MotifOpensource], *rasterdiv* [@Rocchini2021RasterdivInformation], or *geodiv* [@Smith2021GeodivPackage].
+<!-- BESSpatial: https://github.com/Nowosad/bespatial/ -->
+Packages with a focus on scale include e.g., *multilandr* [@Huais2024MultilandrPackage] or *grainchanger* [@Graham2024GrainchangerMovingwindow].
+Connectivity analysis can be conducted using packages including *lconnect* [@Mestre2023LconnectPackage], or *grainscape* [@Chubaty2020ToolboxGrainscape].
+Furthermore, neutral landscapes can be simulated using *NLMR* [@Sciaini2018NLMRLandscapetools], and the *graph4lg* [@Savary2021Graph4lgPackage] provides functions for landscape genetics.
 
-<!-- 
-landscapemetrics
-geodiv
-vectormetrics (also: VecLI -- they mention open source in their paper a few times, but the software seems to be closed source -- https://urbancomp.net/archives/vecliv3beta)
-bespatial
-LandComp
-motif
-?GLCMTextures
-?rasterdiv
-?terra 
--->
-
-<!-- Python: Bosch -->
-
-<!--Julia: Schouten -->
-
-<!--Others: -->
-
-<!-- 
-GuidosToolbox
-?GRASS GIS
-?QGIS: LecoS plugin 
--->
+<!-- Python: Bosch, Jung? -->

### 2024-09-02 Sections/06-spatial-planning.qmd
Minor cosmetic edits to spatial planning

---
bibliography: CLER.bib
---

<!-- Person responsible: Jung -->

# Spatial conservation planning {#sec-spatialplanning}

Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
In the context of landscape-ecology, it provides a pattern-process link <!-- (see section @sec-spatialpattern) -->, and outcomes related to decision making by distinguishing between structure, function and scales [@Hersperger2021LandscapeEcological].
As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology, for example through using simulation model outputs <!-- (see section @sec-simulation) --> as input features  <!-- (see also @daigle2020operationalizing) -->, there have been several recent advances in considering landscape patterns directly in spatial planning.

Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
Furthermore, aggregation can be important because of ecological or practical reasons, for example by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
In particular, recently the direct consideration of connectivity <!-- (see also section @sec-connectivity) --> has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].

Particular noteworthy are two advances that incorporate landscape indices in spatial planning.
One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
These spatial networks approaches have the potential to be provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and indicators to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.

A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathway.
Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatial-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
However, this comes with the drawback of reduced interpretability.

## Open-source software in spatial conservation planning {#sec-spatialplanning-software}

Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
Most software are based on linear or constrained programming approaches requiring a solver to create any outputs.
Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].

The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows to flexible create planning problems in multiple configurations. 
Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
The *CAPTAIN* software suite implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

### 2024-09-03 Manuscript.qmd
Move software to corresponding sections

@@
         \usepackage[noblocks]{authblk}
+        \usepackage{setspace}
         \renewcommand*{\Authsep}{, }
@@
 {{< include Sections/07-genetics.qmd >}}
-{{< include Sections/08-software.qmd >}}
-{{< include Sections/09-discussion.qmd >}}
+{{< include Sections/08-discussion.qmd >}}
 
@@
 
-\begingroup
-\raggedright
+\footnotesize
+\setstretch{1.0}
 ::: {#refs}
 :::
-\endgroup

### 2024-09-03 Sections/01-introduction.qmd
Move software to corresponding sections

@@
 
-# Introduction
+# Introduction {#sec-introduction}
 
@@
 Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984LandscapeEcology; @Turner1989LandscapeEcology; @Turner2005LandscapeEcology].
-Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, linking landscape metrics to ecological processes, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].
+<!-- MH: This sentence needs some editing to be more concise-->
+Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, pattern-process links, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].
 
-Computational science involves analyzing abstracted core mechanisms of research questions using data and algorithms.
-Thus, computational ecology can be defined as computational science that is used to address ecological research questions with focus on complex, adaptive systems and data-driven approaches [@Poisot2019DatabasedSynthesisdriven].
-Computational ecology in general and simulations specifically are of importance for ecology because ecological data is often context- and scale-dependent and unfeasible to study in controllable, reproducible, and replicable experimental settings [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
-Because landscape ecology is a cross-disciplinary field including [e.g., social sciences, geography, or ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the studied complex landscape systems [@Newman2019ScalingComplexity], there is the need for sophisticated computational methods in order to link heterogeneity patterns and ecological processes.
-Thus, computational methods are one of the most important tools of modern scientific research [@Prlic2012TenSimple; @Wilson2014BestPractices].
+Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Prlic2012TenSimple; @Wilson2014BestPractices].
+Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
+Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
+Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
+Thus, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
 
+<!-- MH: Given the limited word count I would cut this
 For all computational methods the natural world must be translated into representative data models.
@@
 The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003ComparisonVector; @Lechner2016RecentProgress; @Nedd2021SynthesisLand].
+-->
 
@@
 For a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
-First, we will introduce recent computational developments of the most important topics of landscape ecology.
-This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations.
+First, we will introduce recent computational developments in the most important topics of landscape ecology.
+<!-- This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations. -->
 Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
-However, due to its many advantages [@Powers2019OpenScience], we will focus on open-source software , and specifically scripting languages, i.e., R, Python, and Julia.
-Last, we will discuss potential future needs developments of computational methods in landscape ecology.
+Lastly, we will discuss potential future directions for the development of computational methods in landscape ecology.
+
+"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields, such as landscape ecology, can massively benefit from this concept [@Hampton2015TaoOpen].
+Using open source software is one key aspect of open science because it facilitates shareability, reproducibility, and transparency and consequently increases trust in the analyses [@Powers2019OpenScience; @Hesselbarth2021OpensourceTools].
+<!-- In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015TaoOpen; @Powers2019OpenScience].
+-->
+Thus, we will focus here on open-source software, specifically scripting languages such as R, Python, and Julia.

### 2024-09-03 Sections/02-spatial-patterns.qmd
Move software to corresponding sections

@@
 
-<!-- jn: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
+<!-- JN: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
 <!-- MH: I think we decided to not reference software here but in a separate section -->
@@
 Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpen].
-Recently also several metrics describing landscape complexity were introduced, e.g., several entropy measures (see following sections).
-
-<!-- https://github.com/r-spatialecology/landcapemetrics -->
-<!-- https://fragstats.org/ -->
+Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
 
@@
 
-<!-- jn: should we use equations here? -->
+<!-- JN: should we use equations here? -->
 <!-- MH: Personally, I think if possible not -->
+<!-- JN: ask T. Stepinski for review of this section later -->
 
-<!-- jn: ask T. Stepinski for review of this section later -->
-
-It is mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
+Entropy measures are mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
 Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
@@
 
-<!-- https://github.com/r-spatialecology/landcapemetrics -->
-
 Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape or @Vranken2015ReviewUse, there was a surge in the development of entropy-based metrics for landscape ecology in the last few years.
@@
 Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
-
-<!-- https://jakubnowosad.com/bespatial/ -->
 
@@
 
+Entropy metrics are available in various software such as *landscapemetrics*, *PyLandStats*, *besspatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], or *LandComp* [@Konrad2023AdaptingMultiscale].
+
 <!-- Kolmogorov, etc. -->
-<!--@Konrad2023; Juhász-Nagy’s information theory-based functions;  https://cran.r-project.org/web/packages/LandComp/index.html -->
 <!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->
@@
 Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].
-
-<!-- https://github.com/Nowosad/motif/ -->
-<!-- https://jakubnowosad.com/spquery/ -->
-<!-- FOSS4G Europe pap? -->
+R packages for such operations include *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource].
+Popular Python libraries for spatial operations include *Geopandas* [@Jordahl2020GeopandasPython], *Shapely*[@Gillies2024ShapelyPython], or *PySal*[@Rey2010PySALPython].
+The *GeoStats.jl* [@Hoffimann2018GeoStatsJl] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.
 
@@
 
+<!-- MH: What about Spectral and Wavelet Analysis? -->
+
 Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
@@
 Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-<!-- MH: What about Spectral and Wavelet Analysis?
-
-<!-- https://www.digitalsurf.com/free-trial/ -->
-<!-- https://github.com/bioXgeo/geodiv -->
+However, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics.
 
@@
 
-<!-- jn: ask T. Matuszek for review of this section later -->
+<!-- JN: ask T. Matuszek for review of this section later -->
 
@@
 Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
-
-<!-- https://github.com/r-spatialecology/vectormetrics -->
+The *vectormetrics* R package, currently under development [@Matuszek2024VectormetricsLandscape], is specifically designed for vector metrics. 
+However, all the general software frameworks mentioned earlier, which support vector data handling, can also be used to calculate basic metrics.

### 2024-09-03 Sections/03-connectivity.qmd
Move software to corresponding sections

@@
 
-<!-- Person responsible: Hesselbarth, ??? -->
+<!-- Person responsible: Hesselbarth, Gerber?, Simpkins? -->
 
@@
 Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
+Software to model connectivity based on resistance surfaces include *ConScape.jl* [@VanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
 

### 2024-09-03 Sections/05-simulations.qmd
Move software to corresponding sections

@@
 In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@VanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
-<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->
+In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand; @Cambui2015GradientLandSoftware]
 

### 2024-09-07 Manuscript.qmd
Adding genetics section by ADF

@@
     affil-id: 2,3
-  - name: Marti Bosch
+  - name: Martí Bosch
     orcid: 0000-0001-8735-9144
@@
     affil-id: 1
+  - name: Alida de Flamingh
+    orcid: 0000-0003-1223-6654
+    email: adeflamingh@gmail.com
+    affil-id: 5
 
@@
     name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
+  - id: 5
+    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
 

### 2024-09-07 Sections/03-connectivity.qmd
Adding genetics section by ADF

--- 2024-09-02 Sections/03-connectivity.qmd
@@
 Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
+Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
 

### 2024-09-07 Sections/05-simulations.qmd
Adding genetics section by ADF

@@
 Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
-In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@VanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
+In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
 In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand; @Cambui2015GradientLandSoftware]
@@
 
-Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @DeKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
-The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@DeKoning2023DigitalTwins].
+Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
+The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@deKoning2023DigitalTwins].
 Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
@@
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
-Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@DeKoning2023DigitalTwins].
+Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-09-07 Sections/07-genetics.qmd
Move software to corresponding sections
Adding genetics section by ADF

@@
 
-<!-- Person responsible: De Flamingh -->
+<!-- Person responsible: de Flamingh -->
 
-# Landscape genetics
+# Landscape genetics {#sec-genetics}
+
+Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns with the ecological processes and patterns typically studied in landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+The field of landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modeling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
+Using holistic approaches that consider evolutionary processes and patterns in addition to only ecological data can fortify results, for example, by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], by allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], and by understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
+
+Advances in molecular methodologies and the synergistic development of bioinformatic and computational approaches to analyze large-scale genomic data have now enabled the integration of genome-wide data with spatial ecological data.
+This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers may apply genomic data across spatiotemporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape] or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
+Analytical frameworks such as the FOLDS model [gene flow, genetic offsets, genetic load, dispersal and SDMs, @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA, @Capblancq2021RedundancyAnalysis] and spatial simulations provide promising potential for future landscape genomic studies.
+Such studies may explore interactions between complex evolutionary processes, e.g., demography, multidimensional (e.g., strength and direction) gene flow and migration across spatiotemporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

### 2024-09-09 Sections/07-genetics.qmd
Minor cosmetic changes genetics

--- 2024-09-02 Sections/07-genetics.qmd
@@
 
-<!-- Person responsible: De Flamingh -->
+<!-- Person responsible: de Flamingh -->
 
-# Landscape genetics
+# Landscape genetics {#sec-genetics}
+Genetic data can be applied to complement ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+Landscape genetic approaches have been used traditionally to inform landscape ecology applications, e.g., through connectivity modeling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
+Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
+
+Advances in molecular methodologies and  synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
+<!-- MH: Ref missing -->.
+This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
+Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs, @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA, @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
+Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

### 2024-09-13 Manuscript.qmd
adds some comments, changes, etc

@@
 papersize: a4paper
-fontsize: 10pt
+fontsize: 12pt
 geometry: margin=25mm
@@
 {{< include Sections/05-simulations.qmd >}}
+<!-- JN: I would suggest switching the order of spatial planning and genetics -->
 {{< include Sections/06-spatial-planning.qmd >}}

### 2024-09-13 Sections/03-connectivity.qmd
adds some comments, changes, etc

--- 2024-09-02 Sections/03-connectivity.qmd
@@
 Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
-Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
+Thus, in the past multiple approaches how to computationally measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
 
@@
 
-## Landscape metrics
+## Landscape metrics related to connectivity
 
@@
 More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].
+<!--JN: software??-->
 
@@
 Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
-The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
-The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
+This concept uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
+The latest developments of resistance surfaces have recently been reviewed by @Dutta2022OverviewComputational.
+
 The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
@@
 Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
+
+<!--maybe add "Julia's", "R's", "Python's" before the software names?-->
+Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
 

### 2024-09-13 Sections/05-simulations.qmd
adds some comments, changes, etc

@@
 Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
-The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
+The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
 Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
@@
 In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
-In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand; @Cambui2015GradientLandSoftware]
+In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python (NLMpy\; @Etherington2015NLMpyPython), R (NLMR; @Sciaini2018NLMRLandscapetools), Julia (NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia), or stand-alone software (GradientLand; @Cambui2015GradientLandSoftware).
 
@@
 More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
+<!--JN: software??-->
 
@@
 
-## Digital twins
+## Digital Twins
 
-Digital Twins include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
+The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
 The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@deKoning2023DigitalTwins].
@@
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
+<!-- JN: "Digital Twins can" or "Digital Twins will" or "Digital Twins aims to"? -->
 Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].
+<!--JN: software??-->

### 2024-09-13 Sections/07-genetics.qmd
adds some comments, changes, etc

@@
 
-Advances in molecular methodologies and  synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
-<!-- MH: Ref missing -->.
+Advances in molecular methodologies and  synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.<!-- MH: Ref missing -->.
 This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
@@
 Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].
+<!--JN: software??-->

### 2024-09-13 Sections/01-introduction.qmd

@@
 
-Landscapes are typically defined as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology; @Forman1995LandMosaics] - with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
+Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology; @Forman1995LandMosaics], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
 Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984LandscapeEcology; @Turner1989LandscapeEcology; @Turner2005LandscapeEcology].
 <!-- MH: This sentence needs some editing to be more concise-->
+<!-- JN: the examples in the following section should be ordered in a similar way as the topics in the paper -->
 Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, pattern-process links, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].
@@
 However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
-For a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
-First, we will introduce recent computational developments in the most important topics of landscape ecology.
+Instead, for a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
+<!-- JN: the first and the second point seem to be mixed in the text atm-->
+Here, we first introduce recent computational developments in the most important topics of landscape ecology.
 <!-- This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations. -->
-Second, we will introduce common software tools that implement or are potentially capable of implementing these developments.
-Lastly, we will discuss potential future directions for the development of computational methods in landscape ecology.
+Second, we present common software tools that implement or are potentially capable of implementing them.
+Lastly, we discuss potential future directions for the development of computational methods in landscape ecology.
 
+<!--JN: I am not the biggest fan of the following definition of open science: full transparency is rarely a goal IMO.-->
 "Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields, such as landscape ecology, can massively benefit from this concept [@Hampton2015TaoOpen].

### 2024-09-13 Sections/02-spatial-patterns.qmd

@@
 Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural word [@Frazier2017LandscapeMetrics].
+The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural world [@Frazier2017LandscapeMetrics].
+
 Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpen].
-Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+
+Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
 
@@
 
-<!-- JN: should we use equations here? -->
-<!-- MH: Personally, I think if possible not -->
 <!-- JN: ask T. Stepinski for review of this section later -->
 
-Entropy measures are mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
-Entropy in landscape ecology is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
-However, studies show that insights do not only depend on the formulation of the used entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix [@Zhao2019CalculatingSpatial].
+Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
+Entropy here is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
+However, studies show that insights do not only depend on the formulation of the selected entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
 
@@
 Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012SpatiotemporalForm].
-However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory can be adapted.
+However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
 @Nowosad2019InformationTheory proposed to treat categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
 This approach allows us to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
-This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (representing the degree of spatial autocorrelation).
+This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).
 
-Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape or @Vranken2015ReviewUse, there was a surge in the development of entropy-based metrics for landscape ecology in the last few years.
-In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
-<!-- REF -->
+Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) (@Gao2019ComputationBoltzmann).
 @Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
-This approach was generalized for calculations based on raster's surface model and point patterns [@Cushman2021EntropyLandscape].
+This approach was later generalized for calculations based on raster's surface model and point patterns [@Cushman2021EntropyLandscape].
 Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
@@
 
-<!--
+<!--JN: I would suggest to include the following paragraph, which I think is important to show the current state of the affairs.-->
 The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
@@
 Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.
--->
 
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
+The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies<!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
 The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRao], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
@@
 
-Entropy metrics are available in various software such as *landscapemetrics*, *PyLandStats*, *besspatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], or *LandComp* [@Konrad2023AdaptingMultiscale].
+Entropy metrics are available in various software such as R's *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or Python's *PyLandStats*.
 
@@
 Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990SpatialTemporal; @Wickham1994MappingAnalyzing; @Jasiewicz2015GeoPATToolbox; @Remmel2020DistributionsHyperlocal; @Nowosad2021MotifOpensource].
-These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between the signatures.
+These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between these signatures.
 
@@
 Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].
+
 R packages for such operations include *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource].
@@
 
+<!--JN: I would restructure the above paragraph to highlight what can and cannot be done with the current software. You may decided to replace the above paragraph with the following one:-->
+<!--
+R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], *rasterio* [@gillies2019] in Python, and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
+-->
+<!--raster package in Julia??-->
+<!--
+These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations, such as spatial clustering, spatial regression, or spatial interpolation.
+Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
+-->
+
 ## Landscape mosaic
 <!-- Bosch -->
+
+<!-- JN: this section seems a bit disconnected from the rest of the text. One option may be to integrate it a bit with the landscape metrics section, and just highlight the most important points.-->
 
@@
 
+<!-- JN: I would suggest to move all of the metrics subsections (landscape, entropy, surface, vector) to the beginning of the section, and just then do the operations on spatial patterns.
+This would make the text more coherent.
+-->
+
 <!-- MH: What about Spectral and Wavelet Analysis? -->
@@
 Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
+
 Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-However, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics.
+Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics.
 
@@
 This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
-Yet, the vector data model has been widely used in other fields, such as geography or urban studies.
+At the same, the vector data model has been widely used in other fields, such as geography or urban studies.
 For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
-Earlier approaches included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
+
+Earlier approaches of quantifying shape complexity  included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
 More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness]
@@
 The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
-<!-- MH: What does topologically correct mean?-->
-Another technical issue is the need for the data to be topologically correct, which is often problematic especially for data from different sources.
+Another technical issue is the need for the data to be topologically correct (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
 Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
-The *vectormetrics* R package, currently under development [@Matuszek2024VectormetricsLandscape], is specifically designed for vector metrics. 
+
+The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics. 
 However, all the general software frameworks mentioned earlier, which support vector data handling, can also be used to calculate basic metrics.

### 2024-09-13 Sections/05-simulations.qmd

@@
 Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
-More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
+More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of an open-pit mine and increase in croplands using Ising model [@stepinski2023kinetic], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
 <!--JN: software??-->

### 2024-09-13 Sections/01-introduction.qmd
Minor cosmetic changes and starting to work on discussion

@@
 Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
-Thus, future progress within the research field will highly depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
-
-<!-- MH: Given the limited word count I would cut this
-For all computational methods the natural world must be translated into representative data models.
-The spatial vector data model is based on points, lines, and polygons features that are defined by their geometry and additional attributes related to them.
-Contrastingly, the raster data model is based on evenly-spaced grid cells and discrete or continuous values related to each cell [@Lovelace2019Geocomputation; @With2019EssentialsLandscape].
-The choice of the data models, spatial and temporal scales, or the thematic resolution are crucial for the analysis [@Wade2003ComparisonVector; @Lechner2016RecentProgress; @Nedd2021SynthesisLand].
--->
 
@@
 Instead, for a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
-<!-- JN: the first and the second point seem to be mixed in the text atm-->
-Here, we first introduce recent computational developments in the most important topics of landscape ecology.
-<!-- This includes, for example, spatial patterns quantification, connectivity and movement, or model simulations. -->
-Second, we present common software tools that implement or are potentially capable of implementing them.
+Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing these developments 
+In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
 Lastly, we discuss potential future directions for the development of computational methods in landscape ecology.
-
-<!--JN: I am not the biggest fan of the following definition of open science: full transparency is rarely a goal IMO.-->
-"Open science" involves full transparency and data access at all stages of the research processes and integrative and multi-disciplinary fields, such as landscape ecology, can massively benefit from this concept [@Hampton2015TaoOpen].
-Using open source software is one key aspect of open science because it facilitates shareability, reproducibility, and transparency and consequently increases trust in the analyses [@Powers2019OpenScience; @Hesselbarth2021OpensourceTools].
-<!-- In this context, script-based workflows have an advantage over workflows involving graphical user interfaces because the former are more shareable and reproducible [@Hampton2015TaoOpen; @Powers2019OpenScience].
--->
-Thus, we will focus here on open-source software, specifically scripting languages such as R, Python, and Julia.

### 2024-09-13 Sections/02-spatial-patterns.qmd
Minor cosmetic changes and starting to work on discussion

@@
 
-Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpen].
+Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpensource].
 
-Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
 
@@
 The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
-@Stepinski2022 and @Huckeba2023 showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
+@Stepinski2022CurbYour and @Huckeba2023MultiscaleSpatial showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
 They argue that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
@@
 The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies<!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
-The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRao], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
+The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
 Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
@@
 Popular Python libraries for spatial operations include *Geopandas* [@Jordahl2020GeopandasPython], *Shapely*[@Gillies2024ShapelyPython], or *PySal*[@Rey2010PySALPython].
-The *GeoStats.jl* [@Hoffimann2018GeoStatsJl] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.
+The *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.
 

### 2024-09-13 Sections/03-connectivity.qmd
Minor cosmetic changes and starting to work on discussion

@@
 However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
-Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
+Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
 
 <!--maybe add "Julia's", "R's", "Python's" before the software names?-->
-Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
+Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapejlSoftware], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
 

### 2024-09-13 Sections/08-discussion.qmd
Minor cosmetic changes and starting to work on discussion

---
bibliography: CLER.bib
---

<!-- Person responsible: Hesselbarth -->

# Discussion {#sec-disc}

Since it emerged in the 1980s [@Turner2005LandscapeEcology], the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
To maximize the benefits of "Open science", such as shareability, reproducibility, and transparency, the research community will gain the most if both existing and new software are made available under open-source licenses and support script-based workflows [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

Furthermore, landscape ecology will benefit and adapt solutions from technological advancements in other fields.
This includes advancements in remote sensing technology, with increased data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems. 
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR [@Crowley2020RemoteSensings].
In combination with other increasingly available data sources, e.g, citizen science data, ecology enters the era of big data requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modeling, or prediction and simulations [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Quantum computing offers a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].

<!-- MH: This is missing final conclusion paragraph -->
<!-- Increasing complexity of pressing issues on different scales, eg. climate change, biodiversity loss etc. -->
<!-- LE can contribute to solutions by connection several fields and scales multi-disciplicanry -->
<!-- Thats cool -->

### 2024-09-14 Sections/03-connectivity.qmd
Just keep typing

--- 2024-09-07 Sections/03-connectivity.qmd
@@
 
-Connectivity describes how movement among resources patches is facilitated or hindered by landscape features such as types, amounts, and arrangement of LULC [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
-While structural connectivity focuses on physical and spatial relationships among resource patches, functional connectivity also includes the movement or flow of individuals and populations through the landscape [@Taylor2006LandscapeConnectivity].
+Landscape connectivity is a fundamental concept in spatial ecology.
+It refers to the degree to which the landscape the landscape facilitates or impedes movement among resource patches [@@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
+Landscape connectivity can be divided into two fundamental types, structural and functional.
+Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
+It is solely a landscape feature and is independent of any particular species' characteristics [@Tischendorf2000LandscapeConnectivity; @CorreaAyram2016HabitatConnectivity].
+Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of a particular species, or set of species, making it both species- and landscape-specific [@Adriaensen2003LCP; @Taylor2006LandscapeConnectivity].
+An often conflated term, habitat connectivity, specifically refers to the degree of functional connectivity between patches of optimal habitat for a species.
+
 Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
@@
 
+## Estimating connectivity
+While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging (Calabrese & Fagan 2004; Taylor et al.
+2006).
+If we restrict ourselves to thinking about the dispersal of organisms from a patch in search of new breeding habitat (e.g., natal dispersal) then what do we need to measure to quantify connectivity?
+Ideally, we would have information for many organisms about the location they started from, high-resolution information about their movement trajectory and the location they ‘finished’ in (or their point of mortality if that occurred).
+Depending on the question of interest, the endpoint of movement could represent a range of alternatives from the short-term movement to access food resources to finding and establishing a territory and successfully breeding.
+From this information, we may be able to infer the elements in the landscape that organisms prefer to move through and how their behaviour changes in them drawing on behavioural landscape ecology (Lima & Zollner 1996) and statistical tools such as statistical state-space models (Jonsen et al.
+2013).
+While high-resolution biotelemetry has made tracking animal movement at high-resolution possible (Kays et al.
+2015), it is still logistically challenging to track many animals over long periods at high spatial resolution.
+Furthermore, we are often interested in predicting the consequences of changes in structural connectivity before they occur.
+Many disciplines (e.g., geography, archaeology, ecology, physics and chemistry) are interested in how matter and energy flow through heterogeneous media (including landscapes), and hence several methods have been developed to quantify connectivity.
+As described above, because dispersal is a fundamental component of metapopulation, metacommunity and metaecosystem theory, adequately describing connectivity is vital for their empirical evaluation.
+
 <!--
-Connectivity modelling in conservation science https://doi.org/10.1038/s41598-022-20370-w
--->
-
 ## Landscape metrics
@@
 More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].
+-->
 
+<!--
 ## Resistance surfaces
@@
 Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
+-->
 
@@
 
+<!--
 ## Graph theory
@@
 Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001LandscapeConnectivity].
+
+-->
 
@@
 
+<!--
 ## Other approaches
+-->
 

### 2024-09-14 Sections/03-connectivity.qmd
Draft V1

@@
 ## Estimating connectivity
-While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging (Calabrese & Fagan 2004; Taylor et al.
-2006).
-If we restrict ourselves to thinking about the dispersal of organisms from a patch in search of new breeding habitat (e.g., natal dispersal) then what do we need to measure to quantify connectivity?
-Ideally, we would have information for many organisms about the location they started from, high-resolution information about their movement trajectory and the location they ‘finished’ in (or their point of mortality if that occurred).
-Depending on the question of interest, the endpoint of movement could represent a range of alternatives from the short-term movement to access food resources to finding and establishing a territory and successfully breeding.
-From this information, we may be able to infer the elements in the landscape that organisms prefer to move through and how their behaviour changes in them drawing on behavioural landscape ecology (Lima & Zollner 1996) and statistical tools such as statistical state-space models (Jonsen et al.
-2013).
-While high-resolution biotelemetry has made tracking animal movement at high-resolution possible (Kays et al.
-2015), it is still logistically challenging to track many animals over long periods at high spatial resolution.
-Furthermore, we are often interested in predicting the consequences of changes in structural connectivity before they occur.
-Many disciplines (e.g., geography, archaeology, ecology, physics and chemistry) are interested in how matter and energy flow through heterogeneous media (including landscapes), and hence several methods have been developed to quantify connectivity.
-As described above, because dispersal is a fundamental component of metapopulation, metacommunity and metaecosystem theory, adequately describing connectivity is vital for their empirical evaluation.
+While connectivity is key to many modern conservation theories and practices, directly quantifying it has proven both technically and conceptually challenging [@Calabrese2004ConnectivityMetrics; @Taylor2006LandscapeConnectivity).
+Data from many individuals recording the start location, trajectory, and endpoint of a movement is required to directly observe connectivity [@Simpkins2018AssessingPerformance].
+Using this data one may be able to infer the elements in the landscape that individuals preferentially move through.
+While constantly improving technlogies has made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking], it is still logistically challenging to track high numbers of individuals for the required time periods and spatial resolution.
+As such in practice connectivity is primarily determined by indirect estimation.
+@Calabrese2004ConnectivityMetrics distinguised three types of connectivity estimates:
+- Structural connectivity: determined by the physical attributes of the landscape (as described above).
+- Potential connectivity: a combination of the physical attributes of the landscape and limited information about the dispersal capabilities of the target species, to estimate how connected a given landscape or patch is for that species.
+- Actual connectivity: Related to observations of individuals moving through a landscape or between patches, providing solid estimates of the linkages between landscpae elements or habitat patches.
 
-<!--
-## Landscape metrics
+### Structural Connectivity Estimation Methods
+Landscape metrics, produced using software such as `landscapemetrics` [@Hesselbarth2019LandscapemetricsOpen], are commonly used to quantify various aspects of landscape structure, decomposed into composition and configuration.
+These metrics describe structural connectivity in terms of:
+- Amount of habitat: Total area of habitat types present.
+- Patch size: Average size of habitat patches.
+- Patch isolation: Average nearest-neighbour distance between patches.
+- Connectance index: the percentage of all possible connections between habitat patches that actually exist in a landscape given a pre-defined maximum distance between patches.
 
-There are several landscape metrics that are traditionally used to measure landscape connectivity.
-For example, the habitat amount within a buffer zone of focal patches, the nearest-neighbor distance, or the connectance index [@Moilanen2002SimpleConnectivity; @Prugh2009EvaluationPatch; @Spanowicz2019MeasuringLandscape].
-However, these metrics are frequently criticized for being to simple and especially the link to functional connectivity is often ambiguous. 
-Thus, more functionality needs to be included using additional ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
-More recently there is also an increasing focus in within-patch connectivity using metrics such as the effective mesh size in addition to the usual measure between-patch connectivity [@Spanowicz2019MeasuringLandscape].
--->
+More recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size in addition to the usual metrics [@Spanowicz2019MeasuringLandscape].
 
-<!--
-## Resistance surfaces
+### Estimating Functional Connectivity Using Resistance Surfaces
+Functional connectivity estimation involves integrating species-specific movement data with landscape features.
+This estimation normally involves the usage of graph-theoretic approaches, though these details are often obscured for convenience by software [@McRae2008UsingCircuit].
+Two primary base approaches are usually applied in modern connectivity estimation:
+- Least-cost paths (LCP)
+- Circuit theory
 
-Resistance surfaces are one of the most popular approaches to model connectivity and developed from simple isolation-by-distance approaches. 
-The approach uses environmental layers and biodiversity data or expert opinions to estimate the cost and of movement through landscapes [@Zeller2012EstimatingLandscape].
-The latest developments of resistance surfaces have recently been reviewed [@Dutta2022OverviewComputational].
-The resistance surface can be used to model movement based on different approaches such as leas-cost paths [@Adriaensen2003ApplicationLeastcost; @Etherington2016LeastcostModelling] or circuit theory [@McRae2008UsingCircuit].
-However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
+A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
+
+These methods all rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
+A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell [@Etherington2016LeastcostModelling].
+The values are based on landscape features and represent factors influencing movement, such as habitat type, topography, or barriers.
+
+Creating a resistance surface involves:
+- Collecting data: Obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
+- Categorisation: Grouping landscape grids into thematically similar categories to capture key elements impacting species movements.
+- Assigning resistance or cost values: Quantifying the value for each cell in the grid using empirical movement data, which can include:
+  - Expert opinion: Very widely used due to its ubiquity and low cost, however the accuracy of using this approach is difficult to measure resulting in uncertainties [@Krueger2012ExpertOpinion].
+  - Detection data: Single point locations of unknown individuals.
+  - Relocation data: Multiple sequential locations of the same individual but at low frequency (e.g. mark recapture).
+  - Pathway data: High-frequency relocation data allowing for movement track inference (e.g., GPS tracking).
+  - Genetic data: Genetic samples used to calculate genetic distances between populations.
+
+Resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
 Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
-Software to model connectivity based on resistance surfaces include *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
--->
 
-<!-- What about multi-species approaches? -->
+#### Least-cost paths (LCP)
+LCP is among the most widely used methods for quantifying functional connectivity.
+The underlying assumption is that organisms will move through landscapes in ways that minimize the cumulative cost of traversing different landscape elements.
+LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003LCP].
+The creation of LCPs is a well established technique and as such there are highly optimised algorithms available for its calculation, often making the approach more computationally effient than others
 
-<!--
-## Graph theory
+By calculating the LCP between two points using a resistance surface, researchers can estimate connectivity between these points based on the accumulated cost along the path.
+This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a cost threshold.
 
-Graph theory describes a landscape as a set of nodes representing habitat patches and edges as potential links between the nodes.
-Thus, graph theory is a powerful tool to model landscape connectivity, especially in a metapopulation context [@Urban2001LandscapeConnectivity].
+#### Circuit theory
+Circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
+In this model, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
+Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
 
--->
-
-<!--
-Coupling landscape graph modeling and biological data https://doi.org/10.1007/s10980-020-00998-7
-Temporal dynamics of landscape connectivity in ever-changing landscapes: a literature review https://doi.org/10.1007/s10980-021-01277-9
-Some methodological considerations to improve landscape planning https://doi.org/10.1016/j.ecolind.2017.05.052
-Comparison of landscape graph modelling methods https://doi.org/10.1007/s10980-020-01164-9
-
--->
-
-<!--
-## Other approaches
--->
-
-<!--
-A stochastic movement simulator improves estimates of landscape connectivity https://doi.org/10.1890/14-1690.1
-Landscape connectivity modeling from the perspective of animal dispersal  https://doi.org/10.1007/s10980-019-00935-3
-Big-data approaches https://doi.org/10.1126/science.abg1780
--->
+Key measures from circuit theory are:
+- Resistance distance: The isolation between pairs of locations measured by incorporating multiple pathways. Locations connected by many low-resistance pathways have lower resistance distances.
+- Current flow: Identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape.

### 2024-09-15 Sections/02-spatial-patterns.qmd
feat: fundamental components of landscape metrics

@@
 Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural world [@Frazier2017LandscapeMetrics].
+<!-- The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural word [@Frazier2017LandscapeMetrics].
+-->
+The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, major limitations have been identified regarding their ability to quantify the spatial structure of landscapes as well as their sensitivity to both spatial scale and thematic resolution [@Gustafson1998QuantifyingLandscape; @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology; @Lausch2015UnderstandingQuantifying; @Frazier2017LandscapeMetrics].
 
@@
 
-Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+<!-- MB: I think we can put the part below within the "landscape metrics" section, otherwise we could add a subsection or even subsubsection entitled "Fundamental components of landscape patterns" or "redundancy and parsimony in landscape metrics"-->
+While recent years have seen the development of a large body of landscape metrics, the growing number of available metrics poses a major challenge in terms of their correlation and redundancy.
+<!-- [@Frazier2017LandscapeMetrics].
+-->
+To address these limitations, researchers have used techniques such as factor analysis, classification trees [@Breiman2017] and principal component analysis (PCA).
+For instance, the seminal study of Riitters et al.
+[@Riitters1995] computed multiple metrics across a variety of landscapes and applied multivariate factor analysis to identify six main components of landscape pattern.
+<!-- namely average patch compaction, image texture, average patch shape, patch perimeter-area scaling, number of attribute classes and large-patch density-area scaling.
+-->
+Another comprehensive work by Cushman et al.
+[@Cushman2008Parsimony] performed a PCA to isolate the structural components of a set of diverse landscapes, and extended the analysis to assess their strength (proportion of variance explained, which can be measured by the component eigenvalue), universality (how present they are accross landscapes) and consistency (how stable they are accross landscapes).
+<!-- At the class level, they find seven "highly universal and consistent structure components" (page 700), namely edge contrast, patch shape complexity, aggregation, nearest neighbor distance, patch dispersion, large patch dominance and neighborhood similarity.
+At the landscape level, five components are identified, namely contagion/diversity (similar to image texture in Riitters et al.
+[@Riitters1995]), large patch dominance, interspersion/juxtaposition, edge contrast, patch shape variability, proximity and nearest neighbor distance.
+Overall, components are more consistent and universal at the landscape level than at the class level.
+-->
+<!-- MB: I suppose we can ommit these:
+- Uuemaa et al.
+[@Uuemaa2011] find four main components: dominance (also identified by Riitters et al.
+[@Riitters1995] and Cushman et al.
+[@Cushman2008Parsimony]), contrast, shape complexity and composition.
+- Additionally, multicollinearity in landscape metrics can be assessed using the Variance Inflation Factor (VIF) [@Plexida2014].
+-->
+<!-- More recently, the study of [@Nowosad2018] divided the entire Earth's land cover map into landscape blocks and performed PCA over a set of landscape metrics. The results suggest that there -->
+More recently, Nowosad and Stepinski [@Nowosad2018] conducted PCA over a set of landscape blocks covering the entire terrestial landmass and resolved two main components of landscape configuration, namely complexity and aggregation (which together explain a 70% of the dataset variance).
+The findings are in line with recent views [@Gustafson2019; @Riitters2019] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency - which are respectively connected to the complexity and aggregation components of Nowosad and Stepinski [@Nowosad2018].
+<!-- Actually, some suggest that maybe only amount matters (Fahrig 2013; 2017). -->
+<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
+
+Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+<!-- Software: Besides FRAGSTATS (to compute metrics), most studies use old-school software such as SAS and CANOCO4 (e.g., [@Cushman2008Parsimony]), STATISTICA (e.g., [@Uuemaa2011]), SPSS (e.g., [@Plexida2014]) -->
+<!-- MB: should we ommit SAS and SPSS? -->
+Given the computed landscape metrics, PCA and other dimension reduction techniques can be performed with stand-alone software such as SAS or SPSS, the R programming language or Python libraries such as statsmodels [@Seabold2010] or scikit-learn [@Pedregosa2011].
+<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology?
+I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
 
@@
 <!-- https://doi.org/10.1371/journal.pone.0304215 -->
-Despite the extensive use of the landscape mosaic model and metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998QuantifyingLandscape; @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology; @Lausch2015UnderstandingQuantifying; @Frazier2017LandscapeMetrics].
-<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
-<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
-<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation.
-Let us go through them in a different order: -->
-Here, we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation [@Dramstad2009SpatialMetrics].
 
@@
 
-### Redundancy
-<!-- (iii) selecting metrics -->
-
-<!-- MH: I think most of this is either not relevant here because not computational focused or already covered above -->
-
-<!-- TODO
-
-- Factor analysis of Riitters et al.
-[@Riitters1995] found six main components: average patch compaction, image texture, average patch shape, patch perimeter-area scaling, number of attribute classes and large-patch density-area scaling.
-
-- Cushman et al.
-[@Cushman2008Parsimony] not only isolate the components but also assess their strength (proportion of variance explained), universality (how present they are accross landscapes) and consistency (how stable they are accross landscapes).
-At the class level, they find seven "highly universal and consistent structure components" (page 700), namely edge contrast, patch shape complexity, aggregation, nearest neighbor distance, patch dispersion, large patch dominance and neighborhood similarity.
-At the landscape level, five components are identified, namely contagion/diversity (similar to image texture in Riitters et al.
-[@Riitters1995], large patch dominance, interspersion/juxtaposition, edge contrast, patch shape variability, proximity and nearest neighbor distance.
-Overall, components are more consistent and universal at the landscape level than at the class level.
-
-- Uuemaa et al.
-[@Uuemaa2011] find four main components: dominance (also identified by Riitters et al.
-[@Riitters1995] and Cushman et al.
-[@Cushman2008Parsimony]), contrast, shape complexity and composition.
-
-- Amount (composition) and adjacency (configuration) are the fundamental elements of pattern [@Riitters2019]
-
-- Entropy [@Nowosad2019a]
-
-- Spatial signatures [@Nowosad2021a]
-
--->
-
 ### Interpretability

### 2024-09-16 Sections/03-connectivity.qmd
One sentence per line

@@
 
-<!-- Person responsible: Hesselbarth, Gerber?, Simpkins? -->
+<!-- Person responsible: Gerber?, Simpkins? -->
 
@@
 Landscape connectivity is a fundamental concept in spatial ecology. 
-It refers to the degree to which the landscape the landscape facilitates or impedes movement among resource patches [@@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
+It refers to the degree to which the landscape the landscape facilitates or impedes movement among resource patches [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity]. 
 Landscape connectivity can be divided into two fundamental types, structural and functional. 
@@
 While constantly improving technlogies has made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking], it is still logistically challenging to track high numbers of individuals for the required time periods and spatial resolution.
-As such in practice connectivity is primarily determined by indirect estimation.
-@Calabrese2004ConnectivityMetrics distinguised three types of connectivity estimates:
+As such in practice connectivity is primarily determined by indirect estimation. @Calabrese2004ConnectivityMetrics distinguised three types of connectivity estimates:
 - Structural connectivity: determined by the physical attributes of the landscape (as described above).
 - Potential connectivity: a combination of the physical attributes of the landscape and limited information about the dispersal capabilities of the target species, to estimate how connected a given landscape or patch is for that species.
-- Actual connectivity: Related to observations of individuals moving through a landscape or between patches, providing solid estimates of the linkages between landscpae elements or habitat patches.
+- Actual connectivity: Related to observations of individuals moving through a landscape or between patches, providing solid estimates of the linkages between landscape elements or habitat patches.
 
@@
 LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003LCP]. 
-The creation of LCPs is a well established technique and as such there are highly optimised algorithms available for its calculation, often making the approach more computationally effient than others
+The creation of LCPs is a well established technique and as such there are highly optimised algorithms available for its calculation, often making the approach more computationally effcient than others
 

### 2024-09-16 Sections/05-simulations.qmd

--- 2024-09-13 Sections/05-simulations.qmd
@@
 In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
-In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python (NLMpy\; @Etherington2015NLMpyPython), R (NLMR; @Sciaini2018NLMRLandscapetools), Julia (NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia), or stand-alone software (GradientLand; @Cambui2015GradientLandSoftware).
+In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
 
@@
 Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
-More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
+More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of an open-pit mine and increase in croplands using Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
 <!--JN: software??-->

### 2024-09-16 Sections/06-spatial-planning.qmd
Move software to corresponding sections

@@
 Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
-In the context of landscape-ecology, it provides a pattern-process link <!-- (see section @sec-spatialpattern) -->, and outcomes related to decision making by distinguishing between structure, function and scales [@Hersperger2021LandscapeEcological].
+In the context of landscape-ecology, it provides a pattern-process link<!-- (see section @sec-spatialpattern) -->, and outcomes related to decision making by distinguishing between structure, function and scales [@Hersperger2021LandscapeEcological].
 As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology, for example through using simulation model outputs <!-- (see section @sec-simulation) --> as input features  <!-- (see also @daigle2020operationalizing) -->, there have been several recent advances in considering landscape patterns directly in spatial planning.
@@
 
-Particular noteworthy are two advances that incorporate landscape indices in spatial planning.
+Particular noteworthy are two advances that incorporate landscape metrics in spatial planning.
 One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
 These spatial networks approaches have the potential to be provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
-For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and indicators to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.
+For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.
 
@@
 
-## Open-source software in spatial conservation planning {#sec-spatialplanning-software}
+<!-- ## Open-source software in spatial conservation planning -->
 
@@
 Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
-The *CAPTAIN* software suite implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
+The *CAPTAIN* Python software suite implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
 However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

### 2024-09-16 Sections/07-genetics.qmd

@@
 # Landscape genetics {#sec-genetics}
+
 Genetic data can be applied to complement ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
@@
 This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
-Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs, @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA, @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
+Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs\; @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA\; @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
 Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

### 2024-09-16 Sections/08-discussion.qmd

@@
 
-<!-- MH: This is missing final conclusion paragraph -->
-<!-- Increasing complexity of pressing issues on different scales, eg. climate change, biodiversity loss etc. -->
-<!-- LE can contribute to solutions by connection several fields and scales multi-disciplicanry -->
-<!-- Thats cool -->
+<!-- MH: Add refs to this -->
+Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them.
+Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution.
+However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions. 
+Here we provided a review of the latest developments, nevertheless, a review of these methods can always be only a snapshot in time and hopefully new computational approaches will emerge in the future.

### 2024-09-17 Manuscript.qmd
review of the discussion by jn

@@
   - id: 2
-    name: Adam Mickiewicz University, Poznan, Poland
+    name: Adam Mickiewicz University, Poznań, Poland
   - id: 3
@@
 papersize: a4paper
-fontsize: 12pt
+fontsize: 10pt
 geometry: margin=25mm

### 2024-09-17 Sections/08-discussion.qmd
review of the discussion by jn

@@
 
-Since it emerged in the 1980s [@Turner2005LandscapeEcology], the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
+Since emerging in the 1980s [@Turner2005LandscapeEcology], the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
 Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
-To maximize the benefits of "Open science", such as shareability, reproducibility, and transparency, the research community will gain the most if both existing and new software are made available under open-source licenses and support script-based workflows [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
+To maximize the benefits of "Open science", such as shareability, reproducibility, and transparency, the research community will gain the most if both existing and new software support script-based workflows and are made available under open-source licenses [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
+<!--JN: maybe add a sentence or two summarizing the previous sections (e.g., that this paper presents various computational methods used in landscape ecology)-->
 
+<!-- JN: the next sentence seems to be broken -->
 Furthermore, landscape ecology will benefit and adapt solutions from technological advancements in other fields.
 This includes advancements in remote sensing technology, with increased data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
-These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR [@Crowley2020RemoteSensings].
-In combination with other increasingly available data sources, e.g, citizen science data, ecology enters the era of big data requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
-Due to its high flexibility and performance artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modeling, or prediction and simulations [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
-Quantum computing offers a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
+These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR[@Crowley2020RemoteSensings].
+In combination with other increasingly available data sources, e.g, citizen science data, ecology is in the era of increasingly large data requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
+Due to its high flexibility and performance, artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modeling, or prediction and simulations [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
+Quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
 
@@
 Here we provided a review of the latest developments, nevertheless, a review of these methods can always be only a snapshot in time and hopefully new computational approaches will emerge in the future.
+
+<!-- JN: additional possible topics to discuss here is that the computational methods are also related to the developments in computer science; thus, new frameworks, tools, good practices, etc.
+are constantly emerging, and not everything is about new methods, but also about new ways to implement them, new ways to share them, better ways to document them, more consistent syntaxes, etc.
+-->

### 2024-09-17 Sections/03-connectivity.qmd
First edits on connectivy section (WIP)

@@
 
-<!-- Person responsible: Gerber?, Simpkins? -->
+<!-- Person responsible: Simpkins, Gerber -->
 
@@
 
-Landscape connectivity is a fundamental concept in spatial ecology. 
-It refers to the degree to which the landscape the landscape facilitates or impedes movement among resource patches [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity]. 
-Landscape connectivity can be divided into two fundamental types, structural and functional. 
+Landscape connectivity describes how landscapes facilitate or impede movement among resource patches [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
+It underpins ecosystem functionality, maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving] and multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
+Connectivity can be divided into structural and functional connectivity.
 Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency. 
-It is solely a landscape feature and is independent of any particular species' characteristics [@Tischendorf2000LandscapeConnectivity; @CorreaAyram2016HabitatConnectivity]. 
-Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of a particular species, or set of species, making it both species- and landscape-specific [@Adriaensen2003LCP; @Taylor2006LandscapeConnectivity].
-An often conflated term, habitat connectivity, specifically refers to the degree of functional connectivity between patches of optimal habitat for a species.
-
-Connectivity underpins ecosystem functionality and maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
-Thus, in the past multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
+It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000LandscapeConnectivity; @CorreaAyram2016HabitatConnectivity]. 
+Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003LCP; @Taylor2006LandscapeConnectivity].
+<!-- An often conflated term, habitat connectivity, specifically refers to the degree of functional connectivity between patches of suitable species habitat. -->
 
 ## Estimating connectivity
-While connectivity is key to many modern conservation theories and practices, directly quantifying it has proven both technically and conceptually challenging [@Calabrese2004ConnectivityMetrics; @Taylor2006LandscapeConnectivity).
-Data from many individuals recording the start location, trajectory, and endpoint of a movement is required to directly observe connectivity [@Simpkins2018AssessingPerformance]. 
-Using this data one may be able to infer the elements in the landscape that individuals preferentially move through. 
-While constantly improving technlogies has made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking], it is still logistically challenging to track high numbers of individuals for the required time periods and spatial resolution.
-As such in practice connectivity is primarily determined by indirect estimation. @Calabrese2004ConnectivityMetrics distinguised three types of connectivity estimates:
-- Structural connectivity: determined by the physical attributes of the landscape (as described above).
-- Potential connectivity: a combination of the physical attributes of the landscape and limited information about the dispersal capabilities of the target species, to estimate how connected a given landscape or patch is for that species.
-- Actual connectivity: Related to observations of individuals moving through a landscape or between patches, providing solid estimates of the linkages between landscape elements or habitat patches.
+
+<!-- MH: Potential shorten here -->
+
+While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging [@Calabrese2004ConnectivityMetrics; @Taylor2006LandscapeConnectivity].
+Data from many individuals recording the start location, trajectory, and endpoint of movement is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
+While improving technologies have made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals for the required tempo-spatial resolutions.
+Thus, in practice connectivity is primarily determined by indirect estimation. 
+@Calabrese2004ConnectivityMetrics distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of the physical attributes of the landscape and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
 
 ### Structural Connectivity Estimation Methods
-Landscape metrics, produced using software such as `landscapemetrics` [@Hesselbarth2019LandscapemetricsOpen], are commonly used to quantify various aspects of landscape structure, decomposed into composition and configuration.
-These metrics describe structural connectivity in terms of:
-- Amount of habitat: Total area of habitat types present.
-- Patch size: Average size of habitat patches.
-- Patch isolation: Average nearest-neighbour distance between patches.
-- Connectance index: the percentage of all possible connections between habitat patches that actually exist in a landscape given a pre-defined maximum distance between patches.
 
-More recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size in addition to the usual metrics [@Spanowicz2019MeasuringLandscape].
+Landscape metrics, produced using software such as `landscapemetrics`, are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration. 
+These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
+Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size. [@Spanowicz2019MeasuringLandscape].
 
 ### Estimating Functional Connectivity Using Resistance Surfaces
+
 Functional connectivity estimation involves integrating species-specific movement data with landscape features. 
 This estimation normally involves the usage of graph-theoretic approaches, though these details are often obscured for convenience by software [@McRae2008UsingCircuit]. 
-Two primary base approaches are usually applied in modern connectivity estimation:
-- Least-cost paths (LCP)
-- Circuit theory
+Two primary base approaches are usually applied in modern connectivity estimation, namely least-cost paths (LCP) and circuit theory.
 
-A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
-
-These methods all rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape]. 
+The methods all rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape]. 
 A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell [@Etherington2016LeastcostModelling].
@@
 
-Creating a resistance surface involves:
-- Collecting data: Obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
-- Categorisation: Grouping landscape grids into thematically similar categories to capture key elements impacting species movements.
-- Assigning resistance or cost values: Quantifying the value for each cell in the grid using empirical movement data, which can include:
-  - Expert opinion: Very widely used due to its ubiquity and low cost, however the accuracy of using this approach is difficult to measure resulting in uncertainties [@Krueger2012ExpertOpinion].
-  - Detection data: Single point locations of unknown individuals.
-  - Relocation data: Multiple sequential locations of the same individual but at low frequency (e.g. mark recapture).
-  - Pathway data: High-frequency relocation data allowing for movement track inference (e.g., GPS tracking).
-  - Genetic data: Genetic samples used to calculate genetic distances between populations.
+Creating a resistance surface involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational]. 
+Next, the cost value for each cell need to be quantified using empirical movement data. 
+This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012ExpertOpinion], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (Genetic samples used to calculate genetic distances between populations).
+
+A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
 
@@
 #### Least-cost paths (LCP)
-LCP is among the most widely used methods for quantifying functional connectivity. 
-The underlying assumption is that organisms will move through landscapes in ways that minimize the cumulative cost of traversing different landscape elements. 
+
 LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003LCP]. 
-The creation of LCPs is a well established technique and as such there are highly optimised algorithms available for its calculation, often making the approach more computationally effcient than others
-
 By calculating the LCP between two points using a resistance surface, researchers can estimate connectivity between these points based on the accumulated cost along the path. 
-This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a cost threshold.
+This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
+The creation of LCPs is a well established technique and there are highly optimised and efficient algorithms available for its calculation.
 
 #### Circuit theory
+
 Circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit]. 
-In this model, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
+In this approach, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
 Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
-
-Key measures from circuit theory are:
-- Resistance distance: The isolation between pairs of locations measured by incorporating multiple pathways. Locations connected by many low-resistance pathways have lower resistance distances.
-- Current flow: Identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape.
+Key measures from circuit theory are the resistance distance and the current flow. 
+The former can describes isolation by measuring multiple low- or high resistance pathways, the later identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape

### 2024-09-18 Manuscript.qmd
Update gitignore and manuscript.qmd

@@
     affil-id: 5
+  - name: Gemma Gerber
+    orcid: 0000-0002-8722-3113
+    email: gerber@iiasa.ac.at
+    affil-id: 1
+  - name: Craig E. Simpkins
+    orcid: 0000-0003-3212-1379
+    email: simpkinscraig063@gmail.com
+    affil-id: 6
 
@@
     name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
+  - id: 6
+    name: Auckland Council, Environmental Services Department, Auckland, New Zealand
 

### 2024-09-18 Sections/01-introduction.qmd
Some cleaning intro and patterns

@@
 
-Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology; @Forman1995LandMosaics], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
-Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Risser1984LandscapeEcology; @Turner1989LandscapeEcology; @Turner2005LandscapeEcology].
-<!-- MH: This sentence needs some editing to be more concise-->
-<!-- JN: the examples in the following section should be ordered in a similar way as the topics in the paper -->
-Typical research topics include ecological flows in landscape mosaics, drivers and consequences of land use and land cover (LULC) change, nonlinear dynamics and landscape complexity, scale and scaling, spatial analysis and landscape modeling, pattern-process links, integrating humans and their activities into landscape ecology, landscape conservation and sustainability, or accuracy assessment and uncertainty analysis [@Wu2002KeyIssues; @Wu2013KeyConcepts].
+Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
+Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
+Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2002KeyIssues; @Wu2013KeyConcepts].
 
-Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Prlic2012TenSimple; @Wilson2014BestPractices].
+Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
 Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
 Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
-Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @Wiens1997MetapopulationDynamics; @With2019EssentialsLandscape], the general increase of data availability [@Chi2016BigData; @Jaric2020IEcologyHarnessing; @Wuest2020MacroecologyAge; @Nathan2022BigdataApproaches], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
+Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
 
@@
 However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
-Instead, for a more comprehensive, general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
-Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing these developments 
+For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
+Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them. 
 In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
-Lastly, we discuss potential future directions for the development of computational methods in landscape ecology.

### 2024-09-18 Sections/02-spatial-patterns.qmd
Some cleaning intro and patterns

--- 2024-09-13 Sections/02-spatial-patterns.qmd
@@
 
-<!-- JN: should we reference software packages here (or in Software) when they implement the methods we discuss? -->
-<!-- MH: I think we decided to not reference software here but in a separate section -->
-<!-- MH: Alternatively, if we move the software section to the beginning, we could reference software here -->
-
 # Spatial patterns {#sec-spatialpattern}
@@
 Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
-However, spatial patterns are not only a result of processes, but could also be a driver of them.
-Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner1989LandscapeEcology].
+However, spatial patterns are not only a result of processes, but could also be drivers of them.
+Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner2015LandscapeEcology].
 
 ## Landscape metrics
-<!-- Hesselbarth -->
 
 Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, they are also critiqued for not being able to realistically represent continuous features of the natural world [@Frazier2017LandscapeMetrics].
+<!-- Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpensource].
+-->
+The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified regarding their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 
-Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpensource].
+<!-- MH: Make sure this includes all similar papers, e.g., Uuemaa2011 -->
+To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995], principal component analysis [@Cushman2008Parsimony], the variance inflation factor [@Plexida2014], or classification trees [@Breiman2017] were used to identify main components of landscape patterns.
+<!-- MH: Add one sentence summarising studies above if possible -->
+More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
+This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018.
+<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
 
-Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology?
+I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
+<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->
+
+<!-- Short section about Landscape mosaic here based on https://doi.org/10.1371/journal.pone.0304215 -->
 
 ## Entropy
-<!-- Nowosad, Bosch -->
 
 <!-- JN: ask T. Stepinski for review of this section later -->
+<!-- MH: Maybe we need to shorten this a bit -->
 
@@
 
-<!--JN: I would suggest to include the following paragraph, which I think is important to show the current state of the affairs.-->
 The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
@@
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies<!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
+The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
 The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
@@
 
-<!-- Kolmogorov, etc. -->
-<!-- "Fractal dimension seems directly related to entropy." @cushmanThermodynamicsLandscapeEcology2015a -->
+## Surface metrics
+
+Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
+The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
+Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
+Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
+
+Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
+Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics.
+
+## Vector metrics
+
+<!-- JN: ask T. Matuszek for review of this section later -->
+
+Many landscape ecology studies are based on the raster data model.
+However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
+For example, many LULC maps are provided as raster as they often related to underlying remote sensing products.
+Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.
+
+Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
+This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
+At the same, the vector data model has been widely used in other fields, such as geography or urban studies.
+For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
+
+Earlier approaches of quantifying shape complexity  included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
+More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness]
+
+Vector-based metrics also have their limitations.
+The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
+Another technical issue is the need for the data to be topologically correct (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
+Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
+
+The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics. 
+However, all the general software frameworks mentioned earlier, which support vector data handling, can also be used to calculate basic metrics.
 
@@
 
-Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping [@Turner1990SpatialTemporal; @Wickham1994MappingAnalyzing; @Jasiewicz2015GeoPATToolbox; @Remmel2020DistributionsHyperlocal; @Nowosad2021MotifOpensource].
-These operations are based on spatial signatures which are multi-numerical representations of landscape pattern, and dissimilarity measures which are functions that quantify differences between these signatures.
+Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping.
+These operations are based on spatial signatures which are multi-numerical representations of landscape pattern and dissimilarity measures which are functions that quantify differences between the signatures.
+Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
 
-Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
 Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
@@
 
-R packages for such operations include *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource].
-Popular Python libraries for spatial operations include *Geopandas* [@Jordahl2020GeopandasPython], *Shapely*[@Gillies2024ShapelyPython], or *PySal*[@Rey2010PySALPython].
-The *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] software provide general spatial data operations in the Julia programming language.
+R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@gillies2019] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
+These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
+Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
 
-<!--JN: I would restructure the above paragraph to highlight what can and cannot be done with the current software. You may decided to replace the above paragraph with the following one:-->
-<!--
-R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], *rasterio* [@gillies2019] in Python, and *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
--->
-<!--raster package in Julia??-->
-<!--
-These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations, such as spatial clustering, spatial regression, or spatial interpolation.
-Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
--->
-
-## Landscape mosaic
-<!-- Bosch -->
-
-<!-- JN: this section seems a bit disconnected from the rest of the text. One option may be to integrate it a bit with the landscape metrics section, and just highlight the most important points.-->
-
-<!-- MH: Is the landscape mosaic after e.g., Vogt et al.
-10.1371/journal.pone.0304215 the same as the patch mosaic model and landscape metrics?
-I think there is some confusion here.
-We already discuss metrics above.
-I think it would be more valuable to discuss the LLM and the most recent additions after Vogt et al.
-here?
--->
-
-<!-- https://doi.org/10.1371/journal.pone.0304215 -->
-Despite the extensive use of the landscape mosaic model and metrics based on it, major limitations have been identified regarding both its ability to quantify the spatial structure of landscapes and their relationship to ecological processes [@Gustafson1998QuantifyingLandscape; @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology; @Lausch2015UnderstandingQuantifying; @Frazier2017LandscapeMetrics].
-<!-- two key considerations to select metrics: (i) ability to consistently identify structural landscape components and (ii) ability to explain relationships with ecological processes (ecological relevance), e.g., biodiversity, often referred to as functional metrics - the challenge of matching the scales of observation to the ecological scales affecting organisms or processes of interest, i.e., the characteristic scales (from "Spatial dependence of predictions from image segmentation: a variogram-based method to determine appropriate scales for producing land-management information") -->
-<!-- Here we focus on the first issue, namely the ability of landscape metrics to consistently capture the spatial structure of landscapes. -->
-<!-- Key issues of structural metrics [@Dramstad2009]: sensitivity to scale, i.e., resolution and extent, sensitivity to the choice of LULC scheme, redundancy and interpretation.
-Let us go through them in a different order: -->
-Here, we review four major issues of landscape metrics, namely their sensitivity to both spatial scale and thematic resolution, redundancy and interpretation [@Dramstad2009SpatialMetrics].
-
-### Sensitivity to spatial scale and thematic resolution
+## Sensitivity to spatial scale and thematic resolution
 
@@
 
-### Redundancy
-<!-- (iii) selecting metrics -->
-
-<!-- MH: I think most of this is either not relevant here because not computational focused or already covered above -->
-
-<!-- TODO
-
-- Factor analysis of Riitters et al.
-[@Riitters1995] found six main components: average patch compaction, image texture, average patch shape, patch perimeter-area scaling, number of attribute classes and large-patch density-area scaling.
-
-- Cushman et al.
-[@Cushman2008Parsimony] not only isolate the components but also assess their strength (proportion of variance explained), universality (how present they are accross landscapes) and consistency (how stable they are accross landscapes).
-At the class level, they find seven "highly universal and consistent structure components" (page 700), namely edge contrast, patch shape complexity, aggregation, nearest neighbor distance, patch dispersion, large patch dominance and neighborhood similarity.
-At the landscape level, five components are identified, namely contagion/diversity (similar to image texture in Riitters et al.
-[@Riitters1995], large patch dominance, interspersion/juxtaposition, edge contrast, patch shape variability, proximity and nearest neighbor distance.
-Overall, components are more consistent and universal at the landscape level than at the class level.
-
-- Uuemaa et al.
-[@Uuemaa2011] find four main components: dominance (also identified by Riitters et al.
-[@Riitters1995] and Cushman et al.
-[@Cushman2008Parsimony]), contrast, shape complexity and composition.
-
-- Amount (composition) and adjacency (configuration) are the fundamental elements of pattern [@Riitters2019]
-
-- Entropy [@Nowosad2019a]
-
-- Spatial signatures [@Nowosad2021a]
-
--->
-
-### Interpretability
+## Interpretability
 <!-- (iv) interpretatability -->
@@
 -->
-
-## Surface metrics
-<!-- Bosch, Hesselbarth -->
-
-<!-- JN: I would suggest to move all of the metrics subsections (landscape, entropy, surface, vector) to the beginning of the section, and just then do the operations on spatial patterns.
-This would make the text more coherent.
--->
-
-<!-- MH: What about Spectral and Wavelet Analysis? -->
-
-Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
-The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
-Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
-Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
-
-Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics.
-
-## Vector metrics
-<!-- Nowosad, Hesselbarth -->
-
-<!-- JN: ask T. Matuszek for review of this section later -->
-
-Many landscape ecology studies are based on the raster data model.
-However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
-For example, many LULC maps are provided as raster as they often related to underlying remote sensing products.
-Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.
-
-Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
-This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
-At the same, the vector data model has been widely used in other fields, such as geography or urban studies.
-For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
-
-Earlier approaches of quantifying shape complexity  included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
-More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness]
-
-Vector-based metrics also have their limitations.
-The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
-Another technical issue is the need for the data to be topologically correct (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
-Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
-
-The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics. 
-However, all the general software frameworks mentioned earlier, which support vector data handling, can also be used to calculate basic metrics.

### 2024-09-18 Sections/02-spatial-patterns.qmd
Update BIB

@@
 <!-- MH: Make sure this includes all similar papers, e.g., Uuemaa2011 -->
-To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995], principal component analysis [@Cushman2008Parsimony], the variance inflation factor [@Plexida2014], or classification trees [@Breiman2017] were used to identify main components of landscape patterns.
+To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify main components of landscape patterns.
 <!-- MH: Add one sentence summarising studies above if possible -->
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018.
+This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
 <!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
 
-Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpen], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
 <!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology?
@@
 
-R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@gillies2019] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
+R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
 These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.

### 2024-09-18 Sections/08-discussion.qmd
Update BIB

@@
 
-Since emerging in the 1980s [@Turner2005LandscapeEcology], the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
+Since emerging in the 1980s <!-- REF -->, the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
 Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].

### 2024-09-18 Sections/03-connectivity.qmd
Adding graph theory sections

@@
 Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency. 
-It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000LandscapeConnectivity; @CorreaAyram2016HabitatConnectivity]. 
-Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003LCP; @Taylor2006LandscapeConnectivity].
-<!-- An often conflated term, habitat connectivity, specifically refers to the degree of functional connectivity between patches of suitable species habitat. -->
+It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity]. 
+Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].
 
@@
 
-While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging [@Calabrese2004ConnectivityMetrics; @Taylor2006LandscapeConnectivity].
+While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
 Data from many individuals recording the start location, trajectory, and endpoint of movement is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
-While improving technologies have made tracking animal movement at high-resolution possible [@Kays2015AnimalTracking; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals for the required tempo-spatial resolutions.
+While improving technologies have made tracking animal movement at high-resolution possible [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals for the required tempo-spatial resolutions.
 Thus, in practice connectivity is primarily determined by indirect estimation. 
-@Calabrese2004ConnectivityMetrics distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of the physical attributes of the landscape and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
+@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of the physical attributes of the landscape and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
 
@@
 Next, the cost value for each cell need to be quantified using empirical movement data. 
-This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012ExpertOpinion], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (Genetic samples used to calculate genetic distances between populations).
+This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (Genetic samples used to calculate genetic distances between populations).
 
-A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapeJl], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
+A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapejlSoftware], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
 
 Resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape]. 
-Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifterExtended; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
+Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
 
@@
 
-LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003LCP]. 
+LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost]. 
 By calculating the LCP between two points using a resistance surface, researchers can estimate connectivity between these points based on the accumulated cost along the path. 
@@
 The former can describes isolation by measuring multiple low- or high resistance pathways, the later identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape
+
+#### Graph theory
+
+Following the formative work of @Urban2001LandscapeConnectivity, graph theory has rapidly developed into a cornerstone approach to studying landscape connectivity by integrating changes in landscape spatial structure and species movement or dispersal in a landscape graph [or habitat network\; @Cushman2013BiologicalCorridors].
+In graph-based landscape connectivity studies, the system boundary is usually delineated as the spatial extent of the modeled landscape. 
+Within the landscape, discrete habitat patches (suitable areas for a species) are modeled as nodes, and the potential dispersal pathways or species movements between habitat patches are modeled as edges.
+The edges can be binary (or "unweighted"), representing whether a connection between two nodes exists or weighted by a value representing individual dispersal effort from one habitat patch to another [@Petsas2021LandscapeConnectivity].
+
+From the modeled landscape graphs, landscape connectivity can be quantified through various graph-theoretical metrics. 
+Each metric provides complementary connectivity information, and some can be used to determine the relative importance of a patch in maintaining connectivity. 
+Currently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for a species’s spatial use and movement [@Petsas2021LandscapeConnectivity].
+Widely reported landscape connectivity metrics include the probability of connectivity index [PC\; for details see @Saura2007NewHabitat], a habitat availability metric quantifying potential and functional connectivity within a landscape [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], and the integral index of connectivity [IIC\; @Pascual-Hortal2006ComparisonDevelopment], which is based on a simple, binary connection model in which two habitat patches are either connected by an edge or not [@Saura2010CommonCurrency].
+
+Graph theory and connectivity metrics have broad utility in landscape connectivity studies, and can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], and require very little data input [@Godet2021ComparisonLandscape].
+This approach can provide key information for conservation planning and management. 
+For example, it can be used to identify critical habitat patches and their edges that facilitate movement or dispersal within a landscape. 
+This information can assist in planning corridors that facilitate wildlife movement [e.g., @Wang2023OptimalConfiguration], planning infrastructure (e.g., roads) placement to minimize lost connectivity due to habitat fragmentation [@Pietsch2018ContributionConnectivity], and identifying which habitat patches could be prioritized for protection and restoration to maintain overall landscape connectivity [@Liu2020UsingLandscape].
+
+Graph theory and associated connectivity metrics in landscape connectivity studies are constantly evolving. 
+The modeling framework, initially only applied to one species at a time, is increasingly being applied to model multispecies landscapes [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach], which account for multiple species’ different movement abilities and habitat preferences, and can be used to design conservation strategies that benefit a more comprehensive range of species.
+In addition, newer methodologies offer promising novel avenues to model not only spatial graphs, but spatiotemporal graphs, where the nodes and edges are represented in space and time through a multiple-layer graph approach [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
+Modeling the temporal dynamics of connectivity helps researchers identify habitat patches that have changed in connectivity, or not, over time, and how these changes affect(ed) biodiversity patterns [@Uroy2021HowAssess].
+In addition, new connectivity metrics are frequently being developed [@Hashemi2022ReviewEcological], often building upon and addressing the caveats of previous metrics.
+
+Despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, there are many ways to construct a habitat graph with different sources of information, which could lead to different results.
+For example, @Godet2021ComparisonLandscape modeled the same landscape using three different types of data, all of which resulted in large differences in the statistical distribution of connectivity values.
+In many cases, data availability limits the models' representation of the studied landscape, with subsequent implications for the calculated connectivity metrics and inferences of landscape connectivity [Foltete2020CouplingLandscape].
+
+Currently, graph-theoretical landscape graphs and connectivity indices can be calculated using several open-source software, including *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], R packages *lconnect* [@Mestre2023LconnectPackage], *GRAINSCAPE* [@Chubaty2020ToolboxGrainscape], and Julia’s *ConScape* library [@vanMoorter2023AcceleratingAdvances], among others.

### 2024-09-18 Sections/03-connectivity.qmd
Remove trailing whitespaces

@@
 Landscape connectivity describes how landscapes facilitate or impede movement among resource patches [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
-It underpins ecosystem functionality, maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving] and multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
+It underpins ecosystem functionality, maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
+Thus, multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
 Connectivity can be divided into structural and functional connectivity.

### 2024-09-18 Sections/03-connectivity.qmd
WIP: Shorten connectiviy section

@@
 
-Landscape connectivity describes how landscapes facilitate or impede movement among resource patches [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
+Landscape connectivity describes how landscape featuers facilitate or impede movement [@Taylor2006LandscapeConnectivity].
 It underpins ecosystem functionality, maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
-Thus, multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
 Connectivity can be divided into structural and functional connectivity.
@@
 
-## Estimating connectivity
+Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, simultaneously quantifying it has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
+Data from many individuals recording the start location, trajectory, and endpoint of movement is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
+While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
+Thus, in practice connectivity is primarily determined by indirect estimations.
+@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
 
-<!-- MH: Potential shorten here -->
+### Estimating structural connectivity
 
-While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
-Data from many individuals recording the start location, trajectory, and endpoint of movement is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
-While improving technologies have made tracking animal movement at high-resolution possible [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals for the required tempo-spatial resolutions.
-Thus, in practice connectivity is primarily determined by indirect estimation.
-@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of the physical attributes of the landscape and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
-
-### Structural Connectivity Estimation Methods
-
-Landscape metrics, produced using software such as `landscapemetrics`, are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
+Landscape metrics, are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
 These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
@@
 
-### Estimating Functional Connectivity Using Resistance Surfaces
+### Estimating functional connectivity
+
+#### Resistance surfaces
 
 Functional connectivity estimation involves integrating species-specific movement data with landscape features.
-This estimation normally involves the usage of graph-theoretic approaches, though these details are often obscured for convenience by software [@McRae2008UsingCircuit].
-Two primary base approaches are usually applied in modern connectivity estimation, namely least-cost paths (LCP) and circuit theory.
+Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
+A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
 
-The methods all rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
-A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell [@Etherington2016LeastcostModelling].
-The values are based on landscape features and represent factors influencing movement, such as habitat type, topography, or barriers.
-
-Creating a resistance surface involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
-Next, the cost value for each cell need to be quantified using empirical movement data.
-This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (Genetic samples used to calculate genetic distances between populations).
+Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
+Next, cost values for each cell need to be quantified using movement data.
+This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
 
@@
 
-Resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
+However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
 Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
 
-#### Least-cost paths (LCP)
+Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
+By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
+This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
+The creation of least-cost paths is a well established technique and there are highly optimised and efficient algorithms available for its calculation.
 
-LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
-By calculating the LCP between two points using a resistance surface, researchers can estimate connectivity between these points based on the accumulated cost along the path.
-This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
-The creation of LCPs is a well established technique and there are highly optimised and efficient algorithms available for its calculation.
-
-#### Circuit theory
-
-Circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
+<!-- This is missing some refs -->
+<!-- Move to next section -->
+Contrastingly, circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
 In this approach, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.

### 2024-09-19 Sections/03-connectivity.qmd
Update and shorten connectivity

@@
 
-Landscape connectivity describes how landscape featuers facilitate or impede movement [@Taylor2006LandscapeConnectivity].
-It underpins ecosystem functionality, maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
+Landscape connectivity describes how landscape features facilitate or impede movement [@Taylor2006LandscapeConnectivity].
+It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
 Connectivity can be divided into structural and functional connectivity.
@@
 It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
-Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].
+Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioral, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].
 
-Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, simultaneously quantifying it has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
-Data from many individuals recording the start location, trajectory, and endpoint of movement is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
+Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
+Movement data from many individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
 While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
@@
 
-Landscape metrics, are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
+Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
 These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
-Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size. [@Spanowicz2019MeasuringLandscape].
+Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
+However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, however, functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012].
 
@@
 
-Functional connectivity estimation involves integrating species-specific movement data with landscape features.
 Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
 A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
-
-Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
-Next, cost values for each cell need to be quantified using movement data.
-This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
+Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational] and quantifying cost values for each cell using movement data.
+This can included expert opinion [widely used due to low effort, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
 
 A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapejlSoftware], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
+
+Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
+By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
+This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
+The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.
+
+Contrastingly, circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
+In this approach, the landscape is represented as a network of electrical nodes connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
+Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
+It can describes isolation by measuring multiple low- or high resistance pathways and identify areas of high movement probability highlighting important corridors and bottlenecks in the landscape [@Cushman2013BiologicalCorridors].
 
@@
 
-Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
-By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
-This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
-The creation of least-cost paths is a well established technique and there are highly optimised and efficient algorithms available for its calculation.
-
-<!-- This is missing some refs -->
-<!-- Move to next section -->
-Contrastingly, circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
-In this approach, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
-Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
-Key measures from circuit theory are the resistance distance and the current flow.
-The former can describes isolation by measuring multiple low- or high resistance pathways, the later identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape
-
 #### Graph theory
 
-Following the formative work of @Urban2001LandscapeConnectivity, graph theory has rapidly developed into a cornerstone approach to studying landscape connectivity by integrating changes in landscape spatial structure and species movement or dispersal in a landscape graph [or habitat network\; @Cushman2013BiologicalCorridors].
-In graph-based landscape connectivity studies, the system boundary is usually delineated as the spatial extent of the modeled landscape.
-Within the landscape, discrete habitat patches (suitable areas for a species) are modeled as nodes, and the potential dispersal pathways or species movements between habitat patches are modeled as edges.
-The edges can be binary (or "unweighted"), representing whether a connection between two nodes exists or weighted by a value representing individual dispersal effort from one habitat patch to another [@Petsas2021LandscapeConnectivity].
+Following the formative work of @Urban2001LandscapeConnectivity, graph theory has developed into a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
+Graph theory and connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
+For example, it can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity for protection and restoration [@Liu2020UsingLandscape].
 
-From the modeled landscape graphs, landscape connectivity can be quantified through various graph-theoretical metrics.
-Each metric provides complementary connectivity information, and some can be used to determine the relative importance of a patch in maintaining connectivity.
-Currently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for a species’s spatial use and movement [@Petsas2021LandscapeConnectivity].
-Widely reported landscape connectivity metrics include the probability of connectivity index [PC\; for details see @Saura2007NewHabitat], a habitat availability metric quantifying potential and functional connectivity within a landscape [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], and the integral index of connectivity [IIC\; @Pascual-Hortal2006ComparisonDevelopment], which is based on a simple, binary connection model in which two habitat patches are either connected by an edge or not [@Saura2010CommonCurrency].
+Within the landscape delineating the system boundary, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
+The edges can be binary, representing whether a connection between two nodes exists or not, or weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
+Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
+Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment] which is based on a binary connection model [@Saura2010CommonCurrency].
 
-Graph theory and connectivity metrics have broad utility in landscape connectivity studies, and can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], and require very little data input [@Godet2021ComparisonLandscape].
-This approach can provide key information for conservation planning and management.
-For example, it can be used to identify critical habitat patches and their edges that facilitate movement or dispersal within a landscape.
-This information can assist in planning corridors that facilitate wildlife movement [e.g., @Wang2023OptimalConfiguration], planning infrastructure (e.g., roads) placement to minimize lost connectivity due to habitat fragmentation [@Pietsch2018ContributionConnectivity], and identifying which habitat patches could be prioritized for protection and restoration to maintain overall landscape connectivity [@Liu2020UsingLandscape].
+Recently, a new generation of graph-theoretical metrics, commonly reported in combination and addressing caveats of previous metrics [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
+The modeling framework, initially only applied to single species is increasingly being used to model multispecies landscapes [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach], which account for interspecific movement abilities and habitat preferences.
+In addition, modern multiple-layer graph approach offer to model not only spatial but spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
+Modeling the temporal dynamics helps to identify habitat patches impact connectivity over time, and how these impacts affect biodiversity patterns [@Uroy2021HowAssess].
 
-Graph theory and associated connectivity metrics in landscape connectivity studies are constantly evolving.
-The modeling framework, initially only applied to one species at a time, is increasingly being applied to model multispecies landscapes [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach], which account for multiple species’ different movement abilities and habitat preferences, and can be used to design conservation strategies that benefit a more comprehensive range of species.
-In addition, newer methodologies offer promising novel avenues to model not only spatial graphs, but spatiotemporal graphs, where the nodes and edges are represented in space and time through a multiple-layer graph approach [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
-Modeling the temporal dynamics of connectivity helps researchers identify habitat patches that have changed in connectivity, or not, over time, and how these changes affect(ed) biodiversity patterns [@Uroy2021HowAssess].
-In addition, new connectivity metrics are frequently being developed [@Hashemi2022ReviewEcological], often building upon and addressing the caveats of previous metrics.
+However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge due to various ways to construct a graph and different data sources.
+In many cases, data availability limits the representation of the modelled landscape, with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
+For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types resulting in different distribution of connectivity values.
 
-Despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, there are many ways to construct a habitat graph with different sources of information, which could lead to different results.
-For example, @Godet2021ComparisonLandscape modeled the same landscape using three different types of data, all of which resulted in large differences in the statistical distribution of connectivity values.
-In many cases, data availability limits the models' representation of the studied landscape, with subsequent implications for the calculated connectivity metrics and inferences of landscape connectivity [Foltete2020CouplingLandscape].
-
-Currently, graph-theoretical landscape graphs and connectivity indices can be calculated using several open-source software, including *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], R packages *lconnect* [@Mestre2023LconnectPackage], *GRAINSCAPE* [@Chubaty2020ToolboxGrainscape], and Julia’s *ConScape* library [@vanMoorter2023AcceleratingAdvances], among others.
+Currently, graph-theoretical landscape graphs and connectivity indices can be calculated using several open-source software, including *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], the R packages *lconnect* [@Mestre2023LconnectPackage] and *grainscape* [@Chubaty2020ToolboxGrainscape], or Julia’s *ConScape* library [@vanMoorter2023AcceleratingAdvances], among others.

### 2024-09-19 Manuscript.qmd
Remove trailing whitespaces
Updates to general MS structure

--- 2024-09-17 Manuscript.qmd
@@
 
-keywords: keyword1, keyword2
+keywords: spatial data, spatial patterns, open-source, scientific software
 
@@
     affil-id: 1
-    attributes:
-      - corresponding: boolean
   - name: Jakub Nowosad
@@
     affil-id: 2,3
-  - name: Martí Bosch
-    orcid: 0000-0001-8735-9144
-    email: marti.bosch@epfl.ch
+  - name: Gemma Gerber
+    orcid: 0000-0002-8722-3113
+    email: gerber@iiasa.ac.at
+    affil-id: 1
+  - name: Alida de Flamingh
+    orcid: 0000-0003-1223-6654
+    email: adeflamingh@gmail.com
     affil-id: 4
@@
     affil-id: 1
-  - name: Alida de Flamingh
-    orcid: 0000-0003-1223-6654
-    email: adeflamingh@gmail.com
+  - name: Craig E. Simpkins
+    orcid: 0000-0003-3212-1379
+    email: simpkinscraig063@gmail.com
     affil-id: 5
+  - name: Martí Bosch
+    orcid: 0000-0001-8735-9144
+    email: marti.bosch@epfl.ch
+    affil-id: 6
 
@@
   - id: 4
+    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
+  - id: 5
+    name: Auckland Council, Environmental Services Department, Auckland, New Zealand
+  - id: 6
     name: École polytechnique fédérale de Lausanne, Lausanne, Switzerland
-  - id: 5
-    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
 
@@
 
-<!-- MH: Author order is not fixed yet. I just started to list everyone for now. -->
+# Abstract {.unnumbered}
+
+This is a test!
 
@@
 {{< include Sections/03-connectivity.qmd >}}
-{{< include Sections/04-sampling.qmd >}}
+<!-- {{< include Sections/04-sampling.qmd >}} -->
 {{< include Sections/05-simulations.qmd >}}
-<!-- JN: I would suggest switching the order of spatial planning and genetics -->
-{{< include Sections/06-spatial-planning.qmd >}}
-{{< include Sections/07-genetics.qmd >}}
+{{< include Sections/06-genetics.qmd >}}
+{{< include Sections/07-spatial-planning.qmd >}}
 {{< include Sections/08-discussion.qmd >}}
+
+# Acknowledgments {.unnumbered}
+
+We are thankful to the anonymous reviewers and the Editors who provided valuable comments to improve the manuscript.
+
+## Funding {.unnumbered}
+
+No external funding was received.
+
+## Author contributions {.unnumbered}
+
+All authors contributed particular sections to the manuscript coordinated by MHKH.
+Correspondence to [Maximilian H.K. Hesselbarth](mailto:mhk.hesselbarth@gmail.com)
+
+## Conflict of Interest {.unnumbered}
+
+The authors declare no competing interests.
+
+## Human and Animal Rights and Informed Consent {.unnumbered}
+
+This article does not contain any studies with human or animal subjects performed by any of the authors.
 
@@
 
-### References {.unnumbered}
+## References {.unnumbered}
 

### 2024-09-19 Sections/03-connectivity.qmd
Updates to general MS structure

@@
 Movement data from many individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
-While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
+While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
 Thus, in practice connectivity is primarily determined by indirect estimations.
@@
 
-### Estimating structural connectivity
+## Estimating structural connectivity
 
@@
 Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
-However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, however, functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012].
+However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, however, functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
 
-### Estimating functional connectivity
-
-#### Resistance surfaces
+## Resistance surfaces
 
@@
 
-#### Graph theory
+## Graph theory
 

### 2024-09-19 Sections/07-spatial-planning.qmd
Updates to general MS structure

--- 2024-07-01 Sections/07-spatial-planning.qmd
@@
 
-# Spatial planning
+# Spatial conservation planning {#sec-spatialplanning}
 
-Landscape ecology is fun [@Turner1989].
-There are many things to explore [@With2019].
+Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
+In the context of landscape-ecology, it provides a pattern-process link<!-- (see section @sec-spatialpattern) -->, and outcomes related to decision making by distinguishing between structure, function and scales [@Hersperger2021LandscapeEcological].
+As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology, for example through using simulation model outputs <!-- (see section @sec-simulation) --> as input features  <!-- (see also @daigle2020operationalizing) -->, there have been several recent advances in considering landscape patterns directly in spatial planning.
+
+Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
+Furthermore, aggregation can be important because of ecological or practical reasons, for example by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
+Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
+Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
+In particular, recently the direct consideration of connectivity <!-- (see also section @sec-connectivity) --> has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].
+
+Particular noteworthy are two advances that incorporate landscape metrics in spatial planning.
+One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
+These spatial networks approaches have the potential to be provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
+For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.
+
+A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathway.
+Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
+Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
+Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatial-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
+However, this comes with the drawback of reduced interpretability.
+
+<!-- ## Open-source software in spatial conservation planning -->
+
+Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
+Most software are based on linear or constrained programming approaches requiring a solver to create any outputs.
+Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].
+
+The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows to flexible create planning problems in multiple configurations.
+Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
+The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
+Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
+The *CAPTAIN* Python software suite implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
+However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

### 2024-09-19 Manuscript.qmd
Adding abstract

@@
 
-This is a test!
+## Purpose of Review {.unnumbered}
+
+Landscapes can be defined as mosaics of different land covers, habitats, ecosystems, or land-use systems. 
+The link between spatial heterogeneity and ecological processes is the core concept in the research field of landscape ecology. 
+Due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems, advanced computational methods are essential.
+
+## Recent findings {.unnumbered}
+
+This review provides an overview of recent developments in computational methods that have advanced landscape ecology research. 
+We focus on key topics such as spatial patterns, connectivity, simulations and modeling, landscape genetics, and spatial planning.
+
+## Summary {.unnumbered}
+
+The review highlights key innovations, challenges, and potential future directions in the field, emphasizing the role of computational methods in addressing complex ecological questions.
+
+{{< pagebreak >}}
 

### 2024-09-19 Sections/08-discussion.qmd
Remove trailing whitespaces
Update Discussion

@@
 
-Since emerging in the 1980s <!-- REF -->, the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
+Since emerging in the 1980s [@With2019EssentialsLandscape], the field of landscape ecology has constantly evolved and is now highly depending on computational methods.
 Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
-To maximize the benefits of "Open science", such as shareability, reproducibility, and transparency, the research community will gain the most if both existing and new software support script-based workflows and are made available under open-source licenses [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
-<!--JN: maybe add a sentence or two summarizing the previous sections (e.g., that this paper presents various computational methods used in landscape ecology)-->
+Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
+In this context, we highlight open-source software as a cornerstone of Open Science, offering key advantages like shareability, reproducibility, and transparency, which provide the greatest benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
 
-<!-- JN: the next sentence seems to be broken -->
-Furthermore, landscape ecology will benefit and adapt solutions from technological advancements in other fields.
-This includes advancements in remote sensing technology, with increased data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
+Alongside methodological advancements specific to landscape ecology, the research field will also benefit from and integrate technological innovations from related disciplines.
+This includes advancements in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
 These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR[@Crowley2020RemoteSensings].
-In combination with other increasingly available data sources, e.g, citizen science data, ecology is in the era of increasingly large data requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
-Due to its high flexibility and performance, artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modeling, or prediction and simulations [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
+In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
+Due to its high flexibility and performance, artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
 Quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
+Furthermore, landscape ecology will benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].
 
-<!-- MH: Add refs to this -->
-Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them.
+Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
 Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution.
@@
 Here we provided a review of the latest developments, nevertheless, a review of these methods can always be only a snapshot in time and hopefully new computational approaches will emerge in the future.
-
-<!-- JN: additional possible topics to discuss here is that the computational methods are also related to the developments in computer science; thus, new frameworks, tools, good practices, etc.
-are constantly emerging, and not everything is about new methods, but also about new ways to implement them, new ways to share them, better ways to document them, more consistent syntaxes, etc.
--->

### 2024-09-19 Sections/06-genetics.qmd
Updates to general MS structure
Adding software to genetics

--- 2024-09-16 Sections/07-genetics.qmd
@@
 Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
-Landscape genetic approaches have been used traditionally to inform landscape ecology applications, e.g., through connectivity modeling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
+Landscape genetic approaches have been used traditionally to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
 Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
@@
 Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].
-<!--JN: software??-->
+
+Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
+Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], besides others.
+Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

### 2024-09-19 Sections/02-spatial-patterns.qmd
Remove trailing whitespaces
Clean and update spatial patterns

@@
 Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-<!-- Metrics can be calculated on patch-level (describing each patch), class-level (describing all patches of the same class), or landscape level (describing the entire landscape) and are able to quantify characteristics related to e.g., patch areas and edge lengths, patch shapes, or class aggregation and diversity [@Hesselbarth2019LandscapemetricsOpensource].
--->
 The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified regarding their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 
-<!-- MH: Make sure this includes all similar papers, e.g., Uuemaa2011 -->
-To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify main components of landscape patterns.
-<!-- MH: Add one sentence summarising studies above if possible -->
+To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify main components of landscape patterns.
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
@@
 
-Software to calculate landscape metrics include the *landscapemetrics* R package [@Hesselbarth2019LandscapemetricsOpensource], the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
 <!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology?
@@
 
-<!-- Short section about Landscape mosaic here based on https://doi.org/10.1371/journal.pone.0304215 -->
+## Landscape Mosaic
+
+The Landscape Mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes.
+The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
+These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class. 
+This allows the assessment of content, context, and interface zones of LULC classes in a scale-dependent manner. 
+Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].
 
@@
 
-Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics, such as Shannon entropy (and its analogous forms), Boltzmann entropy, or Renyi entropy.
-Entropy here is mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity) and scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
+Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics.
+They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
 However, studies show that insights do not only depend on the formulation of the selected entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
 
-The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape omitting spatial configuration.
-The index is based on information theory and is an adaptation of the Shannon entropy which measures uncertainty of a random variable.
+The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape omitting the spatial configuration.
 Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012SpatiotemporalForm].
 However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
-@Nowosad2019InformationTheory proposed to treat categorical raster as a bivariate random variable $(x,y)$ where $x$ is the category of the focal cell and $y$ is the category of the neighboring cell.
-This approach allows us to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
+@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
 This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).
@@
 Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
-In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) (@Gao2019ComputationBoltzmann).
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
 @Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
@@
 Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
-
-The relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion.
-@Stepinski2022CurbYour and @Huckeba2023MultiscaleSpatial showed that the Shannon entropy is a general form of entropy, and other entropies, including the Boltzmann entropy, could be derived as specific applications of the Shannon entropy.
-They argue that the newly proposed entropy measures inspired by the Boltzmann equation have no connotation to statistical mechanics and thermodynamics.
-Thus, more research is needed to understand the relations between the entropy measures and how they relate to the environmental processes.
+However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial]
 
@@
 Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics.
+Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
+Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].
 
@@
 Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
-
-## Sensitivity to spatial scale and thematic resolution
-
-<!-- MH: Suggestion is to focus much more on computational issues here.
-Currently, this is more about landscape metrics and their behavior in general.
-I like the point about growing availability of remote sensing products.
-Maybe we can focus a bit more on that and cut a lot of the more general, theoretical behavior parts?
--->
-
-<!-- (i) scale: https://doi.org/10.1007/s10980-014-0066-3 "Behavior of class-level landscape metrics across gradients of class aggregation and area"-->
-<!-- "identifying the characteristic scales of relevant patterns and processes is a critical first step in designing a successful research project" [@Wu2007, pp. 119]. -->
-Identifying scales at which the ecological processes of interest operate is a critical first step towards establishing relationships between pattern and process [@Levin1992ProblemPattern; @Wu2007ScaleScaling; @Cushman2008LandscapeMetrics].
-To that end, a key requirement is to understand the sensitivity of landscape metrics to spatial scale in terms of both grain and extent.
-
-A key concept to consider regarding grain size is the minimum mapping unit (MMU), namely the smallest area entity that can be mapped and analyzed as a discrete patch.
-<!-- Accordingly, a grain size several times smaller than the smallest patch in order to grasp (even schematically) its shape. -->
-Since the input maps should represent not only the location of patches but also their shape, the grain size should be several times smaller than the smallest patch (i.e., the MMU) in order to quantify shapes.
-Even though the MMU is often assumed as a natural characteristic of the dataset, it has an impact on the computed landscape metrics, especially on the number of patches and their size distribution [@Saura2002EffectsMinimum].
-<!-- On the other hand, landscape metrics such as the landscape division index, largest patch index and area-weighted mean shape indices show more predictable responses to changing the MMU. -->
-
-<!-- Since the applications of landscape metrics are manifold, the choices of grain and especially extent vary largely depending on the context of the study. -->
-<!-- The growing availability of global remote sensing products at homogenized resolution has resulted in a myriad of studies sharing the same resolution, e.g., based on Landsat scenes at 30 m.
-At the same time, the choice of spatial extent varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient.
--->
-While the choice of grain is often constrained by the available data, e.g., Landsat scenes at 30 m resolution, the selection of spatial extent varies largely depending on the phenomenon of interest, e.g., from large transects of forests in natural parks to small buffers along the urban-rural gradient.
-<!-- REF -->
-The spatial extent and sampling strategy is often subject to many arbitrary criteria that can largely influence results.
-This issue has long been recognized in physical geography as the modifiable areal unit problem [MAUP; @Openshaw1979MillionCorrelated; @Openshaw1984EcologicalFallacies].
-<!-- The MAUP has two interrelated components, namely the scale problem and the zoning problem -->
-<!-- A first approach to dealing with the MAUP is to identify individual entities that are ecologically meaninful and not modifiable [@Jelinski1996].
-In the landscape matrix model, such entities could be target patches of interest.
-However, it is not always possible to identify basic entities, especially since ecological processes of interest typically involve the integration of spatially distributed data.
--->
-Dealing with the MAUP in the context of landscape metrics include the identification of individual entities that are ecologically meaningful and not modifiable, the evaluation of "optimal" zoning systems that maximize interzonal variations and minimize intrazonal variation, and the analysis of the sensitivity of the zoning configuration [@Jelinski1996ModifiableAreal; @Duque2018SmaupStatistical; @Fletcher2023AddressingProblem].
-<!-- TODO: use a moving window measurement device to "postpone" the MAUP [@Riitters2019] -->
-
-<!-- MH: Can we move this to the sampling section maybe? -->
-<!-- TODO: section about approaches to sample landscapes to compute metrics and available software tools, e.g., zonal analysis with polygons (e.g., administrative limits), grids, transects or buffers along a landscape gradient, moving windows [@Bosch2019; @Hesselbarth2019; @Huais2024]?
--->
-
-<!-- While selecting the adequate scale should be based on the phenomenon of interest, researchers are often constrained by the availability of data. -->
-<!-- Therefore, a thorough understanding of these relationships is essential in order to compare spatial patterns measured at different scales. -->
-<!-- Several types of responses of landscape metrics to changing scale have been identified in computational experiments by changing both the spatial resolution (grain) and extent. -->
-The scale effects of landscape metrics can be assessed by computing scalograms, i.e., the response curves of metrics to changing scales. 
-Some metrics related to the number of patches, edges show consistent linear and power-law scaling relationships to changing grain, while other metrics exhibit less predictable responses [@Turner1989EffectsChanging; @Saura2001SensitivityLandscape; @Wu2002EmpiricalPatterns; @Wu2004EffectsChanging; @Saura2007ScalingFunctions; @Simova2012LandscapeIndices; @Arganaraz2014ScalingFunctions; @Frazier2023ScalingSpatial].
-<!-- Advancing the science of scaling in landscape ecology is key to "permit the prediction and inference of quantities measured at one scale to another" [@Frazier2023ScalingSpatial] -->
-<!-- also multi-scale is key to habitat assessment, see "Multi-scale habitat selection modeling: a review and outlook" -->
-<!-- Predictable scaling relationships can be exploited to estimate spatial patterns at higher resolutions -->
-Nonetheless, the effects of changing the sampling extent can be less predictable than those of changing grain size [@Wu2004EffectsChanging].
-For instance, in landscapes with small extent to grain ratio, boundary effects are likely to dominate and truncate patches resulting in unreliable metric computations.
-In order to address scale issues, studies should explicitly consider the concept of scope, namely the ratio of extent to resolution [@Schneider2001RiseConcept].
-This involves three fundamental steps, i) considering scope guidelines when designing experiments, ii) explicitly reporting grain and extent allowing quantitative scope comparisons with other studies, iii) and comparisons and replications based on scope [@Frazier2023ScalingSpatial].
-
-<!-- TODO: section about landscape metric rescaling [@Saura2007; @Arganaraz2014; @Frazier2014; @Frazier2016; @Frazier2021]? -->
-
-<!-- Overall, robust metrics seem more desirable because (i) they improve comparability between studies and (ii) the relationships between pattern and process occur at multiple scales -->
-<!-- which highlights the importance of considering multiple scales when attempting to measure landscape patterns. -->
-
-<!-- TODO: mention hierarchichal patch dynamics [@Wu1999]? -->
-
-<!-- (ii) choice of LULC scheme/thematic resolution and image processing/classification procedure: https://doi.org/10.1016/j.foreco.2009.12.016 https://doi.org/10.1016/j.isprsjprs.2007.03.004 https://doi.org/10.1007/s10021-005-0119-1 https://doi.org/10.1007/s10980-006-9035-9 https://doi.org/10.1007/s10980-022-01546-1 -->
-<!-- Scale and heterogeneity are inherently linked (Wu 2007), so it is not surprising that the results indicate downscaling accuracy varies with heterogeneity. -->
-<!-- TODO: sensitivity to thematic resolution [@Huang2006; @Bailey2007; @Buyantuyev2007; @Castilla2009]. -->
-
-## Interpretability
-<!-- (iv) interpretatability -->
-
-<!-- TODO
-- do landscape metrics measure fragmentation?
-- normalized metrics [@Wang2011]
-- [@Frazier2021]
-
-CLOSING THOUGHTS AND BRIDGING WITH THE NEXT SECTION 
-Responses:
-- incorporating aspects of ecological function into landscape metrics (e.g., core area, least-cost distances) and moving towards graph-based and circuit theory approaches [@Kupfer2012].
-- point-data analysis [@Gustafson1998] and gradient surface model [@Lausch2015] require fewer assumptions regarding LULC categorization schemes or spatial scale
-
-Note that all approaches (landscape metrics, graph and surface - TODO: what about point?) suffer from correlation and redundancy issues, scale dependency and target problem [@Frazier2017].
--->

### 2024-09-19 Manuscript.qmd
Minor cosmetic changes

@@
 title: "Computational methods in landscape ecology"
-
-keywords: spatial data, spatial patterns, open-source, scientific software
 
@@
 
+\noindent Keywords: (1) spatial data, (2) spatial patterns, (3) pattern-process link (4) open-source, (5) scientific software
+
 # Abstract {.unnumbered}
@@
 The link between spatial heterogeneity and ecological processes is the core concept in the research field of landscape ecology.
-Due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems, advanced computational methods are essential.
+Nowadays, advanced computational methods are essential to the field due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems.
 
@@
 
-This review provides an overview of recent developments in computational methods that have advanced landscape ecology research. 
-We focus on key topics such as spatial patterns, connectivity, simulations and modeling, landscape genetics, and spatial planning.
+This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
+We focus on key topics such as spatial patterns, connectivity, simulations and modelling, landscape genetics, and spatial planning.
 
@@
 
-We are thankful to the anonymous reviewers and the Editors who provided valuable comments to improve the manuscript.
+We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.
 

### 2024-09-19 Sections/01-introduction.qmd
Remove trailing whitespaces
Minor cosmetic changes

@@
 Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
-Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2002KeyIssues; @Wu2013KeyConcepts].
+Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].
 
@@
 Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
-Computational ecology is crucial for the field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
+Computational ecology is crucial for the research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
 Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
 
-Here, we aim to introduce the latest developments of computational methods in landscape ecology.
+Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them.
+In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
 However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
 For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
-Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them. 
-In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
+![Overview of landscape ecology topics for which computational developments are discussed.](Fig-1.png)

### 2024-09-19 Sections/02-spatial-patterns.qmd
Minor cosmetic changes

@@
 
-Traditionally and still the most prominent approaches to quantify spatial patterns revolve around raster data using categorical categories based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified regarding their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
+Traditionally and still the most prominent approaches to quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
+The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
+This includes shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 
-To address some limitations, such as correlation and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify main components of landscape patterns.
+To address some limitations, such as correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify core metrics that capture main components of landscape patterns.
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-This is in line with various reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency, which are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
+This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency.
+These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
 <!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
+<!-- MH: I moved sections around quite a bit unfortunately -->
 
@@
 These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
-This allows the assessment of content, context, and interface zones of LULC classes in a scale-dependent manner. 
+This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
 Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].
@@
 <!-- JN: ask T. Stepinski for review of this section later -->
-<!-- MH: Maybe we need to shorten this a bit -->
 
@@
 The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape omitting the spatial configuration.
-Shannon's entropy formula can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and inter-class distances [@Claramunt2012SpatiotemporalForm].
+Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
 However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
@@
 @Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
-This approach was later generalized for calculations based on raster's surface model and point patterns [@Cushman2021EntropyLandscape].
+This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
 Subsequently, @Gao2017HierarchybasedSolution proposed to use the Boltzmann entropy to quantify the complexity of a landscape surface by transforming the input raster into a series of landscape surfaces with different levels of detail (microstate) and calculating the Boltzmann entropy based on the total number of microstates that are able to generate the observed macrostate.
 Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
-However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial]
+However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].
 
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the complexity of the landscape.
-The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
-Thus, it can be useful in cases, when the dissimilarities between LULC types, and not only their categories, are relevant.
+The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the landscape complexity.
+The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
+Thus, it can be useful in cases, when the dissimilarities between LULC classes are relevant.
 Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].
 
-Entropy metrics are available in various software such as R's *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or Python's *PyLandStats*.
+Entropy metrics are available in various software such as the R packages *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or the Python package *PyLandStats*.
 
@@
 
-Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm] and were adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
-The gradient surface model can increase the resemblance of the data model and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
+Surface metrics are based on the gradient surface model using gridded raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
+These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
+The gradient surface model can increase the resemblance of the data and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
 Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
@@
 Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
-Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].
+Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
+Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].
 
@@
 However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
-For example, many LULC maps are provided as raster as they often related to underlying remote sensing products.
-Yet, nowadays, vector data is increasingly available, and most geospatial software supports it.
+For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products
+Yet, nowadays, vector data is increasingly available and has been widely used in other fields, such as geography or urban studies.
 
@@
 This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
-At the same, the vector data model has been widely used in other fields, such as geography or urban studies.
 For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
+Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
+More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].
 
-Earlier approaches of quantifying shape complexity  included four categories of compactness measures for vector data, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
-More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness]
-
-Vector-based metrics also have their limitations.
+However, also vector-based metrics have their limitations.
 The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
-Another technical issue is the need for the data to be topologically correct (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
+Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
 Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
@@
 The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics.
-However, all the general software frameworks mentioned earlier, which support vector data handling, can also be used to calculate basic metrics.
+However, all general software frameworks designed for spatial data handling which support vector data can theoretically be used to calculate basic vector metrics.
 
@@
 Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping.
-These operations are based on spatial signatures which are multi-numerical representations of landscape pattern and dissimilarity measures which are functions that quantify differences between the signatures.
+These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures).
 Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
@@
 
-R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
+R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
 These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.

### 2024-09-19 Sections/03-connectivity.qmd
Minor cosmetic changes

@@
 
-Landscape connectivity describes how landscape features facilitate or impede movement [@Taylor2006LandscapeConnectivity].
+Landscape connectivity describes how landscape features facilitate or impede movement [@Taylor2006LandscapeConnectivity] and can be divided into structural and functional connectivity.
 It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
-Connectivity can be divided into structural and functional connectivity.
 Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
-It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
+It is solely a landscape feature and independent of species characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
 Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioral, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].
@@
 Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
-Movement data from many individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
+Generally, movement data from individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
 While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
 Thus, in practice connectivity is primarily determined by indirect estimations.
-@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
+@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal characteristics of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
 
-## Estimating structural connectivity
+## Structural connectivity
 
@@
 Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
-However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, however, functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
+However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, yet functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
 
@@
 
-A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [@vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [@Landau2021OmniscapejlSoftware], *lconnect* [@Mestre2023LconnectPackage], *grainscape* [@Chubaty2020ToolboxGrainscape], or *Circuitscape* [@Anantharaman2019CircuitscapeJulia].
-
 Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
 By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
-This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
+This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping to identify reachable areas within a threshold.
 The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.
@@
 
-However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
+However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
 Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
@@
 Graph theory and connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
-For example, it can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity for protection and restoration [@Liu2020UsingLandscape].
+For example, they can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].
 
-Within the landscape delineating the system boundary, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
+Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
 The edges can be binary, representing whether a connection between two nodes exists or not, or weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
@@
 
-Recently, a new generation of graph-theoretical metrics, commonly reported in combination and addressing caveats of previous metrics [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
-The modeling framework, initially only applied to single species is increasingly being used to model multispecies landscapes [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach], which account for interspecific movement abilities and habitat preferences.
-In addition, modern multiple-layer graph approach offer to model not only spatial but spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
-Modeling the temporal dynamics helps to identify habitat patches impact connectivity over time, and how these impacts affect biodiversity patterns [@Uroy2021HowAssess].
+Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
+Furthermore, the modelling framework initially only applied to single species is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
+In addition, modern multiple-layer graph approaches offer to model not only spatial, but spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
+Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and how these impacts affect biodiversity patterns [@Uroy2021HowAssess].
 
-However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge due to various ways to construct a graph and different data sources.
-In many cases, data availability limits the representation of the modelled landscape, with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
-For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types resulting in different distribution of connectivity values.
+However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge due to various ways how to construct a graph and different data sources.
+In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
+For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.
 
-Currently, graph-theoretical landscape graphs and connectivity indices can be calculated using several open-source software, including *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], the R packages *lconnect* [@Mestre2023LconnectPackage] and *grainscape* [@Chubaty2020ToolboxGrainscape], or Julia’s *ConScape* library [@vanMoorter2023AcceleratingAdvances], among others.
+A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.

### 2024-09-19 Sections/05-simulations.qmd
Remove trailing whitespaces
Minor cosmetic changes

--- 2024-09-13 Sections/05-simulations.qmd
@@
 
-Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Pascual2005ComputationalEcology; @Petrovskii2012ComputationalEcology].
+Simulation models are a powerful tool to study complex, adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
 Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
-Due to the usual temporal and spatial scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
+Due to the usual spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
 In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
@@
 
-Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
+Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes], and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
 Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].
@@
 Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
-The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
+The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels], or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
 Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
-Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
-Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024LeastcostNetwork].
-In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
-In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python (NLMpy\; @Etherington2015NLMpyPython), R (NLMR; @Sciaini2018NLMRLandscapetools), Julia (NeutralLandscapes.jl; @Poisot2023NeutralLandscapesJulia), or stand-alone software (GradientLand; @Cambui2015GradientLandSoftware).
+Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise], or binary space partitioning [@Etherington2022BinarySpace].
+Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
+In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
+In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
 
 Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
-Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
-More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
-<!--JN: software??-->
+Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular], or deforestation [@Soares-Filho2002DinamicaStochastic].
+More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
 
@@
 
-Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
-In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling] or connectivity [e.g., @Day2020IndividualbasedModeling].
-In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building block [@Berger2024ReusableBuilding].
-Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
+Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behavior.
+Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
+In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [e.g., @Day2020IndividualbasedModeling].
+In order to facilitate future model developments, recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling], or reusable building blocks [@Berger2024ReusableBuilding].
+Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
 
@@
 
-The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
-The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for ecology [@deKoning2023DigitalTwins].
+The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins], and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
+The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
 Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
@@
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
-<!-- JN: "Digital Twins can" or "Digital Twins will" or "Digital Twins aims to"? -->
-Thus, Digital Twins can provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].
-<!--JN: software??-->
+Thus, Digital Twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-09-19 Sections/06-genetics.qmd
Minor cosmetic changes

@@
 Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
-Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], besides others.
+Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], among others.
 Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

### 2024-09-19 Sections/07-spatial-planning.qmd
Minor cosmetic changes

@@
 Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
-In the context of landscape-ecology, it provides a pattern-process link<!-- (see section @sec-spatialpattern) -->, and outcomes related to decision making by distinguishing between structure, function and scales [@Hersperger2021LandscapeEcological].
-As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology, for example through using simulation model outputs <!-- (see section @sec-simulation) --> as input features  <!-- (see also @daigle2020operationalizing) -->, there have been several recent advances in considering landscape patterns directly in spatial planning.
+In the context of landscape-ecology, it provides a pattern-process link, and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
+As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
+For example, there have been several recent advances in considering landscape patterns directly in spatial planning, or through using simulation model outputs as input features [e.g., @Daigle2020OperationalizingEcological].
 
@@
 Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
-In particular, recently the direct consideration of connectivity <!-- (see also section @sec-connectivity) --> has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].
+In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].
 
@@
 
-A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathway.
+A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
 Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
 Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
-Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatial-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
+Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
 However, this comes with the drawback of reduced interpretability.
 
-<!-- ## Open-source software in spatial conservation planning -->
-
 Multiple open-source software solutions have been developed to support integration of landscape ecology theory and indicators into spatial planning.
-Most software are based on linear or constrained programming approaches requiring a solver to create any outputs.
+Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
 Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].
@@
 Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
-The *CAPTAIN* Python software suite implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
+The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
 However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

### 2024-09-19 Sections/08-discussion.qmd
Minor cosmetic changes

@@
 
-Since emerging in the 1980s [@With2019EssentialsLandscape], the field of landscape ecology has constantly evolved and is now highly depending on computational methods.
-Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
+Since it emerged in the 1980s [@With2019EssentialsLandscape], the research field of landscape ecology has constantly evolved and is now highly depending on computational methods.
+Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
 Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
-In this context, we highlight open-source software as a cornerstone of Open Science, offering key advantages like shareability, reproducibility, and transparency, which provide the greatest benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
+In this context, we highlight open-source software as a cornerstone of "Open Science" which offers key advantages like shareability, reproducibility, and transparency, providing great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
 
-Alongside methodological advancements specific to landscape ecology, the research field will also benefit from and integrate technological innovations from related disciplines.
-This includes advancements in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
-These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR[@Crowley2020RemoteSensings].
-In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
-Due to its high flexibility and performance, artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
-Quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
-Furthermore, landscape ecology will benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].
+Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
+This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
+These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
+Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
+In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
+Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
+Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
+Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].
 
 Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
-Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution.
+Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
 However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
-Here we provided a review of the latest developments, nevertheless, a review of these methods can always be only a snapshot in time and hopefully new computational approaches will emerge in the future.
+Nevertheless, the here provided review of the latest computational methods in landscape ecology can hopefully only be a snapshot in time because new approaches will emerge in the future.

### 2024-09-19 Manuscript.qmd
Update authors

@@
     affil-id: 2,3
+  - name: Craig E. Simpkins
+    orcid: 0000-0003-3212-1379
+    email: simpkinscraig063@gmail.com
+    affil-id: 4
+  - name: Martin Jung
+    orcid: 0000-0002-7569-1390
+    email: jung@iiasa.ac.at
+    affil-id: 1
   - name: Gemma Gerber
@@
     email: adeflamingh@gmail.com
-    affil-id: 4
-  - name: Martin Jung
-    orcid: 0000-0002-7569-1390
-    email: jung@iiasa.ac.at
-    affil-id: 1
-  - name: Craig E. Simpkins
-    orcid: 0000-0003-3212-1379
-    email: simpkinscraig063@gmail.com
     affil-id: 5
@@
   - id: 4
+    name: Auckland Council, Environmental Services Department, Auckland, New Zealand
+  - id: 5
     name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
-  - id: 5
-    name: Auckland Council, Environmental Services Department, Auckland, New Zealand
   - id: 6

### 2024-09-19 Sections/04-sampling.qmd
Process to connectivity section
Update references
Move software to corresponding sections
Remove trailing whitespaces
feat: sampling section

@@
 
-# Sampling in landscape ecology
+# Sampling in landscape ecology {#sec-sampling}
+
+<!--MB: the following comments were for the introduction I first wrote when I was thinking of also including zonal analysis (e.g., pattern-based regionalization in this section.
+But at the end I think it is better to focus on sampling for pattern-process inference only.
+->
+<!-- MB: As introduced in previous sections, spatial patterns can be represented by spatial signatures such as vectors of computed landcape metrics.
+Nevertheless, computing spatial signatures in large landscapes can be problematic, not only because of the incurred computational costs, but also because involves reducing large complex spatial patterns into a few scalar values that can thus be hard to interpret.
+Instead, large spatial extents are usually divided into a set of zones, which can collectively cover the entire spatial extent or alternatively represent a sample of it.
+-->
+<!-- MB: we could make the keywords "synoptic" or "wall-to-wall" [@Ramezani2013] explicit when referring to computing a single metric for the (large) landscape  -->
+<!-- MB: The definition of landscape zones can be generally divided into two main categories, namely regular tesselations (e.g., grids) and irregular zones (e.g., administrative boundaries). -->
+<!-- MB: I think it be interesting to also talk about other sampling devices such as gradient/transect analysis -->
+
+<!-- When designing landscape ecology studies, the selection of landscapes is a critical step. -->
+<!-- Overall, the criteria considered at this stage depend largely on the objectives of the study. -->
+Establishing relationships between landscape patterns and ecological (e.g., species abundance) usually requires fieldwork and data collection.
+Given the considerable cost of collecting field observations accross multiple locations, computational sampling tools have been designed to optimize the study design [@Brennan2002; @Eigenbrod2011; @Pasher2013; @Zuckerberg2020].
+<!-- MB: feel free to uncomment the two sentences below -->
+<!-- Sampling methods can also been used as a cost-effective estimation of landscape metrics for large maps [@Hunsaker1994; @Ramezani2013].
+Nevertheless, given the increasing availability of computational power and global remotely sensed data, variables describing landscape pattern are in most cases easier to obtain than those measuring ecological processes.
+-->
+<!-- MB: regardless of whether the two previous sentences are included, we can resume here -->
+In order to improve the statistical significance of the relationships between field observations and its corresponding patch and landscape pattern, several key aspects need to be carefully considered when selecting landscape sites for data collection.
+
+## Characteristic scales of landscape patterns and ecological processes
+
+A central element of the study design is the definition of the size and spatial extent of each landscape.
+To this end, identifying scales at which the ecological processes of interest operate, i.e., characteristic scales, is a crucial first step towards establishing relationships between pattern and process [@Levin1992ProblemPattern; @Brennan2002; @Holland2004; @Wu2007ScaleScaling; @Cushman2008LandscapeMetrics; @Eigenbrod2011; @Jackson2015; @Zuckerberg2020].
+<!-- For each heterogeneity metric and for each extent (1 km × 1 km up to 7.5 km × 7.5 km), we calculated the coefficient of variation (CV) across all potential study landscapes.
+Based on a plot of CV versus landscape extent, we selected a threshold extent beyond which insufficient variability existed among potential sample landscapes.
+{@Pasher2013] -->
+<!-- The scale of effect would indicate the spatial scale at which the response variable exhibits the strongest relationship with the evaluated landscape attribute, supported by the strongest statistical model (e.g.
+the model with the highest R2 value or the lowest value for the Akaike Information Criterion) [@Huais2024] -->
+However, in practice, the appropriate landscape scale may not obvious or the data at such a scale may not be avaiable.
+Under these circumstances, analysis can be conducted at multiple scales (e.g., landscape extent sizes) so that it is possible to select the spatial scales of interest.
+For instance, scalograms can be used to determine thresholds or characteristic scales that maximize landscape heterogeneity [@Pasher2013] or show the strongest relationships with the response variables [@Holland2004; @Jackson2015; @Huais2024].
+<!-- MB: maybe we could also mention here the block size effect in regionalization/pattern-based spatial analysis, e.g., Table 2 of [@Nowosad2018]. -->
+
+Software to perform multi-scale analysis include the R package multilandr [@Huais2024], ZonalMetrics [@Adamczyk2017] or PyLandStats [@Bosch2019PyLandStatsOpensource] libraries in Python.
+TODO: add more?
+
+## Common statistical pitfalls in landscape selection
+
+The design of the landscape sampling strategy should address several statistical considerations [@Brennan2002; @Eigenbrod2011; @Pasher2013].
+First, with the aim of maximizing the variance within the independent variables, the selected sites should cover the maximum possible range of landscape heterogeneity.
+<!-- Additionally, statistical power is improved by ensuring independence between predictor variables. As reviewed in the spatial patterns section, this -->
+Secondly, statistical power is improved by ensuring independence between samples and predictor variables.
+This concerns both the selection of landscape extents and landscape metrics used to characterize them (see the section on spatial patterns above regarding landscape metrics redundancy and spatial signatures).
+Finally, since many environmental factors that influence variables of interest are spatially dependent, it is essential to evaluate spatial autocorrelation.
+
+The choice of landscape locations and extent sizes is critical because it can lead to overlapping landscapes with lower range of variation and lack of statistical independence [@Brennan2002; @Eigenbrod2011; @Pasher2013].
+In fact, many landscape ecology researchers consider the issue of overlapping landscapes to be a form of pseudoreplication [@Hurlbert1984], which can increase the chance of Type I errors [@Legendre1993].
+<!-- MB: should we add the definition of pseudoreplication when citing [@Hulrbert1984], e.g., something like "namely when statistical inference is employed to test for relationships when samples are not statistically independent"?
+-->
+<!-- more broad (less landscape ecology centric references about pseudoreplication: [@Oksanen2001; @Davies2015]) -->
+Nonetheless, the lack of independence between landscape samples is likely more related to spatial autocorrelation and thus not prevented by using non-overlapping extents [@Schooley2006; @Zuckerberg2012; @Zuckerberg2020].
+Additionally, preventing overlapping landscapes can be logistically challenging and complicate experimental design, leading to reduced sample sizes, failure to capture sufficient landscape heterogeneity, omision of characteristic scales (e.g., by excluding large buffers around sites) and even increase spatial autocorrelation in the model residuals [@Zuckerberg2012].
+
+<!-- While systematic sampling designs can limit spatial autocorrelation [@Pasher2013; @Zuckerberg2020] -->
+Instead, spatial autocorrelation can be evaluated at the study design stage by computing a measure of similarity, e.g., Moran's I [@Moran1950], on the landscape metrics that correspond to each sampling approach, e.g., combination of sample locations and extent sizes [@Pasher2013].
+However, selecting landscapes that are not spatially autocorrelated is not always feasible based on logistics or environmental constraints of the study region.
+<!-- In presence of spatial autocorrelation, its influence in model residuals can be diagnosed using correlograms, i.e., a measure of similarity between two samples plotted as a function of the distance between them -->
+The presence of spatial autocorrelation in the model residuals can be diagnosed using correlograms, i.e., a measure of similarity between two samples plotted as a function of the distance between them [@Dormann2007].
+As suggested by Zuckerberg et al.
+[@Zuckerberg2020], when residual spatial autocorrelation is detected, researchers may either consider further data collection and check whether that reduces spatial autocorrelation, or use modelling methods to accomodate spaital dependencies such as mixed models [@Dormann2007; @Beale2010; @Arnqvist2020] or apply smoothing kernels that compute distance-weighted averages of landscape patterns surrounding the sampling sites [@Heaton2011; @Chandler2016].
+
+TODO: I do not know if we want to measure software that is less landscape ecology centric/more general spatial analysis.
+If so, please complete the packages (based on [@Beale2010] I found glmm for R as well as SAS and Stata, but you probably know more about it, especially for R).
+Software to conduct diagnostics of spatial autocorrelation include PySAL [@Rey2007].
+Mixed effects models are available in the statsmodels [@Seabold2010] and PySAL [@Rey2007] libraries in Python.
+Spatial weights based on kernel functions can be computed using the PySAL [@Rey2007] library in Python.

### 2024-09-20 Manuscript.qmd
Update sampling section

@@
 {{< include Sections/03-connectivity.qmd >}}
-<!-- {{< include Sections/04-sampling.qmd >}} -->
+{{< include Sections/04-sampling.qmd >}}
 {{< include Sections/05-simulations.qmd >}}

### 2024-09-20 Sections/04-sampling.qmd
Update sampling section

--- 2024-07-12 Sections/04-sampling.qmd
@@
 
-# Sampling in landscape ecology
+# Sampling in landscape ecology {#sec-sampling}
+
+Establishing a pattern-process link usually requires fieldwork and data collection.
+Given the cost of collecting field observations across landscapes, computational sampling tools have been designed to optimize the study designs [@Pasher2013OptimizingLandscape; @Zuckerberg2020ReviewOverlapping].
+To improve the statistical significance of the relationships between landscape features and field observations, several aspects need to be considered when selecting sampling sites for data collection.
+First, selected sites should cover the maximum possible range of landscape heterogeneity to maximize the variance of response variables.
+Second, statistical power can be improved by ensuring independence between sites and consequently response variables.
+Finally, since many landscape features that influence response variables are spatially dependent, it is essential to evaluate spatial autocorrelation.
+
+Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
+However, in practice, the appropriate scales may not be obvious, or data at such scales may not be available.
+Nevertheless, analyses at multiple spatial scales can help to identify the scales of interest.
+For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape], or show the strongest relationships with response variables [@Huais2024MultilandrPackage].
+
+Directly related to the location and extent of sampling sites is the issue of overlapping landscapes which can result in a lower range of landscape heterogeneity, lack of statistical independence, or pseudo-replications [@Hurlbert1984PseudoreplicationDesign; @Zuckerberg2020ReviewOverlapping].
+Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
+Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures (e.g., Moran's I) between sampling approaches or spatial scales [@Pasher2013OptimizingLandscape].
+Furthermore spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
+If spatial autocorrelation is detected, researchers may either consider further data collection, or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
+
+Software to perform multi-scale analysis include the R package *multilandr*, or the Python libraries *ZonalMetrics* [@Adamczyk2017ZonalMetricsPython], and *PyLandStats*.

### 2024-09-20 Manuscript.qmd
adds a few comments/changes/etc

@@
 {{< include Sections/03-connectivity.qmd >}}
-{{< include Sections/04-sampling.qmd >}}
+<!-- JN: Max, should this be uncommented? -->
+<!-- {{< include Sections/04-sampling.qmd >}} -->
 {{< include Sections/05-simulations.qmd >}}
@@
 
+<!-- JN: Are we proactively thankful? I would suggest to uncomment this sentence and reinstate it when the final version of the paper is accepted -->
 We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.

### 2024-09-20 Sections/01-introduction.qmd
Update authors
adds a few comments/changes/etc

@@
 Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
-Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].
+Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, including human activities [@Wu2013KeyConcepts].
 
@@
 Computational ecology is crucial for the research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
+<!--JN: why do you use "but see" here?
+is the Wiersma2022 paper contradicting Petrovskii2012?-->
 Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
@@
 For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
-![Overview of landscape ecology topics for which computational developments are discussed.](Fig-1.png)
+
+![Overview of landscape ecology topics for which computational developments are discussed.](Figures/Fig-1.png){width=80%}

### 2024-09-20 Sections/02-spatial-patterns.qmd
feat: sampling section
adds a few comments/changes/etc

@@
 Traditionally and still the most prominent approaches to quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
+The strength of such landscape metrics is that they are easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
 This includes shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
@@
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency.
-These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
-<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
-<!-- MH: I moved sections around quite a bit unfortunately -->
+This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundaments of landscape patterns, namely amount and adjacency.
+These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.
 
@@
 <!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->
+<!-- JN: Agree. There are many R tools related to dimensionality reduction, that could be easily connected to landscape metrics, but I would not mention them here. -->
 
 ## Landscape Mosaic
+
+<!-- JN: why the capitalization of Landscape Mosaic? -->
 
@@
 This allows the assessment of content, context, and interface zones of LULC data in a scale-dependent manner.
+<!-- JN: you mentioned that the method has been improved, but there is no reference to the original paper. -->
+<!-- JN: also -- it this idea important enough to have its own section? I assume that there are many papers that propose new ways to classify LULC data. -->
 Recently, the method has been improved by expanding the classification to 103 classes and incorporating heatmap visualizations and summaries [@Vogt2024RevisitingLandscape].
@@
 ## Entropy
-
-<!-- JN: ask T. Stepinski for review of this section later -->
 
@@
 
-<!-- JN: ask T. Matuszek for review of this section later -->
-
 Many landscape ecology studies are based on the raster data model.
@@
 
-R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
+R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures] or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
 These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
-Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
+Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.

### 2024-09-20 Sections/06-genetics.qmd
adds a few comments/changes/etc

@@
 Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], among others.
-Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].
+Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., @Brown2017SDMtoolbox20].

### 2024-09-20 Sections/08-discussion.qmd
adds a few comments/changes/etc

@@
 Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
+<!-- JN: what are "large satellites"?  -->
 This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
@@
 Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
+<!-- JN: "is entering" or "has entered"? -->
 In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
@@
 Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
-Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].
+Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].
 
@@
 Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
-However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
-Nevertheless, the here provided review of the latest computational methods in landscape ecology can hopefully only be a snapshot in time because new approaches will emerge in the future.
+Up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
+Nevertheless, the here provided review of the latest computational methods in landscape ecology can only be a snapshot in time because new approaches will emerge in the future.

### 2024-09-20 Sections/04-sampling.qmd
cleans samping

--- 2024-09-19 Sections/04-sampling.qmd
@@
 Software to perform multi-scale analysis include the R package multilandr [@Huais2024], ZonalMetrics [@Adamczyk2017] or PyLandStats [@Bosch2019PyLandStatsOpensource] libraries in Python. 
-TODO: add more?
+<!-- TODO: add more? -->
 
@@
 
-TODO: I do not know if we want to measure software that is less landscape ecology centric/more general spatial analysis.
+<!-- TODO: I do not know if we want to measure software that is less landscape ecology centric/more general spatial analysis.
 If so, please complete the packages (based on [@Beale2010] I found glmm for R as well as SAS and Stata, but you probably know more about it, especially for R).
-Software to conduct diagnostics of spatial autocorrelation include PySAL [@Rey2007].
-Mixed effects models are available in the statsmodels [@Seabold2010] and PySAL [@Rey2007] libraries in Python.
+-->
+Software to conduct diagnostics of spatial autocorrelation include PySAL [@Rey2007]. Mixed effects models are available in the statsmodels [@Seabold2010] and PySAL [@Rey2007] libraries in Python. 
 Spatial weights based on kernel functions can be computed using the PySAL [@Rey2007] library in Python.

### 2024-09-20 Manuscript.qmd
Adding comments AdF

--- 2024-09-20 Manuscript.qmd
@@
     affil-id: 1
+  - name: Alida de Flamingh
+    orcid: 0000-0003-1223-6654
+    email: adeflamingh@gmail.com
+    affil-id: 5
   - name: Gemma Gerber
@@
     affil-id: 1
-  - name: Alida de Flamingh
-    orcid: 0000-0003-1223-6654
-    email: adeflamingh@gmail.com
-    affil-id: 5
   - name: Martí Bosch
@@
   - id: 4
-    name: Auckland Council, Environmental Services Department, Auckland, New Zealand
+    name: Auckland Council, Environmental Services, Auckland, New Zealand
   - id: 5
@@
 
-## References {.unnumbered}
+# References {.unnumbered}
 

### 2024-09-20 Sections/01-introduction.qmd
Adding comments AdF

--- 2024-09-19 Sections/01-introduction.qmd
@@
 
-Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with empathize on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
-Linking spatial heterogeneity and ecological processes, including potential feedbacks, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
-Typical research topics include, besides others, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].
+Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
+Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
+Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].
 
@@
 Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
-Computational ecology is crucial for the research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
-Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], or the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
+Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
+Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
 
@@
 However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
-For a more comprehensive and general introductions to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
-![Overview of landscape ecology topics for which computational developments are discussed.](Fig-1.png)
+For a more comprehensive and general introduction to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
+
+![Overview of landscape ecology topics for which computational developments are discussed.](Figures/Fig-1.png){width=85%}

### 2024-09-20 Sections/02-spatial-patterns.qmd
Adding comments AdF

--- 2024-09-19 Sections/02-spatial-patterns.qmd
@@
 
-Traditionally and still the most prominent approaches to quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of landscape metrics is that they are comparable easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, also limitations have been identified. 
-This includes shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, or correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
+Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
+The strength of landscape metrics is that they are uncomplicated to apply and communicate [@Lausch2015UnderstandingQuantifying], however, limitations have also been identified. 
+These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 
-To address some limitations, such as correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees were used to identify core metrics that capture main components of landscape patterns.
+To address some limitations, for example correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-This is in line with several reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] suggesting that there are two fundamental components of landscape patterns, namely amount and adjacency.
-These are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
-<!-- MB: I wonder if we could already mention [@Nowosad2019InformationTheory] here since it presents the same two-component view of landscapes and links with the following section quite naturally -->
-<!-- MH: I moved sections around quite a bit unfortunately -->
+These components of landscape patterns are in line with revent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency, and that these components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
 
@@
 They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
-However, studies show that insights do not only depend on the formulation of the selected entropy, but also on the underlining data model, for example the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
+However, studies show that insights gained from entropy measures depend on the formulation of the selected entropy, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
 
@@
 
-Based on notions from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
+Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
 In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
@@
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies <!--[REF]-->, which are both generalizations of the Shannon entropy, are applied to quantify the landscape complexity.
+The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies, which are both generalizations of the Shannon entropy, have been applied to quantify the landscape complexity.
 The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
-Thus, it can be useful in cases, when the dissimilarities between LULC classes are relevant.
+Thus, it can be useful in cases where the dissimilarities between LULC classes are relevant.
 Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].
@@
 These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
-The gradient surface model can increase the resemblance of the data and the natural world because it allows to include more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
+The gradient surface model can increase the resemblance of the data to the natural world because it allows for the inclusion of more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
 Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
-Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows to reveal different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
+Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
 
-Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecological meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
+Yet, software to calculate these metrics are still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
 Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
@@
 Many landscape ecology studies are based on the raster data model.
-However, the decision on the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
+However, the choice of the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
 For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products
-Yet, nowadays, vector data is increasingly available and has been widely used in other fields, such as geography or urban studies.
+However, nowadays, vector data has become increased in availability and has been widely used in other fields, such as geography or urban studies.
 
-Hundreds of metrics have already been developed for raster data, so the most straightforward approach is to reimplement the same metrics for vector data.
-This is possible for most of the metrics, with only several exceptions of metrics that are specifically related to the raster data model [@Yao2022VecLIFramework].
-For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
+Hundreds of metrics have already been developed for raster data, and the most straightforward approach may therefore be to reimplement these same metrics for vector data.
+Although this is possible for most of the metrics, there are a few exceptions where metrics are specifically related to the raster data model [@Yao2022VecLIFramework].
+For example, in urban planning, vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
 Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
@@
 
-However, also vector-based metrics have their limitations.
-The most important is computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
-Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often problematic especially for data from different sources.
-Last, the pattern-process link is still less explored for vector-based metrics compared to raster-based metrics.
+However, also vector-based metrics also have limitations.
+The most important limitations relate to computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
+Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often especially problematic for data from different sources.
+Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.
 
@@
 
-Spatial patterns can be analyzed through a range of computational operations, such as comparing, searching, or grouping.
+Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
 These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures).
@@
 Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
-This allows to find areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
+This allows for the identification of areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
 Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].

### 2024-09-20 Sections/03-connectivity.qmd
Adding comments AdF

@@
 
-Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, quantifying connectivity has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
+Although multiple approaches that measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], the technical and conceptual quantification of connectivity has proven challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
 Generally, movement data from individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
@@
 These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
-Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
+Additionally, there has been a recent increase in focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
 However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, yet functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
@@
 Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational] and quantifying cost values for each cell using movement data.
-This can included expert opinion [widely used due to low effort, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
+This can include expert opinion [widely used due to low effort, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
 
@@
 Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
-It can describes isolation by measuring multiple low- or high resistance pathways and identify areas of high movement probability highlighting important corridors and bottlenecks in the landscape [@Cushman2013BiologicalCorridors].
+It can describe isolation by measuring multiple low- or high resistance pathways and identify areas of high movement probability highlighting important corridors and bottlenecks in the landscape [@Cushman2013BiologicalCorridors].
 
 However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
-Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
+Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
 
@@
 
-Following the formative work of @Urban2001LandscapeConnectivity, graph theory has developed into a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
-Graph theory and connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
-For example, they can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].
+Following the formative work of @Urban2001LandscapeConnectivity, graph theory has become a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
+Graph theory and related connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
+For example, graph theory based metrics can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].
 
 Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
-The edges can be binary, representing whether a connection between two nodes exists or not, or weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
+The edges can be binary, i.e., representing whether a connection between two nodes either exists or does not exists, or edges can be weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
 Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
@@
 
-Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
-Furthermore, the modelling framework initially only applied to single species is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
-In addition, modern multiple-layer graph approaches offer to model not only spatial, but spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
-Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and how these impacts affect biodiversity patterns [@Uroy2021HowAssess].
+Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
+Furthermore, the modelling framework was initially only applied to single species but is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
+In addition, modern multiple-layer graph approaches enable the modelling of not only spatial, but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
+Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].
 
-However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge due to various ways how to construct a graph and different data sources.
+However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge from different methods used to construct a graph or from different data sources.
 In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
@@
 
-A large number of software applications have been developed, with these approaches as a base, to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.
+A large number of graph theory based software applications have been developed to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.

### 2024-09-20 Sections/04-sampling.qmd
Adding comments AdF

--- 2024-09-20 Sections/04-sampling.qmd
@@
 Furthermore spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
-If spatial autocorrelation is detected, researchers may either consider further data collection, or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
+If spatial autocorrelation is detected, researchers may either consider further data collection, or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
 

### 2024-09-20 Sections/05-simulations.qmd
Adding comments AdF

@@
 Simulation models are a powerful tool to study complex, adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
-Thus, they can be seen as experimental systems that allow, unlike to the natural world, all imaginable manipulations in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
-Due to the usual spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
-In general, simulations models can be classified using two major divisions, namely as *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
-Here, we are going to focus mainly on exploratory models, but include both pattern- and process-based models.
+Thus, simulations can be seen as experimental systems that allow all imaginable manipulations in order to advance theoretical developments or test hypotheses, which would be impossible in natural systems [@Peck2004SimulationExperiment].
+Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
+In general, simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
+Here, we focus mainly on exploratory models, but include both pattern- and process-based models.
 
@@
 
-Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes], and are generally simpler than ecological simulations models [@Hess2020PioLaGPiosphere].
+Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow the users to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes], and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
 Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].
@@
 Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
-In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
+In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
 In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
@@
 Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular], or deforestation [@Soares-Filho2002DinamicaStochastic].
-More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
+More recent landscape generators have allowed for the simulation of patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
 
@@
 In order to facilitate future model developments, recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling], or reusable building blocks [@Berger2024ReusableBuilding].
-Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
+Furthermore, hybrid approaches that couple or even integrate different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
 
@@
 
-The Digital Twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins], and there is an increasing interest and use of them in both the industry and academia [@Jones2020CharacterisingDigital].
-The need for timely and evidence-based decision making in combination with increasing data availability makes them a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
-Digital twins are increasingly applied in agricultural landscapes research, including topics such as crop and land cultivation, livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
+The digital twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins], and there is an increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
+The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
+Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
 Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
-The use of general data processing, statistical and mechanistic models, or artificial intelligence allows them to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
-Thus, Digital Twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or identify uncertainties and information gaps [@deKoning2023DigitalTwins].
+The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
+Thus, digital twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-09-20 Sections/06-genetics.qmd
Adding comments AdF

--- 2024-09-19 Sections/06-genetics.qmd
@@
 
-Genetic data can be applied to complement ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
 Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
-Landscape genetic approaches have been used traditionally to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
-Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
+Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
+Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
 
@@
 Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
-Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], or *landgenreport* [@Gruber2015LandgenreportNew], among others.
-Furthermore, series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].
+Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
+Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].

### 2024-09-20 Sections/07-spatial-planning.qmd
Adding comments AdF

@@
 Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
-Furthermore, aggregation can be important because of ecological or practical reasons, for example by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
+Furthermore, aggregation can be important because of ecological or practical reasons, for example, by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
 Novel computational approaches make use of graph theory to identify valuable landscape areas in terms of compact core habitat and reducing boundary exposure [@Weerasena2023SequentialApproach].
-Similar, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
+Similarly, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
 In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].
 
-Particular noteworthy are two advances that incorporate landscape metrics in spatial planning.
+Particularly noteworthy are two advances that incorporate landscape metrics in spatial planning.
 One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
-These spatial networks approaches have the potential to be provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
+These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
 For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.
@@
 Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
-Additionally, it can be applied to identify best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
+Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
 Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
@@
 
-The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows to flexible create planning problems in multiple configurations.
+The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows flexibility to create planning problems in multiple configurations.
 Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
@@
 The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
-However, many spatial planning software still requires substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.
+However, many spatial planning software still require substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.

### 2024-09-20 Sections/08-discussion.qmd
Adding comments AdF

--- 2024-09-19 Sections/08-discussion.qmd
@@
 
-Since it emerged in the 1980s [@With2019EssentialsLandscape], the research field of landscape ecology has constantly evolved and is now highly depending on computational methods.
+Since it emerged in the 1980s [@With2019EssentialsLandscape], the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
 Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
@@
 Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
-Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].
+Last, landscape ecology will, as many other research fields, benefit from a general paradigm shift related to data sharing, management, and documentation [@Tenopir2020DataSharing].
 
-Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
+Climate change as well as the biodiversity crisis are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
 Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
 However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
-Nevertheless, the here provided review of the latest computational methods in landscape ecology can hopefully only be a snapshot in time because new approaches will emerge in the future.
+Nevertheless, the review provided here that showcases current computational methods in landscape ecology will hopefully only be a snapshot in time because of the development and emergence of future analytical approaches.

### 2024-09-20 Sections/01-introduction.qmd
Comments JN

@@
 Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
-Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change including human activities [@Wu2013KeyConcepts].
+Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, including human activities [@Wu2013KeyConcepts].
 
@@
 Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
-Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape].
+Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of landscape ecology experiments].
 Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.

### 2024-09-20 Sections/02-spatial-patterns.qmd
Comments JN

@@
 Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of landscape metrics is that they are uncomplicated to apply and communicate [@Lausch2015UnderstandingQuantifying], however, limitations have also been identified. 
+The strength of such landscape metrics is that they are easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, limitations have also been identified. 
 These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
@@
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-These components of landscape patterns are in line with revent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency, and that these components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory.
+These components of landscape patterns are in line with recent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency.
+These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.
 
 Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
-<!-- MB: Are there any libraries to perform factor analysis in R that are kind of specific to landscape ecology?
-I have had a draft on how to perform factor analysis with pylandstats https://github.com/martibosch/pylandstats-notebooks/blob/main/notebooks/05-cluster-analysis.ipynb, hopefully I will include it in the next release before -->
-<!-- MH: Would not mention PCA or similar here. Thats a general statistical issue and outside scope imho -->
 
-## Landscape Mosaic
-
-The Landscape Mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes.
+The landscape mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes [@Riitters2009IndicatorForest].
 The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
@@
 
-R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], *terra* [@Hijmans2024TerraSpatial], or *motif* [@Nowosad2021MotifOpensource] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance] or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
+R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance], or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
 These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
-Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for landscape ecology.
+Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.

### 2024-09-20 Sections/06-genetics.qmd
Comments JN

@@
 Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
-Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., Brown2017SDMtoolbox20].
+Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., @Brown2017SDMtoolbox20].

### 2024-09-20 Sections/08-discussion.qmd
Comments JN

@@
 Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
-This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
+This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
 These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
 Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
-In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
+In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
 Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
@@
 Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].
-However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
-Nevertheless, the review provided here that showcases current computational methods in landscape ecology will hopefully only be a snapshot in time because of the development and emergence of future analytical approaches.
+Up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
+Nevertheless, the review provided here that showcases current computational methods in landscape ecology will only be a snapshot in time because of the development and emergence of future analytical approaches.

### 2024-09-20 Manuscript.qmd
Remove numbering

@@
         \renewcommand\Affilfont{\small}
-
 papersize: a4paper
@@
 linestretch: 1.5
-number-sections: true
+number-sections: false
 
@@
 
-# Abstract {.unnumbered}
+# Abstract
 
-## Purpose of Review {.unnumbered}
+## Purpose of Review
 
@@
 
-## Recent findings {.unnumbered}
+## Recent findings
 
@@
 
-## Summary {.unnumbered}
+## Summary
 
@@
 
-# Acknowledgments {.unnumbered}
+# Acknowledgments
 
@@
 
-## Funding {.unnumbered}
+## Funding
 
@@
 
-## Author contributions {.unnumbered}
+## Author contributions
 
@@
 
-## Conflict of Interest {.unnumbered}
+## Conflict of Interest
 
@@
 
-## Human and Animal Rights and Informed Consent {.unnumbered}
+## Human and Animal Rights and Informed Consent
 
@@
 
-# References {.unnumbered}
+# References
 

### 2024-09-20 Manuscript.qmd
Update layout

@@
   - name: Maximilian H.K. Hesselbarth
-    orcid: 0000-0003-1125-9918
-    email: mhk.hesselbarth@gmail.com
     affil-id: 1
   - name: Jakub Nowosad
-    orcid: 0000-0002-1057-3721
-    email: nowosad.jakub@gmail.com
     affil-id: 2,3
+  - name: Alida de Flamingh
+    affil-id: 4
   - name: Craig E. Simpkins
-    orcid: 0000-0003-3212-1379
-    email: simpkinscraig063@gmail.com
-    affil-id: 4
+    affil-id: 5
   - name: Martin Jung
-    orcid: 0000-0002-7569-1390
-    email: jung@iiasa.ac.at
     affil-id: 1
-  - name: Alida de Flamingh
-    orcid: 0000-0003-1223-6654
-    email: adeflamingh@gmail.com
-    affil-id: 5
   - name: Gemma Gerber
-    orcid: 0000-0002-8722-3113
-    email: gerber@iiasa.ac.at
     affil-id: 1
   - name: Martí Bosch
-    orcid: 0000-0001-8735-9144
-    email: marti.bosch@epfl.ch
     affil-id: 6
@@
   - id: 4
+    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA 
+  - id: 5
     name: Auckland Council, Environmental Services, Auckland, New Zealand
-  - id: 5
-    name: Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA
   - id: 6
@@
       text: |
+        \usepackage{academicons}
+        \usepackage{fontawesome5}
         \usepackage{lineno}\linenumbers
@@
         \usepackage{setspace}
-        \renewcommand*{\Authsep}{, }
-        \renewcommand*{\Authand}{, }
-        \renewcommand*{\Authands}{, }
-        \renewcommand\Affilfont{\small}
+        \renewcommand\Affilfont{\footnotesize}
+        \newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\textcolor[HTML]{A6CE39}{\aiOrcid#1}}}
+        \newcommand{\mail}[1]{\href{mailto:#1}{\textcolor[HTML]{052E81}{\faEnvelope#1}}}
+
 papersize: a4paper
-fontsize: 10pt
+mainfont: Garamond
+fontsize: 11pt
 geometry: margin=25mm
@@
 
-\noindent Keywords: (1) spatial data, (2) spatial patterns, (3) pattern-process link (4) open-source, (5) scientific software
+\noindent \footnotesize M.H.K. Hesselbarth \orcid{0000-0003-1125-9918} \mail{mhk.hesselbarth@gmail.com}
+
+\noindent \footnotesize J. Nowosad \orcid{0000-0002-1057-3721}
+
+\noindent \footnotesize A. de Flamingh \orcid{0000-0003-1223-6654}
+
+\noindent \footnotesize C.E. Simpkins \orcid{0000-0003-3212-1379}
+
+\noindent \footnotesize M. Jung \orcid{0000-0002-7569-1390}
+
+\noindent \footnotesize G. Gerber \orcid{0000-0002-8722-3113}
+
+\noindent \footnotesize M. Bosch \orcid{0000-0001-8735-9144}
+
+\noindent \small \textit{Keywords: (1) spatial data, (2) spatial patterns, (3) pattern-process link (4) open-source, (5) scientific software}
 
@@
 The review highlights key innovations, challenges, and potential future directions in the field, emphasizing the role of computational methods in addressing complex ecological questions.
-
-{{< pagebreak >}}
 
@@
 
-{{< pagebreak >}}
-
 # References

### 2024-09-20 Sections/01-introduction.qmd
Update Figure

@@
 
-Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them.
+Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them (Fig. 1).
 In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
@@
 
-![Overview of landscape ecology topics for which computational developments are discussed.](Figures/Fig-1.png){width=85%}
+![Overview of landscape ecology topics for which computational methods are discussed.](Figures/Fig-1.png){width=100%}

### 2024-09-20 R-Scripts/dummy.R
Update README and Outline

@@ -1,4 +1,0 @@
-# This is just a dummy script
-
-x <- 5
-x * 2

### 2024-09-20 Manuscript.qmd
Fix format issues

@@
 papersize: a4paper
-mainfont: Garamond
 fontsize: 11pt
@@
 bibliography: CLER.bib
-csl: springer-basic-author-date.csl
+csl: springer-basic-brackets-no-et-al.csl
 ---
@@
 \noindent \small \textit{Keywords: (1) spatial data, (2) spatial patterns, (3) pattern-process link (4) open-source, (5) scientific software}
+
+\noindent \textbf{Acknowledgments}
+
+We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.
+
+\noindent \textbf{Funding}
+
+No external funding was received.
+
+\noindent \textbf{Author contributions}
+
+All authors contributed particular sections to the manuscript coordinated by MHKH.
+
+Correspondence to [Maximilian H.K. Hesselbarth](mailto:mhk.hesselbarth@gmail.com)
+
+\noindent \textbf{Conflict of Interest}
+
+The authors declare no competing interests.
+
+\noindent \textbf{Human and Animal Rights and Informed Consent}
+
+This article does not contain any studies with human or animal subjects performed by any of the authors.
 
@@
 {{< include Sections/07-spatial-planning.qmd >}}
-{{< include Sections/08-discussion.qmd >}}
+{{< include Sections/08-conclusion.qmd >}}
 
-# Acknowledgments
+# Key references
 
-We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.
+**Frazier (2019) Emerging trajectories for spatial pattern analysis in landscape ecology.
+Landscape Ecology 34:2073–2082.** The paper describes how interdisciplinary perspectives have contributed to spatial pattern analysis and introduces new innovative developments.
 
-## Funding
+**Unnithan Kumar et al.
+(2022) Moving beyond landscape resistance: Considerations for the future of connectivity modelling and conservation science.
+Landscape Ecology 37:2465–2480.** The paper provides a historical overview of connectivity modelling with a focus on resistance surfaces and discusses current missing aspects and how to address them.
 
-No external funding was received.
+**Zuckerberg et al.
+(2020) A review of overlapping landscapes: Pseudoreplication or a red herring in landscape ecology?
+Current Landscape Ecology Reports 5:140–148.** The paper discusses theoretical and practical implications over overlapping landscapes in the context of landscape buffers of varying sizes surrounding sampling sites.
 
-## Author contributions
+**Synes et al.
+(2016) Emerging opportunities for landscape ecological modelling.
+Current Landscape Ecology Reports 1:146–167.** The paper identifies existing gaps in landscape ecological modelling while highlighting potential emerging opportunities.
 
-All authors contributed particular sections to the manuscript coordinated by MHKH.
-Correspondence to [Maximilian H.K. Hesselbarth](mailto:mhk.hesselbarth@gmail.com)
+**Manel and Holderegger (2013) Ten years of landscape genetics.
+Trends in Ecology & Evolution 28:614–621.** Ten years after their first paper introducing landscape genetics, this paper reviews the state-of-the-art and describes main topics that have contributed to the progress of landscape genetics.
 
-## Conflict of Interest
-
-The authors declare no competing interests.
-
-## Human and Animal Rights and Informed Consent
-
-This article does not contain any studies with human or animal subjects performed by any of the authors.
+**Jung et al.
+(2024) An assessment of the state of conservation planning in europe.
+Philosophical Transactions of the Royal Society B 379:20230015.** The paper assesses where and how spatial conservation planning have been applied across Europe.
 

### 2024-09-20 Sections/06-genetics.qmd
Fix format issues

@@
 
-Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.<!-- MH: Ref missing -->.
+Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
 This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].

### 2024-10-11 Manuscript.qmd
Correspondance
Starting to work on Review

@@
   pdf:
-    keep-tex: true
     template-partials:
@@
 
-We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.
+\noindent We are thankful to the anonymous reviewers and the editors who provided valuable comments to improve the manuscript.
 
@@
 
-No external funding was received.
+\noindent No external funding was received.
 
@@
 
-All authors contributed particular sections to the manuscript coordinated by MHKH.
-
-Correspondence to [Maximilian H.K. Hesselbarth](mailto:mhk.hesselbarth@gmail.com)
+\noindent All authors contributed particular sections to the manuscript coordinated by MHKH.
+Correspondence to [mhk.hesselbarth@gmail.com](mailto:mhk.hesselbarth@gmail.com)
 
@@
 
-The authors declare no competing interests.
+\noindent The authors declare no competing interests.
 
@@
 
-This article does not contain any studies with human or animal subjects performed by any of the authors.
+\noindent This article does not contain any studies with human or animal subjects performed by any of the authors.
 
@@
 
+<!-- R1: Figure 1.
+The links between Spatial sampling, Spatial planning, and Genetics, towards Landscape mean?
+The figure would be more meaningful with an explanation of these links, either in the caption or directly on the figure.
+Also, for other topics (e.g.
+connectivity), a keyword (or short sentence) summarizing the contribution of the topic to Landscape Ecology would be great.
+-->
+
 {{< include Sections/01-introduction.qmd >}}
@@
 
-**Frazier (2019) Emerging trajectories for spatial pattern analysis in landscape ecology.
-Landscape Ecology 34:2073–2082.** The paper describes how interdisciplinary perspectives have contributed to spatial pattern analysis and introduces new innovative developments.
+\noindent _Frazier (2019) Emerging trajectories for spatial pattern analysis in landscape ecology.
+Landscape Ecology 34:2073–2082._ The paper describes how interdisciplinary perspectives have contributed to spatial pattern analysis and introduces new innovative developments.
 
-**Unnithan Kumar et al.
+\noindent _Unnithan Kumar et al.
 (2022) Moving beyond landscape resistance: Considerations for the future of connectivity modelling and conservation science.
-Landscape Ecology 37:2465–2480.** The paper provides a historical overview of connectivity modelling with a focus on resistance surfaces and discusses current missing aspects and how to address them.
+Landscape Ecology 37:2465–2480._ The paper provides a historical overview of connectivity modelling with a focus on resistance surfaces and discusses current missing aspects and how to address them.
 
-**Zuckerberg et al.
+\noindent _Zuckerberg et al.
 (2020) A review of overlapping landscapes: Pseudoreplication or a red herring in landscape ecology?
-Current Landscape Ecology Reports 5:140–148.** The paper discusses theoretical and practical implications over overlapping landscapes in the context of landscape buffers of varying sizes surrounding sampling sites.
+Current Landscape Ecology Reports 5:140–148._ The paper discusses theoretical and practical implications over overlapping landscapes in the context of landscape buffers of varying sizes surrounding sampling sites.
 
-**Synes et al.
+\noindent _Synes et al.
 (2016) Emerging opportunities for landscape ecological modelling.
-Current Landscape Ecology Reports 1:146–167.** The paper identifies existing gaps in landscape ecological modelling while highlighting potential emerging opportunities.
+Current Landscape Ecology Reports 1:146–167._ The paper identifies existing gaps in landscape ecological modelling while highlighting potential emerging opportunities.
 
-**Manel and Holderegger (2013) Ten years of landscape genetics.
-Trends in Ecology & Evolution 28:614–621.** Ten years after their first paper introducing landscape genetics, this paper reviews the state-of-the-art and describes main topics that have contributed to the progress of landscape genetics.
+\noindent _Manel and Holderegger (2013) Ten years of landscape genetics.
+Trends in Ecology & Evolution 28:614–621._ Ten years after their first paper introducing landscape genetics, this paper reviews the state-of-the-art and describes main topics that have contributed to the progress of landscape genetics.
 
-**Jung et al.
+\noindent _Jung et al.
 (2024) An assessment of the state of conservation planning in europe.
-Philosophical Transactions of the Royal Society B 379:20230015.** The paper assesses where and how spatial conservation planning have been applied across Europe.
+Philosophical Transactions of the Royal Society B 379:20230015._ The paper assesses where and how spatial conservation planning have been applied across Europe.
 

### 2024-10-11 Review.qmd
Starting to work on Review

---
title: "Reviewer Comments"
format:
  pdf:
    keep-tex: false
    include-in-header:
      text: |
        \setlength\parindent{0pt}

papersize: a4paper
fontsize: 11pt
geometry: margin=25mm
---

**Reviewer 1**

This manuscript provides a timely overview of recent developments in computational methods applied to landscape ecology. It is organized around 6 main topics: spatial patterns, simulation models, connectivity, genetics, spatial planning, and spatial sampling. Overall, I found the manuscript's content very comprehensive, well-written, and well-documented with numerous references. Some parts could be made more accessible to readers unfamiliar with computing methods since they could be among the main audience for such a review. Apart from a few minor details listed below, I do not have any major issues to report in the content.

\textcolor[HTML]{0BB67D}{Reply: Thank you very much for the encouraging words. We will use the valuable and helpful feedback to improve the manuscript.}

On the other hand, I think that the structure of the manuscript could be improved to make it clearer, and better organized, and to give it a logical thread which is somehow missing at the moment. Indeed, the current structure seems like a ‘listing’ of topics, with no obvious link between them. This sometimes leads to redundancy and a lack of clarity. As an example, I found the subsection "Vector metrics" a bit confusing, mainly because, as you say in L159-160, most landscape metrics can be computed on both raster and vector data. Another example: I did not find the subsection "Structural connectivity" very consistent with the other two ("Resistance surface" and "Graph theory"). In addition to being very short, I wonder why this section was included without another on functional connectivity, and why it was distinguished from "Graph theory" (in terms of manuscript structure), since structural (as long as functional) connectivity can be modelled with graph theory.

\textcolor[HTML]{0BB67D}{Reply:}

In conclusion, my main recommendation is to revise the structure of the manuscript, without needing to change the content and richness of what it contains. One possibility could be to start with a section focused on data: which types of data structures are available to represent a landscape (e.g. raster, hexagonal grids, vectors, graphs)? Then, you could introduce spatial patterns, emphasizing the fact that their measurement needs to be supported by a suitable data structure, and then other topics.

\textcolor[HTML]{0BB67D}{Reply:}

**Other comments:**

L54. Data-driven and model-driven

\textcolor[HTML]{0BB67D}{Reply: We added the term model-driven as suggested.}

L62. Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table.

\textcolor[HTML]{0BB67D}{Reply:}

L77. They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time.

\textcolor[HTML]{0BB67D}{Reply: We agree that landscape metrics can be extracted straightforwardly from raster data, and we are adding this as another advantage of using them.}

L92. Missing subsection title?

\textcolor[HTML]{0BB67D}{Reply: Indeed we forgot to add a subsection title which is added now.}

L104. "of the selected entropy [measure]"?

\textcolor[HTML]{0BB67D}{Reply: Added as suggested.}

L119. "the edge length" -> Could you specify a bit?

\textcolor[HTML]{0BB67D}{Reply:}

L143. I guess that such metrics are also very scale dependent

\textcolor[HTML]{0BB67D}{Reply: Similar to landscape metrics also surface metrics are scale-dependent because both are designed to quantify landscape heterogeneity - which is inherently scale-dependent. We added this explanation including an according literature references.}

L182. As defined here, it looks to me that this operation is not necessarily linked to a query area, but rather to a spatial signature which can be directly specified or extracted from a query area.

\textcolor[HTML]{0BB67D}{Reply:}

L175-190. I think that this could be clearer by introducing spatial patterns operation and queries as an extension of the relational data model to spatial data. I'm not sure if it is useful to use database jargon such as "one-to-many" or "many-to-many" without defining them, as it could be confusing to readers not familiar with databases. You could either avoid using these terms, define them, or illustrate the operations with a figure.

\textcolor[HTML]{0BB67D}{Reply:}

L175. This subsection is missing in the summary Figure 1.

\textcolor[HTML]{0BB67D}{Reply:}

Figure 1. The links between Spatial sampling, Spatial planning, and Genetics, towards Landscape mean?  The figure would be more meaningful with an explanation of these links, either in the caption or directly on the figure. Also, for other topics (e.g. connectivity), a keyword (or short sentence) summarizing the contribution of the topic to Landscape Ecology would be great.

\textcolor[HTML]{0BB67D}{Reply:}

L192. "facilitate or impeded movement [and flow]"

\textcolor[HTML]{0BB67D}{Reply:}

L200. "movement [and dispersal] data"

\textcolor[HTML]{0BB67D}{Reply:}

L201. "that individuals preferentially move [or disperse] through"

\textcolor[HTML]{0BB67D}{Reply:}

L192-207. Here connectivity seems to be described through the prism of animals, you should modify it slightly to include plants, which do not move, but disperse through different vectors, e.g. animals, gravity, wind, or water.

\textcolor[HTML]{0BB67D}{Reply:}

L219. "to traverse [or disperse through] that cell"

\textcolor[HTML]{0BB67D}{Reply:}

L221-222. Please explain what "edge artifacts" are. I could not find any explanation in the cited reference.

\textcolor[HTML]{0BB67D}{Reply:}

L227. Least cost path is just the result of a mathematically formulated problem, so I don't think that it assumes anything. Assumptions are rather in how such results are used in an ecological context.

\textcolor[HTML]{0BB67D}{Reply:}

L255-257. Edges can be weighted or unweighted. They also can be directed or undirected.

\textcolor[HTML]{0BB67D}{Reply:}

L273-274. I don't think that it is specific to graphs. Any data-dependant model will be sensitive to input data. This highlights the importance of data.

\textcolor[HTML]{0BB67D}{Reply:}

L326. Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e. the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied).

\textcolor[HTML]{0BB67D}{Reply:}

L354-358. You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g. drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models.

\textcolor[HTML]{0BB67D}{Reply:}

L388. Spatial conservation planning -> spatial planning

\textcolor[HTML]{0BB67D}{Reply:}

L405. Reference 204 relies on constraint programming, not linear programming

\textcolor[HTML]{0BB67D}{Reply:}

L412. Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found. Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions.

\textcolor[HTML]{0BB67D}{Reply:}

L418-419. Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one. Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others. Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances. They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers. A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem. There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems.

\textcolor[HTML]{0BB67D}{Reply:}

### 2024-10-11 Sections/01-introduction.qmd
Fix format issues
Rename some files
Starting to work on Review

@@
 Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
-Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
+<!-- R1: Data-driven and model-driven -->
+Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
 Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of landscape ecology experiments].
@@
 
+<!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
 Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them (Fig. 1).
@@
 
-![Overview of landscape ecology topics for which computational methods are discussed.](Figures/Fig-1.png){width=100%}
+![Overview of landscape ecology topics for which computational methods are discussed. Contains modified Copernicus Sentinel data 2024-09-07 for Sentinel data.](Figures/Fig-1.jpg)

### 2024-10-11 Sections/02-spatial-patterns.qmd
Starting to work on Review

@@
 Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
-The strength of such landscape metrics is that they are easy to apply and communicate [@Lausch2015UnderstandingQuantifying], however, limitations have also been identified. 
+<!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
+The strength of such landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data [e.g., remote sensing products\; @Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
+However, also limitations have been identified. 
 These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
@@
 Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+
+<!-- R1: Missing subsection title -->
+## Landscape mosaic method
 
@@
 They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
-However, studies show that insights gained from entropy measures depend on the formulation of the selected entropy, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
+<!-- R1: of the selected entropy measure -->
+However, studies show that insights gained from entropy measures depend on the formulation of the selected entropy measure, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
 
@@
 In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
+<!-- R1: the edge length -> Could you specify a bit -->
 @Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
@@
 These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
+<!-- R1: I guess that such metrics are also very scale dependent -->
 The gradient surface model can increase the resemblance of the data to the natural world because it allows for the inclusion of more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
@@
 Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
+Nevertheless, similar to landscape metrics, surface metrics are also scale-dependent because they are used to quantify landscape heterogeneity, which is itself scale-dependent [@Frazier2016SurfaceMetrics].
 
@@
 ## Operations on spatial patterns 
+<!-- R1: I think that this could be clearer by introducing spatial patterns operation and queries as an extension of the relational data model to spatial data.
+I'm not sure if it is useful to use database jargon such as "one-to-many" or "many-to-many" without defining them, as it could be confusing to readers not familiar with databases.
+You could either avoid using these terms, define them, or illustrate the operations with a figure -->
+<!-- R1: This section is missing in the summary Figure 1 -->
 
@@
 Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
+<!-- R1: As defined here, it looks to me that this operation is not necessarily linked to a query area, but rather to a spatial signature which can be directly specified or extracted from a query area -->
 This allows for the identification of areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].

### 2024-10-11 Sections/03-connectivity.qmd
Starting to work on Review

@@
 
+<!-- R1: Here connectivity seems to be described through the prism of animals, you should modify it slightly to include plants, which do not move, but disperse through different vectors, e.g.
+animals, gravity, wind, or water -->
+<!-- R1: facilitate or impeded movement and flow -->
 Landscape connectivity describes how landscape features facilitate or impede movement [@Taylor2006LandscapeConnectivity] and can be divided into structural and functional connectivity.
@@
 Although multiple approaches that measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], the technical and conceptual quantification of connectivity has proven challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
+<!-- R1: movement and dispersal data -->
+<!-- R1: that individuals preferentially move [or disperse] through -->
 Generally, movement data from individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
@@
 Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
+<!-- R1: to traverse or disperse through that cell -->
 A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
+<!-- R1: Please explain what "edge artifacts" are. I could not find any explanation in the cited reference -->
 Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational] and quantifying cost values for each cell using movement data.
@@
 
+<!-- R!: Least cost path is just the result of a mathematically formulated problem, so I don't think that it assumes anything.
+Assumptions are rather in how such results are used in an ecological context -->
 Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
@@
 Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
+<!-- R1: Edges can be weighted or unweighted. They also can be directed or undirected -->
 The edges can be binary, i.e., representing whether a connection between two nodes either exists or does not exists, or edges can be weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
@@
 However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge from different methods used to construct a graph or from different data sources.
-In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [Foltete2020CouplingLandscape].
+<!-- R1: I don't think that it is specific to graphs. Any data-dependent model will be sensitive to input data. This highlights the importance of data -->
+In many cases, data availability limits the representation of the modelled landscape with subsequent implications for the calculated connectivity metrics and inferences of connectivity [@Foltete2020CouplingLandscape].
 For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.

### 2024-10-11 Sections/05-simulations.qmd
Starting to work on Review

@@
 Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
+<!-- R1: Note that there, although the methods in reference 146 effectively rely on target values optimization (i.e.
+the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
 In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
@@
 Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
+<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g.
+drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.

### 2024-10-11 Sections/07-spatial-planning.qmd
Starting to work on Review

@@
 
+<!-- R1: Spatial conservation planning -> spatial planning -->
 # Spatial conservation planning {#sec-spatialplanning}
@@
 Particularly noteworthy are two advances that incorporate landscape metrics in spatial planning.
+<!-- R1: Reference 204 relies on constraint programming, not linear programming; Justeau-Allaire2021ConstrainedOptimization -->
 One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
@@
 A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
+<!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found.
+Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
 Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
@@
 Most software are based on linear or constrained programming approaches and require a solver to create any outputs.
+<!-- R1: Be careful with such statements, it is extremely complex to objectively compare two solvers, and almost impossible to state the general superiority of one state-of-the-art solver against another one.
+Indeed, some solvers may be very efficient for some kinds of problems while very inefficient for others.
+Also, modelling and solver parametrization can make a lot of differences in solving efficiency and has often a lot of impact on performances.
+They are solver competitions, however, they are based on benchmarks, therefore classic and well-known problems that do not always reflect common uses of solvers.
+A solver fine-tuned to be good on such benchmarks will not necessarily perform well on any problem.
+There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
 Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].

### 2024-10-15 Manuscript.qmd
Further progress on review

@@
 {{< include Sections/01-introduction.qmd >}}
-{{< include Sections/02-spatial-patterns.qmd >}}
-{{< include Sections/03-connectivity.qmd >}}
-{{< include Sections/04-sampling.qmd >}}
-{{< include Sections/05-simulations.qmd >}}
-{{< include Sections/06-genetics.qmd >}}
-{{< include Sections/07-spatial-planning.qmd >}}
-{{< include Sections/08-conclusion.qmd >}}
+{{< include Sections/02-data.qmd >}}
+{{< include Sections/03-spatial-patterns.qmd >}}
+{{< include Sections/04-connectivity.qmd >}}
+{{< include Sections/05-sampling.qmd >}}
+{{< include Sections/06-simulations.qmd >}}
+{{< include Sections/07-genetics.qmd >}}
+{{< include Sections/08-spatial-planning.qmd >}}
+{{< include Sections/09-conclusion.qmd >}}
 

### 2024-10-15 Review.qmd
Further progress on review

@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0025ED}{Reply:}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We agree that a general section on data could be useful for readers. Thus we added such a section at the beginning of the manuscript.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: Indeed such a table would be very helpful.
+Thus, we removed all software-only paragraphs from the specific sections to avoid repetitions and added them to a table (Tab.
+1).}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We added a brief definition of edge length in this context.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We removed the term 'query' from the sentence because it potentially adds confusion.
+However, we believe that the sentence using areas instead is clear enough for readers to follow.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We agree that the whole section was confusing for readers unfamiliar with the topic.
+Thus, we reworded it to make it easier to follow also for all readers by removing a lot of the database jargon.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: Thanks for pointing this out. We added it to the figure.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: Added as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We added this correct specification that movement also includes plants.
+Furthermore, we believe that this aspect is also highlighted by the other suggestions of the reviewer that we implemented in this section.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We are unsure why the point about edge artifacts was added to the manuscript. Thus, we completely removed it here.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We agree that the original statement about underlying assumptions was unclear and reworded the sentence.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We added this specification.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We agree that this statement is true for most if not all methods presented in this manuscript.
+However, we believe it is also helpful to add here specifically for this section.
+Thus, we did not change it.
+However, we added a new short section at the beginning of the manuscript focusing on data specifically.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: Thank you very much for adding this insightful remark about the references. We reworded the sentence to ensure they are reflected in the manuscript.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We added that digital twins benefit from the increasing rate of data generation in ecology as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: Done as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We reworded the sentence being more specific about the references.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: Thank you for pointing out this detail. We added this as another potentail drawback of the method.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{0BB67D}{Reply: We reworded the sentence stating that there are differences between solvers and that the optimal solver depends on the specific problem, however, could potentially not be accessible due to software licenses.}

### 2024-10-15 Sections/01-introduction.qmd
Further progress on review

@@
 <!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
-Here, we introduce recent computational developments in some important topics of landscape ecology including software that implements or is potentially capable of implementing them (Fig. 1).
+Here, we introduce recent computational developments in some important topics of landscape ecology (Fig. 1) including software that implements or is potentially capable of implementing them (Tab. 1).
 In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.

### 2024-10-15 Sections/02-data.qmd
Further progress on review

---
bibliography: CLER.bib
---

<!-- Person responsible: Hesselbarth, Nowosad -->

# Data {#sec-data}

<!-- The choice of the data type, model, and scale are crucial for the analysis's results because every type is a simplified model of the natural world. -->

There are two fundamental data models used to represents landscapes, namely the raster data and the vector data model. 
Raster data uses regularly spaced grid cells, while vector data uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
The issue of scale is closely related to the used data model. 
This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories.
Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
Finally, also temporal scales are relevant both in terms of extent and resolution.

Less frequently discussed, but often critical, is the spatial reference system of the data. 
Cartesian coordinate reference systems measure distances between two points in Euclidean distances, typically in meters. 
In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
Numerous coordinate reference systems exist, each with distinct properties that may distort areas, distances, angles, or all three. 
Therefore, it is essential to choose the appropriate spatial reference system based on the available data and the specific research question

Data quality is generally variable, and data products are not flawless and can contain errors and biases.
For example, global land cover products typically have an overall accuracy of 70 to 80%, indicating that 20 to 30% of the grid cells may be misclassified.
Furthermore, these misclassification is not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping].
Despite this, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

### 2024-10-15 Sections/03-spatial-patterns.qmd
Further progress on review

--- 2024-10-11 Sections/02-spatial-patterns.qmd
@@
 
+<!-- 
 Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
+-->
 
@@
 <!-- R1: the edge length -> Could you specify a bit -->
-@Cushman2016CalculatingConfigurational proposed to relate the edge length to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
+@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the length of neighboring cells with different LULC classes) to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
 This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
@@
 
+<!-- 
 Entropy metrics are available in various software such as the R packages *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or the Python package *PyLandStats*.
+-->
 
@@
 Nevertheless, similar to landscape metrics, surface metrics are also scale-dependent because they are used to quantify landscape heterogeneity, which is itself scale-dependent [@Frazier2016SurfaceMetrics].
+Yet, software to calculate these metrics are still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
+Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
 
-Yet, software to calculate these metrics are still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
+<!--
 Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
 Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].
+-->
 
@@
 
+<!--
 The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics.
 However, all general software frameworks designed for spatial data handling which support vector data can theoretically be used to calculate basic vector metrics.
+-->
 
@@
 Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
-These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures).
-Spatial patterns can be compared between different areas, or for the same area at two different moments in time.
+These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures), and can be calculated for different areas or for the same area at two different moments in time.
 
-Comparing spatial patterns is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
-Furthermore, pattern-based search is a one-to-many operation in which the spatial signature of a query area is compared to the signatures of all other areas.
+Comparing spatial signatures is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
 <!-- R1: As defined here, it looks to me that this operation is not necessarily linked to a query area, but rather to a spatial signature which can be directly specified or extracted from a query area -->
-This allows for the identification of areas with similar patterns compared to the query area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
-Finally, the grouping of spatial patterns is a many-to-many operation in which spatial signatures are calculated for all areas and clustered based on their similarity [@Nowosad2021PatternbasedIdentification].
+Furthermore, signature-based search can be used to compare the spatial signature of a focal area to the spatial signatures of other areas.
+This allows for the identification of areas with similar signatures compared to the focal area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
+Additionally, it is possible to calculate spatial signatures for multiple areas and group them to clusters based on their similarity of signatures [@Nowosad2021PatternbasedIdentification].
 
+<!--
 R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance], or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
@@
 Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.
+-->

### 2024-10-15 Sections/04-connectivity.qmd
Further progress on review

--- 2024-10-11 Sections/03-connectivity.qmd
@@
 <!-- R1: facilitate or impeded movement and flow -->
-Landscape connectivity describes how landscape features facilitate or impede movement [@Taylor2006LandscapeConnectivity] and can be divided into structural and functional connectivity.
+Landscape connectivity describes how landscape features facilitate or impede movement, flow, and dispersal of organisms, e.g., direct movement of animals, but dispersal of sessile plants by wind or water, and can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
 It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
@@
 <!-- R1: that individuals preferentially move [or disperse] through -->
-Generally, movement data from individuals is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
+Generally, movement and dispersal data from individuals is required to infer the elements in the landscape that individuals preferentially move or disperse through [@Simpkins2018AssessingPerformance].
 While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
@@
 <!-- R1: to traverse or disperse through that cell -->
-A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
+A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse or disperse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
 <!-- R1: Please explain what "edge artifacts" are. I could not find any explanation in the cited reference -->
-Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational] and quantifying cost values for each cell using movement data.
+Creating resistance surfaces involves obtaining landscape data for the area of interest, quantifying cost values for each cell using movement and dispersal data, and finally analyzing the surfaces [@Dutta2022OverviewComputational].
 This can include expert opinion [widely used due to low effort, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
@@
 Assumptions are rather in how such results are used in an ecological context -->
-Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
+Least-cost modelling identifies potential paths between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
 By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
@@
 
-Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement between patches is modeled as edges.
+Within the landscape, discrete habitat patches are modeled as nodes, and the potential movement and dispersal between patches is modeled as edges.
 <!-- R1: Edges can be weighted or unweighted. They also can be directed or undirected -->
-The edges can be binary, i.e., representing whether a connection between two nodes either exists or does not exists, or edges can be weighted by cost values representing movement efforts from one patch to another [@Petsas2021LandscapeConnectivity].
+The edges can be binary representing whether a connection between two nodes either exists or does not exists, weighted by cost values representing movement efforts from one patch to another, or undirected or directed allowing movement and dispersal in both or only one direction [@Urban2001LandscapeConnectivity; @Petsas2021LandscapeConnectivity].
 Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
@@
 
+<!--
 A large number of graph theory based software applications have been developed to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.
+-->

### 2024-10-15 Sections/05-sampling.qmd
Further progress on review

--- 2024-09-20 Sections/04-sampling.qmd
@@
 
+<!--
 Software to perform multi-scale analysis include the R package *multilandr*, or the Python libraries *ZonalMetrics* [@Adamczyk2017ZonalMetricsPython], and *PyLandStats*.
+-->

### 2024-10-15 Sections/06-simulations.qmd
Further progress on review

--- 2024-10-11 Sections/05-simulations.qmd
@@
 the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
-In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [@vanStrien2016ImprovedNeutral; @Justeau-Allaire2022FragmentedLandscape].
-In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
+In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [target values are optimized as close as possible\; @vanStrien2016ImprovedNeutral] or target value satisfaction [target values are strictly satisfied, or cannot be satisfied\; @Justeau-Allaire2022FragmentedLandscape].
+<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
+-->
 
@@
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
+This is further facilitated by the increasing rate of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].
 Thus, digital twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-10-15 Sections/07-genetics.qmd
Further progress on review

@@
 
-Genetic data can be applied to complement ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
 Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
-Landscape genetic approaches have been used traditionally to inform landscape ecology applications, e.g., through connectivity modeling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
-Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
+Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
+Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
 
-Advances in molecular methodologies and  synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.<!-- MH: Ref missing -->.
+Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
 This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
@@
 Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].
-<!--JN: software??-->
+
+<!--
+Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
+Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
+Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., @Brown2017SDMtoolbox20].
+-->

### 2024-10-15 Sections/08-spatial-planning.qmd
Further progress on review

--- 2024-10-11 Sections/07-spatial-planning.qmd
@@
 <!-- R1: Spatial conservation planning -> spatial planning -->
-# Spatial conservation planning {#sec-spatialplanning}
+# Spatial planning {#sec-spatialplanning}
 
@@
 <!-- R1: Reference 204 relies on constraint programming, not linear programming; Justeau-Allaire2021ConstrainedOptimization -->
-One is the use of graph theory in linear programming to guide conservation and restoration efforts [@Justeau-Allaire2021ConstrainedOptimization; @Hamonic2023CumulativeEffects].
+One is the use of graph theory in linear [@Hamonic2023CumulativeEffects] or constraint programming [@Justeau-Allaire2021ConstrainedOptimization] to guide conservation and restoration efforts.
 These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
@@
 Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
-However, this comes with the drawback of reduced interpretability.
+However, this comes with the drawbacks of reduced interpretability and no guarantees that the best possible solution can been found because it is not an exact approach.
 
@@
 There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
-Unfortunately there is still a gap between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing].
+Unfortunately there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing], and the optimal solver will depend on the specific problem.
+However, due to software licenses and related cost, the optimal solver might not be always accessible.
 
+<!--
 The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows flexibility to create planning problems in multiple configurations.
@@
 However, many spatial planning software still require substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.
+-->

### 2024-10-15 Sections/09-conclusion.qmd
Further progress on review

--- 2024-09-20 Sections/08-discussion.qmd
@@
 
-# Discussion {#sec-disc}
+# Conclusion {#sec-disc}
 
@@
 In this context, we highlight open-source software as a cornerstone of "Open Science" which offers key advantages like shareability, reproducibility, and transparency, providing great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
+
+<!-- MH: Check connectivity, simulations sections again for 'hidden' software -->
+
+| Topic | Programming language | Software | Reference |
+|:---:|:---:|:---:|:---:|
+| Spatial patterns | R | sf | [@Pebesma2018SimpleFeatures] |
+| Spatial patterns | R | terra | [@Hijmans2024TerraSpatial] |
+| Spatial patterns | R | landscapemetrics | [@Hesselbarth2019LandscapemetricsOpensource] |
+| Spatial patterns | R | multilandr | [@Huais2024MultilandrPackage] |
+| Spatial patterns | R | bespatial | [@Nowosad2024BespatialBoltzmann] |
+| Spatial patterns | R | rasterdiv | [@Rocchini2021RasterdivInformation] |
+| Spatial patterns | R | LandComp | [@Konrad2023AdaptingMultiscale] |
+| Spatial patterns | R | geodiv | [@Smith2021GeodivPackage] |
+| Spatial patterns | R | glcm | [@Zvoleff2020GlcmCalculate] |
+| Spatial patterns | R | GraySpatCon | [@Riitters2023MappingLandscape] |
+| Spatial patterns | R | vectormetrics | [@Matuszek2024VectormetricsLandscape] |
+| Spatial patterns | R | motif | [@Nowosad2021MotifOpensource] |
+| Spatial patterns | Python | GeoPandas | [@Jordahl2020GeopandasPython] |
+| Spatial patterns | Python | rasterio | [@Gillies2013RasterioGeospatial] |
+| Spatial patterns | Python | PyLandStats | [@Bosch2019PyLandStatsOpensource] |
+| Spatial patterns | Julia | GeoStats | [@Hoffimann2018GeoStatsjlHighperformance] |
+| Spatial patterns | Julia | GeoInterface | [@JuliaGeoandcontributors2024GeoInterfaceJulia] |
+| Spatial patterns | Stand-alone software | FRAGSTATS | [@McGarigal2023FRAGSTATSV4] |
+| Spatial patterns | Stand-alone software | GuidosToolbox | [@Vogt2017GuidosToolboxUniversal] |
+| Connectivity | R | ResistanceGA | [@Peterman2018ResistanceGAPackage] |
+| Connectivity | R | lconnect | [@Mestre2023LconnectPackage] |
+| Connectivity | R | grainscape | [@Chubaty2020ToolboxGrainscape] |
+| Connectivity | Julia | Circuitscape | [@Anantharaman2019CircuitscapeJulia] |
+| Connectivity | Julia | ConScape | [@vanMoorter2023AcceleratingAdvances] |
+| Connectivity | Julia | Omniscape | [@Landau2021OmniscapejlSoftware] |
+| Connectivity | Stand-alone software | Graphab | [@Foltete2012SoftwareTool] |
+| Connectivity | Stand-alone software | Conefor | [@Saura2009ConeforSensinode] |
+| Connectivity | Stand-alone software | lscorridors | [@Ribeiro2017LandScapeCorridors] |
+| Sampling | Python | ZonalMetrics | [@Adamczyk2017ZonalMetricsPython] |
+| Simulations | R | NLMR | [@Sciaini2018NLMRLandscapetools] |
+| Simulations | Python | NLMpy | [@Etherington2015NLMpyPython] |
+| Simulations | Julia | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
+| Simulations | Stand-alone software | GradientLand | [@Cambui2015GradientLandSoftware] |
+| Simulations | Stand-alone software | HexSim | [@Schumaker2018HexSimModeling] |
+| Genetics | R | graph4lg | [@Savary2021Graph4lgPackage] |
+| Genetics | R | landgenreport | [@Gruber2015LandgenreportNew] |
+| Genetics | R | adegenet | [@Jombart2008AdegenetPackage] |
+| Genetics | R | ape | [@Paradis2019Ape50] |
+| Genetics | Python | Python based scripts | [@Etherington2011PythonBased] |
+| Genetics | Python | SDMtoolbox 2.0 | [@Brown2017SDMtoolbox20] |
+| Planning | R | prioritizr | [@Hanson2024PrioritizrSystematic] |
+| Planning | R | coco | [@vanMantgem2023CocoConservation] |
+| Planning | R | restoptr | [@Justeau-Allaire2023RestoptrPackage] |
+| Planning | Python | CAPTAIN | [@Silvestro2022ImprovingBiodiversity] |
+
+: Software implementing methodological developments in landscace ecology.
 

### 2024-10-16 Review.qmd
Updates review

--- 2024-10-11 Review.qmd
@@
 ---
-title: "Reviewer Comments"
+title: ""
 format:
@@
 
-\textcolor[HTML]{0BB67D}{Reply: Thank you very much for the encouraging words. We will use the valuable and helpful feedback to improve the manuscript.}
+\textcolor[HTML]{007895}{Reply: Thank you very much for the encouraging words. We will use the valuable and helpful feedback to improve the manuscript.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
-
 In conclusion, my main recommendation is to revise the structure of the manuscript, without needing to change the content and richness of what it contains.
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: This is a very helpful comment about the general structure of the manuscript and we tried to address it as much as possible.
+This included several points.
+First, we agree that a general section on data is useful for readers.
+Thus, we added such a section at the beginning of the manuscript.
+Second, we modified especially the first paragraphs of the "Spatial patterns" (sub-)sections in order to improve the logical flow of the manuscript.
+Third, we rearranged and modified the "Connectivity" section as well.
+We removed the "Structural connectivity" sub-section by including the key statements into the general paragraph about connectivity.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply: We added the term model-driven as suggested.}
+\textcolor[HTML]{007895}{Reply: We added the term "model-driven" as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: Indeed, such a table is very helpful.
+Thus, we removed all paragraphs that only list software from the specific sections and now list all software in a central table (Tab.
+1).}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply: We agree that landscape metrics can be extracted straightforwardly from raster data, and we are adding this as another advantage of using them.}
+\textcolor[HTML]{007895}{Reply: We agree that landscape metrics can be extracted straightforwardly from raster data, and we are adding this as an advantage related to them.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply: Indeed we forgot to add a subsection title which is added now.}
+\textcolor[HTML]{007895}{Reply: Indeed, we forgot to add a sub-section title which is added now.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply: Added as suggested.}
+\textcolor[HTML]{007895}{Reply: We replaced "entropy" with "measure".}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added a brief definition of edge length in this context.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply: Similar to landscape metrics also surface metrics are scale-dependent because both are designed to quantify landscape heterogeneity - which is inherently scale-dependent.
-We added this explanation including an according literature references.}
+\textcolor[HTML]{007895}{Reply: Similar to landscape metrics, also surface metrics are scale-dependent because both are designed to quantify landscape heterogeneity - which is inherently scale-dependent.
+We added this explanation including an according literature reference.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We removed the term "query" from the sentence because we agree that it is a source of confusion.
+However, we believe that the sentence using only "areas" instead is clear enough for readers to follow.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We agree that the whole section was confusing for readers unfamiliar with the topic.
+Thus, we reworded it to make it easier to follow by removing a lot of the database jargon.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{e93c26}{Reply: TODO Thanks for pointing this out. We added it to the figure.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{e93c26}{Reply: TODO}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added "flow" as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added "dispersal" as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added dispersal as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added the specification that movement also includes dispersal of plants.
+Furthermore, we believe that this aspect is now also highlighted due to the other suggestions of the reviewer that we implemented in this section.
+Generally, we added this each time movement is mentioned in the section.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added "disperse" as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We are unsure why the point about edge artifacts was added to the manuscript.
+Thus, we completely removed it here since we think it adds more confusion than helping the readers.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We agree that the original statement about underlying assumptions was unclear and reworded the sentence.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added this specification as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We agree that this statement is true for most if not all methods presented in this manuscript.
+However, we believe it is also helpful to keep it in this section specifically.
+However, we added a new short "Data" section at the beginning of the manuscript in which we also highlight importance of input data specifically.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: Thank you very much for adding this insightful remark about the references. We reworded the sentence to ensure the remarks are reflected in the manuscript.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We added that digital twins benefit from the increasing rate of data generation in ecology.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We renamed the section as suggested.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We reworded the sentence being more specific about the corresponding references.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: Thank you for pointing out this detail. We added this as another potentail drawback of the method.}
 
@@
 
-\textcolor[HTML]{0BB67D}{Reply:}
+\textcolor[HTML]{007895}{Reply: We reworded the sentence stating that there are differences between solvers and that the optimal solver depends on the specific problem.
+However, the best solver could potentially not be accessible due to software licenses.}

### 2024-10-16 Sections/01-introduction.qmd
Updates review

@@
 
-Landscapes are typically defined in landscape ecology as mosaics of different land covers, habitats, ecosystems, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
+Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology] with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
 Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
-Typical research topics include include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, including human activities [@Wu2013KeyConcepts].
+Typical research topics include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].
 
@@
 <!-- R1: Data-driven and model-driven -->
-Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches and complex, adaptive systems [@Poisot2019DatabasedSynthesisdriven].
-Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of landscape ecology experiments].
-Additionally, because landscape ecology is a cross-disciplinary field [including e.g., social sciences, geography, and ecology and evolution\; @With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
+Following, computational ecology can be defined as computational science that is used to address ecological research questions with focus on data-driven and model-driven approaches [@Poisot2019DatabasedSynthesisdriven].
+Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
+Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
 
@@
 Here, we introduce recent computational developments in some important topics of landscape ecology (Fig. 1) including software that implements or is potentially capable of implementing them (Tab. 1).
-In this context, we are focusing on open-source software and specifically scripting languages such as R, Python, and Julia.
-However, we do not provide a systematic literature review, or a general introduction to (computational) landscape ecology.
-For a more comprehensive and general introduction to landscape ecology, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
+In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
+However, we do not provide a systematic literature review or a general introduction to (computational) landscape ecology.
+For such, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
+
+<!-- R1: The links between Spatial sampling, Spatial planning, and Genetics, towards Landscape mean?
+The figure would be more meaningful with an explanation of these links, either in the caption or directly on the figure.
+Also, for other topics (e.g.
+connectivity), a keyword (or short sentence) summarizing the contribution of the topic to Landscape Ecology would be great.
+-->
 

### 2024-10-16 Sections/02-data.qmd
Updates review

@@
 
-# Data {#sec-data}
+# Data models {#sec-data}
 
@@
 Raster data uses regularly spaced grid cells, while vector data uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
+Often, the choice of the data model is driven by data and software availability or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
+For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products.
+
 The issue of scale is closely related to the used data model. 
@@
 Numerous coordinate reference systems exist, each with distinct properties that may distort areas, distances, angles, or all three. 
-Therefore, it is essential to choose the appropriate spatial reference system based on the available data and the specific research question
+Therefore, it is essential to choose the appropriate spatial reference system based on the available data and the specific research question.
 
@@
 For example, global land cover products typically have an overall accuracy of 70 to 80%, indicating that 20 to 30% of the grid cells may be misclassified.
-Furthermore, these misclassification is not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping].
+Furthermore, these misclassifications are not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping].
 Despite this, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

### 2024-10-16 Sections/03-spatial-patterns.qmd
Updates review

@@
 
-Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999SpatialPattern], or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996PatternorientedModelling].
+Spatial patterns can be defined as the scale-dependent predictability of the physical arrangement of observations [@Dale1999SpatialPattern] or as clearly identifiable structures in nature itself or data extracted from nature [@Grimm1996PatternorientedModelling].
 Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
@@
 
-Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around gridded raster data using categorical LULC data on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
+Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
 <!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
-The strength of such landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data [e.g., remote sensing products\; @Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
+The strength of such landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
 However, also limitations have been identified. 
-These limitations include shortcomings in their ability to quantify the spatial structure of landscapes, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [e.g., @Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
+These limitations include shortcomings in their ability to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 
-To address some limitations, for example correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
+To address  limitations related to correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-These components of landscape patterns are in line with recent reviews [e.g., @Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are also two fundamental components of landscape patterns, namely amount and adjacency.
+These components of landscape patterns are in line with recent reviews [@Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are two fundamental components of landscape patterns, namely amount and adjacency.
 These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.
@@
 The landscape mosaic method offers a way to quantify spatial patterns through a tri-polar classification model involving three LULC classes [@Riitters2009IndicatorForest]. 
-The approach uses a moving window to determine the proportions of these three classes within each central grid cell.
+The approach uses a moving window to determine the proportions of these three classes within each focal grid cell.
 These proportions are then classified into 19 mosaic classes based on thresholds that describe the presence, dominance, or uniqueness of each class.
@@
 Entropy measures in landscape ecology are mainly derived from information theory and thermodynamics.
-They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence [spatio-temporal heterogeneity\; @Vranken2015ReviewUse].
+They are mainly used to quantify the complexity of the landscape (spatial heterogeneity), and less often unpredictability (temporal heterogeneity), as well as scale dependence (spatio-temporal heterogeneity) [@Vranken2015ReviewUse].
 <!-- R1: of the selected entropy measure -->
-However, studies show that insights gained from entropy measures depend on the formulation of the selected entropy measure, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
+However, studies show that insights gained from entropy measures depend on the formulation of the selected measure, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
 
@@
 Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
-In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate [the configuration of the system elements\; @Gao2019ComputationBoltzmann].
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
 <!-- R1: the edge length -> Could you specify a bit -->
-@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the length of neighboring cells with different LULC classes) to the microstate of the landscape, and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
+@Cushman2016CalculatingConfigurational proposed to relate the edge length (defined as the side lengths of neighboring cells with different LULC classes) to the microstate of the landscape and use the proportion of microstates to compute the relative Boltzmann entropy of a landscape mosaic.
 This approach was later generalized for calculations based on the raster surface model and point patterns [@Cushman2021EntropyLandscape].
@@
 Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
-However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [e.g., @Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].
+However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].
 
@@
 The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies, which are both generalizations of the Shannon entropy, have been applied to quantify the landscape complexity.
-The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements, but also the pairwise dissimilarities or distances between them.
+The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements but also the pairwise dissimilarities or distances between them.
 Thus, it can be useful in cases where the dissimilarities between LULC classes are relevant.
-Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy), which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].
+Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy) which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].
 
@@
 
-Surface metrics are based on the gradient surface model using gridded raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
+Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
 These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
@@
 Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
-Nevertheless, similar to landscape metrics, surface metrics are also scale-dependent because they are used to quantify landscape heterogeneity, which is itself scale-dependent [@Frazier2016SurfaceMetrics].
-Yet, software to calculate these metrics are still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
+Nevertheless, similar to landscape metrics, surface metrics are also scale-dependent because they are used to quantify landscape heterogeneity which is itself scale-dependent [@Frazier2016SurfaceMetrics].
+Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
+Yet, another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
 
@@
 
-Many landscape ecology studies are based on the raster data model.
-However, the choice of the data model is often driven by data and software availability, or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
-For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products
-However, nowadays, vector data has become increased in availability and has been widely used in other fields, such as geography or urban studies.
+<!-- R1: As an example, I found the subsection "Vector metrics" a bit confusing, mainly because, as you say in L159-160, most landscape metrics can be computed on both raster and vector data. -->
 
@@
 Although this is possible for most of the metrics, there are a few exceptions where metrics are specifically related to the raster data model [@Yao2022VecLIFramework].
-For example, in urban planning, vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
+For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
 Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
@@
 
-However, also vector-based metrics also have limitations.
-The most important limitations relate to computational complexity, which makes calculations of vector metrics slower than their raster equivalents.
-Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap), which is often especially problematic for data from different sources.
+However, also vector-based metrics have limitations.
+The most important limitations relate to computational complexity which makes calculations of vector metrics slower than their raster equivalents.
+Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap) which is often especially problematic for data from different sources.
 Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.
@@
 Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
-These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures), and can be calculated for different areas or for the same area at two different moments in time.
+These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures) and can be calculated for different areas or for the same area at two different moments in time.
 

### 2024-10-16 Sections/04-connectivity.qmd
Updates review

@@
 
+<!-- R1: I did not find the subsection "Structural connectivity" very consistent with the other two ("Resistance surface" and "Graph theory").
+In addition to being very short, I wonder why this section was included without another on functional connectivity, and why it was distinguished from "Graph theory" (in terms of manuscript structure), since structural (as long as functional) connectivity can be modelled with graph theory.
+-->
+
 <!-- R1: Here connectivity seems to be described through the prism of animals, you should modify it slightly to include plants, which do not move, but disperse through different vectors, e.g.
@@
 <!-- R1: facilitate or impeded movement and flow -->
-Landscape connectivity describes how landscape features facilitate or impede movement, flow, and dispersal of organisms, e.g., direct movement of animals, but dispersal of sessile plants by wind or water, and can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
+Landscape connectivity describes how landscape features facilitate or impede movement, flow, and dispersal of organisms, e.g., active movement of animals, or dispersal of sessile plants by wind or water, and can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
 It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
@@
 It is solely a landscape feature and independent of species characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
+Landscape metrics are commonly used to describe structural connectivity in terms of e.g., total habitat amount, patch size and isolation, or possible connections between habitat patches.
+Recently, there has been an increase in focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
+However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity [@Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioral, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].
@@
 <!-- R1: that individuals preferentially move [or disperse] through -->
-Generally, movement and dispersal data from individuals is required to infer the elements in the landscape that individuals preferentially move or disperse through [@Simpkins2018AssessingPerformance].
+Generally, movement and dispersal data from individuals is required to infer the elements in the landscape that organisms preferentially move or disperse through [@Simpkins2018AssessingPerformance].
 While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; @Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
 Thus, in practice connectivity is primarily determined by indirect estimations.
-@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal characteristics of species, and iii) actual connectivity related to observations of individuals moving through a landscape.
-
-## Structural connectivity
-
-Landscape metrics are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
-These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
-Additionally, there has been a recent increase in focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
-However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity, yet functionality can be partly included by using ecological meaningful metrics, such as the core area [e.g., @Kupfer2012LandscapeEcology].
+@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: *i)* structural connectivity determined by the physical attributes of the landscape, *ii)* potential connectivity as a combination of physical landscape attributes and limited information about dispersal characteristics of species, and *iii)* actual connectivity related to observations of individuals moving through a landscape.
 
@@
 <!-- R1: to traverse or disperse through that cell -->
-A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse or disperse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
+A resistance surface is a raster-based representation of a landscape where each cell is assigned a value representing the cost for an individual of a particular species to traverse or disperse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
 <!-- R1: Please explain what "edge artifacts" are. I could not find any explanation in the cited reference -->
 Creating resistance surfaces involves obtaining landscape data for the area of interest, quantifying cost values for each cell using movement and dispersal data, and finally analyzing the surfaces [@Dutta2022OverviewComputational].
-This can include expert opinion [widely used due to low effort, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
+This can include expert opinion (widely used due to low effort, however, difficult to measure accuracy) [@Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).
 
@@
 By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
-This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping to identify reachable areas within a threshold.
+This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations helping to identify reachable areas within a threshold.
 The creation of least-cost paths is a well established technique and there are highly optimized and efficient algorithms available for its calculation.
@@
 In this approach, the landscape is represented as a network of electrical nodes connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
-Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
+Circuit theory is particularly useful when multiple alternative pathways are available.
 It can describe isolation by measuring multiple low- or high resistance pathways and identify areas of high movement probability highlighting important corridors and bottlenecks in the landscape [@Cushman2013BiologicalCorridors].
@@
 Graph theory and related connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
-For example, graph theory based metrics can assist wildlife movement corridor planning [e.g., @Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].
+For example, graph theory based metrics can assist wildlife movement corridor planning [@Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].
 
@@
 <!-- R1: Edges can be weighted or unweighted. They also can be directed or undirected -->
-The edges can be binary representing whether a connection between two nodes either exists or does not exists, weighted by cost values representing movement efforts from one patch to another, or undirected or directed allowing movement and dispersal in both or only one direction [@Urban2001LandscapeConnectivity; @Petsas2021LandscapeConnectivity].
+The edges can be binary representing whether a connection between two nodes either exists or not, weighted by cost values representing movement efforts from one patch to another, or undirected or directed allowing movement and dispersal in both or only one direction [@Urban2001LandscapeConnectivity; @Petsas2021LandscapeConnectivity].
 Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.

### 2024-10-16 Sections/05-sampling.qmd
Updates review

@@
 Establishing a pattern-process link usually requires fieldwork and data collection.
-Given the cost of collecting field observations across landscapes, computational sampling tools have been designed to optimize the study designs [@Pasher2013OptimizingLandscape; @Zuckerberg2020ReviewOverlapping].
-To improve the statistical significance of the relationships between landscape features and field observations, several aspects need to be considered when selecting sampling sites for data collection.
+Given the cost of collecting field observations across landscapes computational sampling tools have been designed to optimize the study designs [@Pasher2013OptimizingLandscape; @Zuckerberg2020ReviewOverlapping].
+To improve the statistical significance of the relationships between landscape features and field observations several aspects need to be considered when selecting sampling sites for data collection.
 First, selected sites should cover the maximum possible range of landscape heterogeneity to maximize the variance of response variables.
 Second, statistical power can be improved by ensuring independence between sites and consequently response variables.
-Finally, since many landscape features that influence response variables are spatially dependent, it is essential to evaluate spatial autocorrelation.
+Finally, since many landscape features that influence response variables are spatially dependent it is essential to evaluate spatial autocorrelation.
 
 Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
-However, in practice, the appropriate scales may not be obvious, or data at such scales may not be available.
+However, in practice, the appropriate scales may not be obvious or data at such scales may not be available.
 Nevertheless, analyses at multiple spatial scales can help to identify the scales of interest.
-For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape], or show the strongest relationships with response variables [@Huais2024MultilandrPackage].
+For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape] or show the strongest relationships with response variables [@Huais2024MultilandrPackage].
 
@@
 Furthermore spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
-If spatial autocorrelation is detected, researchers may either consider further data collection, or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
+If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
 

### 2024-10-16 Sections/06-simulations.qmd
Updates review

@@
 
-Simulation models are a powerful tool to study complex, adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
-Thus, simulations can be seen as experimental systems that allow all imaginable manipulations in order to advance theoretical developments or test hypotheses, which would be impossible in natural systems [@Peck2004SimulationExperiment].
+Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
+Thus, simulations can be seen as experimental systems that allow all imaginable manipulations in order to advance theoretical developments or test hypotheses which would be impossible in natural systems [@Peck2004SimulationExperiment].
 Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
-In general, simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
-Here, we focus mainly on exploratory models, but include both pattern- and process-based models.
+In general simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
+Here, we focus mainly on exploratory models but include both pattern- and process-based models.
 
@@
 
-Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow the users to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes], and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
+Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
 Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].
 
-Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987NeutralModels].
-The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels], or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
+Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes (i.e., neutral landscape model) [@Gardner1987NeutralModels].
+The earliest neutral models are based on percolation theory and assign LULC classes randomly to cells in the landscape [@Gardner1987NeutralModels] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992HierarchicalNeutral].
 Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992CoexistenceSpecies].
-Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022PerlinNoise], or binary space partitioning [@Etherington2022BinarySpace].
+Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis (e.g., Perlin noise) [@Etherington2022PerlinNoise] or binary space partitioning [@Etherington2022BinarySpace].
 Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities based on least-cost networks [@Etherington2024LeastcostNetwork].
@@
 the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
-In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], or are based on target values optimization [target values are optimized as close as possible\; @vanStrien2016ImprovedNeutral] or target value satisfaction [target values are strictly satisfied, or cannot be satisfied\; @Justeau-Allaire2022FragmentedLandscape].
+In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral] or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
 <!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
@@
 
-Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
-Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular], or deforestation [@Soares-Filho2002DinamicaStochastic].
-More recent landscape generators have allowed for the simulation of patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
+Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes (i.e., landscape generators) [@Langhammer2019AgriculturalLandscape; @Salecker2019EFForTSLGrafLandscape].
+Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
+More recent landscape generators have allowed to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
 
@@
 Patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014IndividualbasedModels].
-In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [e.g., @Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [e.g., @Day2020IndividualbasedModeling].
-In order to facilitate future model developments, recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling], or reusable building blocks [@Berger2024ReusableBuilding].
+In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
+In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
 Furthermore, hybrid approaches that couple or even integrate different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
@@
 
-The digital twins concept include three connected elements, namely a physical object or system, its digital representation, and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins], and there is an increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
+The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
 The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
 Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
-Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin], or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
+Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
 <!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g.

### 2024-10-16 Sections/07-genetics.qmd
Updates review

@@
 Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
-Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology, and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
 Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
-Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [e.g., @Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [e.g., @Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [e.g., @Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
+Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [@Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [@Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [@Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].
 
 Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
-This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape], or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
-Analytical frameworks such as the *FOLDS* model [gene flow, genetic offsets, genetic load, dispersal and SDMs\; @Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis [RDA\; @Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
+This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape] or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
+Analytical frameworks such as the *FOLDS* model (gene flow, genetic offsets, genetic load, dispersal and SDMs) [@Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis (RDA) [@Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.
 Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].

### 2024-10-16 Sections/08-spatial-planning.qmd
Updates review

@@
 Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
-In the context of landscape-ecology, it provides a pattern-process link, and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
+In the context of landscape-ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
 As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
-For example, there have been several recent advances in considering landscape patterns directly in spatial planning, or through using simulation model outputs as input features [e.g., @Daigle2020OperationalizingEcological].
+For example, there have been several recent advances in considering landscape patterns directly in spatial planning or through using simulation model outputs as input features [@Daigle2020OperationalizingEcological].
 
-Generally, spatial patterns influence planning outcomes because of their compactness, contiguity and connectedness.
+Generally, spatial patterns influence planning outcomes because of their compactness, contiguity, and connectedness.
 Furthermore, aggregation can be important because of ecological or practical reasons, for example, by ensuring that selected areas satisfy minimum patch size constraints [@Smith2010ApproachEnsuring; @Justeau-Allaire2021ConstrainedOptimization].
@@
 These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
-For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts, aiming to maximize broader landscape connectivity in New Caledonia.
+For example, @Justeau-Allaire2021ConstrainedOptimization used landscape ecology theory and metrics to spatially optimize restoration efforts aiming to maximize broader landscape connectivity in New Caledonia.
 
@@
 Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
-Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
-Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions, as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
+Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation] such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
+Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
 However, this comes with the drawbacks of reduced interpretability and no guarantees that the best possible solution can been found because it is not an exact approach.
@@
 Unfortunately there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing], and the optimal solver will depend on the specific problem.
-However, due to software licenses and related cost, the optimal solver might not be always accessible.
+However, due to software licenses and related cost the optimal solver might not be always accessible.
 

### 2024-10-16 Sections/09-conclusion.qmd
Updates review

@@
 
-Since it emerged in the 1980s [@With2019EssentialsLandscape], the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
+Since it emerged in the 1980s [@With2019EssentialsLandscape] the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
 Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
@@
 Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines.
-This includes advances in remote sensing technology, with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
+This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
 These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
@@
 In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
-Due to its high flexibility and performance, artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
+Due to its high flexibility and performance artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for, e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
 Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].

### 2024-10-17 Review.qmd
further minor changes

@@
 Thus, we added such a section at the beginning of the manuscript.
-Second, we modified especially the first paragraphs of the "Spatial patterns" (sub-)sections in order to improve the logical flow of the manuscript.
+Second, we modified especially the first paragraphs of most (sub-)sections in the "Spatial patterns" section to improve the logical flow of the manuscript.
 Third, we rearranged and modified the "Connectivity" section as well.

### 2024-10-17 Sections/01-introduction.qmd
further minor changes

@@
 
-Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology] with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
+Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
 Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
-Typical research topics include but are not limited to pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].
+Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].
 
@@
 <!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
-Here, we introduce recent computational developments in some important topics of landscape ecology (Fig. 1) including software that implements or is potentially capable of implementing them (Tab. 1).
+Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of implementing them (Tab. 1).
 In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.

### 2024-10-17 Sections/02-data.qmd
further minor changes

@@
 There are two fundamental data models used to represents landscapes, namely the raster data and the vector data model. 
-Raster data uses regularly spaced grid cells, while vector data uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
-Often, the choice of the data model is driven by data and software availability or the familiarity with the approach [@Costanza2019DescribingAnalyzing].
+The raster data model uses regularly spaced grid cells, while the vector data model uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
+Often, the choice of the data model is driven by data and software availability or by familiarity with the approach [@Costanza2019DescribingAnalyzing].
 For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products.
@@
 
-Data quality is generally variable, and data products are not flawless and can contain errors and biases.
+Data quality is generally variable and data products are not flawless potentially containing errors and biases.
 For example, global land cover products typically have an overall accuracy of 70 to 80%, indicating that 20 to 30% of the grid cells may be misclassified.

### 2024-10-17 Sections/03-spatial-patterns.qmd
further minor changes

@@
 Importantly, patterns observed in nature contain information about the history of the system, such as demographic processes, dispersal characteristics, or climatic patterns [@Wiegand2003UsingPatternoriented].
-However, spatial patterns are not only a result of processes, but could also be drivers of them.
+However, spatial patterns are not only a result of processes but could also be drivers of them.
 Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner2015LandscapeEcology].
@@
 
-Traditionally and at present, most of the prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
+Traditionally and at present, many prominent approaches that quantify spatial patterns revolve around raster data using categorical values based on the patch mosaic model [@Frazier2017LandscapeMetrics; @Costanza2019DescribingAnalyzing].
 <!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
-The strength of such landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
+The strength of landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
 However, also limitations have been identified. 
@@
 
-To address  limitations related to correlations and consistency, multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], the variance inflation factor [@Plexida2014SelectingLandscape], or classification trees have been used to identify core metrics that capture main components of landscape patterns.
+To address limitations related to correlation and redundancy several approaches have been used to identify core metrics that capture main components of landscape patterns, such as multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], or the variance inflation factor [@Plexida2014SelectingLandscape].
 More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-These components of landscape patterns are in line with recent reviews [@Gustafson2019HowHas; @Riitters2019PatternMetrics] that have similarly suggested that there are two fundamental components of landscape patterns, namely amount and adjacency.
-These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory, and later formalized in @Nowosad2019InformationTheory.
+These components of landscape patterns are in line with recent reviews that have similarly suggested that there are two fundamental components of landscape patterns, namely amount and adjacency [@Gustafson2019HowHas; @Riitters2019PatternMetrics].
+These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory and later formalized in @Nowosad2019InformationTheory.
 
@@
 <!-- R1: of the selected entropy measure -->
-However, studies show that insights gained from entropy measures depend on the formulation of the selected measure, and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
+However, studies show that insights gained from entropy measures depend on the formulation of the selected measure and on the underlining data model, e.g., the composition of categories or the co-occurrence matrix representation [@Zhao2019CalculatingSpatial].
 
@@
 Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
-However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
-@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix, which could be used to calculate various entropy measures.
+However, to quantify the spatial configuration and the total complexity of the landscape other measures from information theory must be adapted.
+@Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
 This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).
 
-Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, there has been a surge in the development of entropy-based metrics for landscape ecology in the last few years.
+Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years there has been a surge in the development of entropy-based metrics for landscape ecology.
 In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
@@
 Moreover, many other entropy-based metrics have been proposed for use in landscape ecology.
-The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies, which are both generalizations of the Shannon entropy, have been applied to quantify the landscape complexity.
-The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos] as it measures not only the relative abundances of elements but also the pairwise dissimilarities or distances between them.
+The Renyi [@Renyi1961MeasuresEntropy] and Gibbs entropies, which are both generalizations of the Shannon entropy, have been applied to quantify landscape complexity.
+The Rao quadratic entropy [@Rao1982DiversityDissimilarity] has also been applied recently [@Rocchini2017MeasuringRaos], as it measures not only the relative abundances of elements but also the pairwise dissimilarities or distances between them.
 Thus, it can be useful in cases where the dissimilarities between LULC classes are relevant.
-Another recent development is the use of Kullback-Leibler divergence (also known as relative entropy) which is a measure of differences between two probability distributions to describe patterns across scales [@Huckeba2023MultiscaleSpatial].
+Another recent development to describe patterns across scales is the use of Kullback-Leibler divergence (also known as relative entropy) which is a measure of differences between two probability distributions [@Huckeba2023MultiscaleSpatial].
 
@@
 Surface metrics are based on the gradient surface model using raster data and continuous values [@McGarigal2009SurfaceMetrics; @Cushman2010GradientParadigm]. 
-These metrics were mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
+These metrics are mostly adapted from microscopy and molecular physics [@McGarigal2009SurfaceMetrics; @Kedron2018SurfaceMetrics].
+The gradient surface model can increase the resemblance of the data to the natural world because it allows for the inclusion of more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
+Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
+The metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
 <!-- R1: I guess that such metrics are also very scale dependent -->
-The gradient surface model can increase the resemblance of the data to the natural world because it allows for the inclusion of more heterogeneity within each grid cell [@Cushman2010GradientParadigm; @Kedron2018SurfaceMetrics].
-Surface metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
-Many surface metrics have analogous landscape metrics [@Cushman2010GradientParadigm], however, using surface metrics allows for the exploration of different or additional patterns and potentially pattern-process links [@McGarigal2009SurfaceMetrics; @Borthwick2020AlternativeQuantifications].
-Nevertheless, similar to landscape metrics, surface metrics are also scale-dependent because they are used to quantify landscape heterogeneity which is itself scale-dependent [@Frazier2016SurfaceMetrics].
-Yet, software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-Yet, another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
+Nevertheless, similar to landscape metrics, also surface metrics are scale-dependent because they are used to quantify landscape heterogeneity which is itself scale-dependent [@Frazier2016SurfaceMetrics].
+Software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
+Related to surface metrics, another recent approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
 
@@
 
-Hundreds of metrics have already been developed for raster data, and the most straightforward approach may therefore be to reimplement these same metrics for vector data.
-Although this is possible for most of the metrics, there are a few exceptions where metrics are specifically related to the raster data model [@Yao2022VecLIFramework].
+As hundreds of metrics have already been developed for raster data, the most straightforward approach may be to reimplement these same metrics for vector data.
+Although this is possible for many metrics, there are also metrics specifically related to the corresponding data models [@Yao2022VecLIFramework].
 For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
@@
 More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].
-
 However, also vector-based metrics have limitations.
 The most important limitations relate to computational complexity which makes calculations of vector metrics slower than their raster equivalents.
-Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap) which is often especially problematic for data from different sources.
+Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap) which is often problematic especially for data from different sources.
 Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.
@@
 Spatial patterns can be analyzed through a range of computational operations, e.g., comparing, searching, or grouping.
-These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures) and can be calculated for different areas or for the same area at two different moments in time.
-
+These operations are based on spatial signatures (multi-numerical representations of landscape pattern) and dissimilarity measures (functions that quantify differences between the signatures), and can be calculated for different areas or for the same area at two different moments in time.
 Comparing spatial signatures is often used to analyze landscape dynamics, e.g., to detect changes in landscape structure over time [@Duncan2023MonitoringSpatial].
 <!-- R1: As defined here, it looks to me that this operation is not necessarily linked to a query area, but rather to a spatial signature which can be directly specified or extracted from a query area -->
-Furthermore, signature-based search can be used to compare the spatial signature of a focal area to the spatial signatures of other areas.
+Furthermore, signature-based searches can be used to compare the spatial signature of a focal area to the spatial signatures of multiple other areas.
 This allows for the identification of areas with similar signatures compared to the focal area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
-Additionally, it is possible to calculate spatial signatures for multiple areas and group them to clusters based on their similarity of signatures [@Nowosad2021PatternbasedIdentification].
+Additionally, it is possible to calculate spatial signatures for multiple areas and group them to similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].
 

### 2024-10-17 Sections/04-connectivity.qmd
further minor changes

@@
 <!-- R1: facilitate or impeded movement and flow -->
-Landscape connectivity describes how landscape features facilitate or impede movement, flow, and dispersal of organisms, e.g., active movement of animals, or dispersal of sessile plants by wind or water, and can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
+Landscape connectivity describes how landscape features facilitate or impede movement, flow, and dispersal of organisms (e.g., active movement of animals, or dispersal of sessile plants by wind or water).
 It underpins ecosystem functionality, maintains biodiversity and populations, and plays an important role in many conservation actions [@CorreaAyram2016HabitatConnectivity].
+Connectivity can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
 Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
 It is solely a landscape feature and independent of species characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
-Landscape metrics are commonly used to describe structural connectivity in terms of e.g., total habitat amount, patch size and isolation, or possible connections between habitat patches.
-Recently, there has been an increase in focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
+Landscape metrics are commonly used to describe structural connectivity in terms of, e.g., total habitat amount, patch size, or patch isolation.
+Recently, there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size [@Spanowicz2019MeasuringLandscape].
 However, landscape metrics are frequently criticized for an ambiguous link to functional connectivity [@Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
@@
 
-Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
+Many modern connectivity approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
 <!-- R1: to traverse or disperse through that cell -->
-A resistance surface is a raster-based representation of a landscape where each cell is assigned a value representing the cost for an individual of a particular species to traverse or disperse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
+A resistance surface is a raster-based representation of a landscape where each cell is assigned a value reflecting the species-specific cost for an individual to traverse or disperse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].
 <!-- R1: Please explain what "edge artifacts" are. I could not find any explanation in the cited reference -->
@@
 Assumptions are rather in how such results are used in an ecological context -->
-Least-cost modelling identifies potential paths between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
-By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
+Least-cost modelling identifies potential pathways between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
+By calculating pathways between two points it is possible to estimate connectivity between these points based on the accumulated cost along the path.
 This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations helping to identify reachable areas within a threshold.
@@
 However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
-Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].
+Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or general computational improvements [@Leonard2017GlowSoftware].
 
@@
 
-Following the formative work of @Urban2001LandscapeConnectivity, graph theory has become a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement in landscape graphs [@Cushman2013BiologicalCorridors].
-Graph theory and related connectivity metrics based on it can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
+Following the formative work of @Urban2001LandscapeConnectivity, graph theory has become a cornerstone approach to studying landscape connectivity by integrating landscape features and species movement and dispersal in landscape graphs [@Cushman2013BiologicalCorridors].
+Graph theory and related connectivity metrics can be applied over various geographical contexts and for various species [@Foltete2020CouplingLandscape], require very little data inputs [@Godet2021ComparisonLandscape], and can provide key information for conservation planning and management.
 For example, graph theory based metrics can assist wildlife movement corridor planning [@Wang2023OptimalConfiguration], assist infrastructure placement while minimizing habitat fragmentation [@Pietsch2018ContributionConnectivity], or identify habitat patches that maintain overall landscape connectivity [@Liu2020UsingLandscape].
@@
 <!-- R1: Edges can be weighted or unweighted. They also can be directed or undirected -->
-The edges can be binary representing whether a connection between two nodes either exists or not, weighted by cost values representing movement efforts from one patch to another, or undirected or directed allowing movement and dispersal in both or only one direction [@Urban2001LandscapeConnectivity; @Petsas2021LandscapeConnectivity].
+The edges can be binary representing whether a connection between two nodes exists or not, weighted by cost values representing movement or dispersal efforts from one patch to another, or undirected or directed allowing movement or dispersal in both or only one direction [@Urban2001LandscapeConnectivity; @Petsas2021LandscapeConnectivity].
 Connectivity can be quantified through various graph-theoretical metrics describing different aspects, such as the importance of specific patches in maintaining connectivity.
-Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment] which is based on a binary connection model [@Saura2010CommonCurrency].
+Widely reported connectivity metrics include the probability of connectivity index [@Saura2007NewHabitat], habitat availability metrics quantifying potential and functional connectivity [@Saura2010CommonCurrency; @Pietsch2018ContributionConnectivity], or the integral index of connectivity which is based on a binary connection model [@Pascual-Hortal2006ComparisonDevelopment; @Saura2010CommonCurrency].
 
-Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and movement of species [@Petsas2021LandscapeConnectivity].
-Furthermore, the modelling framework was initially only applied to single species but is increasingly being used to model multispecies landscapes to account for interspecific movement abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
-In addition, modern multiple-layer graph approaches enable the modelling of not only spatial, but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
+Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and species-specific movement or dispersal [@Petsas2021LandscapeConnectivity].
+Furthermore, the modelling framework was initially only applied to single species, but is increasingly being used to model multispecies landscapes to account for interspecific movement and dispersal abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
+In addition, modern multiple-layer graph approaches enable the modelling of not only spatial but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
 Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].
 
-However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge from different methods used to construct a graph or from different data sources.
+However, despite the robust theoretical underpinnings of graph theory and its powerful application in landscape connectivity research, contradictory results could emerge from different methods used to construct a graph or various data sources.
 <!-- R1: I don't think that it is specific to graphs. Any data-dependent model will be sensitive to input data. This highlights the importance of data -->

### 2024-10-17 Sections/05-sampling.qmd
further minor changes

@@
 Establishing a pattern-process link usually requires fieldwork and data collection.
-Given the cost of collecting field observations across landscapes computational sampling tools have been designed to optimize the study designs [@Pasher2013OptimizingLandscape; @Zuckerberg2020ReviewOverlapping].
+Given the cost of collecting field observations across landscapes, computational sampling tools have been designed to optimize the study designs [@Pasher2013OptimizingLandscape; @Zuckerberg2020ReviewOverlapping].
 To improve the statistical significance of the relationships between landscape features and field observations several aspects need to be considered when selecting sampling sites for data collection.
@@
 Second, statistical power can be improved by ensuring independence between sites and consequently response variables.
-Finally, since many landscape features that influence response variables are spatially dependent it is essential to evaluate spatial autocorrelation.
+Finally, since many landscape features that influence response variables are spatially dependent, it is essential to evaluate spatial autocorrelation.
 
@@
 Nonetheless, the lack of independence between sites is likely more related to spatial autocorrelation and thus not necessarily prevented by using non-overlapping landscapes [@Zuckerberg2020ReviewOverlapping].
-Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures (e.g., Moran's I) between sampling approaches or spatial scales [@Pasher2013OptimizingLandscape].
-Furthermore spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
-If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount], or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
+Instead, spatial autocorrelation between sampling sites can be evaluated using comparisons of similarity measures between sampling approaches or spatial scales (e.g., Moran's I) [@Pasher2013OptimizingLandscape].
+Furthermore, spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
+If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount] or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
 

### 2024-10-17 Sections/06-simulations.qmd
further minor changes

@@
 Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
-Thus, simulations can be seen as experimental systems that allow all imaginable manipulations in order to advance theoretical developments or test hypotheses which would be impossible in natural systems [@Peck2004SimulationExperiment].
+Thus, simulations can be seen as experimental systems that allow all imaginable manipulations which would be impossible in natural systems in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
 Due to the spatio-temporal scales, complex interactions and feedbacks, or scale mismatches between patterns and processes, simulation models are one of the major approaches in landscape ecology [@Schroder2006AnalysisPattern; @Synes2016EmergingOpportunities].
-In general simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
-Here, we focus mainly on exploratory models but include both pattern- and process-based models.
+In general, simulation models can be classified using two major divisions, namely *i)* predictive or exploratory models, and *ii)* pattern- or process-based models [@Peck2004SimulationExperiment; @Synes2016EmergingOpportunities].
+Here, we focus mainly on exploratory models, but include both pattern- and process-based models.
 
@@
 
-Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow to control certain aspects of the landscape [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
+Landscape simulators are typically used to generate null hypotheses, baselines, or scenarios that allow to control certain aspects of the landscape using the raster data model [@Li2004EvaluatingEffectiveness; @Wang2008NeutralLandscapes] and are generally simpler than ecological simulation models [@Hess2020PioLaGPiosphere].
 Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013SimpleProcessbased; @Langhammer2019AgriculturalLandscape].
@@
 the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
-In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral] or are based on target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral] or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
+In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
 <!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
@@
 Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998LoosecouplingCellular] or deforestation [@Soares-Filho2002DinamicaStochastic].
-More recent landscape generators have allowed to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020PioLaGPiosphere].
+More recent, landscape generators have allowed to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006NeutralModels], deforestation based on road and agricultural fields access [@Peer2013SimpleProcessbased], loss of wetlands, expansion of mining and croplands using the Ising model [@Stepinski2023KineticIsing], agricultural areas dominated by smallholders [@Salecker2019EFForTSLGrafLandscape], or vegetation surrounding watering points in semi-arid savanna rangelands (combining pattern- and process-bases approaches) [@Hess2020PioLaGPiosphere].
 
@@
 In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
-Furthermore, hybrid approaches that couple or even integrate different model types with individual-based models could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
+Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].
 
@@
 
-The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins] and there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
+The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins].
 The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
+They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
+Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
 Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
 Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
+The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
 <!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g.
 drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
-The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
 This is further facilitated by the increasing rate of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].
-Thus, digital twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].

### 2024-10-17 Sections/07-genetics.qmd
further minor changes

@@
 Genetic data can be applied in complement with ecological data to integrate evolutionary processes and patterns into landscape ecology [@Holderegger2008LandscapeGenetics; @Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
-Landscape genetics unites molecular population genetics, spatial statistics, and landscape ecology and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
+Landscape genetics unite molecular population genetics, spatial statistics, and landscape ecology and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
 Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
@@
 Advances in molecular methodologies and synergistic developments of bioinformatic and computational approaches to analyze large-scale genomic data have recently enabled the integration of genome-wide data with spatial ecological data.
-This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape] or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
+This shift towards landscape genomics [@Cushman2018EditorialLeast; @Storfer2018NavigatingInterface] comes with novel utilities through which researchers can apply genomic data across spatio-temporal scales, including the integration of historic or ancient DNA [@Fenderson2020SpatiotemporalLandscape] or simulated present-day and future population genomic data [@House2018EvaluatingMethods; @Chevy2024PopulationGenetics].
 Analytical frameworks such as the *FOLDS* model (gene flow, genetic offsets, genetic load, dispersal and SDMs) [@Aguirre-Liguori2021EvolutionaryGenomics], constrained coordination where covarying sets of genotypes are correlated with multivariate environments through redundancy analysis (RDA) [@Capblancq2021RedundancyAnalysis], and spatial simulations provide promising potential for future landscape genomic studies.

### 2024-10-17 Sections/08-spatial-planning.qmd
further minor changes

@@
 Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
-In the context of landscape-ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
+In the context of landscape ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
 As an integrative approach most spatial planning approaches can benefit from various computational advances in landscape ecology.
@@
 Similarly, contiguity and connectivity are generally considered important to ensure that landscapes or habitat patches are connected, such as river networks [@Beger2010IncorporatingAsymmetric].
-In particular, recently the direct consideration of connectivity has received increasing attention in spatial conservation planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].
+In particular, recently the direct consideration of connectivity has received increasing attention in spatial planning and new computational methods are developed [@Hanson2019ConventionalMethods; @Daigle2020OperationalizingEcological; @Hanson2022ComparisonApproaches; @Beger2022DemystifyingEcological].
 
@@
 <!-- R1: Reference 204 relies on constraint programming, not linear programming; Justeau-Allaire2021ConstrainedOptimization -->
-One is the use of graph theory in linear [@Hamonic2023CumulativeEffects] or constraint programming [@Justeau-Allaire2021ConstrainedOptimization] to guide conservation and restoration efforts.
+One is the use of graph theory to guide conservation and restoration efforts in linear [@Hamonic2023CumulativeEffects] or constraint programming [@Justeau-Allaire2021ConstrainedOptimization].
 These spatial networks approaches have the potential to provide more cost-effective and precise solutions to design reserve networks [@Hamonic2023CumulativeEffects].
@@
 
-A second main development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
+A second development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
+Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
+Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
+Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
 <!-- R1: Beware, deep reinforcement learning is not an exact approach, and thus cannot provide guarantees that the best possible solution has been found.
 Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
-Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
-Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation] such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
-Advantages of reinforcement learning include the ease of incorporating both linear and non-linear functions as well as the scalability of spatio-temporal processes to larger spatial extents [@Equihua2024ConnectivityConservation].
-However, this comes with the drawbacks of reduced interpretability and no guarantees that the best possible solution can been found because it is not an exact approach.
+However, this comes with the drawbacks of reduced interpretability and because it is not an exact approach no guarantee that the best possible solution can been found.
 
@@
 There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
-Unfortunately there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing], and the optimal solver will depend on the specific problem.
+Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the optimal solver will depend on the specific problem.
 However, due to software licenses and related cost the optimal solver might not be always accessible.

### 2024-10-17 Sections/09-conclusion.qmd
further minor changes

@@
 Since it emerged in the 1980s [@With2019EssentialsLandscape] the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
-Thus, future progress within the field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
+Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
 Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
-In this context, we highlight open-source software as a cornerstone of "Open Science" which offers key advantages like shareability, reproducibility, and transparency, providing great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
+In this context, we highlight open-source software as a cornerstone of "Open Science" offering key advantages like shareability, reproducibility, and transparency, which provide great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
 
@@
 These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
-Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible for research studies [@Markham2023ReviewMethods].
+Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
 In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
-Due to its high flexibility and performance artificial intelligence has become popular in ecology and deep learning and machine learning algorithms can be used for, e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
+Due to its high flexibility and performance artificial intelligence has become popular in ecology, and deep learning and machine learning algorithms can be used for, e.g., mapping, classifying and extracting features, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
 Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].

### 2024-10-18 Sections/01-introduction.qmd
adds tiny improvements to the intro

@@
 <!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
-Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of implementing them (Tab. 1).
+Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of executing them (Tab. 1).
 In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
@@
 
-![Overview of landscape ecology topics for which computational methods are discussed. Contains modified Copernicus Sentinel data 2024-09-07 for Sentinel data.](Figures/Fig-1.jpg)
+![Overview of landscape ecology topics for which computational methods are discussed. The image uses modified Copernicus Sentinel-2 data from 2024-09-07.](Figures/Fig-1.jpg)

### 2024-10-18 Sections/01-introduction.qmd
Minor updates figure and table related

--- 2024-10-17 Sections/01-introduction.qmd
@@
 
-![Overview of landscape ecology topics for which computational methods are discussed. Contains modified Copernicus Sentinel data 2024-09-07 for Sentinel data.](Figures/Fig-1.jpg)
+![Overview of landscape ecology topics discussed in relation to computational methods.
+All topics are highly interconnected.
+For example, the quantification of spatial patterns often relies on sampling approaches and can serve as target values for simulation models.
+Simulated neutral landscapes may act as null hypotheses for connectivity analyses.
+Landscape genetic analyses frequently depend on connectivity estimates, which can, in turn, inform spatial planning decisions.
+The figure incorporates modified Copernicus Sentinel data (Date: 2024-09-07).](Figures/Fig-1.jpg)

### 2024-10-18 Sections/03-spatial-patterns.qmd
Minor updates figure and table related

@@
 These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory and later formalized in @Nowosad2019InformationTheory.
-
-<!-- 
-Software to calculate landscape metrics include the *landscapemetrics* [@Hesselbarth2019LandscapemetricsOpensource] and *multilandr* [@Huais2024MultilandrPackage] R packages, the *PyLandStats* Python library [@Bosch2019PyLandStatsOpensource], or the stand-alone software *FRAGSTATS* [@McGarigal2023FRAGSTATSV4].
--->
 
@@
 
-<!-- 
-Entropy metrics are available in various software such as the R packages *landscapemetrics*, *bespatial* [@Nowosad2024BespatialBoltzmann], *rasterdiv* [@Rocchini2021RasterdivInformation], *LandComp* [@Konrad2023AdaptingMultiscale], or the Python package *PyLandStats*.
--->
-
 ## Surface metrics
@@
 Related to surface metrics, another recent approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
-
-<!--
-Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
-Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].
--->
 
@@
 
-<!--
-The *vectormetrics* R package [@Matuszek2024VectormetricsLandscape] is specifically designed for vector metrics.
-However, all general software frameworks designed for spatial data handling which support vector data can theoretically be used to calculate basic vector metrics.
--->
-
 ## Operations on spatial patterns 
@@
 Additionally, it is possible to calculate spatial signatures for multiple areas and group them to similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].
-
-<!--
-R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance], or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
-These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
-Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.
--->

### 2024-10-18 Sections/08-spatial-planning.qmd
Minor updates figure and table related

@@
 However, due to software licenses and related cost the optimal solver might not be always accessible.
-
-<!--
-The *prioritizr* R package [@Hanson2024PrioritizrSystematic] supports a range of compactness and contiguity constraints, penalties, and generally allows flexibility to create planning problems in multiple configurations.
-Similarly, the *coco* R package uses connectivity constraints in an integer linear planning framework to identify potential outcomes that minimize further fragmentation of the landscape [@vanMantgem2023CocoConservation].
-The *restoptr* R package uses constrained programming to identify restoration areas that would most effectively contribute to improved landscape connectivity [@Justeau-Allaire2023RestoptrPackage].
-Innovatively, it uses landscape ecology indices not as constraints, but as specific targets, e.g., where one could optimize the effective mesh size in a region.
-The *CAPTAIN* Python software library implements reinforcement learning for area-based conservation efforts [@Silvestro2022ImprovingBiodiversity].
-However, many spatial planning software still require substantial prior knowledge on spatial optimization and statistics, which can create a barrier to more widespread adoption.
--->

### 2024-10-18 Sections/09-conclusion.qmd
Minor updates figure and table related

@@
 
-<!-- MH: Check connectivity, simulations sections again for 'hidden' software -->
+<!-- MH: Double check programming languages -->
 
@@
 | Connectivity | Stand-alone software | lscorridors | [@Ribeiro2017LandScapeCorridors] |
+| Connectivity | Stand-alone software | GFLOW | [@Leonard2017GlowSoftware] |
 | Sampling | Python | ZonalMetrics | [@Adamczyk2017ZonalMetricsPython] |
 | Simulations | R | NLMR | [@Sciaini2018NLMRLandscapetools] |
-| Simulations | Python | NLMpy | [@Etherington2015NLMpyPython] |
+| Simulations | R | rflsgen | [@Justeau-Allaire2022FragmentedLandscape] |
+| Simulations | Python | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
+| Simulations | Python | Pathwalker | [@UnnithanKumar2022PathwalkerNew] |
 | Simulations | Julia | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
+| Simulations | NetLogo | EFForTS-LGraf | [@Salecker2019EFForTSLGrafLandscape] |
 | Simulations | Stand-alone software | GradientLand | [@Cambui2015GradientLandSoftware] |
+| Simulations | Stand-alone software | Landscape Generator | [@vanStrien2016ImprovedNeutral] |
 | Simulations | Stand-alone software | HexSim | [@Schumaker2018HexSimModeling] |
+| Simulations | Stand-alone software | RangeShifter2.0 | [@Bocedi2021RangeShifter20] |
 | Genetics | R | graph4lg | [@Savary2021Graph4lgPackage] |
@@
 
-: Software implementing methodological developments in landscace ecology.
+: Software and models that implement recent methodological advances in landscace ecology.
 

### 2024-10-18 Sections/02-data.qmd
improves data part

@@
 Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
-Finally, also temporal scales are relevant both in terms of extent and resolution.
+Lastly, temporal scales are relevant in terms of extent and resolution.
 
-Less frequently discussed, but often critical, is the spatial reference system of the data. 
+The spatial reference system of the data is less frequently discussed but often critical. 
 Cartesian coordinate reference systems measure distances between two points in Euclidean distances, typically in meters. 
 In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
-Numerous coordinate reference systems exist, each with distinct properties that may distort areas, distances, angles, or all three. 
-Therefore, it is essential to choose the appropriate spatial reference system based on the available data and the specific research question.
+Numerous coordinate reference systems exist, each with distinct properties that may distort areas, distances, angles, and aimed to represent specific regions or countries.
+Therefore, it is essential to choose the appropriate spatial reference system based on the available data, area of interest, and the specific research question.
 
-Data quality is generally variable and data products are not flawless potentially containing errors and biases.
+Data quality is generally variable, and data products are not flawless, potentially containing errors and biases.
 For example, global land cover products typically have an overall accuracy of 70 to 80%, indicating that 20 to 30% of the grid cells may be misclassified.

### 2024-10-18 Sections/03-spatial-patterns.qmd
improves spat patterns

--- 2024-10-17 Sections/03-spatial-patterns.qmd
@@
 However, spatial patterns are not only a result of processes but could also be drivers of them.
-Thus, untangling the history of a landscape and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner2015LandscapeEcology].
+Thus, untangling a landscape's history and linking its spatial patterns to ecological processes is one of the core concepts of landscape ecology [@Turner2015LandscapeEcology].
 
@@
 To address limitations related to correlation and redundancy several approaches have been used to identify core metrics that capture main components of landscape patterns, such as multivariate factor analysis [@Riitters1995FactorAnalysis], multivariate statistics [@Lustig2015MorePredictable], principal component analysis [@Cushman2008ParsimonyLandscape], or the variance inflation factor [@Plexida2014SelectingLandscape].
-More recently, principal component analysis over a set of landscapes blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
+More recently, principal component analysis over a set of landscape blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
 These components of landscape patterns are in line with recent reviews that have similarly suggested that there are two fundamental components of landscape patterns, namely amount and adjacency [@Gustafson2019HowHas; @Riitters2019PatternMetrics].
@@
 
-The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape omitting the spatial configuration.
+The Shannon diversity index [@Shannon1948MathematicalTheory] quantifies the richness and evenness of categories in the landscape, omitting the spatial configuration.
 Shannon's entropy can also be modified to include the landscape's spatial configuration, e.g., by weights calculated from intra- and interclass distances [@Claramunt2012SpatiotemporalForm].
-However, to quantify the spatial configuration and the total complexity of the landscape other measures from information theory must be adapted.
+However, to quantify the spatial configuration and the total complexity of the landscape, other measures from information theory must be adapted.
 @Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
@@
 
-Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years there has been a surge in the development of entropy-based metrics for landscape ecology.
+Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years, there has been a surge in the development of entropy-based metrics for landscape ecology.
 In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
@@
 Recently, @Zhang2020CalculatingWasserstein extended the original definition of Boltzmann entropy to incorporate information about the adjacency of the same categories in the landscape mosaic by using the number of contiguous patches of the same category.
-However, the relations between the Shannon entropy and the Boltzmann entropy and their interpretations in landscape ecology are the subject of ongoing discussion [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial].
+The relationship between Shannon and Boltzmann entropies in landscape ecology remains contentious, with recent studies challenging the thermodynamic interpretation of Boltzmann-inspired measures and advocating for Shannon entropy as a more general form [@Stepinski2022CurbYour; @Huckeba2023MultiscaleSpatial], highlighting the need for further research to elucidate their connections to environmental processes.
 
@@
 Software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
-Related to surface metrics, another recent approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
+Related to surface metrics, another recent approach to quantifying the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
 
@@
 
-## Vector metrics
+## Vector-based metrics
 
@@
 Although this is possible for many metrics, there are also metrics specifically related to the corresponding data models [@Yao2022VecLIFramework].
-For example, in urban planning vector metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
-Earlier approaches to quantify shape complexity included four categories of compactness measures, namely perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
+For example, in urban planning vector-based metrics are applied to quantify the shapes of urban areas and characterize the complexity of building footprints [@Basaraner2017PerformanceShape].
+Earlier approaches to quantifying shape complexity included four categories of compactness measures: perimeter-area, single parameters of related circles, dispersion of elements of the area around a centroid, and direct comparison to standard shapes [@Maceachren1985CompactnessGeographic].
 More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].
 However, also vector-based metrics have limitations.
-The most important limitations relate to computational complexity which makes calculations of vector metrics slower than their raster equivalents.
-Another technical issue is the requirement of topologically correct data (e.g., geometries cannot overlap) which is often problematic especially for data from different sources.
+The most important limitation relates to computational complexity, which makes calculations of vector-based metrics slower than their raster equivalents.
+Another technical issue is the requirement for topologically correct data (e.g., geometries cannot overlap), which is often problematic, especially for data from different sources.
 Last, the pattern-process link for vector-based metrics remains underexplored compared to raster-based metrics.
@@
 This allows for the identification of areas with similar signatures compared to the focal area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
-Additionally, it is possible to calculate spatial signatures for multiple areas and group them to similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].
+Additionally, it is possible to calculate spatial signatures for multiple areas and group them into similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].
 

### 2024-10-18 Sections/06-simulations.qmd
adds tiny changes

@@
 the model aims to get as close as possible to target values), the methods in reference 147 are rather based on target value satisfaction (the model guarantees that target values are strictly satisfied, or guarantees that they cannot be satisfied) -->
-In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
+In order to ensure realistic neutral landscapes, several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
 <!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].

### 2024-10-18 Sections/09-conclusion.qmd
adds tiny changes

--- 2024-10-17 Sections/09-conclusion.qmd
@@
 
-Since it emerged in the 1980s [@With2019EssentialsLandscape] the research field of landscape ecology has constantly evolved and is now highly dependent on computational methods.
+Since its emergence in the 1980s [@With2019EssentialsLandscape], landscape ecology research has constantly evolved and is now highly dependent on computational methods.
 Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
@@
 This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
-These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR [@Crowley2020RemoteSensings].
+These systems provide both passive sensors, such as multispectral, hyperspectral, and thermal sensors and active optical sensors, such as LiDAR and SAR [@Crowley2020RemoteSensings].
 Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
@@
 
-Climate change as well as the biodiversity crisis are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
+Climate change, as well as the biodiversity crisis, are two intervening and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
 Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution [@With2019EssentialsLandscape].

### 2024-10-18 Sections/03-spatial-patterns.qmd

@@
 
-<!--
-Nonetheless, the *geodiv* R package [@Smith2021GeodivPackage] provides various surface metrics, and related textures from grey-level co-occurrence matrices are provided by the *glcm* R package [@Zvoleff2020GlcmCalculate].
-Another approach to quantify the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values as implemented in the *GraySpatCon* software [@Riitters2023MappingLandscape].
--->
-
-## Vector-based metrics
+## Vector metrics
 
@@
 This allows for the identification of areas with similar signatures compared to the focal area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
-Additionally, it is possible to calculate spatial signatures for multiple areas and group them into similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].
 
-<!--
-R, Python, and Julia have a wide range of packages for spatial operations, including *sf* [@Pebesma2018SimpleFeatures], or *terra* [@Hijmans2024TerraSpatial] in R, *GeoPandas* [@Jordahl2020GeopandasPython], or *rasterio* [@Gillies2013RasterioGeospatial] in Python, and *GeoStats.jl* [@Hoffimann2018GeoStatsjlHighperformance], or *GeoInterface.jl* [@JuliaGeoandcontributors2024GeoInterfaceJulia] in Julia.
-These packages together with additional statistical and machine learning libraries allow for a wide range of spatial operations.
-Additionally, the *motif* R package [@Nowosad2021MotifOpensource] provides a set of functions for spatial pattern operations, such as comparing, searching, or grouping spatial patterns customized for the use in landscape ecology.
--->
+Additionally, it is possible to calculate spatial signatures for multiple areas and group them to similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].

### 2024-10-18 Sections/04-connectivity.qmd

@@
 For example, @Godet2021ComparisonLandscape modelled the same landscape using three different data types which resulted in different distribution of connectivity values.
-
-<!--
-A large number of graph theory based software applications have been developed to estimate connectivity including *ConScape.jl* [Julia\; @vanMoorter2023AcceleratingAdvances], *Omniscape.jl* [Julia\;@Landau2021OmniscapejlSoftware], *lconnect* [R\; @Mestre2023LconnectPackage], *grainscape* [R\; @Chubaty2020ToolboxGrainscape], *Circuitscape* [@Anantharaman2019CircuitscapeJulia], *Conefor* [@Saura2009ConeforSensinode], *Graphab* [@Foltete2012SoftwareTool], among others.
--->

### 2024-10-18 Sections/06-simulations.qmd

--- 2024-10-17 Sections/06-simulations.qmd
@@
 In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017DesigningNeutral], target values optimization (target values are optimized as close as possible) [@vanStrien2016ImprovedNeutral], or target value satisfaction (target values are strictly satisfied, or cannot be satisfied) [@Justeau-Allaire2022FragmentedLandscape].
-<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available including Python [NLMpy\; @Etherington2015NLMpyPython], R [NLMR\; @Sciaini2018NLMRLandscapetools], Julia [NeutralLandscapes.jl\; @Poisot2023NeutralLandscapesJulia], or stand-alone software [GradientLand\; @Cambui2015GradientLandSoftware].
--->
 

### 2024-10-18 Sections/07-genetics.qmd

@@
 Such studies may explore interactions between complex evolutionary processes, including demography, multidimensional (e.g., strength and direction) gene flow and migration across spatio-temporal scales, genetic differentiation across landscapes, genomic load and adaptive potential, and the interpretation of these evolutionary processes in context of landscape ecology [@Balkenhol2017LandscapeGenomics; @House2018EvaluatingMethods; @Fenderson2020SpatiotemporalLandscape; @Marcus2021FastFlexible; @Chevy2024PopulationGenetics].
-
-<!--
-Given the importance of connectivity, all previously discussed connectivity software are also relevant for landscape genetics.
-Additionally, popular software for landscape genetics includes the R packages *graph4lg* [@Savary2021Graph4lgPackage], *landgenreport* [@Gruber2015LandgenreportNew], *adegenet* [@Jombart2008AdegenetPackage], or *ape* [@Paradis2019Ape50], among others.
-Furthermore, a series of Python scripts have been developed for use in landscape genetics studies [@Etherington2011PythonBased], or GIS-based toolboxes written in Python exist that provide landscape genetics functionality [e.g., @Brown2017SDMtoolbox20].
--->

### 2024-10-18 Sections/01-introduction.qmd
Make sure conflict changes are included

@@
 Landscape genetic analyses frequently depend on connectivity estimates, which can, in turn, inform spatial planning decisions.
-The figure incorporates modified Copernicus Sentinel data (Date: 2024-09-07).](Figures/Fig-1.jpg)
+The image uses modified Copernicus Sentinel-2 data from 2024-09-07.](Figures/Fig-1.jpg)

### 2024-10-18 Manuscript.qmd
Updates review
Small ✍️ updates to intro

@@
         \renewcommand\Affilfont{\footnotesize}
-        \newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\textcolor[HTML]{A6CE39}{\aiOrcid#1}}}
-        \newcommand{\mail}[1]{\href{mailto:#1}{\textcolor[HTML]{052E81}{\faEnvelope#1}}}
+        \newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\textcolor[HTML]{A6CE39}{\aiOrcid\hspace{1mm}#1}}}
+        \newcommand{\mail}[1]{\href{mailto:#1}{\textcolor[HTML]{052E81}{\faEnvelope\hspace{1mm}#1}}}
 
@@
 
-\noindent \footnotesize M.H.K. Hesselbarth \orcid{0000-0003-1125-9918} \mail{mhk.hesselbarth@gmail.com}
+\noindent \footnotesize M.H.K. Hesselbarth \orcid{0000-0003-1125-9918} \hspace{5mm} \mail{mhk.hesselbarth@gmail.com}
 
@@
 
-\noindent All authors contributed particular sections to the manuscript coordinated by MHKH.
+\noindent All authors contributed particular sections to the manuscript and assisted with the revision. Overall coordination, first and revised draft was led by MHKH.
 Correspondence to [mhk.hesselbarth@gmail.com](mailto:mhk.hesselbarth@gmail.com)
@@
 Landscapes can be defined as mosaics of different land covers, habitats, ecosystems, or land-use systems.
-The link between spatial heterogeneity and ecological processes is the core concept in the research field of landscape ecology.
+The link between spatial heterogeneous patterns and ecological processes is the core concept in the research field of landscape ecology.
 Nowadays, advanced computational methods are essential to the field due to its cross-disciplinary nature, the increasing availability of data, and the complexity of landscape systems.
@@
 
-This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
+This review provides an overview of recent developments in computational methods that in our eyes have advanced the research field of landscape ecology.
 We focus on key topics such as spatial patterns, connectivity, simulations and modelling, landscape genetics, and spatial planning.

### 2024-10-18 Sections/01-introduction.qmd
Small ✍️ updates to intro

@@
 
-Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
+Landscapes are typically defined in landscape ecology as mosaics of different ecosystems, habitats, land covers, or land-use systems [@Forman1986LandscapeEcology], with emphasis on heterogeneity of at least one factor of interest [@Turner2015LandscapeEcology].
 Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
@@
 
-Computational science analyzes abstracted core mechanisms of research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices].
+The use of computational science to abstract core mechanisms and research questions using data and algorithms and is one of the most important tools of modern science [@Wilson2014BestPractices] and by extension landscape ecology as such.
 <!-- R1: Data-driven and model-driven -->
@@
 Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
-Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
+Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the availability of data increases steadily [@Jaric2020IEcologyHarnessing], and because of the complexity of landscape systems [@Newman2019ScalingComplexity], there is often a need to more advanced computational methods.
 
 <!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
-Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of executing them (Tab. 1).
+Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig.
+1), including software that implements or is potentially capable of creating novel insights in these topics (Tab.
+1).
 In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
-However, we do not provide a systematic literature review or a general introduction to (computational) landscape ecology.
-For such, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
+We do not aim to provide a systematic literature review or a general introduction to (computational) landscape ecology, but rather a perspective on recent developments and advances in computational methods.
+For a set of more comprehensive reviews we refer to @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.
 

### 2024-10-18 Sections/09-conclusion.qmd
Fix table

--- 2024-10-18 Sections/09-conclusion.qmd
@@
 | Spatial patterns | R | glcm | [@Zvoleff2020GlcmCalculate] |
-| Spatial patterns | R | GraySpatCon | [@Riitters2023MappingLandscape] |
 | Spatial patterns | R | vectormetrics | [@Matuszek2024VectormetricsLandscape] |
@@
 | Spatial patterns | Python | GeoPandas | [@Jordahl2020GeopandasPython] |
-| Spatial patterns | Python | rasterio | [@Gillies2013RasterioGeospatial] |
+| Spatial patterns | Python | Rasterio | [@Gillies2013RasterioGeospatial] |
 | Spatial patterns | Python | PyLandStats | [@Bosch2019PyLandStatsOpensource] |
@@
 | Connectivity | Stand-alone software | Conefor | [@Saura2009ConeforSensinode] |
-| Connectivity | Stand-alone software | lscorridors | [@Ribeiro2017LandScapeCorridors] |
+| Connectivity | Stand-alone software | LSCorridors | [@Ribeiro2017LandScapeCorridors] |
 | Sampling | Python | ZonalMetrics | [@Adamczyk2017ZonalMetricsPython] |
 | Simulations | R | NLMR | [@Sciaini2018NLMRLandscapetools] |
+| Simulations | R | rflsgen | [@Justeau-Allaire2022FragmentedLandscape] |
 | Simulations | Python | NLMpy | [@Etherington2015NLMpyPython] |
+| Simulations | Python | Pathwalker | [@UnnithanKumar2022PathwalkerNew] |
 | Simulations | Julia | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
+| Simulations | NetLogo | EFForTS-LGraf | [@Salecker2019EFForTSLGrafLandscape] |
 | Simulations | Stand-alone software | GradientLand | [@Cambui2015GradientLandSoftware] |
+| Simulations | Stand-alone software | Landscape Generator | [@vanStrien2016ImprovedNeutral] |
 | Simulations | Stand-alone software | HexSim | [@Schumaker2018HexSimModeling] |
+| Simulations | Stand-alone software | RangeShifter2.0 | [@Bocedi2021RangeShifter20] |
 | Genetics | R | graph4lg | [@Savary2021Graph4lgPackage] |
@@
 | Planning | R | prioritizr | [@Hanson2024PrioritizrSystematic] |
-| Planning | R | coco | [@vanMantgem2023CocoConservation] |
+| Planning | R | CoCo | [@vanMantgem2023CocoConservation] |
 | Planning | R | restoptr | [@Justeau-Allaire2023RestoptrPackage] |

### 2024-10-18 Sections/02-data.qmd
Data updates with some suggestions of literature ✍️

@@
 
-There are two fundamental data models used to represents landscapes, namely the raster data and the vector data model. 
-The raster data model uses regularly spaced grid cells, while the vector data model uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
+There are two fundamental data models used to represents landscapes, namely the raster and the vector data model. 
+The raster data model typically uses regularly spaced grid cells, while the vector data model uses points, lines, and polygons to represent landscape features [@With2019EssentialsLandscape].
 Often, the choice of the data model is driven by data and software availability or by familiarity with the approach [@Costanza2019DescribingAnalyzing].
-For example, many LULC maps are provided as raster data as they often relate to underlying remote sensing products.
+For example, many LULC maps are provided as gridded raster data as they often relate to underlying remote sensing products.
 
-The issue of scale is closely related to the used data model. 
-This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
-Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories.
+Similarly the issue of scale is closely related to the used data model. This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
+Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories or habitats [@Jung2020global].
 Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
-Lastly, temporal scales are relevant in terms of extent and resolution.
+Lastly, temporal scales are relevant in terms of extent and resolution, but also with regards to the detail of captured processes acting at the landscape scale.
 
@@
 In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
-Numerous coordinate reference systems exist, each with distinct properties that may distort areas, distances, angles, and aimed to represent specific regions or countries.
-Therefore, it is essential to choose the appropriate spatial reference system based on the available data, area of interest, and the specific research question.
+Numerous coordinate reference systems exist and novel projections are constantly being developed to improve spatial cohesion, efficiency and accuracy [@bauer2014optimisation].
+However each reference systems comes with distinct properties that may distort areas, distances, angles, and aimed to represent specific regions or countries.
+Therefore, for analysts it is essential to choose the appropriate spatial reference system based on the available data, area of interest, and the specific research question.
 
 Data quality is generally variable, and data products are not flawless, potentially containing errors and biases.
-For example, global land cover products typically have an overall accuracy of 70 to 80%, indicating that 20 to 30% of the grid cells may be misclassified.
+For example, global land-cover products typically have an overall accuracy of 70 to 80% [@xu2024comparative], indicating that 20 to 30% of the grid cells may be misclassified.
 Furthermore, these misclassifications are not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping].
-Despite this, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.
+Rather than relying on global products, there are however also increasingly voices that advocate for the production and use of regional land-cover products as alternative [@tulbure2022regional].
+Nevertheless, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

### 2024-10-18 Sections/05-sampling.qmd
Further edits ✍️

@@
 Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
-However, in practice, the appropriate scales may not be obvious or data at such scales may not be available.
-Nevertheless, analyses at multiple spatial scales can help to identify the scales of interest.
+However, in practice, the appropriate scales may not be obvious or data at such scales may not be available and landscape-wide and local process interactions can occur across multiple scales [@miguet2016determines].
+Here analyses at multiple spatial scales can help to identify the scales of interest.
 For instance, scalograms can be used to reveal scale thresholds that maximize landscape heterogeneity [@Pasher2013OptimizingLandscape] or show the strongest relationships with response variables [@Huais2024MultilandrPackage].
@@
 If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount] or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
-
-<!--
-Software to perform multi-scale analysis include the R package *multilandr*, or the Python libraries *ZonalMetrics* [@Adamczyk2017ZonalMetricsPython], and *PyLandStats*.
--->

### 2024-10-18 Sections/06-simulations.qmd
Further edits ✍️

@@
 
-Simulation models are a powerful tool to study complex adaptive ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
+Simulation models are a powerful tool to study complex adaptive socio-ecological systems in controllable, reproducible, and replicable settings [@Petrovskii2012ComputationalEcology].
 Thus, simulations can be seen as experimental systems that allow all imaginable manipulations which would be impossible in natural systems in order to advance theoretical developments or test hypotheses [@Peck2004SimulationExperiment].
@@
 In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
+They can also incorporate a range of socio-economic, political and governance information to influence how agents (such as farmers) interact within a landscape [@murray2014combining].
 In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
@@
 The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins].
+Digital twins are particular useful for real-time workflows that allow now- and forecasting of complex dynamics in a landscape.
 The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
 They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
+<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g.
+drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
+This is further facilitated by the increasing rate of novel sources of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].
+
 Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
@@
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
-<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g.
-drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
-This is further facilitated by the increasing rate of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].

### 2024-10-18 Sections/09-conclusion.qmd
Minor edits in conclusion and edition

--- 2024-10-18 Sections/09-conclusion.qmd
@@
 Since its emergence in the 1980s [@With2019EssentialsLandscape], landscape ecology research has constantly evolved and is now highly dependent on computational methods.
-Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
+Thus, future progress within the field will largely depend on integration of novel data sources and available software and computational tools [@Costanza2019DescribingAnalyzing].
 Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
@@
 | Spatial patterns | R | motif | [@Nowosad2021MotifOpensource] |
+| Spatial patterns | Python | LecoS | [@jung2016lecos] |
 | Spatial patterns | Python | GeoPandas | [@Jordahl2020GeopandasPython] |
@@
 
-: Software implementing methodological developments in landscace ecology.
+: Software and models that implement recent methodological advances in landscape ecology.
 
@@
 These systems provide both passive sensors, such as multispectral, hyperspectral, and thermal sensors and active optical sensors, such as LiDAR and SAR [@Crowley2020RemoteSensings].
-Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
+Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible to scientists and spatial planners [@Markham2023ReviewMethods].
 In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].

### 2024-10-18 Manuscript.qmd
use modeling spelling

--- 2024-10-15 Manuscript.qmd
@@
 This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
-We focus on key topics such as spatial patterns, connectivity, simulations and modelling, landscape genetics, and spatial planning.
+We focus on key topics such as spatial patterns, connectivity, simulations and modeling, landscape genetics, and spatial planning.
 
@@
 \noindent _Unnithan Kumar et al.
-(2022) Moving beyond landscape resistance: Considerations for the future of connectivity modelling and conservation science.
-Landscape Ecology 37:2465–2480._ The paper provides a historical overview of connectivity modelling with a focus on resistance surfaces and discusses current missing aspects and how to address them.
+(2022) Moving beyond landscape resistance: Considerations for the future of connectivity modeling and conservation science.
+Landscape Ecology 37:2465–2480._ The paper provides a historical overview of connectivity modeling with a focus on resistance surfaces and discusses current missing aspects and how to address them.
 
@@
 \noindent _Synes et al.
-(2016) Emerging opportunities for landscape ecological modelling.
-Current Landscape Ecology Reports 1:146–167._ The paper identifies existing gaps in landscape ecological modelling while highlighting potential emerging opportunities.
+(2016) Emerging opportunities for landscape ecological modeling.
+Current Landscape Ecology Reports 1:146–167._ The paper identifies existing gaps in landscape ecological modeling while highlighting potential emerging opportunities.
 

### 2024-10-18 Sections/01-introduction.qmd
use modeling spelling

--- 2024-10-18 Sections/01-introduction.qmd
@@
 Linking spatial heterogeneity and ecological processes, including potential interactions between heterogeneity and processes, is the fundamental concept of landscape ecology [@Turner2015LandscapeEcology].
-Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modelling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].
+Typical research topics include, but are not limited to, pattern-process links, landscape complexity, ecological flows, scale effects, landscape modeling, conservation, drivers and consequences of land use and land cover (LULC) change, and human activities within landscapes [@Wu2013KeyConcepts].
 

### 2024-10-18 Sections/04-connectivity.qmd
use modeling spelling

@@
 Assumptions are rather in how such results are used in an ecological context -->
-Least-cost modelling identifies potential pathways between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
+Least-cost modeling identifies potential pathways between two points that minimizes the related movement or dispersal costs based on the resistance surface [@Adriaensen2003ApplicationLeastcost].
 By calculating pathways between two points it is possible to estimate connectivity between these points based on the accumulated cost along the path.
@@
 However, resistance surfaces are not free of critique including missing spatio-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
-Thus, recent developments have shifted towards a combination of both resistance and processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or general computational improvements [@Leonard2017GlowSoftware].
+Thus, recent developments have shifted towards a combination of both resistance and processes-based modeling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modeling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or general computational improvements [@Leonard2017GlowSoftware].
 
@@
 Recently, a new generation of graph-theoretical metrics, commonly reported in combination [@Hashemi2022ReviewEcological], have been developed to account for landscape features that are critical for space use and species-specific movement or dispersal [@Petsas2021LandscapeConnectivity].
-Furthermore, the modelling framework was initially only applied to single species, but is increasingly being used to model multispecies landscapes to account for interspecific movement and dispersal abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
-In addition, modern multiple-layer graph approaches enable the modelling of not only spatial but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
-Modelling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].
+Furthermore, the modeling framework was initially only applied to single species, but is increasingly being used to model multispecies landscapes to account for interspecific movement and dispersal abilities and habitat preferences [@Clauzel2018CouplingLandscapebased; @Modica2021ImplementationMultispecies; @Liang2023MultispeciesApproach].
+In addition, modern multiple-layer graph approaches enable the modeling of not only spatial but also spatio-temporal graphs [@Dale2017SpatiotemporalGraphs; @Uroy2021HowAssess].
+Modeling the temporal dynamics helps to identify habitat patches that impact connectivity over time and the corresponding effects on biodiversity patterns [@Uroy2021HowAssess].
 

### 2024-10-18 Sections/05-sampling.qmd
use modeling spelling

--- 2024-10-17 Sections/05-sampling.qmd
@@
 Furthermore, spatial autocorrelation in model residuals can be diagnosed using the similarity between two sites as a function of the distance between them, i.e., correlograms [@Dormann2007MethodsAccount].
-If spatial autocorrelation is detected researchers may either consider further data collection or use modelling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount] or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].
-
-<!--
-Software to perform multi-scale analysis include the R package *multilandr*, or the Python libraries *ZonalMetrics* [@Adamczyk2017ZonalMetricsPython], and *PyLandStats*.
--->
+If spatial autocorrelation is detected researchers may either consider further data collection or use modeling methods to accommodate for spatial dependencies [@Zuckerberg2020ReviewOverlapping], such as mixed models [@Dormann2007MethodsAccount] or apply smoothing kernels that compute distance-weighted averages surrounding the sites [@Heaton2011SpatialRegression].

### 2024-10-18 Sections/06-simulations.qmd
use modeling spelling

--- 2024-10-18 Sections/06-simulations.qmd
@@
 In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
-In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
+In order to facilitate future model developments recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
 Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

### 2024-10-18 Sections/09-conclusion.qmd
use modeling spelling

--- 2024-10-18 Sections/09-conclusion.qmd
@@
 Thus, future progress within the field will largely depend on available software and computational tools [@Costanza2019DescribingAnalyzing].
-Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
+Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modeling, landscape genetics and spatial planning.
 In this context, we highlight open-source software as a cornerstone of "Open Science" offering key advantages like shareability, reproducibility, and transparency, which provide great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
@@
 In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].
-Due to its high flexibility and performance artificial intelligence has become popular in ecology, and deep learning and machine learning algorithms can be used for, e.g., mapping, classifying and extracting features, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
+Due to its high flexibility and performance artificial intelligence has become popular in ecology, and deep learning and machine learning algorithms can be used for, e.g., mapping, classifying and extracting features, modeling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
 Furthermore, quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].

### 2024-10-18 Sections/07-genetics.qmd

@@
 Landscape genetics unite molecular population genetics, spatial statistics, and landscape ecology and emerged from the goal to study the interaction between landscape features and microevolutionary processes, such as gene flow, genetic drift and selection [@Manel2003LandscapeGenetics; @Balkenhol2017LandscapeGenomics].
-Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modelling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
+Landscape genetic approaches have traditionally been used to inform landscape ecology applications, e.g., through connectivity modeling and spatial conservation planning [@Manel2003LandscapeGenetics; @Holderegger2008LandscapeGenetics; @Manel2013TenYears; @Bolliger2014LandscapeGenetics].
 Using holistic approaches that consider evolutionary processes and patterns in addition to ecological data can fortify results, e.g., by improving predictive models of species range shifts in response to climate change [@Aguirre-Liguori2021EvolutionaryGenomics], improving the identification and delineation of landscape connections among populations [@Homola2019LandscapeGenetics; @deFlamingh2024IntegratingHabitat], allowing for the interpretation of spatial structuring in context of socio-cultural connections [@Henson2021ConvergentGeographic], or understanding local adaptation associated with specific environments [@Sork2013PuttingLandscape; @Rolland2015CombiningNiche; @Aguirre-Liguori2021EvolutionaryGenomics].

### 2024-10-18 Sections/09-conclusion.qmd

--- 2024-10-18 Sections/09-conclusion.qmd
@@
 | Connectivity | Stand-alone software | Conefor | [@Saura2009ConeforSensinode] |
-| Connectivity | Stand-alone software | lscorridors | [@Ribeiro2017LandScapeCorridors] |
-| Connectivity | Stand-alone software | GFLOW | [@Leonard2017GlowSoftware] |
+| Connectivity | Stand-alone software | LSCorridors | [@Ribeiro2017LandScapeCorridors] |
 | Sampling | Python | ZonalMetrics | [@Adamczyk2017ZonalMetricsPython] |
@@
 | Simulations | R | rflsgen | [@Justeau-Allaire2022FragmentedLandscape] |
-| Simulations | Python | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
+| Simulations | Python | NLMpy | [@Etherington2015NLMpyPython] |
 | Simulations | Python | Pathwalker | [@UnnithanKumar2022PathwalkerNew] |

### 2024-10-18 Sections/03-spatial-patterns.qmd

@@
 <!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
-The strength of landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
-However, also limitations have been identified. 
-These limitations include shortcomings in their ability to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
+The strength of landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing derived products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
+However, also limitations have been identified for many landscape metrics. 
+These limitations include shortcomings to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 
@@
 @Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
-This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).
+This includes conditional entropy (representing complexity in configuration), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).
 
 Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years, there has been a surge in the development of entropy-based metrics for landscape ecology.
-In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of macrostate (the general state of a system) and microstate (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of 'macrostate' (the general state of a system) and 'microstate' (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
 <!-- R1: the edge length -> Could you specify a bit -->
@@
 
-As hundreds of metrics have already been developed for raster data, the most straightforward approach may be to reimplement these same metrics for vector data.
+As hundreds of metrics have already been developed for gridded raster data, the most straightforward approach may be to reimplement these same metrics for vector data.
 Although this is possible for many metrics, there are also metrics specifically related to the corresponding data models [@Yao2022VecLIFramework].

### 2024-10-18 Sections/08-spatial-planning.qmd

@@
 
-Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
+Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState;@hanson2024systematic].
 In the context of landscape ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
@@
 
-A second development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.
-Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
+A second development is the increasing use of deep reinforcement learning, a machine learning approach based not on patterns but agents and pathways.Reinforcement learning can be used to identify cost-efficient solutions to area-based conservation planning [@Silvestro2022ImprovingBiodiversity].
 Additionally, it can be applied to identify the best possible solutions to improve landscape connectivity indices [@Equihua2024ConnectivityConservation], such as the integral index of connectivity [@Pascual-Hortal2006ComparisonDevelopment].
@@
 Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
-However, this comes with the drawbacks of reduced interpretability and because it is not an exact approach no guarantee that the best possible solution can been found.
+<!-- MJ: Broadly agree (although similar can be said for many other heuristical approaches).
+The main benefit is in the use of dynamic and non-linear functions which would not be computationally feasible with exact algorithms.
+-->
+However, this comes with the drawbacks of reduced interpretability, and similar to traditional spatial planning approaches such as Marxan, it relies on heuristics that can not guarantee that a best possible solution can been found [@Hanson2019OptimalityPrioritizing].
 
@@
 There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
-Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the optimal solver will depend on the specific problem.
-However, due to software licenses and related cost the optimal solver might not be always accessible.
+<!-- MJ: Agree to some extent, although again there are many comparisons that have shown that commercial solvers will usually outcompete open-source solvers.--->
+Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the performance of any individual solvers can depend on the specific problem, data and computational resources.
+Nevertheless, existing comparisons often show that proprietary solvers can be used to obtain solutions to complex problems in reasonable time periods [@hanson2024systematic], however, due to software licenses and related cost certain solvers might not always be openly accessible.

### 2024-10-18 Sections/06-simulations.qmd
Making sure PR didnt add modelling

--- 2024-10-18 Sections/06-simulations.qmd
@@
 They can also incorporate a range of socio-economic, political and governance information to influence how agents (such as farmers) interact within a landscape [@murray2014combining].
-In order to facilitate future model developments recent progress includes adaptable modelling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
+In order to facilitate future model developments recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
 Furthermore, hybrid approaches, coupling or integrating different model types with individual-based models, could benefit from rich development histories or facilitate corresponding strengths [@OSullivan2016StrategicDirections], e.g., linking an individual-based and ecosystem model [@Polhill2011AgentbasedModelling].

### 2024-10-18 Sections/09-conclusion.qmd
Making sure PR didnt add modelling

@@
 Thus, future progress within the field will largely depend on integration of novel data sources and available software and computational tools [@Costanza2019DescribingAnalyzing].
-Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
+Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modeling, landscape genetics and spatial planning.
 In this context, we highlight open-source software as a cornerstone of "Open Science" offering key advantages like shareability, reproducibility, and transparency, which provide great benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

### 2024-10-19 Sections/01-introduction.qmd
Clean changes MJ

--- 2024-10-18 Sections/01-introduction.qmd
@@
 Computational ecology is crucial for landscape ecology as a research field because data is often context- and scale-dependent, making it challenging to design controllable, reproducible, and replicable experiments [@Petrovskii2012ComputationalEcology; but see @Wiersma2022ReviewLandscape for a review of experimental studies].
-Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the general increase of data availability [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is the need for advanced computational methods.
+Additionally, because landscape ecology is a cross-disciplinary field [@With2019EssentialsLandscape], the availability of data increases steadily [@Jaric2020IEcologyHarnessing], and the complexity of landscape systems [@Newman2019ScalingComplexity], there is a need for advanced computational methods.
 
 <!-- R1: Very good idea to suggest software implementing methodological developments. It would be even better with a recapitulative table -->
-Here, we introduce recent computational developments in multiple key topics of landscape ecology (Fig. 1), including software that implements or is potentially capable of executing them (Tab. 1).
+Here, we provide a perspective on recent developments and advances in computational methods in multiple key topics of landscape ecology (Fig.
+1), including software that implements them or is potentially capable of creating novel insights in these topics (Tab.
+1).
 In this context, we are focusing on open-source software and scripting languages such as R, Python, and Julia.
-However, we do not provide a systematic literature review or a general introduction to (computational) landscape ecology.
+However, we do not aim to provide a systematic literature review or a general introduction to (computational) landscape ecology.
 For such, please see @Turner2015LandscapeEcology, @Gergel2017LearningLandscape, or @With2019EssentialsLandscape.

### 2024-10-19 Sections/02-data.qmd
Clean changes MJ

@@
 
-Similarly the issue of scale is closely related to the used data model. This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
-Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories or habitats [@Jung2020global].
+Similarly the issue of scale is closely related to the used data model. 
+This includes the extent of the study area and resolution of the data (i.e., the smallest data unit).
+Additionally, thematic resolution in landscape ecology often refers to the values describing landscape features, e.g., the number of discrete LULC categories or habitats
+[@Jung2020GlobalMap].
 Various classification systems are relevant at local, regional, and global extent, such as the National Land Cover Database (NLCD), the Coordination of Information on the Environment (CORINE), or the Food and Agriculture Organization (FAO) systems [@Nedd2021SynthesisLand].
-Lastly, temporal scales are relevant in terms of extent and resolution, but also with regards to the detail of captured processes acting at the landscape scale.
+Lastly, temporal scales are relevant in terms of extent and resolution, but also with regards to the detail of captured ecological processes.
 
@@
 In contrast, geographic coordinate reference systems are based on degrees and measure distances as great-circle distances.
-Numerous coordinate reference systems exist and novel projections are constantly being developed to improve spatial cohesion, efficiency and accuracy [@bauer2014optimisation].
+Numerous coordinate reference systems exist and novel projections are constantly being developed to improve spatial cohesion, efficiency and accuracy [@Bauer-Marschallinger2014OptimisationGlobal].
 However each reference systems comes with distinct properties that may distort areas, distances, angles, and aimed to represent specific regions or countries. 
-Therefore, for analysts it is essential to choose the appropriate spatial reference system based on the available data, area of interest, and the specific research question.
+Therefore, it is essential to choose the appropriate spatial reference system based on the available data, area of interest, and the specific research question.
 
 Data quality is generally variable, and data products are not flawless, potentially containing errors and biases.
-For example, global land-cover products typically have an overall accuracy of 70 to 80% [@xu2024comparative], indicating that 20 to 30% of the grid cells may be misclassified.
+For example, global land-cover products typically have an overall accuracy of 70 to 80% [@Xu2024ComparativeValidation], indicating that 20 to 30% of the grid cells may be misclassified.
 Furthermore, these misclassifications are not random, but are often correlated with specific LULC types, regions, and seasonal variations [@Liu2021FinerresolutionMapping]. 
-Rather than relying on global products, there are however also increasingly voices that advocate for the production and use of regional land-cover products as alternative [@tulbure2022regional].
+Rather than relying on global products, however, there are also increasingly voices that advocate for the production and use of regional land-cover products [@Tulbure2022RegionalMatters].
 Nevertheless, many landscape ecology studies tend to accept the data at face value, neglecting its accuracy and inherent uncertainty.

### 2024-10-19 Sections/03-spatial-patterns.qmd
Clean changes MJ

@@
 <!-- R1: They are also naturally extracted from remote sensing products (raster data), which often makes them easier to monitor over time -->
-The strength of landscape metrics is that they are easy to apply, communicate, and can be extracted straightforwardly from raster data based on remote sensing derived products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
+The strength of landscape metrics is that they are easy to apply, communicate, and can be calculated straightforwardly from raster data based on remote sensing derived products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
 However, also limitations have been identified for many landscape metrics.
@@
 @Nowosad2019InformationTheory proposed to compress information about the landscape's composition and configuration into a co-occurrence matrix which could be used to calculate various entropy measures.
-This includes conditional entropy (representing complexity in configuration), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).
+This includes conditional entropy (representing configurational complexity), joint entropy (representing overall spatial-thematic complexity), mutual information, and relative mutual information (both representing the degree of spatial autocorrelation).
 
 Based on concepts from thermodynamics initiated by the work of @Cushman2015ThermodynamicsLandscape and @Vranken2015ReviewUse, in the last few years, there has been a surge in the development of entropy-based metrics for landscape ecology.
-In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of 'macrostate' (the general state of a system) and 'microstate' (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
+In the late 1800s, Boltzmann formulated a probabilistic interpretation of the second law of thermodynamics using the concepts of "macrostate" (the general state of a system) and "microstate" (the configuration of the system elements) [@Gao2019ComputationBoltzmann].
 <!-- R1: the edge length -> Could you specify a bit -->

### 2024-10-19 Sections/05-sampling.qmd
Clean changes MJ

@@
 Furthermore, identifying scales at which the ecological processes of interest operate is a crucial step [@Jackson2015AreEcologists].
-However, in practice, the appropriate scales may not be obvious or data at such scales may not be available and landscape-wide and local process interactions can occur across multiple scales [@miguet2016determines].
+However, in practice, the appropriate scales may not be obvious or data at such scales may not be available and landscape-wide and local process interactions can occur across multiple scales [@Miguet2016WhatDetermines].
 Here analyses at multiple spatial scales can help to identify the scales of interest.

### 2024-10-19 Sections/06-simulations.qmd
Clean changes MJ

@@
 In landscape ecology individual-based models are increasingly used to model social-ecological systems [@Carauta2018CanPreferential; @Dislich2018LanduseChange; @Gonzalez-Redin2019ExploringSustainable], but also disturbances [@Seidl2017ClimateChange; @Spies2017UsingAgentbased; @Kamimura2019AgentbasedModelling], or connectivity [@Day2020IndividualbasedModeling].
-They can also incorporate a range of socio-economic, political and governance information to influence how agents (such as farmers) interact within a landscape [@murray2014combining].
+They can also incorporate a range of socio-economic, political, and governance information to influence how individuals, such as farmers, interact within a landscape [@Murray-Rust2014CombiningAgent].
 In order to facilitate future model developments recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
@@
 Digital twins are particular useful for real-time workflows that allow now- and forecasting of complex dynamics in a landscape.
-The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology [@deKoning2023DigitalTwins].
+The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology.
 They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
-<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g.
-drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
-This is further facilitated by the increasing rate of novel sources of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].
-
 Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
@@
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
+<!-- R1: You could also mention the increasing availability and accessibility of sensors deployed by drones along with data processing tools (e.g.
+drone photogrammetry, LiDAR) which enable high-frequency, high-resolution monitoring of 2D and 3D landscape models -->
+This is further facilitated by the increasing rate of novel sources of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].

### 2024-10-19 Sections/08-spatial-planning.qmd
Clean changes MJ

--- 2024-10-18 Sections/08-spatial-planning.qmd
@@
 
-Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState].
+Spatial planning uses decision theory to identify and allocate areas to specific purposes, such as reserve selection through spatial conservation prioritization [@Margules2000SystematicConservation; @Kukkala2013CoreConcepts; @Jung2024AssessmentState; @Hanson2024SystematicConservation].
 In the context of landscape ecology, it provides a pattern-process link and outcomes related to decision making by distinguishing between structure, function, and scales [@Hersperger2021LandscapeEcological].
@@
 Its strength mainly lies in the fact that, for large problems where exact approaches are currently inefficient, it can approximate high-quality solutions -->
-However, this comes with the drawbacks of reduced interpretability and because it is not an exact approach no guarantee that the best possible solution can been found.
+However, this comes with the drawbacks of reduced interpretability, and similar to traditional spatial planning approaches such as Marxan, it relies on heuristics that can not guarantee that a best possible solution can been found [@Hanson2019OptimalityPrioritizing].
 
@@
 There are many excellent open-source solvers (in MILP, CP, and SAT), and their openness offers a lot more possibilities for extension and fine-tuning to specific problems -->
-Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the optimal solver will depend on the specific problem.
-However, due to software licenses and related cost the optimal solver might not be always accessible.
+<!-- MJ: Agree to some extent, although again there are many comparisons that have shown that commercial solvers will usually outcompete open-source solvers.--->
+Unfortunately, there are differences between open-source and proprietary solvers in terms of computational efficiency [@Hanson2019OptimalityPrioritizing] and the performance of any individual solvers can depend on the specific problem, data and computational resources.
+Nevertheless, existing comparisons often show that proprietary solvers can be used to obtain solutions to complex problems in reasonable time periods [@Hanson2024SystematicConservation], however, due to software licenses and related cost certain solvers might not always be openly accessible.

### 2024-10-19 Sections/09-conclusion.qmd
Clean changes MJ

@@
 | Spatial patterns | R | motif | [@Nowosad2021MotifOpensource] |
-| Spatial patterns | Python | LecoS | [@jung2016lecos] |
 | Spatial patterns | Python | GeoPandas | [@Jordahl2020GeopandasPython] |
@@
 | Spatial patterns | Python | PyLandStats | [@Bosch2019PyLandStatsOpensource] |
+| Spatial patterns | Python | LecoS | [@Jung2016LecoSPython] |
 | Spatial patterns | Julia | GeoStats | [@Hoffimann2018GeoStatsjlHighperformance] |
@@
 These systems provide both passive sensors, such as multispectral, hyperspectral, and thermal sensors and active optical sensors, such as LiDAR and SAR [@Crowley2020RemoteSensings].
-Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible to scientists and spatial planners [@Markham2023ReviewMethods].
+Related to this, there is a growing number of up- and downscaling approaches for remote sensing data which makes a wider range of products accessible [@Markham2023ReviewMethods].
 In combination with other increasingly available data sources, e.g, citizen science data, ecology has entered the era of big data, requiring methods that can handle heterogeneous data and high-throughput computing resources [@Farley2018SituatingEcology].

### 2024-10-19 Sections/09-conclusion.qmd
Include references from issue

@@
 Alongside the previously introduced methodological advances specific to landscape ecology, the research field will also benefit from technological innovations from related disciplines. 
-This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems.
+This includes advances in remote sensing technology with increasing data availability and diversity from a range of sources, including satellites, unmanned aerial vehicles, and ground-based sensor systems [@Yu2019LandscapeEcology; @Crowley2020RemoteSensings].
 These systems provide both passive sensors, such as multispectral, hyperspectral, and thermal sensors and active optical sensors, such as LiDAR and SAR [@Crowley2020RemoteSensings].

### 2024-10-19 Review.qmd
Adding line numbers to replies

--- 2024-10-16 Review.qmd
@@
 
-\textcolor[HTML]{007895}{Reply: This is a very helpful comment about the general structure of the manuscript and we tried to address it as much as possible.
+\textcolor[HTML]{007895}{Reply: This is a very helpful comment about the general structure of the manuscript and we tried to address it the best possible.
 This included several points.
-First, we agree that a general section on data is useful for readers.
-Thus, we added such a section at the beginning of the manuscript.
-Second, we modified especially the first paragraphs of the "Spatial patterns" (sub-)sections in order to improve the logical flow of the manuscript.
+First, we agree that a general section on data models is useful for readers.
+Thus, we added such a section at the beginning of the manuscript (L65 - 92).
+Second, we modified especially the first paragraphs of most (sub-)sections in the "Spatial patterns" section to improve the logical flow of the manuscript (L93 - 202).
 Third, we rearranged and modified the "Connectivity" section as well.
-We removed the "Structural connectivity" sub-section by including the key statements into the general paragraph about connectivity.}
+We removed the "Structural connectivity" sub-section by including the key statements into the general paragraph about connectivity (L207 - 212).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added the term "model-driven" as suggested.}
+\textcolor[HTML]{007895}{Reply: We added the term "model-driven" as suggested (L53 - 55).}
 
@@
 \textcolor[HTML]{007895}{Reply: Indeed, such a table is very helpful.
-Thus, we removed all paragraphs that only list software from the specific sections and now list all software in a central table (Tab.
+Thus, we removed all paragraphs that list software from the specific sections and include all software in a central table (Tab.
 1).}
@@
 
-\textcolor[HTML]{007895}{Reply: We agree that landscape metrics can be extracted straightforwardly from raster data, and we are adding this as an advantage related to them.}
+\textcolor[HTML]{007895}{Reply: We agree that landscape metrics can be extracted straightforwardly from raster data, and we are adding this as an advantage related to them (L102 - 104).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: Indeed, we forgot to add a sub-section title which is added now.}
+\textcolor[HTML]{007895}{Reply: Indeed, we forgot to add a sub-section title which is added now (L116).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We replaced "entropy" with "measure".}
+\textcolor[HTML]{007895}{Reply: We replaced "entropy" with "measure" (L128 - 130).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added a brief definition of edge length in this context.}
+\textcolor[HTML]{007895}{Reply: We added a brief definition of edge length in this context (L144 - 146).}
 
@@
 \textcolor[HTML]{007895}{Reply: Similar to landscape metrics, also surface metrics are scale-dependent because both are designed to quantify landscape heterogeneity - which is inherently scale-dependent.
-We added this explanation including an according literature reference.}
+We added this explanation including an according literature reference (L172 - 173).}
 
@@
 \textcolor[HTML]{007895}{Reply: We removed the term "query" from the sentence because we agree that it is a source of confusion.
-However, we believe that the sentence using only "areas" instead is clear enough for readers to follow.}
+However, we believe that the sentence using only "areas" instead is clear enough for readers to follow (L193-201).}
 
@@
 \textcolor[HTML]{007895}{Reply: We agree that the whole section was confusing for readers unfamiliar with the topic.
-Thus, we reworded it to make it easier to follow by removing a lot of the database jargon.}
+Thus, we reworded it to make it easier to follow by removing a lot of the database jargon (L193-201).}
 
@@
 
-\textcolor[HTML]{e93c26}{Reply: TODO Thanks for pointing this out. We added it to the figure.}
+\textcolor[HTML]{e93c26}{Reply: Thank you for pointing this out. We added it to the figure.}
 
@@
 
-\textcolor[HTML]{e93c26}{Reply: TODO}
+\textcolor[HTML]{007895}{Reply: The figure caption now includes more details how the different topcis of the figure are interlinked.}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added "flow" as suggested.}
+\textcolor[HTML]{007895}{Reply: We added "flow" as suggested (L204 - 205).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added "dispersal" as suggested.}
+\textcolor[HTML]{007895}{Reply: We added "dispersal" as suggested (L217 - 219).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added dispersal as suggested.}
+\textcolor[HTML]{007895}{Reply: We added dispersal as suggested (L217 - 219).}
 
@@
 Furthermore, we believe that this aspect is now also highlighted due to the other suggestions of the reviewer that we implemented in this section.
-Generally, we added this each time movement is mentioned in the section.}
+Generally, we added this each time movement is mentioned in the section (L204 - 284).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added "disperse" as suggested.}
+\textcolor[HTML]{007895}{Reply: We added "disperse" as suggested (L227 - 230).}
 
@@
 \textcolor[HTML]{007895}{Reply: We are unsure why the point about edge artifacts was added to the manuscript.
-Thus, we completely removed it here since we think it adds more confusion than helping the readers.}
+Thus, we completely removed it here since we think it adds more confusion than helping the readers (L230 - 232).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We agree that the original statement about underlying assumptions was unclear and reworded the sentence.}
+\textcolor[HTML]{007895}{Reply: We agree that the original statement about underlying assumptions was unclear and reworded the sentence (L237 - 239).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added this specification as suggested.}
+\textcolor[HTML]{007895}{Reply: We added this specification as suggested (L263 - 266).}
 
@@
 However, we believe it is also helpful to keep it in this section specifically.
-However, we added a new short "Data" section at the beginning of the manuscript in which we also highlight importance of input data specifically.}
+However, we added a new short "Data" section at the beginning of the manuscript in which we also highlight importance of input data specifically (L86 - 92).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: Thank you very much for adding this insightful remark about the references. We reworded the sentence to ensure the remarks are reflected in the manuscript.}
+\textcolor[HTML]{007895}{Reply: Thank you very much for adding this insightful remark about the references. We reworded the sentence to ensure the remarks are reflected in the manuscript (L331 - 334)}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added that digital twins benefit from the increasing rate of data generation in ecology.}
+\textcolor[HTML]{007895}{Reply: We added that digital twins benefit from the increasing rate of data generation in ecology (L365 - 367).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We renamed the section as suggested.}
+\textcolor[HTML]{007895}{Reply: We renamed the section as suggested (L392).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We reworded the sentence being more specific about the corresponding references.}
+\textcolor[HTML]{007895}{Reply: We reworded the sentence being more specific about the corresponding references (L408 - 410).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: Thank you for pointing out this detail. We added this as another potentail drawback of the method.}
+\textcolor[HTML]{007895}{Reply: Thank you for pointing out this detail. We added this as another potentail drawback of the method (L418 - 421).}
 
@@
 \textcolor[HTML]{007895}{Reply: We reworded the sentence stating that there are differences between solvers and that the optimal solver depends on the specific problem.
-However, the best solver could potentially not be accessible due to software licenses.}
+However, the best solver could potentially not be accessible due to software licenses (424 - 429).}

### 2024-10-20 Sections/04-connectivity.qmd
Move around some sentences

@@
 -->
-
 <!-- R1: Here connectivity seems to be described through the prism of animals, you should modify it slightly to include plants, which do not move, but disperse through different vectors, e.g.
 animals, gravity, wind, or water -->
+
 <!-- R1: facilitate or impeded movement and flow -->
@@
 Connectivity can be divided into structural and functional connectivity [@Taylor2006LandscapeConnectivity].
+
 Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.

### 2024-10-20 Sections/06-simulations.qmd
Move around some sentences

@@
 They can also incorporate a range of socio-economic, political, and governance information to influence how individuals, such as farmers, interact within a landscape [@Murray-Rust2014CombiningAgent].
+
 In order to facilitate future model developments recent progress includes adaptable modeling frameworks [@Schumaker2018HexSimModeling] or reusable building blocks [@Berger2024ReusableBuilding].
@@
 The digital twins concept include three connected elements, namely *i)* a physical object or system, *ii)* its digital representation, *iii)* and a data exchange between the former two [@Jones2020CharacterisingDigital; @deKoning2023DigitalTwins].
-Digital twins are particular useful for real-time workflows that allow now- and forecasting of complex dynamics in a landscape.
 The need for timely and evidence-based decision making in combination with increasing data availability make digital twins a powerful tool for landscape ecology.
-They aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
-Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
-Digital twins are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
-Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].
 The use of general data processing, statistical and mechanistic models, or artificial intelligence allows digital twins to constantly update the digital representation and analyze, visualize, or predict the physical counterpart.
@@
 This is further facilitated by the increasing rate of novel sources of data generation in ecology, e.g., due to sensors deployed on airplane, satellites, or unmanned aerial vehicle [@Farley2018SituatingEcology].
+Digital twins are particular useful for real-time workflows that allow now- and forecasting of complex dynamics in a landscape.
+
+Digital twins aim to provide real-time state of nature measurements, early detection of conservation trends, relationships of ecosystem trends and environmental conditions, impact assessments of interventions, or to identify uncertainties and information gaps [@deKoning2023DigitalTwins].
+Thus, there is increasing interest and use of digital twins in both the industry and academia [@Jones2020CharacterisingDigital].
+They  are increasingly applied in agricultural landscapes research, e.g., for livestock farming, controlled environment farming, or fertilization management [@Cesco2023SmartAgriculture; @Purcell2023DigitalTwins].
+Further applications include planning of rural ecological landscapes [@Tan2024DigitalTwin] or exploration of relationships between urban expansion and vegetation coverage [@Zhao2022ApplyingDigital].

### 2024-10-22 Manuscript.qmd
Clean changes MJ
Move around some sentences
Comments Marti

@@
         \newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\textcolor[HTML]{A6CE39}{\aiOrcid\hspace{1mm}#1}}}
-        \newcommand{\mail}[1]{\href{mailto:#1}{\textcolor[HTML]{052E81}{\faEnvelope\hspace{1mm}#1}}}
+        \newcommand{\mail}[1]{\href{mailto:#1}{\textcolor[HTML]{32B2DA}{\faEnvelope\hspace{1mm}#1}}}
 
@@
 
+\newpage
+
 # Abstract
@@
 {{< include Sections/04-connectivity.qmd >}}
-{{< include Sections/05-sampling.qmd >}}
-{{< include Sections/06-simulations.qmd >}}
-{{< include Sections/07-genetics.qmd >}}
+{{< include Sections/05-genetics.qmd >}}
+{{< include Sections/06-sampling.qmd >}}
+{{< include Sections/07-simulations.qmd >}}
 {{< include Sections/08-spatial-planning.qmd >}}
@@
 \footnotesize
-\setstretch{1.0}
 ::: {#refs}

### 2024-10-22 Review.qmd
Move around some sentences
Comments Marti

@@
 Third, we rearranged and modified the "Connectivity" section as well.
-We removed the "Structural connectivity" sub-section by including the key statements into the general paragraph about connectivity (L207 - 212).}
+We removed the "Structural connectivity" sub-section by including the key statements into the general paragraph about connectivity (L207 - 212).
+Last, we moved the "Landscape genetics" section closer to the "Connectivity" section due to the importance of connectivity modelling in landscape genetics.}
 
@@
 
-\textcolor[HTML]{007895}{Reply: Indeed, we forgot to add a sub-section title which is added now (L116).}
+\textcolor[HTML]{007895}{Reply: Indeed, we forgot to add a sub-section title which is added now (L115).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We replaced "entropy" with "measure" (L128 - 130).}
+\textcolor[HTML]{007895}{Reply: We replaced "entropy" with "measure" (L127 - 129).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added a brief definition of edge length in this context (L144 - 146).}
+\textcolor[HTML]{007895}{Reply: We added a brief definition of edge length in this context (L143 - 145).}
 
@@
 \textcolor[HTML]{007895}{Reply: We removed the term "query" from the sentence because we agree that it is a source of confusion.
-However, we believe that the sentence using only "areas" instead is clear enough for readers to follow (L193-201).}
+However, we believe that the sentence using only "areas" instead is clear enough for readers to follow (L195-197).}
 
@@
 \textcolor[HTML]{007895}{Reply: We agree that the whole section was confusing for readers unfamiliar with the topic.
-Thus, we reworded it to make it easier to follow by removing a lot of the database jargon (L193-201).}
+Thus, we reworded it to make it easier to follow by removing a lot of the database jargon (L194-203).}
 
@@
 
-\textcolor[HTML]{e93c26}{Reply: Thank you for pointing this out. We added it to the figure.}
+\textcolor[HTML]{007895}{Reply: Thank you for pointing this out. We added it to the figure.}
 
@@
 
-\textcolor[HTML]{007895}{Reply: The figure caption now includes more details how the different topcis of the figure are interlinked.}
+\textcolor[HTML]{007895}{Reply: The figure caption now includes more details how the different topics of the figure are interlinked.}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added "flow" as suggested (L204 - 205).}
+\textcolor[HTML]{007895}{Reply: We added "flow" as suggested (L205 - 206).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added "dispersal" as suggested (L217 - 219).}
+\textcolor[HTML]{007895}{Reply: We added "dispersal" as suggested (L218 - 220).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added dispersal as suggested (L217 - 219).}
+\textcolor[HTML]{007895}{Reply: We added dispersal as suggested (L218 - 220).}
 
@@
 Furthermore, we believe that this aspect is now also highlighted due to the other suggestions of the reviewer that we implemented in this section.
-Generally, we added this each time movement is mentioned in the section (L204 - 284).}
+Generally, we added this each time movement is mentioned in the section (L205 - 285).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added "disperse" as suggested (L227 - 230).}
+\textcolor[HTML]{007895}{Reply: We added "disperse" as suggested (L228 - 231).}
 
@@
 \textcolor[HTML]{007895}{Reply: We are unsure why the point about edge artifacts was added to the manuscript.
-Thus, we completely removed it here since we think it adds more confusion than helping the readers (L230 - 232).}
+Thus, we completely removed it here since we think it adds more confusion than helping the readers (L231 - 233).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We agree that the original statement about underlying assumptions was unclear and reworded the sentence (L237 - 239).}
+\textcolor[HTML]{007895}{Reply: We agree that the original statement about underlying assumptions was unclear and reworded the sentence (L238 - 239).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added this specification as suggested (L263 - 266).}
+\textcolor[HTML]{007895}{Reply: We added this specification as suggested (L264 - 267).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: Thank you very much for adding this insightful remark about the references. We reworded the sentence to ensure the remarks are reflected in the manuscript (L331 - 334)}
+\textcolor[HTML]{007895}{Reply: Thank you very much for adding this insightful remark about the references. We reworded the sentence to ensure the remarks are reflected in the manuscript (L356 - 359)}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We added that digital twins benefit from the increasing rate of data generation in ecology (L365 - 367).}
+\textcolor[HTML]{007895}{Reply: We added that digital twins benefit from the increasing rate of data generation in ecology (L384 - 386).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We renamed the section as suggested (L392).}
+\textcolor[HTML]{007895}{Reply: We renamed the section as suggested (L395).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: We reworded the sentence being more specific about the corresponding references (L408 - 410).}
+\textcolor[HTML]{007895}{Reply: We reworded the sentence being more specific about the corresponding references (L411 - 413).}
 
@@
 
-\textcolor[HTML]{007895}{Reply: Thank you for pointing out this detail. We added this as another potentail drawback of the method (L418 - 421).}
+\textcolor[HTML]{007895}{Reply: Thank you for pointing out this detail. We added this as another potential drawback of the method (L421 - 424).}
 
@@
 \textcolor[HTML]{007895}{Reply: We reworded the sentence stating that there are differences between solvers and that the optimal solver depends on the specific problem.
-However, the best solver could potentially not be accessible due to software licenses (424 - 429).}
+However, the best solver could potentially not be accessible due to software licenses (427 - 432).}

### 2024-10-22 Sections/03-spatial-patterns.qmd
Minor typos
Adding line numbers to replies
Move around some sentences
Comments Marti

@@
 The strength of landscape metrics is that they are easy to apply, communicate, and can be calculated straightforwardly from raster data based on remote sensing derived products [@Dramstad2009SpatialMetrics; @Lausch2015UnderstandingQuantifying].
-However, also limitations have been identified for many landscape metrics.
-These limitations include shortcomings to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
+However, limitations have also been identified for many landscape metrics.
+These include shortcomings to quantify the spatial structure, sensitivity to both spatial scale and thematic resolution, and correlation and redundancy between metrics [@Li2004UseMisuse; @Dramstad2009SpatialMetrics; @Kupfer2012LandscapeEcology].
 
@@
 More recently, principal component analysis over a set of landscape blocks revealed two main components of landscape configuration, namely complexity and aggregation which together explain about 70% of variance [@Nowosad2018GlobalInventory].
-These components of landscape patterns are in line with recent reviews that have similarly suggested that there are two fundamental components of landscape patterns, namely amount and adjacency [@Gustafson2019HowHas; @Riitters2019PatternMetrics].
+These results are in line with recent reviews that have similarly suggested that there are two fundamental components of landscape patterns, namely amount and adjacency [@Gustafson2019HowHas; @Riitters2019PatternMetrics].
 These components are connected to complexity and aggregation as suggested by @Nowosad2018GlobalInventory and later formalized in @Nowosad2019InformationTheory.
@@
 The metrics are able to quantify various characteristics, such as roughness, skewness and kurtosis, total and relative amplitudes, curvatures of local peaks, or surface bearing area ratios [@Cushman2010GradientParadigm].
+
 <!-- R1: I guess that such metrics are also very scale dependent -->
@@
 Software to calculate these metrics is still rare and further research is needed into the specific pattern-process links and ecologically meaningful interpretations [@Kedron2018SurfaceMetrics; @Frazier2019EmergingTrajectories].
+
 Related to surface metrics, another recent approach to quantifying the gradient surface model is based on frequencies and local adjacencies of continuous input pixel values [@Riitters2023MappingLandscape].
 
-## Vector metrics
+## Vector-based metrics
 
@@
 More recently, a unified theoretical foundation for measuring shape compactness was introduced using a set of ten distinct properties of a circle and metrics associated with each of these properties [@Angel2010TenCompactness].
+
 However, also vector-based metrics have limitations.
@@
 This allows for the identification of areas with similar signatures compared to the focal area, e.g., areas with similar environmental conditions [@Jasiewicz2014LandscapeSimilarity].
-
 Additionally, it is possible to calculate spatial signatures for multiple areas and group them to similar clusters based on their signatures [@Nowosad2021PatternbasedIdentification].

### 2024-10-22 Sections/09-conclusion.qmd
Move around some sentences
Comments Marti

@@
 
-<!-- MH: Double check programming languages -->
+\newpage
+\setstretch{1.0}
 
@@
 | Connectivity | Stand-alone software | LSCorridors | [@Ribeiro2017LandScapeCorridors] |
+| Landscape genetics | R | graph4lg | [@Savary2021Graph4lgPackage] |
+| Landscape genetics | R | landgenreport | [@Gruber2015LandgenreportNew] |
+| Landscape genetics | R | adegenet | [@Jombart2008AdegenetPackage] |
+| Landscape genetics | R | ape | [@Paradis2019Ape50] |
+| Landscape genetics | Python | Python based scripts | [@Etherington2011PythonBased] |
+| Landscape genetics | Python | SDMtoolbox 2.0 | [@Brown2017SDMtoolbox20] |
 | Sampling | Python | ZonalMetrics | [@Adamczyk2017ZonalMetricsPython] |
-| Simulations | R | NLMR | [@Sciaini2018NLMRLandscapetools] |
-| Simulations | R | rflsgen | [@Justeau-Allaire2022FragmentedLandscape] |
-| Simulations | Python | NLMpy | [@Etherington2015NLMpyPython] |
-| Simulations | Python | Pathwalker | [@UnnithanKumar2022PathwalkerNew] |
-| Simulations | Julia | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
-| Simulations | NetLogo | EFForTS-LGraf | [@Salecker2019EFForTSLGrafLandscape] |
-| Simulations | Stand-alone software | GradientLand | [@Cambui2015GradientLandSoftware] |
-| Simulations | Stand-alone software | Landscape Generator | [@vanStrien2016ImprovedNeutral] |
-| Simulations | Stand-alone software | HexSim | [@Schumaker2018HexSimModeling] |
-| Simulations | Stand-alone software | RangeShifter2.0 | [@Bocedi2021RangeShifter20] |
-| Genetics | R | graph4lg | [@Savary2021Graph4lgPackage] |
-| Genetics | R | landgenreport | [@Gruber2015LandgenreportNew] |
-| Genetics | R | adegenet | [@Jombart2008AdegenetPackage] |
-| Genetics | R | ape | [@Paradis2019Ape50] |
-| Genetics | Python | Python based scripts | [@Etherington2011PythonBased] |
-| Genetics | Python | SDMtoolbox 2.0 | [@Brown2017SDMtoolbox20] |
-| Planning | R | prioritizr | [@Hanson2024PrioritizrSystematic] |
-| Planning | R | CoCo | [@vanMantgem2023CocoConservation] |
-| Planning | R | restoptr | [@Justeau-Allaire2023RestoptrPackage] |
-| Planning | Python | CAPTAIN | [@Silvestro2022ImprovingBiodiversity] |
+| Simulation models | R | NLMR | [@Sciaini2018NLMRLandscapetools] |
+| Simulation models | R | rflsgen | [@Justeau-Allaire2022FragmentedLandscape] |
+| Simulation models | Python | NLMpy | [@Etherington2015NLMpyPython] |
+| Simulation models | Python | Pathwalker | [@UnnithanKumar2022PathwalkerNew] |
+| Simulation models | Julia | NeutralLandscapes | [@Poisot2023NeutralLandscapesJulia] |
+| Simulation models | Stand-alone software | GradientLand | [@Cambui2015GradientLandSoftware] |
+| Simulation models | Stand-alone software | Landscape Generator | [@vanStrien2016ImprovedNeutral] |
+| Simulation models | Stand-alone software | HexSim | [@Schumaker2018HexSimModeling] |
+| Simulation models | Stand-alone software | RangeShifter2.0 | [@Bocedi2021RangeShifter20] |
+| Spatial planning | R | prioritizr | [@Hanson2024PrioritizrSystematic] |
+| Spatial planning | R | CoCo | [@vanMantgem2023CocoConservation] |
+| Spatial planning | R | restoptr | [@Justeau-Allaire2023RestoptrPackage] |
+| Spatial planning | Python | CAPTAIN | [@Silvestro2022ImprovingBiodiversity] |
 
 : Software and models that implement recent methodological advances in landscape ecology.
+
+\newpage
+\normalsize
+\setstretch{1.5}
 

### 2024-11-19 Manuscript.qmd
Adjust abstract
Update README

@@
 \noindent All authors contributed particular sections to the manuscript and assisted with the revision. Overall coordination, first and revised draft was led by MHKH.
-Correspondence to [mhk.hesselbarth@gmail.com](mailto:mhk.hesselbarth@gmail.com)
+Correspondence to [mhk.hesselbarth@gmail.com](mailto:mhk.hesselbarth@gmail.com).
 
@@
 This review provides an overview of recent developments in computational methods that have advanced the research field of landscape ecology.
-We focus on key topics such as spatial patterns, connectivity, simulations and modeling, landscape genetics, and spatial planning.
+We focus on key topics such as spatial patterns, connectivity, landscape genetics, sampling, simulations and modeling, and spatial planning.
 
